### 2015-05-12 docs/ms_sageAbundance.Rnw
adds Rnw manuscript file

\documentclass[12pt]{article}
\usepackage{graphics}
\usepackage{graphicx, verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}
\usepackage{authblk}
\textwidth=6.2in
\textheight=8.5in
%\parskip=.3cm
\oddsidemargin=.1in
\evensidemargin=.1in
\headheight=-.3in

\begin{document}
\SweaveOpts{concordance=TRUE}
\SweaveOpts{highlight=TRUE, tidy=TRUE, keep.space=TRUE, keep.blank.space=FALSE, keep.comment=TRUE}
\SweaveOpts{prefix.string=Fig}
%To put figures in subfolder
%\SweaveOpts{prefix.string=figures/fig}
\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}

\title{Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species}
\author[1]{\large Andrew T. Tredennick}
\author[1]{\large Peter B. Adler}
\author[2,3,4,5]{\large Mevin B. Hooten}
\author[5,6,7,8]{\large Cameron L. Aldridge}
\author[9]{\large Colin G. Homer}
\author[1]{\large David T. Iles}
\author[1]{\large Andrew Kleinhesselink}
\author[1]{\large Eric LaMalfa}
\author[1]{\large Rebecca Mann}
\affil[1]{\footnotesize \emph{Department of Wildland Resources and the Ecology Center, Utah State University}}
\affil[2]{\footnotesize \emph{U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit}}
\affil[3]{\footnotesize \emph{Department of Fish, Wildlife, and Conservation Biology, Colorado State University}}
\affil[4]{\footnotesize \emph{Department of Statistics, Colorado State University}}
\affil[5]{\footnotesize \emph{Graduate Degree Program in Ecology, Colorado State University}}
\affil[6]{\footnotesize \emph{Department of Ecosystem Science and Sustainability, Colorado State University}}
\affil[7]{\footnotesize \emph{U.S. Geological Survey, Fort Collins Science Center}}
\affil[8]{\footnotesize \emph{Natural Resource Ecology Laboratory, Colorado State University}}
\affil[9]{\footnotesize \emph{U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center}}
\maketitle

\section*{\large Summary}
\begin{enumerate}
\item Global environmental changes, like climate change, tend to play out at the landscape and regional scales. Thus, it is reasonable that attempts to forecast the impacts of such change do so at similar spatial scales. For plant species, this has meant relying primarily on a single tool, species distribution models, that contain little information on population dynamics and states.

\item Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using “local” scale data from, generally, 1 $\text{m}^{2}$ plots. These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.

\item Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model. At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals. To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (Artemesia spp.) cover in southwestern Wyoming. We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.

\item The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.

\item Our approach...
\end{enumerate}

\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.

%----------------------------------------------------
\section{\large Introduction}
%----------------------------------------------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology. The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions. However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale. Local-scale demographic data make building population projection models an easy task (Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions. Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.

The fundamental limitation of local-scale data is that they are insufficient for estimating the full extent of variation in model parameters possible at larger spatial and temporal extents (Freckleton et al., 2011; Queenborough et al., 2011). It is possible to estimate the variance of population model parameters related to spatial and temporal variability, but the observations used to estimate parameter uncertainty tend to be clustered in space or time. Thus, extrapolating beyond the spatial or temporal resolution of the data used to parameterize models can lead to biased and wrong forecasts of future population states.

Alternatively, large-scale trends in populations are easily detected using widely available monitoring data, but such data are rarely used to project future population states. Given the tradeoffs between the spatial-scale of data and the inference that can be drawn from the data, we suggest combining the best features of local-scale population modeling and large-scale monitoring data. Specifically, we propose a new approach for modeling population dynamics at large spatial extents based on the theory and mechanics of individual-based models (IBMs) as applied to remotely-sensed time series data.

The approach we describe in this paper flows naturally from recent work on density-structured population models. Density-structured models rely on modeling the transitions of a population among discrete states rather than the traditional approach of modeling the transitions of individuals. Relatively few ecologists have taken up the density-structured mantle first proposed by Taylor, C. M. et al. (2004), but recently Freckleton et al. (2011) and Queenborough et al. (2011) have higlighted the potential value the approach.

\section{Materials and Methods}
\subsection{Data: Remotely-sensed time series}
To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]). Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming. The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65. To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011. In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush. For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each full dataset contains 754,596 observations (27 years $\times$ 27,948 pixels).

\begin{figure}[h]
\center{}\includegraphics[width=1\textwidth]{/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/figures/StudyAreasMap_1and2}

\protect\caption{\label{fig:Location-of-the}Location of the study area in southwestern
Wyoming and a snapshot of the percent cover data in 1984 for each
site (a and b). Green shading on the map shows projected distribution
of sagebrush from Kuchler (19xx). Latitude and longitude are in decimal
degrees.}

\end{figure}

\subsection{Data: Climate covariates}
Our approach is based on modeling how changes in plant populations relate to climate variables. We a priori narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model future projections. In all vital rate regressions we focus on the demographic processes driving changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t}, and how these processes are mediated by climate. The four climate variables in our vital rate regressions are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]). Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database. Climate data were extracted for the centroid our study area.

Climate data used to inform model projections of future sagebrush
cover came from average outputs of all models used in the most recent
IPCC report, the Coupled Model Intercomparison Project 5 (CMIP5; http://cmip-pcmdi.llnl.gov/cmip5/).
We downloaded model estimates of historic climate (1950-2000) and
projected climate (2050-2100) for the centroid of our study area.
We wanted to keep the temporal variability from observed climate consistent,
so our aim was to adjust observed climate according to projected changes.
For temperature we simply subtracted the average historic temperature
from the average projected temperature. For precipitation we calculated
the proportional change: we subtracted average projected precipitation
from average historic precipitation, and then divided by average historic
precipitation. We did this for three ``Representative Concentration
Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
http://tntcat.iiasa.ac.at/RcpDb/.
Basically they correspond to stabilization of radiative forcing before
2100, after 2100, and increasing greenhouse gas emissions, respectively.
We implemented projected temperature and precipitation changes in
the model by perturbing observed annual values by the values in Table
\ref{tab:Changes-in-climate}.

\begin{table}[h]
\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
average projections.}

\begin{tabular}{llllll}
\hline
 & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
\hline
RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
\hline
\end{tabular}
\end{table}

\subsection{Spatio-temporal model for sagebrush cover}
We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects. Since the spatial and temporal effects are additive, we are not considering space-time interactions. Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times. For statistical convenience, and with little added bias, we consider discrete values of sagebrush percent cover as counts, giving us the formulation

\begin{align}
[\textbf{C}] &= \text{Poisson}(\boldsymbol{\lambda}) \\
\boldsymbol{\lambda} &= \text{exp}(\boldsymbol{\mu}) 
\end{align}

\noindent which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$). We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):

\begin{align}
\mu_{s,t} = \underbrace{\beta_{0} + \beta_{1}\mu_{s,t-1}}_\text{temporal} + \underbrace{\mathbf{X}_{t}\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{s}}_\text{spatial} + \underbrace{\varepsilon_{s,t}}_\text{error}
\end{align}

Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations. The dimensionality is just too large. To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation. A common strategy is to express high dimensional effects, like $\boldysmbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$. We can then define the spatial effect as

\begin{align}
\boldsymbol{\eta} = \textbf{K}\boldsymbol{\alpha}.
\end{align}

\noindent In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations. Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.

Obviously, the last remaining obstacle is to parameterize \textbf{K}. Here we use kernel convolution to interpolate the spatial random effect between \emph{m} ``knots'' that are nonrandomly distributed across the space of our study area. These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots. We use a simple Gaussian kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are

\begin{align}
K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
\end{align}

\noindent where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth. Based on residual spatial structure we found after fitting a simple model without spatial effects, we set $\phi$ to about one-third of the range of spatial dependence evident in the residual variograms -- 250 meters in this case. In the end, the simplest way to understand our dimension reduction strategy is to consider $\alpha$s as weights for each knot.

\subsection{Computing and model estimation}

\end{document}

### 2015-05-12 R/Conn_util_funcs.R
adding more R sripts and Conn unitiliy functions

@@ -1,0 +1,272 @@
+#' Produce an RW2 structure matrix for a line (e.g. for time series)
+#' @param x length of vector
+#' @return RW2 precision matrix
+#' @export 
+#' @keywords adjacency
+#' @author Paul Conn
+linear_adj_RW2 <- function(x){
+  Q=Matrix(0,x,x)
+  for(i in 1:(x-2)){
+    Q[i,i+2]=1
+    Q[i+2,i]=1
+  }
+  for(i in 1:(x-1)){
+    Q[i,i+1]=-4
+    Q[i+1,i]=-4
+  }
+  Q[1,2]=-2
+  Q[2,1]=-2
+  Q[x,x-1]=-2
+  Q[x-1,x]=-2
+  diag(Q)=-rowSums(Q)
+  Q
+}  
+
+#' Produce an RW1 adjacency matrix for a rectangular grid for use with areal spatial models (queens move)
+#' @param x number of cells on horizontal side of grid
+#' @param y number of cells on vertical side of grid
+#' @param byrow If TRUE, cell indices are filled along rows (default is FALSE)
+#' @return adjacency matrix
+#' @export 
+#' @keywords adjacency
+#' @author Paul Conn \email{paul.conn@@noaa.gov}
+rect_adj <- function(x,y,byrow=FALSE){
+  Ind=matrix(c(1:(x*y)),y,x,byrow=byrow)
+  if(byrow==TRUE)Ind=t(Ind)
+  n.row=nrow(Ind)
+  n.col=ncol(Ind)
+  Adj=matrix(0,x*y,x*y)
+  for(i in 1:n.row){
+    for(j in 1:n.col){
+      if(i==1 & j==1){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+      }
+      if(i==1 & j>1 & j<n.col){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row+1]=1
+      }
[…]

### 2015-05-12 scripts/getKnots_DistanceMatrix.R

@@ -1,0 +1,97 @@
+
+library(reshape2)
+library(raster)
+library(sp)
+library(rgeos)
+library(Matrix)
+
+####
+#### Bring in data
+####
+studyArea="1"
+
+ifelse(studyArea=="1",
+       climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
+       climD<-read.csv("../../studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
+ifelse(studyArea=="1",
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2check.csv"),
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_SA2.csv"))
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+shrub.ras <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/sage/ls3731_00sage_subset_est.img")
+
+years=sort(unique(fullD$Year))
+n.years=length(years)
+sites=sort(unique(fullD$ID))
+n.sites=length(sites)
+
+gray.colors.rev <- function (n, start = 0, end = 1, gamma = 1){ 
+  gray(seq.int(to= start^gamma, from = end^gamma, length.out = n)^(1/gamma))
+}
+
+####
+####  Make Initial Maps 
+####
+subD <- subset(fullD, Year==years[1])
+subC <- dcast(subD[,c('Lon', 'Lat', 'Cover')], formula = Lon~Lat)
+resMat <- as.matrix(subC[, 2:ncol(subC)])
+lon.locs=sort(unique(subD$Lon))
+lat.locs=sort(unique(subD$Lat))
+dat <- subset(subD, select = c("Lon", "Lat", "Cover"))
+coordinates(dat) <- c("Lon", "Lat")
+proj4string(dat)
+image(resMat, x=lon.locs,y=lat.locs,asp=TRUE,col=gray.colors.rev(100))
+
+####
+#### Generate knots and plot
+####
+Coords=coordinates(dat)
+x.min=min(Coords[,1])
+x.max=max(Coords[,1])
+y.min=min(Coords[,2])
+y.max=max(Coords[,2])
+
+splits <- 8
[…]

### 2015-05-12 scripts/getKnots_DistanceMatrix_smallSet.R

@@ -1,0 +1,94 @@
+
+library(reshape2)
+library(raster)
+library(sp)
+library(rgeos)
+library(Matrix)
+
+####
+#### Bring in data
+####
+climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv")
+rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2small.csv")
+
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+shrub.ras <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/sage/ls3731_00sage_subset_est.img")
+shrub.ras <- aggregate(shrub.ras, fact=10, fun="mean", na.rm=TRUE)
+
+years=sort(unique(fullD$Year))
+n.years=length(years)
+sites=sort(unique(fullD$ID))
+n.sites=length(sites)
+
+gray.colors.rev <- function (n, start = 0, end = 1, gamma = 1){ 
+  gray(seq.int(to= start^gamma, from = end^gamma, length.out = n)^(1/gamma))
+}
+
+####
+####  Make Initial Maps 
+####
+subD <- subset(fullD, Year==years[1])
+subC <- dcast(subD[,c('Lon', 'Lat', 'Cover')], formula = Lon~Lat)
+resMat <- as.matrix(subC[, 2:ncol(subC)])
+lon.locs=sort(unique(subD$Lon))
+lat.locs=sort(unique(subD$Lat))
+dat <- subset(subD, select = c("Lon", "Lat", "Cover"))
+coordinates(dat) <- c("Lon", "Lat")
+proj4string(dat)
+image(resMat, x=lon.locs,y=lat.locs,asp=TRUE,col=gray.colors.rev(100))
+
+####
+#### Generate knots and plot
+####
+Coords=coordinates(dat)
+x.min=min(Coords[,1])
+x.max=max(Coords[,1])
+y.min=min(Coords[,2])
+y.max=max(Coords[,2])
+
+splits <- 2
+X=x.min+(x.max-x.min)/splits*c(0:splits)
+Y=y.min+(y.max-y.min)/splits*c(splits:0)
+XY=expand.grid(x=X,y=Y)
+
[…]

### 2015-05-12 scripts/sageAbundance_NIMBLE_smallSet_test.R

@@ -1,0 +1,143 @@
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+####
+#### Load libraries
+####
+library(nimble)
+# library(ggplot2)
+# library(tidyr)
+
+####
+#### Bring in data
+####
+climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv")
+rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2small.csv")
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# get knot data
+load("Knot_cell_distances_smallSet.Rdata")
+
+####
+#### Take subset of data for test fitting
+####
+# rown <- 169
+# rows <- 1
+# ntest <- rown*rows
+# 
+# subD <- subset(fullD, ID<(ntest+1))
+# growD <- subset(subD, is.na(CoverLag)==FALSE)
+growD <- subset(fullD, Year>1984)
+growD$Cover <- round(growD$Cover,0)
+growD$CoverLag <- round(growD$CoverLag,0)
+# growD <- subset(growD, Year<1986)
+
+
+# growD <- subset(fullD, Year>=1985 & Year<=1987)
+# coords.mod <- as.matrix(growD[,c("Lon", "Lat")])
+# D <- as.matrix(dist(coords.mod))
+
+####
+#### Quick glm for initial values ----------------------
+####
+mod <- glm(Cover ~ CoverLag, family="poisson", data=growD)
+summary(mod)
+init_alpha <- as.numeric(coef(mod)[1])
+init_beta <- as.numeric(coef(mod)[2])
+
+####
+#### Set up data structure for NIMBLE ----------------------
+####
+# index <- which(is.na(growD$Cover)==FALSE) #vector of rows of non-missing data
+N.obs <- growD$Cover
+lagObs <- growD$CoverLag
+cell  <- growD$ID
+K <- K.data$K
+nKnots <- ncol(K)
+
+constants <- list(nObs = length(N.obs), 
+                  nKnots = nKnots,
+                  N = lagObs,
+                  K = K,
+                  cell = cell)
+data <- list(C = N.obs)
+inits <- list(intMu=rnorm(1,init_alpha,0.05), 
+              betaMu=rnorm(1,init_beta,0.05), 
+              alpha=rep(0,nKnots))
+
[…]

### 2015-05-12 scripts/sageAbundance_NIMBLEtest.R

@@ -1,0 +1,144 @@
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+####
+#### Load libraries
+####
+library(nimble)
+library(ggplot2)
+library(tidyr)
+
+####
+#### Bring in data
+####
+studyArea="1"
+ifelse(studyArea=="1",
+       climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
+       climD<-read.csv("../../studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
+ifelse(studyArea=="1",
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2check.csv"),
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_SA2.csv"))
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# get knot data
+load("Knot_cell_distances.Rdata")
+
+####
+#### Take subset of data for test fitting
+####
+# rown <- 169
+# rows <- 1
+# ntest <- rown*rows
+# 
+# subD <- subset(fullD, ID<(ntest+1))
+# growD <- subset(subD, is.na(CoverLag)==FALSE)
+growD <- subset(fullD, Year>1984)
+growD <- subset(growD, Year<1986)
+
+
+# growD <- subset(fullD, Year>=1985 & Year<=1987)
+# coords.mod <- as.matrix(growD[,c("Lon", "Lat")])
+# D <- as.matrix(dist(coords.mod))
+
+####
+#### Quick glm for initial values ----------------------
+####
+mod <- glm(Cover ~ CoverLag, family="poisson", data=growD)
+summary(mod)
+init_alpha <- as.numeric(coef(mod)[1])
+init_beta <- as.numeric(coef(mod)[2])
+
+####
+#### Set up data structure for NIMBLE ----------------------
+####
+index <- which(is.na(growD$Cover)==FALSE) #vector of rows of non-missing data
+N.obs <- growD$Cover[index]
+lagObs <- growD$CoverLag[index]
+cell  <- growD$ID[index]
+K <- K.data$K[index,]
+nKnots <- ncol(K)
+
+constants <- list(nObs = length(N.obs), 
+                  nKnots = nKnots,
+                  N = lagObs,
[…]

### 2015-05-12 scripts/sageGompertz_SpatialStructure.R

@@ -1,0 +1,85 @@
+# Bayesian model for sagebrush pixel growth
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#load libraries
+library(rjags)
+library(coda)
+
+####
+#### Bring in data
+####
+studyArea="1"
+
+ifelse(studyArea=="1",
+       climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
+       climD<-read.csv("../../studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
+ifelse(studyArea=="1",
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2check.csv"),
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_SA2.csv"))
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# get knot data
+load("Knot_cell_distances.Rdata")
+
+####
+#### Take subset of data for test fitting
+####
+# rown <- 169
+# rows <- 40
+# ntest <- rown*rows
+# 
+# subD <- subset(fullD, ID<(ntest+1))
+# growD <- subset(subD, is.na(CoverLag)==FALSE)
+growD <- subset(fullD, Year>1984)
+# tmpI <- which(growD$CoverLag==0)
+# growD$CoverLag[tmpI] <- 0.001
+
+# growD <- subset(fullD, Year>=1985 & Year<=1987)
+# coords.mod <- as.matrix(growD[,c("Lon", "Lat")])
+# D <- as.matrix(dist(coords.mod))
+
+####
+#### Set up data structure for JAGS and iteration values
+####
+index <- which(is.na(growD$Cover)==FALSE) #vector of rows of non-missing data
+N.obs <- growD$Cover[index]
+lagObs <- growD$CoverLag[index]
+cell  <- growD$ID[index]
+K <- K.data$K
+nKnots <- ncol(K)
+
+dataJ <- list(C = N.obs,
+              N = lagObs,
+              nObs = length(N.obs),
[…]

### 2015-05-12 scripts/sageGompertz_SpatialStructure_JAGS.R

@@ -1,0 +1,13 @@
+model{
+  for(i in 1:nObs){
+    C[i] ~ dpois(lambda[i])
+    lambda[i] <- exp(intMu + betaMu*N[i] + eta[cell[i]])
+  }
+  eta <- K%*%alpha
+  for(j in 1:nKnots){
+    alpha[j] ~ dnorm(0,tau)
+  }
+  intMu ~ dnorm(0,1e-6)
+  betaMu ~ dnorm(0,1e-6)
+  tau ~ dgamma(1,0.01)
+}

### 2015-05-12 scripts/sageGompertz_StateSpace_AST.R

@@ -1,0 +1,91 @@
+# Bayesian model for sagebrush pixel growth
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+ptm <- proc.time()
+
+#load libraries
+library(rjags)
+library(coda)
+
+####
+#### Bring in data
+####
+studyArea="1"
+
+ifelse(studyArea=="1",
+       climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
+       climD<-read.csv("../../studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
+ifelse(studyArea=="1",
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2check.csv"),
+       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_SA2.csv"))
+
+# merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# get knot data
+load("Knot_cell_distances.Rdata")
+
+####
+#### Take subset of data for test fitting
+####
+rown <- 169
+rows <- 1
+ntest <- rown*rows
+
+growD <- subset(fullD, ID<(ntest+1))
+# growD <- subset(subD, is.na(CoverLag)==FALSE)
+# growD <- subset(subD, Year>1984)
+# tmpI <- which(growD$CoverLag==0)
+# growD$CoverLag[tmpI] <- 0.001
+
+
+####
+#### Set up data structure for JAGS and iteration values
+####
+index <- which(is.na(growD$Cover)==FALSE) #vector of rows of non-missing data
+N.obs <- growD$Cover[index]
+lagObs <- growD$CoverLag[index]
+cell  <- growD$ID[index]
+nCells <- length(unique(cell))
+cellMod <- c(1:nCells)
+years <- growD$Year - 1983
+nYears <- length(unique(years))
+K <- K.data$K
+nKnots <- ncol(K)
+
+dataJ <- list(C = N.obs,
+              lagC = lagObs,
+              nObs = length(N.obs),
+              cell = cell,
+              nCells = nCells,
[…]

### 2015-05-12 scripts/sageGompertz_StateSpace_AST_JAGS.R

@@ -1,0 +1,32 @@
+model{
+  #process model
+  for(t in 2:nYears){
+    for(i in 1:nCells){
+      lambda[i,t] <- exp(intMu + betaMu*lambda[i,t-1] + eta[i])
+    }
+  }
+  
+  #initial conditions
+  for(k in 1:nCells){
+    lambda[k,1] <- exp(intMu + betaMu*lambda0 + eta[k])
+  }
+  
+  #expand alpha effects for full spatial grid
+  eta <- K%*%alpha
+  
+  #likelihood
+  for(i in 1:nObs){
+    C[i] ~ dpois(lambda[cellMod[i], years[i]])
+  }
+  
+  #alpha priors (iid)
+  for(j in 1:nKnots){
+    alpha[j] ~ dnorm(0,tau)
+  }
+  
+  #parameter priors
+  lambda0 ~ dunif(0,10)
+  intMu ~ dnorm(0,0.01)
+  betaMu ~ dnorm(0,0.01)
+  tau ~ dgamma(1,0.01)
+}

### 2015-05-12 scripts/sage_script_mevin.R

@@ -1,0 +1,167 @@
+####
+####  Read in Data
+####
+
+library(raster)
+climD <- read.csv("FormattedClimate_WY_SA1.csv")
+rawD<-read.csv("WY_SAGECoverData_V2check.csv")
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+head(fullD)
+
+years=sort(unique(fullD$Year))
+n.years=length(years)
+sites=sort(unique(fullD$ID))
+n.sites=length(sites)
+
+gray.colors.rev <- function (n, start = 0, end = 1, gamma = 1){ 
+  gray(seq.int(to= start^gamma, from = end^gamma, length.out = n)^(1/gamma))
+}
+
+####
+####  Make Initial Maps 
+####
+
+t=1
+year.tmp=(fullD$Year==years[t])
+cover.tmp=fullD[year.tmp,'Cover']
+lon.locs=sort(unique(fullD[year.tmp,'Lon']))
+lat.locs=sort(unique(fullD[year.tmp,'Lat']))
+n.lon=length(lon.locs)
+n.lat=length(lat.locs)
+#plot(fullD[year.tmp,c('Lon','Lat')],cex=.2)
+image(matrix(cover.tmp,n.lon,n.lat),x=lon.locs,y=lat.locs,asp=TRUE)
+points(sample(x = lon.locs, size = 400), sample(lat.locs, 400), pch=19)
+
+
+
+
+
+cover.mean.vec=rep(0,n.years)
+cover.mat=matrix(0,n.sites,n.years)
+png(file="sage_maps.png",width=10,height=5,units="in",res=200)
+layout(matrix(1:n.years,4,7,byrow=TRUE))
+par(xaxt="n",yaxt="n",mar=c(1.5,1,2,1))
+for(t in 1:n.years){
+  year.tmp=(fullD$Year==years[t])
+  cover.tmp=fullD[year.tmp,'Cover']
+  cover.mat[,t]=cover.tmp
+  cover.mean.vec[t]=mean(cover.tmp,na.rm=TRUE)
+  lon.locs=sort(unique(fullD[year.tmp,'Lon']))
+  lat.locs=sort(unique(fullD[year.tmp,'Lat']))
+  n.lon=length(lon.locs)
+  n.lat=length(lat.locs)
[…]

### 2015-06-18 inst/examples/plot_prelim_mcmc.Rmd
adds plot of prelim results example

---
title: ''
output: html_document
---

Check preliminary results from Rstan MCMC. Took 19 days and 9 hours to fit the full spatial dataset at 2000 iterations with 3 chains in parallel.

```{r plot_mcmc, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
library(ggmcmc)
library(ggplot2)
mcmc <- readRDS("../../results/stanmcmc_sage.RDS")
ggplot(subset(mcmc, Parameter=="beta_mu"),
       aes(x=Iteration, y=value, color=as.factor(Chain)))+
  geom_line()
ggplot(subset(mcmc, Parameter=="int_mu"),
       aes(x=Iteration, y=value, color=as.factor(Chain)))+
  geom_line()
ggplot(subset(mcmc, Parameter=="alpha[1]"),
       aes(x=Iteration, y=value, color=as.factor(Chain)))+
  geom_line()
```

The slope mixed well, but clearly the alphas and the intercept are fighting. Need to work this out with Mevin.

### 2015-06-25 docs/ms_sageAbundance.Rmd
updates ms to RMD rather than Rnw

--- 2015-05-12 docs/ms_sageAbundance.Rnw
@@
-\documentclass[12pt]{article}
-\usepackage{graphics}
-\usepackage{graphicx, verbatim}
-\usepackage{amssymb}
-\usepackage{amsmath}
-% \usepackage[T1]{fontenc}
-% \usepackage[utf8]{inputenc}
-\usepackage{authblk}
-\textwidth=6.2in
-\textheight=8.5in
-%\parskip=.3cm
-\oddsidemargin=.1in
-\evensidemargin=.1in
-\headheight=-.3in
+---
+layout: 12pt
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+   - \usepackage{setspace}
+   - \doublespacing
+bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
+csl: components/ecology.csl
 
-\begin{document}
-\SweaveOpts{concordance=TRUE}
-\SweaveOpts{highlight=TRUE, tidy=TRUE, keep.space=TRUE, keep.blank.space=FALSE, keep.comment=TRUE}
-\SweaveOpts{prefix.string=Fig}
-%To put figures in subfolder
-%\SweaveOpts{prefix.string=figures/fig}
-\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
-\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
-\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
+## rmarkdown render options
+output:
+  pdf_document:
+    fig_caption: true
+    keep_tex: true
+fontsize: 12pt
+geometry: margin=1in
 
-\title{Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species}
-\author[1]{\large Andrew T. Tredennick}
-\author[1]{\large Peter B. Adler}
-\author[2,3,4,5]{\large Mevin B. Hooten}
-\author[5,6,7,8]{\large Cameron L. Aldridge}
-\author[9]{\large Colin G. Homer}
-\author[1]{\large David T. Iles}
-\author[1]{\large Andrew Kleinhesselink}
-\author[1]{\large Eric LaMalfa}
-\author[1]{\large Rebecca Mann}
-\affil[1]{\footnotesize \emph{Department of Wildland Resources and the Ecology Center, Utah State University}}
-\affil[2]{\footnotesize \emph{U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit}}
-\affil[3]{\footnotesize \emph{Department of Fish, Wildlife, and Conservation Biology, Colorado State University}}
-\affil[4]{\footnotesize \emph{Department of Statistics, Colorado State University}}
-\affil[5]{\footnotesize \emph{Graduate Degree Program in Ecology, Colorado State University}}
-\affil[6]{\footnotesize \emph{Department of Ecosystem Science and Sustainability, Colorado State University}}
-\affil[7]{\footnotesize \emph{U.S. Geological Survey, Fort Collins Science Center}}
-\affil[8]{\footnotesize \emph{Natural Resource Ecology Laboratory, Colorado State University}}
-\affil[9]{\footnotesize \emph{U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center}}
-\maketitle
+---
 
-\section*{\large Summary}
+Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species
+===========================================================================================
+
+### Andrew T. Tredennick[^corrauth], Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B. Adler
+
+*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Mevin B.
+Hooten, U.S.
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit; Department of Fish, Wildlife, and Conservation Biology, Colorado State University; Department of Statistics, Colorado State University; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+
+*Cameron L.
+Aldridge, Department of Ecosystem Science and Sustainability, Colorado State University; Natural Resource Ecology Laboratory, Colorado State University; U.S.
+Geological Survey, Fort Collins Science Center; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+
+*Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD*
+
+*David T. Iles, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Andrew Kleinhesselink, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Eric LaMalfa, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Rebecca Mann, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
+
+Summary
+-------
 \begin{enumerate}
@@
 
-\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.
+*Keywords*: population model, forecasting, integral projection model, etc.
 

### 2015-06-29 scripts/run_small_test.R
adds plotting examples like for eta

@@ -39,6 +39,7 @@ load("../results/Knot_cell_distances_smallSet.Rdata")
 model_string <- "
 data{
   int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
   int<lower=0> nknots; // number of interpolation knots
   int<lower=0> ncells; // number of cells
   int<lower=0> cellid[nobs]; // cell id
@@ -47,21 +48,25 @@ data{
   int y[nobs]; // observation vector
   int lag[nobs]; // lag cover vector
   matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
 }
 parameters{
   real int_mu;
   real<lower=0> beta_mu;
-  real<lower=0> sig_a;
-  real<lower=0> sig_mu;
+  real<lower=0.000001> sig_a;
+  real<lower=0.000001> sig_mu;
   vector[nknots] alpha;
   vector[nobs] lambda;
+  vector[ncovs] beta;
 }
 transformed parameters{
   vector[ncells] eta;
   vector[nobs] mu;
+  vector[nobs] climEffs;
   eta <- K*alpha;
+  climEffs <- X*beta;
   for(n in 1:nobs)
-    mu[n] <- int_mu + beta_mu*lag[n] + eta[cellid[n]];
+    mu[n] <- int_mu + beta_mu*lag[n] + climEffs[n] + eta[cellid[n]];
 }
 model{
   // Priors
@@ -70,6 +75,7 @@ model{
   sig_mu ~ uniform(0,10);
   int_mu ~ normal(0,100);
   beta_mu ~ normal(0,10);
+  beta ~ normal(0,10);
   // Likelihood
   lambda ~ normal(mu, sig_mu);
   y ~ poisson(exp(lambda));
@@ -79,47 +85,63 @@ model{
 ####
 ##  Send data to stan function for fitting
 ####
-inits <- list()
-inits[[1]] <- list(int_mu = 1, beta_mu = 0.05, 
-                   alpha = rep(0,ncol(K.data$K)), sigma=0.05, sig_a=0.05)
-inits[[2]] <- list(int_mu = 2, beta_mu = 0.01, 
-                   alpha = rep(0.5,ncol(K.data$K)), sigma=0.02, sig_a=0.005)
-inits[[3]] <- list(int_mu = 1.5, beta_mu = 0.02, 
-                   alpha = rep(0.25,ncol(K.data$K)), sigma=0.04, sig_a=0.025)
-
[…]

### 2015-06-29 scripts/getKnots_DistanceMatrix.R
updates script to default range parameter estimate given number of knots

@@ -4,18 +4,21 @@ library(raster)
 library(sp)
 library(rgeos)
 library(Matrix)
+library(gstat)
+
 
 ####
 #### Bring in data
 ####
+setwd("/Users/atredenn/Dropbox/sageAbundance_data/")
 studyArea="1"
 
 ifelse(studyArea=="1",
-       climD<-read.csv("../../studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
-       climD<-read.csv("../../studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
+       climD<-read.csv("./studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv"),
+       climD<-read.csv("./studyarea2/climate/DAYMET/FormattedClimate_WY_SA2.csv"))
 ifelse(studyArea=="1",
-       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_V2check.csv"),
-       rawD<-read.csv("../../cover_structure/WY_SAGECoverData_SA2.csv"))
+       rawD<-read.csv("./WY_SAGECoverData_V2check.csv"),
+       rawD<-read.csv("./WY_SAGECoverData_SA2.csv"))
 
 # merge in climate data 
 fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
@@ -44,6 +47,25 @@ coordinates(dat) <- c("Lon", "Lat")
 proj4string(dat)
 image(resMat, x=lon.locs,y=lat.locs,asp=TRUE,col=gray.colors.rev(100))
 
+
+####
+####  Construct variograms to estimate range parameter for knots
+####
+# par(xaxt="s",yaxt="s",mar= c(2, 2, 4, 2), mfrow=c(4,7))
+# n.years <- length(unique(fullD$Year))
+# for(t in 1:n.years){
+#   resD <- subset(fullD, Year==years[t])
+#   lon.locs=sort(unique(resD$Lon))
+#   lat.locs=sort(unique(resD$Lat))
+#   dat <- subset(resD, select = c("Lon", "Lat", "Cover"))
+#   dat <- dat[which(is.na(dat$Cover)==FALSE),]
+#   coordinates(dat) <- c("Lon", "Lat")
+#   varMod <- variogram(Cover~1, data=dat, width=50)
[…]

### 2015-06-29 docs/ms_sageAbundance.Rmd
updates ms with kernel rationale

@@
 
-%----------------------------------------------------
-\section{\large Introduction}
-%----------------------------------------------------
+Introduction
+-------------
+
 Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology.
@@
 
-\section{Materials and Methods}
-\subsection{Data: Remotely-sensed time series}
+Materials and Methods
+---------------------
+### Data: Remotely-sensed time series
 To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]).
@@
 
-\begin{figure}[h]
-\center{}\includegraphics[width=1\textwidth]{/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/figures/StudyAreasMap_1and2}
-
-\protect\caption{\label{fig:Location-of-the}Location of the study area in southwestern
-Wyoming and a snapshot of the percent cover data in 1984 for each
-site (a and b). Green shading on the map shows projected distribution
-of sagebrush from Kuchler (19xx). Latitude and longitude are in decimal
-degrees.}
-
-\end{figure}
-
-\subsection{Data: Climate covariates}
+### Data: Climate covariates
 Our approach is based on modeling how changes in plant populations relate to climate variables.
@@
 We implemented projected temperature and precipitation changes in
-the model by perturbing observed annual values by the values in Table
-\ref{tab:Changes-in-climate}.
+the model by perturbing observed annual values by the values in Table #.
 
@@
 average projections.}
-
 \begin{tabular}{llllll}
@@
 
-\subsection{Spatio-temporal model for sagebrush cover}
+### Additive spatio-temporal model for sagebrush cover
 We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects.
@@
 Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times.
-For statistical convenience, and with little added bias, we consider discrete values of sagebrush percent cover as counts, giving us the formulation
+For statistical convenience, we consider discrete values of sagebrush percent cover as counts, giving us the formulation
 
@@
 
-\noindent which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$).
+which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$).
 We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):
@@
 To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
-A common strategy is to express high dimensional effects, like $\boldysmbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$.
+A common strategy is to express high dimensional effects, like $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$.
 We can then define the spatial effect as
@@
 
-\noindent In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations.
+In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations.
 Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.
@@
 Obviously, the last remaining obstacle is to parameterize \textbf{K}.
-Here we use kernel convolution to interpolate the spatial random effect between \emph{m} ``knots'' that are nonrandomly distributed across the space of our study area.
+Here we use kernel convolution to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
 These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
@@
 K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
-w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
+w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma^{2}}),
 \end{align}
 
-\noindent where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
-Based on residual spatial structure we found after fitting a simple model without spatial effects, we set $\phi$ to about one-third of the range of spatial dependence evident in the residual variograms -- 250 meters in this case.
-In the end, the simplest way to understand our dimension reduction strategy is to consider $\alpha$s as weights for each knot.
+where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
+It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$.
+However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around eight knots and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al.
+2014).
 
-\subsection{Computing and model estimation}
-
-\end{document}
+### Computing and model estimation

### 2015-06-29 docs/ms_sageAbundance.tex
updates ms with kernel rationale

--- 2015-06-29 docs/ms_sageAbundance.Rmd
@@
----
-layout: 12pt
-header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-   - \usepackage{setspace}
-   - \doublespacing
-bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
-csl: components/ecology.csl
+\documentclass[12pt,]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1in]{geometry}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-## rmarkdown render options
-output:
-  pdf_document:
-    fig_caption: true
-    keep_tex: true
-fontsize: 12pt
-geometry: margin=1in
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
----
+%%% Change title format to be more compact
+\usepackage{titling}
+\setlength{\droptitle}{-2em}
+  \title{}
+  \pretitle{\vspace{\droptitle}}
+  \posttitle{}
+  \author{}
+  \preauthor{}\postauthor{}
+  \date{}
+  \predate{}\postdate{}
 
-Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species
-===========================================================================================
+\usepackage{lineno}
+\linenumbers
+\usepackage{setspace}
+\doublespacing
 
-### Andrew T. Tredennick[^corrauth], Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B. Adler
+\begin{document}
 
-*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\maketitle
 
-*Mevin B.
-Hooten, U.S.
-Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit; Department of Fish, Wildlife, and Conservation Biology, Colorado State University; Department of Statistics, Colorado State University; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+\section{Forecasting climate change impacts on plant population dynamics
+at large spatial extents: a test case with sagebrush (\emph{Artemisia})
+species}\label{forecasting-climate-change-impacts-on-plant-population-dynamics-at-large-spatial-extents-a-test-case-with-sagebrush-species}
 
-*Cameron L.
-Aldridge, Department of Ecosystem Science and Sustainability, Colorado State University; Natural Resource Ecology Laboratory, Colorado State University; U.S.
-Geological Survey, Fort Collins Science Center; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+\subsubsection[Andrew T. Tredennick, Mevin B. Hooten, Cameron L.
+Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric
+LaMalfa, Rebecca Mann, and Peter B. Adler]{Andrew T.
+Tredennick\footnote{Corresponding author:
+  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}},
+Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles,
+Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B.
+Adler}\label{andrew-t.-tredennickcorrauth-mevin-b.-hooten-cameron-l.-aldridge-colin-g.-homer-david-t.-iles-andrew-kleinhesselink-eric-lamalfa-rebecca-mann-and-peter-b.-adler}
 
-*Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD*
+\emph{Andrew T. Tredennick, Department of Wildland Resources and the
+Ecology Center, Utah State University, Logan, UT}
 
-*David T. Iles, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\emph{Mevin B. Hooten, U.S. Geological Survey, Colorado Cooperative Fish
+and Wildlife Research Unit; Department of Fish, Wildlife, and
+Conservation Biology, Colorado State University; Department of
+Statistics, Colorado State University; Graduate Degree Program in
+Ecology, Colorado State University, Fort Collins, CO}
 
-*Andrew Kleinhesselink, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\emph{Cameron L. Aldridge, Department of Ecosystem Science and
+Sustainability, Colorado State University; Natural Resource Ecology
+Laboratory, Colorado State University; U.S. Geological Survey, Fort
+Collins Science Center; Graduate Degree Program in Ecology, Colorado
+State University, Fort Collins, CO}
 
-*Eric LaMalfa, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\emph{Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources
+Observation and Science (EROS) Center, Sioux Falls, SD}
 
-*Rebecca Mann, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\emph{David T. Iles, Department of Wildland Resources and the Ecology
+Center, Utah State University, Logan, UT}
 
-*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\emph{Andrew Kleinhesselink, Department of Wildland Resources and the
+Ecology Center, Utah State University, Logan, UT}
 
-[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
+\emph{Eric LaMalfa, Department of Wildland Resources and the Ecology
+Center, Utah State University, Logan, UT}
 
-Summary
--------
+\emph{Rebecca Mann, Department of Wildland Resources and the Ecology
+Center, Utah State University, Logan, UT}
+
+\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
+Center, Utah State University, Logan, UT}
+
+\subsection{Summary}\label{summary}
+
 \begin{enumerate}
@@
 
-*Keywords*: population model, forecasting, integral projection model, etc.
+\emph{Keywords}: population model, forecasting, integral projection
+model, etc.
 
-Introduction
--------------
+\subsection{Introduction}\label{introduction}
 
-Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology.
-The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions.
-However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale.
-Local-scale demographic data make building population projection models an easy task (Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions.
-Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.
+Forecasting the impacts of climate change on plant populations and
+communities is a central challenge of modern ecology. The challenge lies
+in faithfully representing population dynamics at spatial and temporal
+scales relevant to policy and management decisions. However, almost
+every study of plant population dynamics relies on demographic
+observations recorded at the meter to sub-meter scale. Local-scale
+demographic data make building population projection models an easy task
+(Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is
+very difficult to extrapolate small-scale studies to large spatial
+extents with any certainty because the data likely only represent a
+small subset of parameter space and environmental conditions. Also,
+given the costliness (in time and money) of collecting detailed
+demographic data, it is unlikely that ``enough'' local-scale data can be
+collected to accurately project large-scale population dynamics.
 
-The fundamental limitation of local-scale data is that they are insufficient for estimating the full extent of variation in model parameters possible at larger spatial and temporal extents (Freckleton et al., 2011; Queenborough et al., 2011).
-It is possible to estimate the variance of population model parameters related to spatial and temporal variability, but the observations used to estimate parameter uncertainty tend to be clustered in space or time.
-Thus, extrapolating beyond the spatial or temporal resolution of the data used to parameterize models can lead to biased and wrong forecasts of future population states.
+The fundamental limitation of local-scale data is that they are
+insufficient for estimating the full extent of variation in model
+parameters possible at larger spatial and temporal extents (Freckleton
+et al., 2011; Queenborough et al., 2011). It is possible to estimate the
+variance of population model parameters related to spatial and temporal
+variability, but the observations used to estimate parameter uncertainty
+tend to be clustered in space or time. Thus, extrapolating beyond the
+spatial or temporal resolution of the data used to parameterize models
+can lead to biased and wrong forecasts of future population states.
 
-Alternatively, large-scale trends in populations are easily detected using widely available monitoring data, but such data are rarely used to project future population states.
-Given the tradeoffs between the spatial-scale of data and the inference that can be drawn from the data, we suggest combining the best features of local-scale population modeling and large-scale monitoring data.
-Specifically, we propose a new approach for modeling population dynamics at large spatial extents based on the theory and mechanics of individual-based models (IBMs) as applied to remotely-sensed time series data.
+Alternatively, large-scale trends in populations are easily detected
+using widely available monitoring data, but such data are rarely used to
+project future population states. Given the tradeoffs between the
+spatial-scale of data and the inference that can be drawn from the data,
+we suggest combining the best features of local-scale population
+modeling and large-scale monitoring data. Specifically, we propose a new
+approach for modeling population dynamics at large spatial extents based
+on the theory and mechanics of individual-based models (IBMs) as applied
+to remotely-sensed time series data.
 
-The approach we describe in this paper flows naturally from recent work on density-structured population models.
-Density-structured models rely on modeling the transitions of a population among discrete states rather than the traditional approach of modeling the transitions of individuals.
-Relatively few ecologists have taken up the density-structured mantle first proposed by Taylor, C.
-M.
-et al.
-(2004), but recently Freckleton et al.
-(2011) and Queenborough et al.
-(2011) have higlighted the potential value the approach.
+The approach we describe in this paper flows naturally from recent work
+on density-structured population models. Density-structured models rely
+on modeling the transitions of a population among discrete states rather
+than the traditional approach of modeling the transitions of
+individuals. Relatively few ecologists have taken up the
+density-structured mantle first proposed by Taylor, C. M. et al. (2004),
+but recently Freckleton et al. (2011) and Queenborough et al. (2011)
+have higlighted the potential value the approach.
 
-Materials and Methods
----------------------
-### Data: Remotely-sensed time series
-To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]).
-Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
-The regression tree model was further validated using another 500 field observations.
-For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
-To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images.
-This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011.
-In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush.
-For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27,948 pixels each.
-Thus, each full dataset contains 754,596 observations (27 years $\times$ 27,948 pixels).
+\subsection{Materials and Methods}\label{materials-and-methods}
 
-### Data: Climate covariates
-Our approach is based on modeling how changes in plant populations relate to climate variables.
-We a priori narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model future projections.
-In all vital rate regressions we focus on the demographic processes driving changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t}, and how these processes are mediated by climate.
-The four climate variables in our vital rate regressions are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]).
-Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database.
-Climate data were extracted for the centroid our study area.
+\subsubsection{Data: Remotely-sensed time
+series}\label{data-remotely-sensed-time-series}
 
-Climate data used to inform model projections of future sagebrush
-cover came from average outputs of all models used in the most recent
-IPCC report, the Coupled Model Intercomparison Project 5 (CMIP5; http://cmip-pcmdi.llnl.gov/cmip5/).
-We downloaded model estimates of historic climate (1950-2000) and
-projected climate (2050-2100) for the centroid of our study area.
-We wanted to keep the temporal variability from observed climate consistent,
-so our aim was to adjust observed climate according to projected changes.
-For temperature we simply subtracted the average historic temperature
-from the average projected temperature. For precipitation we calculated
-the proportional change: we subtracted average projected precipitation
-from average historic precipitation, and then divided by average historic
+To show our pixel-based modeling approach we use a subset of a
+remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub
+percent cover in Wyoming (Homer et al., 2012) (Figure
+{[}fig:Location-of-the{]}). Sagebrush percent cover was estimated using
+a regression tree to relate ground reflectances retrieved by three
+sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780
+field observations of sagebrush cover distributed across Wyoming. The
+regression tree model was further validated using another 500 field
+observations. For Wyoming sagebrush, the model achieved an \(R^{2}\) =
+0.65. To hind-cast sagebrush cover the regression tree model was applied
+to historical remote sensing images. This resulted in an annual time
+series of sagebrush cover at 30 m resolution from 1984 to 2011. In this
+dataset, values represent the percentage of a 30 \(\times\) 30 m pixel
+covered by sagebrush. For our purposes we use two 5 \(\times\) 5 km
+subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each
+full dataset contains 754,596 observations (27 years \(\times\) 27,948
+pixels).
+
+\subsubsection{Data: Climate covariates}\label{data-climate-covariates}
+
+Our approach is based on modeling how changes in plant populations
+relate to climate variables. We a priori narrowed our focus to climate
+covariates we know are important for sagebrush and that could be easily
+calculated from general circulation model future projections. In all
+vital rate regressions we focus on the demographic processes driving
+changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t},
+and how these processes are mediated by climate. The four climate
+variables in our vital rate regressions are: (1) year \emph{t} fall
+through summer precipitation, (2) year \emph{t}-1 fall through summer
+precipitation, (3) year \emph{t} average spring temperature, and (4)
+year \emph{t}-1 average spring temperature (Table
+{[}tab:Definition-of-climate{]}). Climate data to parameterize vital
+rate regressions (`observed' climate) came from the PRISM climate
+database. Climate data were extracted for the centroid our study area.
+
+Climate data used to inform model projections of future sagebrush cover
+came from average outputs of all models used in the most recent IPCC
+report, the Coupled Model Intercomparison Project 5 (CMIP5;
+\url{http://cmip-pcmdi.llnl.gov/cmip5/}). We downloaded model estimates
+of historic climate (1950-2000) and projected climate (2050-2100) for
+the centroid of our study area. We wanted to keep the temporal
+variability from observed climate consistent, so our aim was to adjust
+observed climate according to projected changes. For temperature we
+simply subtracted the average historic temperature from the average
+projected temperature. For precipitation we calculated the proportional
+change: we subtracted average projected precipitation from average
+historic precipitation, and then divided by average historic
 precipitation. We did this for three ``Representative Concentration
 Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
-http://tntcat.iiasa.ac.at/RcpDb/.
-Basically they correspond to stabilization of radiative forcing before
-2100, after 2100, and increasing greenhouse gas emissions, respectively.
-We implemented projected temperature and precipitation changes in
-the model by perturbing observed annual values by the values in Table #.
+\url{http://tntcat.iiasa.ac.at/RcpDb/}. Basically they correspond to
+stabilization of radiative forcing before 2100, after 2100, and
+increasing greenhouse gas emissions, respectively. We implemented
+projected temperature and precipitation changes in the model by
+perturbing observed annual values by the values in Table \#.
 
@@
 
-### Additive spatio-temporal model for sagebrush cover
-We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects.
-Since the spatial and temporal effects are additive, we are not considering space-time interactions.
-Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times.
-For statistical convenience, we consider discrete values of sagebrush percent cover as counts, giving us the formulation
+\subsubsection{Additive spatio-temporal model for sagebrush
+cover}\label{additive-spatio-temporal-model-for-sagebrush-cover}
+
+We use a descriptive model for sagebrush cover that includes additive
+spatial and temporal effects. Since the spatial and temporal effects are
+additive, we are not considering space-time interactions. Our model aims
+to relate observed sagebrush cover, \(C_{s,t}\), at time \emph{t} and
+spatial location \emph{s} to an underlying process \(\mu_{s,t}\) that is
+defined over all spatial locations and observation times. For
+statistical convenience, we consider discrete values of sagebrush
+percent cover as counts, giving us the formulation
 
@@
 
-which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$).
-We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):
+which states that counts (integer percent cover) are Poisson distributed
+with a log link function on abundance intensity
+(\(\boldsymbol{\lambda}\)). We assume the underlying process,
+\(\boldsymbol{\mu}\), is a function of climate covariates (\textbf{X}),
+lagged sagebrush abundace (a temporal process), and a purely spatial
+random effect (\(\boldsymbol{\eta}\)):
 
@@
 
-Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations.
-The dimensionality is just too large.
-To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
-A common strategy is to express high dimensional effects, like $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$.
-We can then define the spatial effect as
+Fitting the spatial random effect (\(\eta\)) is computationally
+demanding when the observations come from 10s of 1,000s locations. The
+dimensionality is just too large. To overcome these computational
+constraints we employed a dimension reduction strategy to reduce the
+number of parameters that need to be estimated to account for spatial
+variation. A common strategy is to express high dimensional effects,
+like \(\boldsymbol{\eta}\), as the product of an expansion matrix,
+\textbf{K}, and a smaller parameter vector, \(\boldsymbol{\alpha}\). We
+can then define the spatial effect as
 
@@
 
-In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations.
-Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.
+In this case, \(\boldsymbol{\alpha}\) is a \(m \times 1\) vector of
+reduced spatial random effects, and \textbf{K} is a \(S \times m\)
+matrix that maps the reduced effects to the full \emph{S}-dimensional
+space, where \emph{S} is the total number of observed locations. Thus,
+we are able to reduce the effective number of parameters from
+\({S} \rightarrow m\).
 
 Obviously, the last remaining obstacle is to parameterize \textbf{K}.
-Here we use kernel convolution to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
-These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
-We use a simple Gaussian kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are
+Here we use kernel convolution to interpolate the spatial random effect
+between \emph{m} ``knots'' that are nonrandomly distributed across the
+space of our study area. These means we are modeling spatial random
+effects at the knot level, and we use \textbf{K} to interpolate those
+effects between knots. We use a simple Gaussian kernel density to define
+the distance-decay function around the knots (\textbf{w}), such that the
+entries of \textbf{K} are
 
@@
 K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
-w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma^{2}}),
+w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
 \end{align}
 
-where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
-It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$.
-However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around eight knots and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al.
-2014).
+where \(d_{s,m}\) is the Euclidean distance between the centroid of
+sample cell \emph{s} and the location of knot \emph{m}, and \(\sigma\)
+is the kernel bandwidth. Based on residual spatial structure we found
+after fitting a simple model without spatial effects, we set \(\phi\) to
+about one-third of the range of spatial dependence evident in the
+residual variograms -- 250 meters in this case. In the end, the simplest
+way to understand our dimension reduction strategy is to consider
+\(\alpha\)s as weights for each knot.
 
-### Computing and model estimation
+\subsubsection*{Computing and model
+estimation}\label{computing-and-model-estimation}
+\addcontentsline{toc}{subsubsection}{Computing and model estimation}
+
+\end{document}

### 2015-06-29 scripts/getKnots_DistanceMatrix.R
adds new knots figure

@@ -110,22 +110,49 @@ K=((2*pi*sigma)^-1)*exp(-Knot.cell.distances/(2*sigma))
 # K=dnorm(Knot.cell.distances,0,sigma)
 K=K/apply(K,1,'sum')
 K.data=list(K=K,Q.knot=Q.knot)
-save(K.data,file="Knot_cell_distances.Rdata")
+# save(K.data,file="Knot_cell_distances.Rdata")
 
 #plot Conn's K
 # load("/Users/atredenn/Desktop/STabundance-0.91/STabundance/data/Knot_cell_distances.rda")
-m <- melt(K.data$K)
-ggplot(data=m, aes(x=Var2, y=Var1))+
-  geom_raster(aes(z=value, fill=value))
+# m <- melt(K.data$K)
+# ggplot(data=m, aes(x=Var2, y=Var1))+
+#   geom_raster(aes(z=value, fill=value))
 
 
-png('SAGE_Grid_wKnots.png', width=4, height=4, units = "in", res=200)
-image(resMat, x=lon.locs,y=lat.locs,asp=TRUE,col=gray.colors.rev(100), ylab="Latitude", xlab="Longitude")
-points(Knots[Which.include,],,col='red',pch=20, add=TRUE)
+
+library(plyr)
+avgD <- ddply(fullD, .(Lon, Lat), summarise,
+              cover = mean(Cover))
+knotsD <- as.data.frame(Knots)
+
+library(RColorBrewer)
+myPalette <- colorRampPalette(brewer.pal(11, "Greens"))
+tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),
+                strip.text=element_text(face="bold"),
+                axis.title=element_text(size=14),text=element_text(size=16),
+                legend.text=element_text(size=12), legend.title=element_text(size=16))
+
+png('SAGE_Grid_wKnots.png', width=6, height=6, units = "in", res=200)
+g <- ggplot()+
+  geom_raster(data=avgD, aes(x=Lon, y=Lat, z=cover, fill=cover))+
+  geom_point(data=knotsD, aes(x,y), size=4, color="white")+
+  geom_point(data=knotsD, aes(x,y), size=3, color="black")+
+  scale_fill_gradientn(colours=myPalette(100), name="Percent \nCover")+
+  tmp.theme+
+  theme(strip.background=element_rect(fill="white"))+
[…]

### 2015-06-29 docs/ms_sageAbundance.Rmd
adds knots figure to ms

@@
 ---
+
+```{r caching, include=FALSE}
+library("methods")
+library("knitr")
+basename <- "manuscript"
+opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
+               cache.path = paste("components/cache/", basename, "/", sep=""))
+opts_chunk$set(cache = 2)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
+               comment = NA, verbose = TRUE, echo=FALSE)
+
+# PDF-based figures
+opts_chunk$set(dev='pdf')
+```
 
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$.
-However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around eight knots and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al.
+However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around 81 knots (Figure #) and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al.
 2014).
@@
 ### Computing and model estimation
+
+Figures
+-------
+```{r figure_1,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6}
+library(png)
+library(grid)
+img <- readPNG("components/figure/SAGE_Grid_wKnots.png")
+grid.raster(img)

### 2015-06-30 R/stanfit_functions.R
updates knot selection based on empirical range parameter

@@ -13,11 +13,12 @@
 #' @export
 #' @return A list with ggs object with MCMC values from rstan and 
 #'          Rhat values for each parameter.
-model_nocovars <- function(y, lag, K, cellid, iters=2000, warmup=1000, nchains=1, inits){
+model_nocovars <- function(y, lag, K, X, cellid, iters=2000, warmup=1000, nchains=1, inits){
   ##  Stan C++ model
   model_string <- "
   data{
     int<lower=0> nobs; // number of observations
+    int<lower=0> ncovs; // number of climate covariates
     int<lower=0> nknots; // number of interpolation knots
     int<lower=0> ncells; // number of cells
     int<lower=0> cellid[nobs]; // cell id
@@ -26,35 +27,43 @@ model_nocovars <- function(y, lag, K, cellid, iters=2000, warmup=1000, nchains=1
     int y[nobs]; // observation vector
     int lag[nobs]; // lag cover vector
     matrix[dK1,dK2] K; // spatial field matrix
+    matrix[nobs,ncovs] X; // spatial field matrix
   }
   parameters{
     real int_mu;
-    real beta_mu;
-    real<lower=0> sig_a;
+    real<lower=0> beta_mu;
+    real<lower=0.000001> sig_a;
+    real<lower=0.000001> sig_mu;
     vector[nknots] alpha;
+    vector[nobs] lambda;
+    vector[ncovs] beta;
   }
   transformed parameters{
     vector[ncells] eta;
-    vector[nobs] lambda;
+    vector[nobs] mu;
+    vector[nobs] climEffs;
     eta <- K*alpha;
+    climEffs <- X*beta;
     for(n in 1:nobs)
-      lambda[n] <- exp(int_mu + beta_mu*lag[n] + eta[cellid[n]]);
+    mu[n] <- int_mu + beta_mu*lag[n] + climEffs[n] + eta[cellid[n]];
   }
   model{
     // Priors
-    for(k in 1:nknots){
-      alpha[k] ~ normal(0,sig_a);
-    }
-    sig_a ~ uniform(0,10);
+    alpha ~ normal(0,sig_a);
+    sig_a ~ cauchy(0, 5);
+    sig_mu ~ cauchy(0, 5);
     int_mu ~ normal(0,100);
     beta_mu ~ normal(0,10);
+    beta ~ normal(0,10);
     // Likelihood
-    y ~ poisson(lambda);
[…]

### 2015-06-30 scripts/getKnots_DistanceMatrix.R

@@ -5,6 +5,8 @@ library(sp)
 library(rgeos)
 library(Matrix)
 library(gstat)
+library(plyr)
+library(ggplot2)
 
 
 ####
@@ -49,18 +51,48 @@ image(resMat, x=lon.locs,y=lat.locs,asp=TRUE,col=gray.colors.rev(100))
 
 
 ####
+#### Run simple Poisson GLM to get residual surface
+####
+modelD <- subset(fullD, Year>1984)
+modelD <- modelD[complete.cases(modelD),]
+outp <- glm(Cover ~ CoverLag+ppt2+ppt1+pptLag+TmeanSpr1+TmeanSpr2,
+            family = "poisson", data = modelD)
+summary(outp)
+modelD$resids <- resid(outp)
+
+
+####
 ####  Construct variograms to estimate range parameter for knots
 ####
+resD <- ddply(modelD, .(Lat, Lon), summarise,
+                  resid = mean(resids))
+lon.locs=sort(unique(resD$Lon))
+lat.locs=sort(unique(resD$Lat))
+dat <- subset(resD, select = c("Lon", "Lat", "resid"))
+coordinates(dat) <- c("Lon", "Lat")
+varMod <- variogram(resid~1, data=dat, width=50)
+
+png("avgresidual_variogram_withrange.png", width = 5, height = 5, 
+    units = "in", res=200)
+plot(varMod$dist, varMod$gamma, ylim=c(0,max(varMod$gamma)), col="dodgerblue", 
+     type="l", main="variogram of model residuals", xlab="distance (m)",
+     ylab="semivariance")
+points(varMod$dist, varMod$gamma, ylim=c(0,max(varMod$gamma)), col="dodgerblue", pch=21, bg="white")
+abline(v = 500/3, lwd=3, lty=2)
+dev.off()
+
+range_parameter <- round(500/3)
+
 # par(xaxt="s",yaxt="s",mar= c(2, 2, 4, 2), mfrow=c(4,7))
-# n.years <- length(unique(fullD$Year))
-# for(t in 1:n.years){
-#   resD <- subset(fullD, Year==years[t])
+# years <- unique(modelD$Year)
+# for(t in years){
+#   resD <- subset(modelD, Year==t)
 #   lon.locs=sort(unique(resD$Lon))
 #   lat.locs=sort(unique(resD$Lat))
[…]

### 2015-06-30 scripts/run_large_test.R

@@ -32,11 +32,6 @@ rms <- which(is.na(growD$Cover) == TRUE)
 rmsK <- which(is.na(subset(growD, Year==1985)["Cover"])==TRUE)
 K <- K[-rmsK,]
 growD <- growD[-rms, ]
-# if(length(which(is.na(growD$CoverLag)))>0){
-#   rms <- which(is.na(growD$CoverLag) == TRUE)
-#   growD <- growD[-rms, ]
-#   K <- K[-rms,]
-# }
 
 # Set up new, continuous cell IDs
 num_ids <- length(unique(growD$ID))
@@ -44,29 +39,89 @@ num_reps <- nrow(growD)/num_ids
 new_ids <- rep(c(1:num_ids), num_reps)
 growD$newID <- new_ids
 
-
-
-####
-## Quick glm for initial values
-####
-# mod <- glm(Cover ~ CoverLag, family="poisson", data=growD)
-# summary(mod)
-
-
 ####
 ##  Send data to stan function for fitting
 ####
-inits <- list()
-inits[[1]] <- list(int_mu = 2, beta_mu = 0.05, 
-                   alpha = rep(0,ncol(K)))
-inits[[2]] <- list(int_mu = 2.5, beta_mu = 0.02, 
-                   alpha = rep(0.5,ncol(K)))
-inits[[3]] <- list(int_mu = 1.5, beta_mu = 0.09, 
-                   alpha = rep(0.35,ncol(K)))
+y = growD$Cover
+lag = growD$CoverLag
+cellid = growD$newID
+X = growD[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
+X = scale(X, center = TRUE, scale = TRUE)
 
-mcmc <- model_nocovars(y = growD$Cover, lag = growD$CoverLag, K = K, 
-                       cellid = growD$newID, iters = 100, warmup = 50, 
-                       nchains = 1, inits=inits)
+inits <- list()
+inits[[1]] <- list(int_mu = 1, beta_mu = 0.05, beta = rep(0, ncol(X)),
+                   alpha = rep(0,ncol(K.data$K)), sigma=0.05, sig_a=0.05,
+                   sig_mu=0.05, lambda=rep(1, length(y)))
+inits[[2]] <- list(int_mu = 2, beta_mu = 0.01, beta = rep(0.5, ncol(X)),
+                   alpha = rep(0.5,ncol(K.data$K)), sigma=0.02, sig_a=0.005,
[…]

### 2015-07-31 docs/ms_sageAbundance.Rmd
updates ms with small sample test

@@
 
-Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species
-===========================================================================================
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
-### Andrew T. Tredennick[^corrauth], Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B. Adler
+\begin{centering}
+\textbf{\large{Remote-sensing based population models for projecting climate change impacts on plant populations at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species}}
 
-*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\textsc{\small{Andrew T.
+Tredennick\footnote{E-mail: atredenn@gmail.com}\textsuperscript{1}, Mevin B.
+Hooten\textsuperscript{2,3,4,5}, Cameron L.
+Aldridge\textsuperscript{5,6,7}, Colin G.
+Homer, David T.
+Iles\textsuperscript{1}, Andrew Kleinhesselink\textsuperscript{1}, Eric LaMalfa\textsuperscript{1}, Rebecca Mann\textsuperscript{1}, and Peter B.
+Adler\textsuperscript{1}}}
 
-*Mevin B.
-Hooten, U.S.
-Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit; Department of Fish, Wildlife, and Conservation Biology, Colorado State University; Department of Statistics, Colorado State University; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA; \textsuperscript{2}U.S.
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University; \textsuperscript{4}Department of Statistics, Colorado State University; \textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO; \textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University; \textsuperscript{7}U.S.
+Geological Survey, Fort Collins Science Center; \textsuperscript{8}U.S.
+Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD}}
 
-*Cameron L.
-Aldridge, Department of Ecosystem Science and Sustainability, Colorado State University; Natural Resource Ecology Laboratory, Colorado State University; U.S.
-Geological Survey, Fort Collins Science Center; Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO*
+\end{centering}
 
-*Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD*
+\renewcommand*{\thefootnote}{\arabic{footnote}}
 
-*David T. Iles, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-*Andrew Kleinhesselink, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-*Eric LaMalfa, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-*Rebecca Mann, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
-
-Summary
--------
-\begin{enumerate}
-\item Global environmental changes, like climate change, tend to play out at the landscape and regional scales.
+Abstract
+--------
+Global environmental changes, like climate change, tend to play out at the landscape and regional scales.
 Thus, it is reasonable that attempts to forecast the impacts of such change do so at similar spatial scales.
 For plant species, this has meant relying primarily on a single tool, species distribution models, that contain little information on population dynamics and states.
+Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using local scale data from, generally, 1 $\text{m}^{2}$ plots.
+These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.
+Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model.
+At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals.
+To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (\emph{Artemesia} spp.) cover in southwestern Wyoming.
+We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.
+The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.
 
-\item Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using “local” scale data from, generally, 1 $\text{m}^{2}$ plots.
-These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.
-
-\item Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model.
-At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals.
-To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (Artemesia spp.) cover in southwestern Wyoming.
-We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.
-
-\item The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.
-
-\item Our approach...
-\end{enumerate}
-
-*Keywords*: population model, forecasting, integral projection model, etc.
+*Key words: population model, climate change, forecasting, remote sensing, sagebrush*, Artemesia
 
@@
 -------------
-
 Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology.
@@
 ### Data: Remotely-sensed time series
-To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]).
+To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure 1).
 Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
-The regression tree model was further validated using another 500 field observations.
-For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
-To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images.
-This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011.
+The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
+To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011.
 In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush.
@@
 Our approach is based on modeling how changes in plant populations relate to climate variables.
-We a priori narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model future projections.
-In all vital rate regressions we focus on the demographic processes driving changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t}, and how these processes are mediated by climate.
-The four climate variables in our vital rate regressions are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]).
-Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database.
-Climate data were extracted for the centroid our study area.
+We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
+The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]).
+Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database. Climate data were extracted for the centroid our study area.
 
@@
 ### Additive spatio-temporal model for sagebrush cover
-We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects.
+We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects (Conn et al. 2014).
 Since the spatial and temporal effects are additive, we are not considering space-time interactions.
-Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times.
-For statistical convenience, we consider discrete values of sagebrush percent cover as counts, giving us the formulation
+We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally negative binomial:
 
 \begin{align}
-[\textbf{C}] &= \text{Poisson}(\boldsymbol{\lambda}) \\
-\boldsymbol{\lambda} &= \text{exp}(\boldsymbol{\mu}) 
+y_{i,t} \sim \text{NegBinom}(\mu_{i,t}, \zeta)
 \end{align}
 
-which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$).
-We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):
+where $\mu_{i,t}$ is the determinstic expected population growth:
 
 \begin{align}
-\mu_{s,t} = \underbrace{\beta_{0} + \beta_{1}\mu_{s,t-1}}_\text{temporal} + \underbrace{\mathbf{X}_{t}\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{s}}_\text{spatial} + \underbrace{\varepsilon_{s,t}}_\text{error}
+\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}
 \end{align}
 
+Our model of population growth includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
 Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations.
@@
 \begin{align}
-\boldsymbol{\eta} = \textbf{K}\boldsymbol{\alpha}.
+\boldsymbol{\eta} &= \textbf{K}\boldsymbol{\alpha} \\
+\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2)
 \end{align}
@@
 Here we use kernel convolution to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
-These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
-We use a simple Gaussian kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are
+This means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
+We use an exponential kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are
 
 \begin{align}
-K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
-w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma^{2}}),
+K_{s,m} &= w_{s,m} / \sum_{s} w_{s,m}, \text{where} \\
+w_{s,m} &= (2 \pi \sigma)^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma})
 \end{align}
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$.
-However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around 81 knots (Figure #) and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al.
-2014).
+However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around 271 knots (Figure 2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix A).
 
-### Computing and model estimation
+The full model can be expressed as:
+
+\begin{align}
+[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2, \zeta] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}, \zeta] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\gamma][\sigma^2][\sigma_{\eta}^2]
+\end{align}
+
+### Model estimation and model projections
+We fit the model using the 'No-U-Turn' MCMC sampler in Stan.
+We obtained posterior distributions of all model parmaters from 1,000 MCMC iterations after discarding an initial 1,000 iterations as burn in.
+We assessed convergence visually and ensured all scale reduction factors were less than 1.01 (Gelman citation).
+
+With the posterior estimates of model parameters in hand, we were able to simulate a matrix of pixels equal to the size of the study area.
+To do so, we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
+We ran the simulation for 2,500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
+We then ran simulations where the projected climate changes in Table 1 were applied to the observed climate time series.
+
+Results
+-------
+Our pixel-based population model was able to reproduce observed percent cover (Figure 3).
+Three of five climate covariates were strong predictors of cover change (Figure 4).
+In particular, mean spring temperature and precipitation at time *t*-2 had strong positive effects on population growth, only slightly offset by a negative effect of precipitation at time  *t*-1.
+The positive effects of precipitation and temperature resulted in projected increases of cover for all climate change scenarios (Figure 5) since precipitation and temperature are both predicted to increase (Table 1).
+
+Discussion
+-----------
+
+\pagebreak{}
 
@@
 -------
-```{r figure_1,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6}
+```{r figure_1,dependson="plot-options", fig.cap="Location of the study area in southwester Wyoming and a snapshot of the percent cover data in 1984 for each site (a and b).
+Green shading on the map shows projected distribution of sagebrush from Kuchler (19xx).
+Latitude and longitude are in decimal degrees.", fig.height=6, eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("components/figure/StudyAreasMap_1and2.png")
+grid.raster(img)
+```
+
+```{r figure_2,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6, eval=TRUE}
 library(png)
@@
 grid.raster(img)
+```
+
+```{r figure_3,dependson="plot-options", fig.cap="Observed and predicted cover, averaged over years for each pixel.
+This histogram shows the distribution of observed cover and then density curve shows the distribution of predicted cover from the model.", fig.height=6, eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("../results/obs_pred_spatial_small.png")
+grid.raster(img)
+```
+
+```{r figure_4,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.", fig.height=6, eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("../results/climeff_dens_small.png")
+grid.raster(img)
+```
+
+```{r figure_5,dependson="plot-options", fig.cap="Projected equilibrium cover under climate change scenarios.
+Top panels show the projections in space, averaged over time for each pixel.
+Lower panel shows distributions of cover under each scenario", fig.height=6, eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("../results/climchange_small.png")
+grid.raster(img)
+```

### 2015-07-31 scripts/run_small_test.R

@@ -125,7 +125,7 @@ outs <- ggs(fit)
 # ggs_traceplot(outs, "alpha")
 # ggs_traceplot(outs, "beta")
 saveRDS(outs, file = "../results/small_test_mcmc.RDS")
-
+outs <- readRDS("../results/small_test_mcmc.RDS")
 
 ## Look at spatial field
 alpha <- outs[grep("alpha", outs$Parameter),]
@@ -213,7 +213,7 @@ g1 <- ggplot(sp.equil2, aes(x=Lon, y=Lat))+
 g2 <- ggplot(sp.equil)+
   geom_histogram(aes(x=Observed, y = ..density..), col="white", fill="grey45", binwidth=1)+
   geom_line(stat="density", aes(x=Predicted, y= ..density..), 
-            col="black", size=1.5, adjust=15, linetype=2)+
+            col="black", size=1.5, linetype=2)+
   xlab("Percent Cover")+
   ylab("Estimated Kernel Density")+
   theme_bw()
@@ -226,11 +226,13 @@ dev.off()
 betas <- outs[grep("beta", outs$Parameter),]
 betas <- as.data.frame(subset(betas, Parameter!="beta_mu"))
 beta.names <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
+cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
+               "#CC79A7", "#0072B2", "#D55E00", "#CC79A7")
 
 png("../results/climeff_dens_small.png", width = 6, height=3.5, units = "in", res=200)
 ggplot(betas)+
   geom_line(stat="density", aes(x=value, y=..density..,color=Parameter), size=1.5)+
-  scale_color_discrete(labels=beta.names, name="Coefficient")+
+  scale_color_manual(labels=beta.names, name="Coefficient", values=cbPalette[1:5])+
   geom_vline(xintercept=0, linetype=2)+
   xlab("Standardized Coefficient Value")+
   ylab("Posterior Density")+
@@ -251,7 +253,13 @@ mean_params <- ddply(outs, .(Parameter), summarise,
[…]

### 2015-07-31 docs/ms_sageAbundance.Rmd
fixes typo

@@
 
-\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA; \textsuperscript{2}U.S.
-Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University; \textsuperscript{4}Department of Statistics, Colorado State University; \textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO; \textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University; \textsuperscript{7}U.S.
-Geological Survey, Fort Collins Science Center; \textsuperscript{8}U.S.
-Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD}}
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}} \\*
+\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit}} \\*
+\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University}} \\*
+\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University}} \\*
+\textit{\small{\textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO}} \\*
+\textit{\small{\textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University}} \\*
+\textit{\small{\textsuperscript{7}U.S. Geological Survey, Fort Collins Science Center}} \\*
+\textit{\small{\textsuperscript{8}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD}}
 

### 2015-08-01 scripts/run_small_test.R
adds varying parameter population sim

@@ -370,3 +370,58 @@ g2 <- ggplot(all4plot)+
   theme_bw()
 gout <- grid.arrange(g1, g2, ncol=1, nrow=2)
 dev.off()
+
+
+
+###################################################
+###################################################
+###################################################
+####
+####  Run population model with parameter uncertainty
+####
+time.steps <- 100
+pixels <- nrow(subset(growD, Year==1985))
+clim_sim <- climD[climD$year %in% unique(growD$Year),]
+X_sim = clim_sim[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
+X_sim = scale(X_sim, center = TRUE, scale = TRUE)
+
+alphasd <- outs[grep("alpha", outs$Parameter),]
+climeffs <- c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]")
+nchains <- 3
+niters <- length(unique(outs$Iteration))
+totsims <- nchains*niters
+ex.arr <- array(NA, dim=c(totsims, time.steps, pixels))
+ex.arr[,1,] <- 1
+pb <- txtProgressBar(min=1, max=totsims, char="+", style=3, width=65)
+counter <- 1
+for(i in 1:nchains){
+  for(j in 1:niters){
+    chains <- i
+    iter <- j
+    int_mu <- as.numeric(subset(outs, Chain==chain & Iteration==iter & Parameter=="int_mu")[,"value"])
+    beta_mu <- as.numeric(subset(outs, Chain==chain & Iteration==iter & Parameter=="beta_mu")[,"value"])
+    betas <- subset(outs, Chain==chain & Iteration==iter & Parameter%in%climeffs)[,"value"]
+    alphas <- subset(alphasd, Chain==chain & Iteration==iter)[,"value"]
+    sizeparam <- as.numeric(subset(outs, Chain==chain & Iteration==iter & Parameter=="phi")[,"value"])
+    eta <- K%*%as.numeric(unlist(alphas))
+    for(t in 2:time.steps){
+      Xtmp <- X_sim[sample(c(1:nrow(X_sim)), 1),]
+      tmp.mu <- exp(int_mu + beta_mu*ex.arr[counter,t-1,]) + sum(betas*Xtmp)
+      tmp.mu <- tmp.mu + eta
[…]

### 2015-08-01 scripts/run_small_test.R
adds parameter vary runs and histogram figure

@@ -425,3 +425,9 @@ ggplot()+
   xlab("Cover (%)")+
   ylab("Density")+
   theme_bw()
+
+
+####
+####  RCP 4.5; parameter vary
+####
+

### 2015-08-04 docs/ms_sageAbundance.Rmd
rms old .Rnw ms file; updates ms with citations from Mevin

@@
    - \doublespacing
-bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
+bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
 csl: components/ecology.csl
@@
 \begin{centering}
-\textbf{\large{Remote-sensing based population models for projecting climate change impacts on plant populations at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species}}
+\textbf{\large{A spatiotemporal modeling framework for forecasting climate change impacts on plant populations over large spatial extents}}
 
@@
 Aldridge\textsuperscript{5,6,7}, Colin G.
-Homer, David T.
+Homer\textsuperscript{8}, David T.
 Iles\textsuperscript{1}, Andrew Kleinhesselink\textsuperscript{1}, Eric LaMalfa\textsuperscript{1}, Rebecca Mann\textsuperscript{1}, and Peter B.
@@
 
-*Key words: population model, climate change, forecasting, remote sensing, sagebrush*, Artemesia
+*Key words: population model, climate change, forecasting, spatiotemporal model, remote sensing, sagebrush*, Artemesia, *dimension reduction*
 
@@
 ### Additive spatio-temporal model for sagebrush cover
-We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects (Conn et al. 2014).
-Since the spatial and temporal effects are additive, we are not considering space-time interactions.
-We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally negative binomial:
+We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects (Conn et al. 2015).
+We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally negative binomial
 
 \begin{align}
-y_{i,t} \sim \text{NegBinom}(\mu_{i,t}, \zeta)
+y_{i,t} \sim \text{NegBinom}(\mu_{i,t}, \zeta),
 \end{align}
 
-where $\mu_{i,t}$ is the determinstic expected population growth:
+where $\mu_{i,t}$ is the expected population growth
 
 \begin{align}
-\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}
+\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
 \end{align}
@@
 Our model of population growth includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
-Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations.
-The dimensionality is just too large.
-To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
-A common strategy is to express high dimensional effects, like $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$.
+Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
+This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations (Wikle 2010).
+To overcome these computational constraints, we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
+Reducing the number of effective spatial random effect parameters speeds up estimation by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
+Fitting models that appropriately account for spatial autocorrelation over large spatial extents would be virtually impossible without these modern techniques.
+We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [@Hooten2007a; @Conn2015].
 We can then define the spatial effect as
@@
 \begin{align}
-\boldsymbol{\eta} &= \textbf{K}\boldsymbol{\alpha} \\
-\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2)
+\boldsymbol{\eta} &= \textbf{K}\boldsymbol{\alpha}, \\
+\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2).
 \end{align}
@@
 In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations.
-Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.
+Thus, we are able to reduce the effective number of parameters from ${S}$ to $m$.
 
-Obviously, the last remaining obstacle is to parameterize \textbf{K}.
-Here we use kernel convolution to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
+The last remaining obstacle is to parameterize \textbf{K}.
+Here we use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
 This means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
@@
 \begin{align}
-K_{s,m} &= w_{s,m} / \sum_{s} w_{s,m}, \text{where} \\
-w_{s,m} &= (2 \pi \sigma)^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma})
+K_{s,m} &= w_{s,m} / \sum_{s} w_{s,m}
+\end{align}
+where
+\begin{align}
+w_{s,m} &= (2 \pi \sigma)^{-1} \text{exp}\left(\frac{-d_{s,m}}{2 \sigma}\right).
 \end{align}
@@
 where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
-It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$.
-However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around 271 knots (Figure 2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix A).
+It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2014].
+However, given the relative largeness of our dataset and computational limitations, we defined kernels around 271 knots (Figure 2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix A).
 
-The full model can be expressed as:
+The Bayesian posterior distribution can be expressed as
 
 \begin{align}
-[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2, \zeta] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}, \zeta] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\gamma][\sigma^2][\sigma_{\eta}^2]
+[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2, \zeta | \textbf{y}] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}, \zeta] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\gamma][\zeta][\sigma_{\eta}^2].
 \end{align}
@@
 ### Model estimation and model projections
-We fit the model using the 'No-U-Turn' MCMC sampler in Stan.
-We obtained posterior distributions of all model parmaters from 1,000 MCMC iterations after discarding an initial 1,000 iterations as burn in.
-We assessed convergence visually and ensured all scale reduction factors were less than 1.01 (Gelman citation).
+We fit the model in R [@R2013] using the 'No-U-Turn' MCMC sampler in Stan [@stan2014] and the RStan package [@rstan2014].
+We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1 000 iterations each, after discarding an initial 1 000 iterations as burn in.
+Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
+However, the Stan sampler is extremely efficient, meaning that compared to other samplers fewer iterations are required to achieve convergence.
+We assessed convergence visually and ensured all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
 
-With the posterior estimates of model parameters in hand, we were able to simulate a matrix of pixels equal to the size of the study area.
+Using the posterior estimates of model parameters, we were able to simulate a matrix of pixels equal to the size of the study area.
 To do so, we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
-We ran the simulation for 2,500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
+We ran the simulation for 2 500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
 We then ran simulations where the projected climate changes in Table 1 were applied to the observed climate time series.
@@
 Green shading on the map shows projected distribution of sagebrush from Kuchler (19xx).
-Latitude and longitude are in decimal degrees.", fig.height=6, eval=TRUE}
+Latitude and longitude are in decimal degrees.", fig.height=6, eval=FALSE}
 library(png)
@@
 
-```{r figure_2,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6, eval=TRUE}
+```{r figure_2,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6, eval=FALSE}
 library(png)
@@
 ```{r figure_3,dependson="plot-options", fig.cap="Observed and predicted cover, averaged over years for each pixel.
-This histogram shows the distribution of observed cover and then density curve shows the distribution of predicted cover from the model.", fig.height=6, eval=TRUE}
+This histogram shows the distribution of observed cover and then density curve shows the distribution of predicted cover from the model.", fig.height=6, eval=FALSE}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.", fig.height=6, eval=TRUE}
+```{r figure_4,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.", fig.height=6, eval=FALSE}
 library(png)
@@
 Top panels show the projections in space, averaged over time for each pixel.
-Lower panel shows distributions of cover under each scenario", fig.height=6, eval=TRUE}
+Lower panel shows distributions of cover under each scenario.", fig.height=6, eval=FALSE}
 library(png)
@@
 ```
+
+\pagebreak{}
+
+References
+----------

### 2015-08-04 docs/ms_sageAbundance.Rmd
updates ms with citations

@@
 \renewcommand*{\thefootnote}{\arabic{footnote}}
+\setcounter{footnote}{0}
 
@@
 --------
-Global environmental changes, like climate change, tend to play out at the landscape and regional scales.
-Thus, it is reasonable that attempts to forecast the impacts of such change do so at similar spatial scales.
-For plant species, this has meant relying primarily on a single tool, species distribution models, that contain little information on population dynamics and states.
-Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using local scale data from, generally, 1 $\text{m}^{2}$ plots.
-These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.
-Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model.
-At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals.
-To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (\emph{Artemesia} spp.) cover in southwestern Wyoming.
-We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.
-The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.
+Population modeling inference has traditionally been limited to small scale observational studies.
+Yet, global environmental changes, like climate change, tend to play out at the landscape and regional scales, creating a mismatch between current tools and current problems.
+New technology is allowing for data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information.
+For example, plant ecologists now have access to species-specific estimates of plant percent cover through time at spatial extents ranging from landscapes to continents.
+The real strength to be gained using this new form of ecological information will be to move beyond simple descriptive summaries of the data, and couple them with sophisticated model-based statistical approaches that can accommodate more realistic ecological mechanisms.
+A major hurdle in properly analyzing these types of data is in building and fitting statistical models that have valid assumptions.
+Rigorously accounting for spatial autocorrelation has been a crux in former analyses at large scales, but we can now fit appropriate models using modern, computationally efficient model reparameterizations and dimension reduction techniques.
+
+Here we advance recent work on spatiotemporal modeling of animal abundance by applying those methods to remotely-sensed estimates of plant percent cover at the landscape scale.
+We use a 27 year time series of sagebrush (*Artemesia* spp.) percent cover over a $5\times5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
+We then use the model to forecast future population states (percent cover) under projected climate scenarios.
+Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to spring temperature.
+However, gains in cover are not homogeneous in space; we forecast some areas will increase in cover while others will decrease.
+Such spatially explicit estimates of increases/decreases in plant abundance have, until now, been elusive.
+The spatiotemporal model we describe can be applied to any spatially explicit time series of plant cover.
+As such datasets become increasingly available and precise, we expect the methods we use here to become an appealing alternative to species distribution models that lack population dynamics and, generally, estimates of species abundance.
 
@@
 -------------
-Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology.
-The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions.
-However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale.
-Local-scale demographic data make building population projection models an easy task (Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions.
+Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
+The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].
+However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [@Salguero-Gomez2015].
+Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
 Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.
 
-The fundamental limitation of local-scale data is that they are insufficient for estimating the full extent of variation in model parameters possible at larger spatial and temporal extents (Freckleton et al., 2011; Queenborough et al., 2011).
-It is possible to estimate the variance of population model parameters related to spatial and temporal variability, but the observations used to estimate parameter uncertainty tend to be clustered in space or time.
-Thus, extrapolating beyond the spatial or temporal resolution of the data used to parameterize models can lead to biased and wrong forecasts of future population states.
-
-Alternatively, large-scale trends in populations are easily detected using widely available monitoring data, but such data are rarely used to project future population states. 
-Given the tradeoffs between the spatial-scale of data and the inference that can be drawn from the data, we suggest combining the best features of local-scale population modeling and large-scale monitoring data.
-Specifically, we propose a new approach for modeling population dynamics at large spatial extents based on the theory and mechanics of individual-based models (IBMs) as applied to remotely-sensed time series data.
-
-The approach we describe in this paper flows naturally from recent work on density-structured population models.
-Density-structured models rely on modeling the transitions of a population among discrete states rather than the traditional approach of modeling the transitions of individuals.
-Relatively few ecologists have taken up the density-structured mantle first proposed by Taylor, C.
-M.
-et al.
-(2004), but recently Freckleton et al.
-(2011) and Queenborough et al.
-(2011) have higlighted the potential value the approach.
+Alternatives to local-scale demographic data are rapidly becoming available.
+Forest Inventory and Analysis (FIA) data have been successfully used to fit sophisticated species distribution models (SDMs) that predict occurance and abundance at large spatial extents [@Clark2014].
+However, all SDMs rely on correlations in space, which have been shown to perform poorly relative to models based on responses to changes in climate over time [@Clark2011].
+Likewise, distributed demographic data, like FIA data, are not continuous in space.
+So creating continuous prediction surfaces requires extrapolating predictions to areas not covered by the data.
 
@@
 ---------------------
-### Data: Remotely-sensed time series
-To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure 1).
-Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
+### Data
+#### Remotely-sensed time series
+To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012] (Figure 1).
+Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1 780 field observations of sagebrush cover distributed across Wyoming.
 The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
@@
 In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush.
-For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27,948 pixels each.
-Thus, each full dataset contains 754,596 observations (27 years $\times$ 27,948 pixels).
+For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27 948 pixels each.
+Thus, each full dataset contains 754 596 observations (27 years $\times$ 27 948 pixels).
 
-### Data: Climate covariates
+#### Climate covariates
 Our approach is based on modeling how changes in plant populations relate to climate variables.
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
-The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]).
-Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database. Climate data were extracted for the centroid our study area.
+The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table #).
+We downloaded daily historic weather data from the parameter-elevation regressions on independent slopes model (PRISM) data set (*available online*)[^link].
+The PRISM weather data is interpolated between coarse observation units and does capture some spatial variation.
+However, instead of further downscaling the 800 meter resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
+We calcualted the four climate variables outline above from the PRISM data for the time period of our remotely sensed data (1984 to 2011).
 
-Climate data used to inform model projections of future sagebrush
-cover came from average outputs of all models used in the most recent
-IPCC report, the Coupled Model Intercomparison Project 5 (CMIP5; http://cmip-pcmdi.llnl.gov/cmip5/).
-We downloaded model estimates of historic climate (1950-2000) and
-projected climate (2050-2100) for the centroid of our study area.
-We wanted to keep the temporal variability from observed climate consistent,
-so our aim was to adjust observed climate according to projected changes.
-For temperature we simply subtracted the average historic temperature
-from the average projected temperature. For precipitation we calculated
-the proportional change: we subtracted average projected precipitation
-from average historic precipitation, and then divided by average historic
-precipitation. We did this for three ``Representative Concentration
-Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
-http://tntcat.iiasa.ac.at/RcpDb/.
-Basically they correspond to stabilization of radiative forcing before
-2100, after 2100, and increasing greenhouse gas emissions, respectively.
-We implemented projected temperature and precipitation changes in
-the model by perturbing observed annual values by the values in Table #.
+[^link]: http://www.prism.oregonstate.edu/
+
+To forecast future population states, we required future projections of climate for our study area.
+We downloaded the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
+The CMIP5 provides projections from a suite of global circulations models.
+For our purposes, we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
+We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
+To keep the temporal variability from observed climate consistent, for climate change scenarios we adjusted observed climate according to projected changes.
+For temperature we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
+For precipitation we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation.
+We did this for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
+We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table #.
+
+[^link2]: http://cmip-pcmdi.llnl.gov/cmip5/
+[^link3]: http://tntcat.iiasa.ac.at/RcpDb/
 
@@
 Our model of population growth includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
+Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
+
 Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
@@
 
-### Model estimation and model projections
+### Computing
+####Fitting the model
 We fit the model in R [@R2013] using the 'No-U-Turn' MCMC sampler in Stan [@stan2014] and the RStan package [@rstan2014].
@@
 
+####Simulating the model
 Using the posterior estimates of model parameters, we were able to simulate a matrix of pixels equal to the size of the study area.

### 2015-08-04 scripts/run_small_test.R

@@ -447,7 +447,7 @@ for(i in 1:nchains){
     eta <- K%*%as.numeric(unlist(alphas))
     for(t in 2:time.steps){
       Xtmp <- X_sim[sample(c(1:nrow(X_sim)), 1),]
-      tmp.mu <- exp(int_mu + beta_mu*ex.arr[counter,t-1,]) + sum(betas*Xtmp)
+      tmp.mu <- exp(int_mu + beta_mu*ex.arr[counter,t-1,] + sum(betas*Xtmp))
       tmp.mu <- tmp.mu + eta
       ex.arr[counter,t,] <- rnbinom(pixels, mu=tmp.mu, size = sizeparam)
     }
@@ -460,12 +460,14 @@ y <- hist(growD$Cover, freq=FALSE, breaks=80)
 yhat <- hist(ex.arr, freq=FALSE)
 hist_ydata <- data.frame(x = y$breaks, y=c(y$density,0))
 hist_yhatdata <- data.frame(x = yhat$breaks, y=c(yhat$density,0))
+pdf("~/Desktop/test_hist.pdf", width = 4, height = 4)
 ggplot()+
   geom_step(data=hist_ydata, aes(x=x, y=y), size=1.5)+
   geom_step(data=hist_yhatdata, aes(x=x, y=y), col="darkorange", size=1)+
   xlab("Cover (%)")+
   ylab("Density")+
   theme_bw()
+dev.off()
 
 
 ####

### 2015-08-08 docs/ms_sageAbundance.Rmd
fixes model description as percent cover change

@@
 
-where $\mu_{i,t}$ is the expected population growth
+where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*
 
@@
 
-Our model of population growth includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
+Our model of percent cover change includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
+We chose a negative binomial likelihood because integer percent cover is a form of count data, and the negative binomial distribution allows for slightly more dispersion than the poisson.
 
@@
 Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
-However, the Stan sampler is extremely efficient, meaning that compared to other samplers fewer iterations are required to achieve convergence.
+However, the Stan sampler is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
 We assessed convergence visually and ensured all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
@@
 ####Simulating the model
-Using the posterior estimates of model parameters, we were able to simulate a matrix of pixels equal to the size of the study area.
-To do so, we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
+Using the posterior distribution of model parameters, we simulate a matrix of pixels equal to the size of the study area (XX XXX pixels or matrix elements).
+We initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
 We ran the simulation for 2 500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.

### 2015-09-08 docs/ms_sageAbundance.Rmd
updates ms with prelim text

@@
 
-Here we advance recent work on spatiotemporal modeling of animal abundance by applying those methods to remotely-sensed estimates of plant percent cover at the landscape scale.
+Here we advance spatiotemporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
 We use a 27 year time series of sagebrush (*Artemesia* spp.) percent cover over a $5\times5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
@@
 \begin{align}
-w_{s,m} &= (2 \pi \sigma)^{-1} \text{exp}\left(\frac{-d_{s,m}}{2 \sigma}\right).
+w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right).
 \end{align}
@@
 However, the Stan sampler is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
-We assessed convergence visually and ensured all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
+We assessed convergence visually and made sure all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
 

### 2015-09-08 devel/stan_chuncktest.R
updates get_mcmcm function to avoid dplyr operations

@@ -5,7 +5,6 @@ rm(list=ls(all=TRUE))
 
 # Load libraries
 library(rstan)
-library(ggmcmc)
 
 ####
 ####  Simulate some data
@@ -52,13 +51,11 @@ get_mcmc <- function(S){
   long <- NULL
   sdf <- as.data.frame(S@sim$samples[[1]])
   sdf$Iteration <- 1:dim(sdf)[1]
-  s <- gather(sdf, Parameter, value, -Iteration) %>%
-    mutate(Chain = 1) %>%
-    dplyr::select(Iteration, Chain, Parameter, value)
-  long <- rbind_list(long, s)
+  s <- melt(sdf, id.vars = "Iteration")
+  colnames(s) <- c("Iteration", "Parameter", "value")
+  long <- rbind(long, s)
   return(long)
 }
-
 
 
 ####

### 2015-09-08 scripts/run_subset_4HPC.R

@@ -100,10 +100,9 @@ get_mcmc <- function(S){
   long <- NULL
   sdf <- as.data.frame(S@sim$samples[[1]])
   sdf$Iteration <- 1:dim(sdf)[1]
-  s <- gather(sdf, Parameter, value, -Iteration) %>%
-    mutate(Chain = 1) %>%
-    dplyr::select(Iteration, Chain, Parameter, value)
-  long <- rbind_list(long, s)
+  s <- melt(sdf, id.vars = "Iteration")
+  colnames(s) <- c("Iteration", "Parameter", "value")
+  long <- rbind(long, s)
   return(long)
 }
 

### 2015-09-08 scripts/run_subset_4HPC.R
makes rstan and reshape2 only required packages

@@ -16,7 +16,7 @@ knotpath <- ""
 ####  Load required libraries
 ####
 library(rstan)
-library(ggmcmc)
+library(reshape2)
 
 
 ####

### 2015-09-10 scripts/run_subset.R
small changes to subset run code

@@ -16,7 +16,7 @@ knotpath <- "../results/"
 ####  Load required libraries
 ####
 library(rstan)
-library(ggmcmc)
+library(reshape2)
 
 
 ####
@@ -26,7 +26,7 @@ args <- commandArgs(trailingOnly = F)
 myargument <- args[length(args)]
 myargument <- sub("-","",myargument)
 chain_id <- as.numeric(myargument)
-
+chain_id <- 1
 
 ####
 ####  Get data
@@ -100,12 +100,12 @@ get_mcmc <- function(S){
   long <- NULL
   sdf <- as.data.frame(S@sim$samples[[1]])
   sdf$Iteration <- 1:dim(sdf)[1]
-  s <- gather(sdf, Parameter, value, -Iteration) %>%
-    mutate(Chain = 1) %>%
-    dplyr::select(Iteration, Chain, Parameter, value)
-  long <- rbind_list(long, s)
+  s <- melt(sdf, id.vars = "Iteration")
+  colnames(s) <- c("Iteration", "Parameter", "value")
+  long <- rbind(long, s)
   return(long)
 }
+
 
 
 ####
@@ -145,9 +145,9 @@ pars <- c("int_mu", "beta_mu",  "alpha", "beta", "phi", "sig_a")
 mcmc_config <- stan(model_code=model_string, data=datalist,
                     pars=pars, chains=0)
 mcmc1 <- stan(fit=mcmc_config, data=datalist, pars=pars, chains=1, 
-              iter = 200, warmup = 100, init=list(inits[[chain_id]]))
+              iter = 20, warmup = 10, init=list(inits[[chain_id]]))
 long <- get_mcmc(mcmc1)
-lastones <- subset(long, Iteration==200)
+lastones <- subset(long, Iteration==20)
 lastmcmc <- mcmc1
 saveRDS(long, paste0("iterchunk1_chain", chain_id, ".RDS"))
 

### 2015-09-10 devel/stan_poisson.R
adds a test stan script for recovering poission regression parameters

@@ -1,0 +1,63 @@
+##  Script to test STAN poisson model
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(rstan)
+library(ggmcmc)
+
+####
+####  Look at difference between poisson and neg binomial
+####
+plot(density(rpois(10000, 10), adjust=2), col="red")
+lines(density(rnbinom(10000, size = 5, mu = 10), adjust=2))
+
+####
+####  Simulate some data
+####
+alpha <- 1
+beta <- 0.1
+x <- rpois(100, 10)
+y <- rpois(100, lambda = exp(alpha+beta*x))
+plot(x, y)
+abline(0,1)
+
+####
+####  Write STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int y[nobs]; // observation vector
+  int x[nobs]; // covariate vector
+}
+parameters{
+  real alpha;
+  real beta;
+}
+transformed parameters{
+  vector[nobs] mu;
+  for(n in 1:nobs)
+    mu[n] <- alpha + beta*x[n];
+}
+model{
+  // Priors
+  alpha ~ normal(0,1000);
+  beta ~ normal(0,1000);
+  // Likelihood
+  y ~ poisson_log(mu);
+}
+"
+
+####
+####  Fit STAN model in chunks of 100 iterations
+####
+datalist <- list(y=y, x=x, nobs=length(y))
+pars <- c("alpha", "beta")
+
+# Compile the model
+mcmc_config <- stan(model_code=model_string, data=datalist, pars=pars, chains=0)
+mcmc1 <- stan(fit=mcmc_config, data=datalist, pars=pars, chains=1, iter = 2000, warmup = 1000)
+traceplot(mcmc1)
+print(mcmc1)

### 2015-09-10 scripts/run_subset_poisson.R
adds poisson versions of the model

@@ -1,0 +1,168 @@
+##  Script to run GLMM fit on subset of sagebrush data
+##  Uses STAN to fit the model
+##
+##  Author:      Andrew Tredennick
+##  Email:       atredenn@gmail.com
+##  Last Update: 9-2-2015
+
+
+####
+####  Set some file paths, etc.
+####
+datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
+knotpath <- "../results/"
+
+####
+####  Load required libraries
+####
+library(rstan)
+library(reshape2)
+
+
+####
+####  Set chain ID from command line prompt
+####
+args <- commandArgs(trailingOnly = F)
+myargument <- args[length(args)]
+myargument <- sub("-","",myargument)
+chain_id <- as.numeric(myargument)
+chain_id <- 1
+
+####
+####  Get data
+####
+fullD <- read.csv(paste0(datapath,"wy_sagecover_subset_noNA.csv"))
+if(length(which(is.na(fullD$Cover))) > 0) stop("data contains NA values")
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load(paste0(knotpath,"Knot_cell_distances_subset.Rdata"))
+K <- K.data$K
+
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  int y[nobs]; // observation vector
+  int lag[nobs]; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
+  real<lower=0> beta_mu;
+  real<lower=0.0001> sig_a;
+  vector[nknots] alpha;
[…]

### 2015-09-10 scripts/run_subset_poisson_4HPC.R

@@ -1,0 +1,165 @@
+##  Script to run GLMM fit on subset of sagebrush data
+##  Uses STAN to fit the model
+##
+##  Author:      Andrew Tredennick
+##  Email:       atredenn@gmail.com
+##  Last Update: 9-2-2015
+
+
+####
+####  Set some file paths, etc.
+####
+datapath <- ""
+knotpath <- ""
+
+####
+####  Load required libraries
+####
+library(rstan)
+library(reshape2)
+
+
+####
+####  Set chain ID from command line prompt
+####
+args <- commandArgs(trailingOnly = F)
+myargument <- args[length(args)]
+myargument <- sub("-","",myargument)
+chain_id <- as.numeric(myargument)
+
+
+####
+####  Get data
+####
+fullD <- read.csv(paste0(datapath,"wy_sagecover_subset_noNA.csv"))
+if(length(which(is.na(fullD$Cover))) > 0) stop("data contains NA values")
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load(paste0(knotpath,"Knot_cell_distances_subset.Rdata"))
+K <- K.data$K
+
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  int y[nobs]; // observation vector
+  int lag[nobs]; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
+  real<lower=0> beta_mu;
+  real<lower=0.0001> sig_a;
+  vector[nknots] alpha;
+  vector[ncovs] beta;
+}
+transformed parameters{
[…]

### 2015-09-10 docs/ms_sageAbundance.Rmd
updates ms with new poisson model version

@@
 \textsc{\small{Andrew T.
-Tredennick\footnote{E-mail: atredenn@gmail.com}\textsuperscript{1}, Mevin B.
+Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B.
 Hooten\textsuperscript{2,3,4,5}, Cameron L.
@@
 
-\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}} \\*
-\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit}} \\*
-\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University}} \\*
-\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University}} \\*
-\textit{\small{\textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO}} \\*
-\textit{\small{\textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University}} \\*
-\textit{\small{\textsuperscript{7}U.S. Geological Survey, Fort Collins Science Center}} \\*
-\textit{\small{\textsuperscript{8}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD}}
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\*
+\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\*
+\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\*
+\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}} \\*
+\textit{\small{\textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80523 USA}} \\*
+\textit{\small{\textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University, Fort Collins, CO 80523 USA}} \\*
+\textit{\small{\textsuperscript{7}U.S. Geological Survey, Fort Collins Science Center, Fort Collins, CO 80526 USA}} \\*
+\textit{\small{\textsuperscript{8}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD 57198 USA}}
 
@@
 We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects (Conn et al. 2015).
-We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally negative binomial
+We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally Poisson
 
 \begin{align}
-y_{i,t} \sim \text{NegBinom}(\mu_{i,t}, \zeta),
+y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
 \end{align}
@@
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
-We chose a negative binomial likelihood because integer percent cover is a form of count data, and the negative binomial distribution allows for slightly more dispersion than the poisson.
+We chose to use a Poisson likelihood because integer percent cover values can be considered a form of count data.
+We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$).
 
@@
 \begin{align}
-[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2, \zeta | \textbf{y}] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}, \zeta] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\gamma][\zeta][\sigma_{\eta}^2].
+[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\eta}^2].
 \end{align}

### 2015-09-11 scripts/traceplots.R
adds traceplot script for mcmc chains

@@ -1,0 +1,33 @@
+
+setwd("/Users/atredenn/Desktop/")
+longchain1 <- readRDS("POIS_iterchunk1_chain2.RDS")
+for(i in 2:3){
+  longchain1 <- rbind(longchain1, readRDS(paste0("POIS_iterchunk",i,"_chain2.RDS")))
+}
+
+longchain2 <- readRDS("POIS_iterchunk1_chain3.RDS")
+for(i in 2:3){
+  longchain2 <- rbind(longchain2, readRDS(paste0("POIS_iterchunk",i,"_chain3.RDS")))
+}
+
+longchain3 <- readRDS("POIS_iterchunk1_chain1.RDS")
+for(i in 2:3){
+  longchain3 <- rbind(longchain3, readRDS(paste0("POIS_iterchunk",i,"_chain1.RDS")))
+}
+
+
+all.params <- unique(longchain1$Parameter)
+iters <- c(100:600)
+iters <- iters[-which(iters %in% c(201:205,401:405))]
+
+pdf("test.pdf", width = 8.5, height = 11)
+par(mfrow=c(6,3))
+for(plot.param in all.params){
+  plot(unlist(longchain1[which(longchain1$Parameter==plot.param),"value"])[iters], 
+       type="l", ylab="value", main=plot.param, xlab="iteration")
+  lines(unlist(longchain2[which(longchain2$Parameter==plot.param),"value"])[iters], col="darkorange")
+  lines(unlist(longchain3[which(longchain3$Parameter==plot.param),"value"])[iters], col="purple")
+}
+dev.off()
+
+

### 2015-09-11 scripts/traceplots.R
updates iter max for traceplots

@@ -1,24 +1,25 @@
 
 setwd("/Users/atredenn/Desktop/")
+maxiter <- 5
 longchain1 <- readRDS("POIS_iterchunk1_chain2.RDS")
-for(i in 2:3){
+for(i in 2:maxiter){
   longchain1 <- rbind(longchain1, readRDS(paste0("POIS_iterchunk",i,"_chain2.RDS")))
 }
 
 longchain2 <- readRDS("POIS_iterchunk1_chain3.RDS")
-for(i in 2:3){
+for(i in 2:maxiter){
   longchain2 <- rbind(longchain2, readRDS(paste0("POIS_iterchunk",i,"_chain3.RDS")))
 }
 
 longchain3 <- readRDS("POIS_iterchunk1_chain1.RDS")
-for(i in 2:3){
+for(i in 2:maxiter){
   longchain3 <- rbind(longchain3, readRDS(paste0("POIS_iterchunk",i,"_chain1.RDS")))
 }
 
 
 all.params <- unique(longchain1$Parameter)
-iters <- c(100:600)
-iters <- iters[-which(iters %in% c(201:205,401:405))]
+iters <- c(100:1000)
+iters <- iters[-which(iters %in% c(201:205,401:405,601:605,801:805))]
 
 pdf("test.pdf", width = 8.5, height = 11)
 par(mfrow=c(6,3))

### 2015-09-14 devel/ms_scripts.R
updates simulation with constraint on predicted values

@@ -32,9 +32,15 @@ growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
 growD$Cover <- round(growD$Cover,0) # round for count-like data
 growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
 
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
 
 tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),
                 strip.text=element_text(face="bold"),
+                axis.title=element_text(size=16),text=element_text(size=16),
+                legend.text=element_text(size=16))
+tmp.theme2=theme(strip.text=element_text(face="bold"),
                 axis.title=element_text(size=16),text=element_text(size=16),
                 legend.text=element_text(size=16))
 
@@ -70,33 +76,56 @@ oneyear$eta <- eta
 etaPalette <- colorRampPalette(rev(brewer.pal(11, "RdBu")))
 spfield_plot <- ggplot(oneyear, aes(x=Lon, y=Lat))+
   geom_raster(aes(z=eta, fill=eta))+
-  scale_fill_gradientn(limits=c(-0.55, 0.55), colours=etaPalette(100))+
+  scale_fill_gradientn(limits=c(-0.55, 0.55), colours=etaPalette(100), name=expression(eta))+
   coord_equal()+
   tmp.theme+
   theme(strip.background=element_rect(fill="white"))+
-  ggtitle(bquote("Spatial field ("*eta*")"))
+  ggtitle(bquote("Posterior spatial field ("*eta*")"))
 
 avg_cover <- ddply(growD, .(Lon, Lat), summarise,
                    Cover = mean(Cover))
+avg_cover$scaled_cov <- scale(avg_cover$Cover, center = TRUE, scale = TRUE)
 avgcover_plot <- ggplot(avg_cover, aes(x=Lon, y=Lat))+
-  geom_raster(aes(z=Cover, fill=Cover))+
-  scale_fill_gradientn(colours=myPalette(100))+
+  geom_raster(aes(z=scaled_cov, fill=scaled_cov))+
[…]

### 2015-09-14 devel/run_small_test_betareg.R
new stan test for beta regression

@@ -1,0 +1,688 @@
+##  Script to run GLMM fit on subset of sagebrush data
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(sageAbundance)
+library(rstan)
+library(ggmcmc)
+library(parallel)
+library(reshape2)
+library(plyr)
+library(gridExtra)
+
+####
+##  Get data
+####
+datapath <-  "/Users/atredenn/Dropbox/sageAbundance_data/"
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
+rawD <- read.csv(paste(datapath, "WY_SAGECoverData_V2small.csv", sep=""))
+
+# Merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load("../results/Knot_cell_distances_smallSet.Rdata")
+
+growD[which(growD$Cover==0), "Cover"] <- 0.1
+growD[which(growD$CoverLag==0), "CoverLag"] <- 0.1
+
+growD$Cover <- growD$Cover/100
+growD$Cover <- growD$CoverLag/100
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  real y[nobs]; // observation vector
+  real lag[nobs]; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
[…]

### 2015-09-15 devel/ms_scripts.R
updates script with graph saves

@@ -21,6 +21,9 @@ library(gridExtra)
 library(RColorBrewer)
 
 
+time.steps <- 500
+
+
 ####
 ####  Get data
 ####
@@ -46,6 +49,8 @@ tmp.theme2=theme(strip.text=element_text(face="bold"),
 
 myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
 #all years cover plot
+
+png("../results/all_years_percCover.png", width = 8.5, height = 5, units="in", res = 100)
 ggplot(growD, aes(x=Lon, y=Lat))+
   geom_raster(aes(z=Cover, fill=Cover))+
   facet_wrap("Year")+
@@ -53,6 +58,7 @@ ggplot(growD, aes(x=Lon, y=Lat))+
   coord_equal()+
   tmp.theme+
   theme(strip.background=element_rect(fill="white"))
+dev.off()
 
 # Load knot data
 load(paste0(knotpath,"Knot_cell_distances_subset.Rdata"))
@@ -73,6 +79,7 @@ oneyear$eta <- eta
 ####
 ####  Plot spatial field and avg cover together
 ####
+png("../results/spatialfield.png", width = 8.5, height = 8.5, units="in", res = 100)
 etaPalette <- colorRampPalette(rev(brewer.pal(11, "RdBu")))
 spfield_plot <- ggplot(oneyear, aes(x=Lon, y=Lat))+
   geom_raster(aes(z=eta, fill=eta))+
@@ -80,7 +87,7 @@ spfield_plot <- ggplot(oneyear, aes(x=Lon, y=Lat))+
   coord_equal()+
   tmp.theme+
   theme(strip.background=element_rect(fill="white"))+
-  ggtitle(bquote("Posterior spatial field ("*eta*")"))
+  ggtitle(bquote("Mean posterior spatial field ("*eta*")"))
 
 avg_cover <- ddply(growD, .(Lon, Lat), summarise,
                    Cover = mean(Cover))
@@ -94,7 +101,7 @@ avgcover_plot <- ggplot(avg_cover, aes(x=Lon, y=Lat))+
   ggtitle("Deviation from mean percent cover")
 
 eta_avgcover_plot <- grid.arrange(avgcover_plot, spfield_plot, nrow=2)
-
+dev.off()
 
 
 ####
[…]

### 2015-09-15 docs/ms_sageAbundance.Rmd
updates ms with new figures from subset runs and new text for discussion

@@
 Here we advance spatiotemporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
-We use a 27 year time series of sagebrush (*Artemesia* spp.) percent cover over a $5\times5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
+We use a 27 year time series of sagebrush (*Artemesia* spp.) percent cover over a $5\times2.5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
 We then use the model to forecast future population states (percent cover) under projected climate scenarios.
 Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to spring temperature.
-However, gains in cover are not homogeneous in space; we forecast some areas will increase in cover while others will decrease.
+However, gains in cover are not homogeneous in space.
 Such spatially explicit estimates of increases/decreases in plant abundance have, until now, been elusive.
@@
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
-Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.
 
@@
 #### Remotely-sensed time series
-To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012] (Figure 1).
-Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1 780 field observations of sagebrush cover distributed across Wyoming.
+To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
+Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
 The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
-To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011.
-In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush.
-For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27 948 pixels each.
-Thus, each full dataset contains 754 596 observations (27 years $\times$ 27 948 pixels).
+To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
+This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Figure 1).
+In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
+For our purposes we use a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year.
+Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).
 
@@
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
-The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table #).
+The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature.
 We downloaded daily historic weather data from the parameter-elevation regressions on independent slopes model (PRISM) data set (*available online*)[^link].
@@
 However, instead of further downscaling the 800 meter resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
-We calcualted the four climate variables outline above from the PRISM data for the time period of our remotely sensed data (1984 to 2011).
+We calculated the four climate variables outline above from the PRISM data for the time period of our remotely sensed data (1984 to 2011).
 
@@
 The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
-We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table #.
+We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.
 
@@
 
-\begin{table}[h]
-\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
-average projections.}
-\begin{tabular}{llllll}
-\hline
- & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
-Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
-\hline
-RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
-RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
-RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
-\hline
-\end{tabular}
-\end{table}
-
 ### Additive spatio-temporal model for sagebrush cover
-We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects (Conn et al. 2015).
+We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects [@Conn2015].
 We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally Poisson
@@
 Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
-This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations (Wikle 2010).
+This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
 To overcome these computational constraints, we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
@@
 ####Simulating the model
-Using the posterior distribution of model parameters, we simulate a matrix of pixels equal to the size of the study area (XX XXX pixels or matrix elements).
+Using the posterior distribution of model parameters, we simulate a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
 We initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
-We ran the simulation for 2 500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
+We ran the simulation for 2,500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
 We then ran simulations where the projected climate changes in Table 1 were applied to the observed climate time series.
@@
 -------
-Our pixel-based population model was able to reproduce observed percent cover (Figure 3).
-Three of five climate covariates were strong predictors of cover change (Figure 4).
-In particular, mean spring temperature and precipitation at time *t*-2 had strong positive effects on population growth, only slightly offset by a negative effect of precipitation at time  *t*-1.
-The positive effects of precipitation and temperature resulted in projected increases of cover for all climate change scenarios (Figure 5) since precipitation and temperature are both predicted to increase (Table 1).
+All climate covariates were statistically important predictors of sagebrush cover change, but their values, on a standardized scale, were small (Figure 2).
+Temperature and precipitation covariates tended to have a positive effect on sagebrush cover change, but *ppt1* had a negative effect (Figure 2).
+The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Figure 3).
+This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
+
+When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover (Figures 4 and 5).
+As expected based on the overall positive effects of temperature on sagebrush cover change (Figure 2), our model predicts increases in sagebrush cover under each climate change scenario (Figure 6).
+In general, increases in percent cover tend to occur in areas with above-average percent cover (red areas in Figure 6A "CURRENT").
+Indeed, density curves of projected cover under the climate change scenarios (Figure 7) show that low cover pixel are generally retained (all curves overlap at low cover), while high cover pixels are added (increased right skew in climate change curves; Figure 7).
 
@@
 -----------
+Landscape scale population models for plant species have been elusive.
+Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
+Here we introduced a new statistical approach to modeling plant populations that overcame two major hurdles for spatially-explicit population models.
+First, we used species-specific remote sensing data as a response variable.
+Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent. 
+In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
+Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation (Figure 3).
+
+The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space (Figure 6). 
+However, several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
+We decided to work at the native spatial resolution of the remote sensing data ($30\times30$ meters). 
+This constrained the extent that we could reasonably model because we needed to place the knots for the spatial ranom effect close enough together to detect the underlying spatial structure. 
+As more and more knots are used, computation time increases substantially [@Wikle2010].
+Had we decided to aggregate the original data to a coarser resolution, we would have been able to model a larger spatial extent using the same number of knots, and therefore computation time.
+
+Our decision on spatial extent and resolution also impacted our incorporation of climate covariates.
+We decided against further downscaling PRISM data to match the spatial resolution of the sagebrush data.
+This means that in each year, all pixels share the same climate covariates.
+In other words, climate only varied through time and not over space.
+This is a of course a limitation of our study, and helps to explain why climate coefficients were so certain (Figure 3).
+In the future we plan to work at a lower spatial resolution but a larger spatial extent so that we can use spatially-explicit climate covariates that match the resolution of the percent cover data.
+
+These decisions ultimately affect our results, in positive and negative ways, but also show the flexibility of the approach we developed.
+The underlying model we described and its subsequent use for projecting population changes remains unchanged in the face of individual choices regarding spatial extent and resolution.
+Likewise, the approach can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
+As such datasets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
+
+\pagebreak{}
+
+Tables
+------
+\begin{table}[h]
+\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
+average projections.}
+\begin{tabular}{llllll}
+\hline
+ & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
+Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
+\hline
+RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
+RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
+RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
+\hline
+\end{tabular}
+\end{table}
 
@@
 -------
-```{r figure_1,dependson="plot-options", fig.cap="Location of the study area in southwester Wyoming and a snapshot of the percent cover data in 1984 for each site (a and b).
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the study area in southwester Wyoming and a snapshot of the percent cover data in 1984 for each site (a and b).
 Green shading on the map shows projected distribution of sagebrush from Kuchler (19xx).
@@
 
-```{r figure_2,dependson="plot-options", fig.cap="Average percent cover or each grid cell of the study area with knots used for spatial convolution in dimension reduction.", fig.height=6, eval=FALSE}
+```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=TRUE}
 library(png)
 library(grid)
-img <- readPNG("components/figure/SAGE_Grid_wKnots.png")
+img <- readPNG("../results/all_years_percCover.png")
 grid.raster(img)
@@
 
-```{r figure_3,dependson="plot-options", fig.cap="Observed and predicted cover, averaged over years for each pixel.
-This histogram shows the distribution of observed cover and then density curve shows the distribution of predicted cover from the model.", fig.height=6, eval=FALSE}
+```{r figure_2,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.
+The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
+All covariates are statistically important since their distributions do not overlap zero.
+However, the effects are relatively small.", fig.height=4, eval=TRUE}
 library(png)
 library(grid)
-img <- readPNG("../results/obs_pred_spatial_small.png")
+img <- readPNG("../results/post_climate_covariates.png")
 grid.raster(img)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.", fig.height=6, eval=FALSE}
+```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=TRUE}
 library(png)
 library(grid)
-img <- readPNG("../results/climeff_dens_small.png")
+img <- readPNG("../results/spatialfield.png")
 grid.raster(img)
@@
 
-```{r figure_5,dependson="plot-options", fig.cap="Projected equilibrium cover under climate change scenarios.
-Top panels show the projections in space, averaged over time for each pixel.
-Lower panel shows distributions of cover under each scenario.", fig.height=6, eval=FALSE}
+```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", eval=TRUE}
 library(png)
 library(grid)
-img <- readPNG("../results/climchange_small.png")
+img <- readPNG("../results/obs_pred_spatial_subset.png")
+grid.raster(img)
+```
+
+```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("../results/obs_pred_hist.png")
+grid.raster(img)
+```
+
+```{r figure_6,dependson="plot-options", fig.cap="Projected equilibrium cover under climate change scenarios.
+Top panels shows equilibrium cover based on simulations using observed cliamte.
+Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=TRUE}
+library(png)
+library(grid)
+img <- readPNG("../results/clim_change_mean_spatial.png")
+grid.raster(img)
+```
+
+```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=TRUE, fig.height=8}
+library(png)
+library(grid)
+img <- readPNG("../results/clim_change_densities.png")
 grid.raster(img)

### 2015-09-15 devel/ms_scripts.R

@@ -21,8 +21,6 @@ library(gridExtra)
 library(RColorBrewer)
 
 
-time.steps <- 500
-
 
 ####
 ####  Get data
@@ -128,6 +126,8 @@ dev.off()
 ####
 ####  Run population simulation
 ####
+time.steps <- 600
+burn.in <- 100
 mean_params <- ddply(outs, .(Parameter), summarise,
                      value = mean(value))
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"value"]
@@ -164,7 +164,7 @@ obs.equil <- ddply(growD, .(Lon, Lat), summarise,
                    cover = mean(Cover))
 obs.equil <- obs.equil[with(obs.equil, order(-Lat, Lon)), ]
 # obs.equil[which(obs.equil[,"cover"]>10), "cover"] <- NA
-eq.pixels <- colMeans(ex.mat)
+eq.pixels <- colMeans(ex.mat[(burn.in+1):time.steps,])
 sp.equil <- data.frame(Lon=subset(growD, Year==1985)$Lon, 
                        Lat=subset(growD, Year==1985)$Lat,
                        pred.cover=eq.pixels,
@@ -253,7 +253,8 @@ mean_params <- ddply(outs, .(Parameter), summarise,
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"..1"]
 betas <- mean_params[grep("beta", mean_params$Parameter),"..1"][2:6]
 eta <- K%*%alphas
-time.steps <- 500
+time.steps <- 600
+burn.in <- 100
 
 p.climD<-climD[climD$year %in% unique(growD$Year),c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
 clim_avg <- apply(X = p.climD, MARGIN = 2, FUN = mean)
@@ -285,7 +286,7 @@ for(t in 2:time.steps){
   tmp.out[which(tmp.out>100)] <- mean(ex.mat[t-1,])
   ex.mat[t,] <- tmp.out
 }
-rcp45 <- ex.mat
+rcp45 <- ex.mat[(burn.in+1):time.steps, ]
 
 
 pixels <- nrow(subset(growD, Year==1985))
@@ -311,7 +312,7 @@ for(t in 2:time.steps){
[…]

### 2015-09-16 docs/ms_sageAbundance.Rmd
updates ms with more intro material

@@
 ---
+
+```{r figures_source, include=FALSE, eval=FALSE}
+## Source this for updated figures
+source("../devel/ms_scripts.R")
+```
 
@@
 The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].
-However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [@Salguero-Gomez2015].
+Inference from population models has been limited to small spatial extents because the data required is difficult to collect across wide species ranges.
+Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
 
-Alternatives to local-scale demographic data are rapidly becoming available.
-Forest Inventory and Analysis (FIA) data have been successfully used to fit sophisticated species distribution models (SDMs) that predict occurance and abundance at large spatial extents [@Clark2014].
-However, all SDMs rely on correlations in space, which have been shown to perform poorly relative to models based on responses to changes in climate over time [@Clark2011].
-Likewise, distributed demographic data, like FIA data, are not continuous in space.
-So creating continuous prediction surfaces requires extrapolating predictions to areas not covered by the data.
+The fact that population models typically face spatial constraints may explain the proliferation of studies using species distribution models (SDMs) to infer future states of populations under climate change (CITATIONS).
+SDMs are extremely useful tools, but they tend to lack information on population dynamics and states beyond probability of presence.
+This means that SDMs provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
+One of the major advantages of SDMs is their use of remotely sensed covariates that allow for extrapolation across broad spatial extents.
+New technologies are now providing information on not just the predictor side of the equation, but also the response side.
+
+Species-specific estimates of plant abundance, usually recorded as percent cover or density, offer new opportunities to model and forecast populations at spatial extents that far exceed those possible using traditional population models.
+But, with new opportunities, come new challenges.
+In particular, spatially-continuous response variables require accounting of spatial autocorrelation to arrive at statistically rigorous model fits (Wikle and Hooten refs.).
+Fortunately, recent advances in the spatiotemporal modeling of animal abundances offer a scaffold for similar work on plant species.
 

### 2015-09-16 scripts/traceplots.R
adds in options for getting negbinom fits

@@ -1,6 +1,15 @@
 ##  Script to read in MCMC results from the DoRC server
 ##  Creates traceplots and save MCMC output for inference (1000 iterations)
 
+#Poisson fit
+# filestart <- "POIS_iterchunk" 
+# traceplotfile <- "poissonSage_traceplots.pdf"
+# outmcmcfile <- "poissonSage_mcmc.RDS"
+
+#Negative binomial fit
+filestart <- "iterchunk" 
+traceplotfile <- "negbinomSage_traceplots.pdf"
+outmcmcfile <- "negbinomSage_mcmc.RDS"
 
 ####
 ####  Load libraries
@@ -11,21 +20,22 @@ library(scales)
 
 ####
 ####  Read in MCMC chunks from server
+####
 setwd("/Volumes/A02046115/sage/")
 maxiter <- 10
-longchain1 <- readRDS("POIS_iterchunk1_chain2.RDS")
+longchain1 <- readRDS(paste0(filestart, "1_chain2.RDS"))
 for(i in 2:maxiter){
-  longchain1 <- rbind(longchain1, readRDS(paste0("POIS_iterchunk",i,"_chain2.RDS")))
+  longchain1 <- rbind(longchain1, readRDS(paste0(filestart,i,"_chain2.RDS")))
 }
 
-longchain2 <- readRDS("POIS_iterchunk1_chain3.RDS")
+longchain2 <- readRDS(paste0(filestart, "1_chain3.RDS"))
 for(i in 2:maxiter){
-  longchain2 <- rbind(longchain2, readRDS(paste0("POIS_iterchunk",i,"_chain3.RDS")))
+  longchain2 <- rbind(longchain2, readRDS(paste0(filestart,i,"_chain3.RDS")))
 }
 
-longchain3 <- readRDS("POIS_iterchunk1_chain1.RDS")
+longchain3 <- readRDS(paste0(filestart, "1_chain1.RDS"))
 for(i in 2:maxiter){
-  longchain3 <- rbind(longchain3, readRDS(paste0("POIS_iterchunk",i,"_chain1.RDS")))
+  longchain3 <- rbind(longchain3, readRDS(paste0(filestart,i,"_chain1.RDS")))
 }
 
 
@@ -46,7 +56,7 @@ setwd("/Users/atredenn/Repos/sageAbundance/results/")
 
 all.params <- unique(longchain1$Parameter)
 
[…]

### 2015-09-16 docs/ms_sageAbundance.Rmd
updates ms intro

@@
 In particular, spatially-continuous response variables require accounting of spatial autocorrelation to arrive at statistically rigorous model fits (Wikle and Hooten refs.).
-Fortunately, recent advances in the spatiotemporal modeling of animal abundances offer a scaffold for similar work on plant species.
+Fortunately, advances in the spatiotemporal modeling of animal abundances offer a scaffold for similar work on plant species.
+
+Spatiotemporal models of animal abundance seek to relate the number of observed counts at location *s* and time *t* ($C_{s,t}$) to a latent process $\mu_{s,t}$ through a probability mass function, a link function relating predictors to the mean response, a set of covarites __X__, and parameters $\boldsymbol{\theta}$.
+A typical formulation takes the form
+
+\begin{align}
+[\textbf{C}] &= \text{Poisson}(\boldsymbol{\mu}) \\
+\boldsymbol{\mu} &= \text{exp}(\boldsymbol{\theta}\textbf{X})
+\end{align}
+
+which states that observed counts are Poisson distributed with a log-link function on the latent abundance intensity ($\boldsymbol{\mu}$), which is itself a function of covariates and estimated parameters.
+The covariates include spatial and temporal autocorrelation effects and any number of additional measured environmental effects. 
+In a population modeling sense, the temporal effect can be defined as the effect of population abundance at time *t-1*.
+More difficult is the definition of the spatial random effects in models that include a large number of locations, such as remote sensing data.
+Here too, advances have been made on strategies for dimension reduction that allow for fitting a reduced parameter set that can be expanded to express high-dimensional random effects [@Wikle2010].
+
+Here we develop a statistical approach for modeling plant populations over large spatial extents based on the above framework,.
+Our approach leverages a unique remotely-sensed time series of *Artemesia* species (sagbrush) to fit a spatiotemporal model of plant population dynamics at the landscape scale.
+We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
+We then use the model to forecast the impacts of climate change on sagebrush percent cover.
+The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but we are most excited about its application in cases where remotely-sensed estimates exist.
+In those cases, we think our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.
 
@@
 ####Simulating the model
-Using the posterior distribution of model parameters, we simulate a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
-We initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series.
-We ran the simulation for 2,500 time steps and then compared the output to the observed time series to see if the model could reproduce the observed cover.
-We then ran simulations where the projected climate changes in Table 1 were applied to the observed climate time series.
+We performed three sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, and (3) to forecast future equilibrium population states under projected climate change.
+Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
+For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
+We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output, after discarding and initial 100 time steps.
+For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step.
+
+One problem with our statistical approach that treats integer cover as counts, is that the Poisson distribution supports values greater than 100, a value that percent plant cover cannot exceed in reality.
+Thus, when simulating the fitted model it is entirely possible to predict values of cover greater than 100%.
+Likewise, once a pixel reaches such a great value, the exponential function $\text{exp}(\beta_{1} y_{i,t-1})$ (Eq.
+#) results in runaway growth and eventually all pixels go to unrealistic values of percent cover.
+We therefore imposed a rule in our simulations that if predicted cover from a pixel exceeded 100%, it's value was set to the mean value of all pixels at that timestep.
+Our main purpose for simulating the model is to assess percent cover at equilibrium, so our focus is not on year-to-year dynamics that are affected by this rule.
+The rule simply ensures that percent cover cannot go over 100% while maintaining any mean changes in percent cover since pixels exceeding 100% are set to the mean matrix value.
+Admittedly, such a rule is not ideal, but we have chosen practicality over idealism in this case.
 
@@
 As such datasets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
+
+Acknowledgments
+---------------
+This work is the outcome of a distributed graduate seminar led by PBA and supported by a National Science Foundation CAREER award (DEB-XXXXXX).
+ATT was supported by a NSF Postdoctoral Research Fellowship in Biology (DBI-XXXXX) and AK was supported by a NSF Graduate Research Fellowship (DEB-XXXXXX).
+We are grateful to Mary LASTNAME at USGS EROS for exctracting the dataset used in this paper.
 
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=TRUE}
+Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=TRUE, fig.height=8}
 library(png)

### 2015-09-16 devel/ms_scripts.R

@@ -373,7 +373,7 @@ ggplot(proj.equil2, aes(x=Lon, y=Lat))+
   coord_equal()+
   tmp.theme+
   theme(strip.background=element_rect(fill="white"))
-ggsave("../results/clim_change_mean_spatial.png", height=8, width=4, dpi=200)
+ggsave("../results/clim_change_mean_spatial.png", height=8, width=4)
 
 ggplot(all4plot)+
   geom_line(stat="density", aes(x=value, y=..density.., col=variable), size=1)+

### 2015-09-16 docs/ms_sageAbundance.Rmd
updates ms intro text; adds new figures for bias and precision

@@
 
-Spatiotemporal models of animal abundance seek to relate the number of observed counts at location *s* and time *t* ($C_{s,t}$) to a latent process $\mu_{s,t}$ through a probability mass function, a link function relating predictors to the mean response, a set of covarites __X__, and parameters $\boldsymbol{\theta}$.
+Spatiotemporal models of animal abundance seek to relate the number of observed counts at location *s* and time *t* ($C_{s,t}$) to a latent process $\mu_{s,t}$ through a probability mass function, a link function relating predictors to the mean response, a set of covarites __X__ and associated parameters $\boldsymbol{\theta}$, and a spatial random effect $\boldsymbol{\eta}$.
 A typical formulation takes the form
@@
 [\textbf{C}] &= \text{Poisson}(\boldsymbol{\mu}) \\
-\boldsymbol{\mu} &= \text{exp}(\boldsymbol{\theta}\textbf{X})
+\boldsymbol{\mu} &= \text{exp}(\boldsymbol{\theta}\textbf{X} + \boldsymbol{\eta})
 \end{align}
@@
 which states that observed counts are Poisson distributed with a log-link function on the latent abundance intensity ($\boldsymbol{\mu}$), which is itself a function of covariates and estimated parameters.
-The covariates include spatial and temporal autocorrelation effects and any number of additional measured environmental effects. 
-In a population modeling sense, the temporal effect can be defined as the effect of population abundance at time *t-1*.
-More difficult is the definition of the spatial random effects in models that include a large number of locations, such as remote sensing data.
-Here too, advances have been made on strategies for dimension reduction that allow for fitting a reduced parameter set that can be expanded to express high-dimensional random effects [@Wikle2010].
+The exact formulation of the latent process model (Eq.
+2) can take many forms, including, for example, those motivated by classic density-dependent growth [@Ross2015] or by resource-selection processes [@Conn2015].
+However, all models like Eq. 2 share the need to estimate the spatial random effect ($\boldsymbol{\eta}$).
+In some cases this is as straightforward as fitting a random effect for each sampling unit.
+But when the number of sampling units is large (1000s), fitting these effects becomes computationally untenable.
+Fortunately, modern model reparameterizations and dimension reduction techniques allow for estimation of spatial random effects by mapping a reduced parameter set (which are estimated in the mode) to the full spatial extent under consideration with an interpolation matrix [@Wikle2010; @Conn2015].
+The general model outlined by Eqs 1-2, and the associated computational strategies for fitting them, have recieved wide usage in the animal abundance literature [@Fink2010; @Conn2015; @Ross2015], but have yet to be applied to the modeling of plant populations.
+This is most likely because the methods are relatively new and because, until now, estimates of plant abundance over large spatial extents have been rare.
 
-Here we develop a statistical approach for modeling plant populations over large spatial extents based on the above framework,.
+Here we develop a statistical approach for modeling plant populations over large spatial extents based on the above framework and a process model motivated by Gompertz density-dependent growth.
 Our approach leverages a unique remotely-sensed time series of *Artemesia* species (sagbrush) to fit a spatiotemporal model of plant population dynamics at the landscape scale.
@@
 
-```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=TRUE}
+```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=F}
 library(png)
@@
 All covariates are statistically important since their distributions do not overlap zero.
-However, the effects are relatively small.", fig.height=4, eval=TRUE}
+However, the effects are relatively small.", fig.height=4, eval=F}
 library(png)
@@
 
-```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=TRUE}
+```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=F}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", eval=TRUE}
+```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", eval=F}
 library(png)
@@
 
-```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=TRUE}
+```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=F}
 library(png)
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=TRUE, fig.height=8}
+Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=F, fig.height=8}
 library(png)
@@
 
-```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=TRUE, fig.height=8}
+```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=F, fig.height=8}
 library(png)

### 2015-09-16 devel/ms_scripts.R

@@ -172,17 +172,54 @@ sp.equil <- data.frame(Lon=subset(growD, Year==1985)$Lon,
 colnames(sp.equil)[3:4] <- c("Predicted", "Observed") 
 sp.equil2 <- melt(sp.equil, id.vars = c("Lon", "Lat"))
 
+##  Calculate prediction bias and precision
+wide_equil <- dcast(sp.equil2, Lon+Lat~variable, value.var = "value")
+wide_equil$bias <- with(wide_equil, Predicted-Observed)
 
+pred.var <- apply(ex.mat, 2, var)
+sp.predvar <- data.frame(Lon=subset(growD, Year==1985)$Lon, 
+                       Lat=subset(growD, Year==1985)$Lat,
+                       pred.prec=1/pred.var)
 
-png("../results/obs_pred_spatial_subset.png", width = 5, height=5, units = "in", res=200)
-g1 <- ggplot(sp.equil2, aes(x=Lon, y=Lat))+
+tmp.theme3=theme(axis.ticks = element_blank(), axis.text = element_blank(),
+                strip.text=element_text(face="bold"),
+                axis.title=element_text(size=12),text=element_text(size=12),
+                legend.text=element_text(size=10))
+
+png("../results/obs_pred_spatial_subset.png", width = 8, height=8, units = "in", res=100)
+g1 <- ggplot(subset(sp.equil2, variable=="Observed"), aes(x=Lon, y=Lat))+
   geom_raster(aes(z=value, fill=value))+
-  scale_fill_gradientn(colours=myPalette(200), name="% Cover")+
-  facet_wrap("variable", nrow=2)+
+  scale_fill_gradientn(colours=myPalette(200), name="% Cover", limit=c(0,25))+
   coord_equal()+
-  tmp.theme+
-  theme(strip.background=element_rect(fill="white"))
-print(g1)
+  tmp.theme3+
+  theme(strip.background=element_rect(fill="white"))+
+  ggtitle("A) Observed cover")
+
+g2 <- ggplot(subset(sp.equil2, variable=="Predicted"), aes(x=Lon, y=Lat))+
+  geom_raster(aes(z=value, fill=value))+
+  scale_fill_gradientn(colours=myPalette(200), name="% Cover", limit=c(0,25))+
+  coord_equal()+
+  tmp.theme3+
[…]

### 2015-09-17 devel/run_small_test_DeltaPois.R
new devel scripts for alternative models

@@ -1,0 +1,143 @@
+##  Script to run GLMM fit on subset of sagebrush data
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(sageAbundance)
+library(rstan)
+library(ggmcmc)
+library(parallel)
+library(reshape2)
+library(plyr)
+library(gridExtra)
+
+####
+##  Get data
+####
+datapath <-  "/Users/atredenn/Dropbox/sageAbundance_data/"
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
+rawD <- read.csv(paste(datapath, "WY_SAGECoverData_V2small.csv", sep=""))
+
+# Merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load("../results/Knot_cell_distances_smallSet.Rdata")
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> npres; // number of observations
+  int<lower=0> nabs; // number of observations
+  int<lower=0> pres;
+  int<lower=0> absd;
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  real y[nobs]; // observation vector
+  //real lag[nobs]; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
+  real a;
+  real b;
+  //real<lower=0> beta_mu;
+  real<lower=0.000001> sig_a;
[…]

### 2015-09-17 devel/run_small_test_Gompertz.R

@@ -1,0 +1,681 @@
+##  Script to run GLMM fit on subset of sagebrush data
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(sageAbundance)
+library(rstan)
+library(ggmcmc)
+library(parallel)
+library(reshape2)
+library(plyr)
+library(gridExtra)
+library(truncnorm)
+
+####
+##  Get data
+####
+datapath <-  "/Users/atredenn/Dropbox/sageAbundance_data/"
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
+rawD <- read.csv(paste(datapath, "WY_SAGECoverData_V2small.csv", sep=""))
+
+# Merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load("../results/Knot_cell_distances_smallSet.Rdata")
+
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  real<lower=-12, upper=5> y[nobs]; // observation vector
+  real<lower=-12, upper=5> lag[nobs]; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
+  real<lower=0> beta_mu;
+  real<lower=0.000001> sig_a;
+  real<lower=0.000001> sig_mu;
[…]

### 2015-09-18 docs/ms_sageAbundance.Rmd
new devel scripts for poisson without 0s

@@
 All covariates are statistically important since their distributions do not overlap zero.
-However, the effects are relatively small.", fig.height=4, eval=F}
+However, the effects are relatively small.", fig.height=4, eval=T}
 library(png)
@@
 
-```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=F}
+```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=T}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", eval=F}
+```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", fig.height=8, eval=T}
 library(png)
@@
 
-```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=F}
+```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=T}
 library(png)
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", eval=F, fig.height=8}
+Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=T}
 library(png)
@@
 
-```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=F, fig.height=8}
+```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", fig.height=8, eval=T}
 library(png)

### 2015-09-18 devel/run_small_test_StateSpacePoisson.R

@@ -1,0 +1,112 @@
+##  Script to run GLMM fit on subset of sagebrush data
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(sageAbundance)
+library(rstan)
+library(ggmcmc)
+library(parallel)
+library(reshape2)
+library(plyr)
+library(gridExtra)
+
+####
+##  Get data
+####
+datapath <-  "/Users/atredenn/Dropbox/sageAbundance_data/"
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
+rawD <- read.csv(paste(datapath, "WY_SAGECoverData_V2small.csv", sep=""))
+
+# Merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load("../results/Knot_cell_distances_smallSet.Rdata")
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nyrs; // number of observation years
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  int y[nyrs,ncells]; // observation matrix
+  matrix[dK1,dK2] K; // spatial field matrix
+  vector[nyrs] X; // climate covariate matrix
+  vector[ncells] startvec;
+}
+parameters{
+  real int_mu;
+  real<lower=0> beta_mu;
+  real<lower=0.0001> sig_a;
+  vector[nknots] alpha;
+  real beta1;
+}
+transformed parameters{
+  vector[ncells] eta;
+  matrix[nyrs, ncells] mu;
+  eta <- K*alpha;
+  for(i in 1:ncells)
+    mu[1,i] <- startvec[i];
+  for(t in 2:nyrs){
+    for(i in 1:ncells){
[…]

### 2015-09-18 devel/run_small_test_impute.R

@@ -1,0 +1,704 @@
+##  Script to run GLMM fit on subset of sagebrush data
+
+# Clear workspace 
+rm(list=ls(all=TRUE))
+
+# Load libraries
+library(sageAbundance)
+library(rstan)
+library(ggmcmc)
+library(parallel)
+library(reshape2)
+library(plyr)
+library(gridExtra)
+
+####
+##  Get data
+####
+datapath <-  "/Users/atredenn/Dropbox/sageAbundance_data/"
+climD <- read.csv(paste(datapath,
+                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
+                        sep=""))
+rawD <- read.csv(paste(datapath, "WY_SAGECoverData_V2small.csv", sep=""))
+
+# Merge in climate data 
+fullD <- merge(rawD,climD,by.x="Year", by.y="year",all.x=T)
+
+# Get data structure right
+growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Load knot data
+load("../results/Knot_cell_distances_smallSet.Rdata")
+
+
+
+####
+####  Write the STAN model
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> nknots; // number of interpolation knots
+  int<lower=0> ncells; // number of cells
+  int<lower=0> cellid[nobs]; // cell id
+  int<lower=0> dK1; // row dim for K
+  int<lower=0> dK2; // column dim for K
+  int y[nobs]; // observation vector
+  vector[nobs] lag; // lag cover vector
+  matrix[dK1,dK2] K; // spatial field matrix
+  matrix[nobs,ncovs] X; // spatial field matrix
+}
+parameters{
+  real int_mu;
+  real<lower=0> beta_mu;
+  real<lower=0.000001> sig_a;
+  real<lower=0.000001> sig_mu;
+  vector[nknots] alpha;
+  vector[ncovs] beta;
[…]

### 2015-09-28 docs/ms_sageAbundance.Rmd
adds new randYear model fits; updates ms with new results

@@
 
-One problem with our statistical approach that treats integer cover as counts, is that the Poisson distribution supports values greater than 100, a value that percent plant cover cannot exceed in reality.
-Thus, when simulating the fitted model it is entirely possible to predict values of cover greater than 100%.
-Likewise, once a pixel reaches such a great value, the exponential function $\text{exp}(\beta_{1} y_{i,t-1})$ (Eq.
-#) results in runaway growth and eventually all pixels go to unrealistic values of percent cover.
-We therefore imposed a rule in our simulations that if predicted cover from a pixel exceeded 100%, it's value was set to the mean value of all pixels at that timestep.
-Our main purpose for simulating the model is to assess percent cover at equilibrium, so our focus is not on year-to-year dynamics that are affected by this rule.
-The rule simply ensures that percent cover cannot go over 100% while maintaining any mean changes in percent cover since pixels exceeding 100% are set to the mean matrix value.
-Admittedly, such a rule is not ideal, but we have chosen practicality over idealism in this case.
-
 Results

### 2015-09-28 devel/ms_scripts.R

@@ -62,7 +62,25 @@ dev.off()
 load(paste0(knotpath,"Knot_cell_distances_subset.Rdata"))
 K <- K.data$K
 
-outs <- readRDS("../results/poissonSage_mcmc.RDS")
+outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
+
+####
+####  Subset data by removing pixels with 0s
+####
+# "%w/o%" <- function(x, y) x[!x %in% y] # x without y
+# # Remove pixels that have 0s in the time series
+# zids <- growD[which(growD$Cover==0),"ID"]
+# zids <- unique(c(zids, growD[which(growD$CoverLag==0),"ID"]))
+# pixels_to_keep <- growD$ID %w/o%  zids
+# growD <- growD[which(growD$ID %in% pixels_to_keep),]
+# 
+# # Take out those pixel rows from the K matrix
+# K <- K.data$K
+# K <- K[-zids,]
+# 
+# # Check to make sure dimensions match
+# if(dim(K)[1] != length(unique(growD$ID))) stop("dimension mismatch between K and data")
+
 
 ## Look at spatial field
 alpha <- outs[grep("alpha", outs$Parameter),]
@@ -108,17 +126,48 @@ dev.off()
 clim_vars <- c("beta.1.", "beta.2.", "beta.3.", "beta.4.", "beta.5.")
 clim_names <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
 clim_posts <- subset(outs, Parameter %in% clim_vars)
+quants <- ddply(clim_posts, .(Parameter), summarise,
+              lower = quantile(value, .05),
+              upper = quantile(value, .95),
+              average = mean(value))
+quants <- quants[with(quants, order(-average)), ]
+quants$rank <- c(1:nrow(quants))
+quants$col <- NA
+
+for(i in 1:nrow(quants)){
+  if(sign(quants$lower[i]) == sign(quants$upper[i]) & quants$average[i] > 0 ){
+    quants$col[i] <- "pos"
+  }
+  if(sign(quants$lower[i]) == sign(quants$upper[i]) & quants$average[i] < 0 ){
+    quants$col[i] <- "neg"
+  }
+  
[…]

### 2015-09-28 scripts/traceplots.R

@@ -3,8 +3,8 @@
 
 #Poisson fit
 filestart <- "POIS_iterchunk" 
-traceplotfile <- "poissonSage_traceplots.pdf"
-outmcmcfile <- "poissonSage_mcmc.RDS"
+traceplotfile <- "poissonSage_randYear_traceplots.pdf"
+outmcmcfile <- "poissonSage_randYear_mcmc.RDS"
 
 #Negative binomial fit
 # filestart <- "iterchunk" 

### 2015-09-28 docs/ms_sageAbundance.Rmd
updates ms with new intro

@@
 Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
-The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].
-Inference from population models has been limited to small spatial extents because the data required is difficult to collect across wide species ranges.
+Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers.
+However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
 Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
+Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].
 
-The fact that population models typically face spatial constraints may explain the proliferation of studies using species distribution models (SDMs) to infer future states of populations under climate change (CITATIONS).
-SDMs are extremely useful tools, but they tend to lack information on population dynamics and states beyond probability of presence.
-This means that SDMs provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
-One of the major advantages of SDMs is their use of remotely sensed covariates that allow for extrapolation across broad spatial extents.
-New technologies are now providing information on not just the predictor side of the equation, but also the response side.
+Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales.
+SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance.
+Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
+Longterm trends in species' occurance can be detected and correlated with climate variables, but these may or may not be representative of a species' latent sensitivity climate.
+Likewise, since SDMs typically rely on occurance data, their projections of habitat suitability or probability of occurance provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
+Despite these limitations, SDMs remain one of the few ways to make population forecasts at large spatial extents.
 
-Species-specific estimates of plant abundance, usually recorded as percent cover or density, offer new opportunities to model and forecast populations at spatial extents that far exceed those possible using traditional population models.
-But, with new opportunities, come new challenges.
-In particular, spatially-continuous response variables require accounting of spatial autocorrelation to arrive at statistically rigorous model fits (Wikle and Hooten refs.).
-Fortunately, advances in the spatiotemporal modeling of animal abundances offer a scaffold for similar work on plant species.
+It seems the ideal tool for forecasting future population states under climate change is a large scale dynamic population model.
+Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary to fit high-dimensional spatial models.
+Fortunately, two recent technological advances put us in the position to develop large scale population models.
+First, remotely-sensed (RS) time series of species-specific plant cover provide data on interannual population dynamics theat can be correlated with yearly climate fluctuations at large spatial scales.
+RS time series have been used to detect long term trends in population abundance, but they have yet to be used to drive a dynamic model.
+Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
+Here, too, these methods have yet to be use in a dynamic setting.
 
-Spatiotemporal models of animal abundance seek to relate the number of observed counts at location *s* and time *t* ($C_{s,t}$) to a latent process $\mu_{s,t}$ through a probability mass function, a link function relating predictors to the mean response, a set of covarites __X__ and associated parameters $\boldsymbol{\theta}$, and a spatial random effect $\boldsymbol{\eta}$.
-A typical formulation takes the form
-
-\begin{align}
-[\textbf{C}] &= \text{Poisson}(\boldsymbol{\mu}) \\
-\boldsymbol{\mu} &= \text{exp}(\boldsymbol{\theta}\textbf{X} + \boldsymbol{\eta})
-\end{align}
-
-which states that observed counts are Poisson distributed with a log-link function on the latent abundance intensity ($\boldsymbol{\mu}$), which is itself a function of covariates and estimated parameters.
-The exact formulation of the latent process model (Eq.
-2) can take many forms, including, for example, those motivated by classic density-dependent growth [@Ross2015] or by resource-selection processes [@Conn2015].
-However, all models like Eq. 2 share the need to estimate the spatial random effect ($\boldsymbol{\eta}$).
-In some cases this is as straightforward as fitting a random effect for each sampling unit.
-But when the number of sampling units is large (1000s), fitting these effects becomes computationally untenable.
-Fortunately, modern model reparameterizations and dimension reduction techniques allow for estimation of spatial random effects by mapping a reduced parameter set (which are estimated in the mode) to the full spatial extent under consideration with an interpolation matrix [@Wikle2010; @Conn2015].
-The general model outlined by Eqs 1-2, and the associated computational strategies for fitting them, have recieved wide usage in the animal abundance literature [@Fink2010; @Conn2015; @Ross2015], but have yet to be applied to the modeling of plant populations.
-This is most likely because the methods are relatively new and because, until now, estimates of plant abundance over large spatial extents have been rare.
-
-Here we develop a statistical approach for modeling plant populations over large spatial extents based on the above framework and a process model motivated by Gompertz density-dependent growth.
+Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent growth.
 Our approach leverages a unique remotely-sensed time series of *Artemesia* species (sagbrush) to fit a spatiotemporal model of plant population dynamics at the landscape scale.
@@
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
-The four climate variables in our population model are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature.
-We downloaded daily historic weather data from the parameter-elevation regressions on independent slopes model (PRISM) data set (*available online*)[^link].
-The PRISM weather data is interpolated between coarse observation units and does capture some spatial variation.
-However, instead of further downscaling the 800 meter resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
+The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}, (2) year \emph{t} fall through summer precipitation, (3) year \emph{t}-1 fall through summer precipitation, (4) year \emph{t} average spring temperature, and (5) year \emph{t}-1 average spring temperature.
+We selected these variables based on previous studies, even though not all emerge as important predictors in our model.
+Several model selection techniques exist that could be applied to our approach [@Hooten2014], but we avoid model selection here to maximize focus on the general approach.
+
+We downloaded daily historic weather data from the NASA Daymet data set (*available online*)[^link].
+The Daymet weather data is interpolated between coarse observation units and does capture some spatial variation.
+However, instead of further downscaling the 1 kilometer resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
 We calculated the four climate variables outline above from the PRISM data for the time period of our remotely sensed data (1984 to 2011).
 
-[^link]: http://www.prism.oregonstate.edu/
+[^link]: http://daymet.ornl.gov/
 
@@
 \begin{align}
-\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
+\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
 \end{align}
@@
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
+The intercept, $\beta{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
 We chose to use a Poisson likelihood because integer percent cover values can be considered a form of count data.
@@
 \begin{align}
-[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] \propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \boldsymbol{\beta}, \boldsymbol{\gamma},\boldsymbol{\alpha}] \right) \prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\beta_{0}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\eta}^2].
+[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
+&\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2].
 \end{align}
@@
 -------
-All climate covariates were statistically important predictors of sagebrush cover change, but their values, on a standardized scale, were small (Figure 2).
-Temperature and precipitation covariates tended to have a positive effect on sagebrush cover change, but *ppt1* had a negative effect (Figure 2).
-The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Figure 3).
+All parameters in our model converged on stable posterior distributions (Appendix A).
+Only the *lagPpt* climate covariate can be considered important at 90% confidence, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 2).
+The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. 3).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
 
-When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover (Figures 4 and 5).
-As expected based on the overall positive effects of temperature on sagebrush cover change (Figure 2), our model predicts increases in sagebrush cover under each climate change scenario (Figure 6).
-In general, increases in percent cover tend to occur in areas with above-average percent cover (red areas in Figure 6A "CURRENT").
-Indeed, density curves of projected cover under the climate change scenarios (Figure 7) show that low cover pixel are generally retained (all curves overlap at low cover), while high cover pixels are added (increased right skew in climate change curves; Figure 7).
+When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover in space (Fig. 4A,B).
+Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
+Nonetheless, observations and predictions are qualitatively similar (Fig. 4A, B).
+Predictions are most confident, even if biased, in low percent cover pixels (Fig. 4D).
+This is due to the fact that the model spatial effect in low cover regions can be so strong that it prevents an increase from zero percent cover even in favorable climate years.
+The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed cover in the previous year for each pixel (Fig. 5)
+
+When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 6, 7).
+The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 6).
+Thus, our model predicts an increase in sagebrush cover heterogeneity since we do not lose low cover pixels and gain higher cover pixels (Fig. 7).
+Without the spatial random effect we would have observed only a directional shift in the distribution of cover, not an expansion as seen in Fig. 7.
 
@@
 In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation (Figure 3).
+Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation (Fig.
+3).
 
-The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space (Figure 6). 
+The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space (Fig. 6). 
 However, several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
@@
 Our decision on spatial extent and resolution also impacted our incorporation of climate covariates.
-We decided against further downscaling PRISM data to match the spatial resolution of the sagebrush data.
+We decided against further downscaling Daymet data to match the spatial resolution of the sagebrush data.
 This means that in each year, all pixels share the same climate covariates.
 In other words, climate only varied through time and not over space.
-This is a of course a limitation of our study, and helps to explain why climate coefficients were so certain (Figure 3).
+This is a of course a limitation of our study, and helps to explain the weak effect of climate covariates when averaged over space (Fig. 3).
 In the future we plan to work at a lower spatial resolution but a larger spatial extent so that we can use spatially-explicit climate covariates that match the resolution of the percent cover data.
@@
 ---------------
-This work is the outcome of a distributed graduate seminar led by PBA and supported by a National Science Foundation CAREER award (DEB-XXXXXX).
-ATT was supported by a NSF Postdoctoral Research Fellowship in Biology (DBI-XXXXX) and AK was supported by a NSF Graduate Research Fellowship (DEB-XXXXXX).
+This work is the outcome of a distributed graduate seminar led by PBA and supported by a National Science Foundation CAREER award (DEB-1054040).
+ATT was supported by an NSF Postdoctoral Research Fellowship in Biology (DBI-1400370) and AK was supported by an NSF Graduate Research Fellowship.
 We are grateful to Mary LASTNAME at USGS EROS for exctracting the dataset used in this paper.
+Compute, storage and other re-sources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
 
@@
 
-```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=F}
+```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=T}
 library(png)
@@
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-All covariates are statistically important since their distributions do not overlap zero.
-However, the effects are relatively small.", fig.height=4, eval=T}
+Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=T}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Projected equilibrium cover from simulations using observed climate (top panel) and average observed percent cover (bottom panel).", fig.height=8, eval=T}
+```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover, and prediction bias and standard deviation (C, D).
+All simulations used mean parameter values and no random year effects.", fig.height=8, eval=T}
 library(png)
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent paneld show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=T}
+Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=T}
 library(png)
@@
 
-```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", fig.height=8, eval=T}
+```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=T}
 library(png)

### 2015-09-29 devel/ms_scripts.R
updates ms with correct figures

@@ -184,22 +184,21 @@ betas <- mean_params[grep("beta", mean_params$Parameter),"value"][2:6]
 eta <- K%*%alphas
 pixels <- nrow(subset(growD, Year==1985))
 ex.mat <- matrix(NA,nrow=time.steps,ncol=pixels)
-ex.mat[1,] <- 10
+ex.mat[1,] <- 1
 clim_sim <- climD[climD$year %in% unique(growD$Year),]
 X_sim = clim_sim[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
 X_sim = scale(X_sim, center = TRUE, scale = TRUE)
 for(t in 2:time.steps){
   Xtmp <- X_sim[sample(c(1:nrow(X_sim)), 1),]
-  tmp.mu <- mean_params[mean_params$Parameter=="int_mu","value"] + mean_params[mean_params$Parameter=="beta_mu","value"]*log(ex.mat[t-1,]) + 
-    sum(betas*Xtmp)
+  dens.dep <- mean_params[mean_params$Parameter=="beta_mu","value"]*log(ex.mat[t-1,])
+  dens.dep[which(dens.dep==-Inf)] <- 0
+  tmp.mu <- mean_params[mean_params$Parameter=="int_mu","value"] + dens.dep + sum(betas*Xtmp)
   tmp.mu <- exp(tmp.mu + eta)
-#   print(max(tmp.mu))
   tmp.out <- rpois(ncol(ex.mat), lambda = tmp.mu)
-#   tmp.out[which(tmp.out>100)] <- mean(ex.mat[t-1,])
   ex.mat[t,] <- tmp.out
 }
 
-matplot(c(1:time.steps), ex.mat, type="l", col="grey")
+# matplot(c(1:time.steps), ex.mat, type="l", col="grey")
 # # hist(y, freq = FALSE)
 # plot(density(ex.mat, adjust = 4), col="red", lwd=2)
 # abline(v = mean(growD$Cover))
@@ -299,8 +298,9 @@ X_sim = scale(X_sim, center = TRUE, scale = TRUE)
 for(t in 1:time.steps){
   Xtmp <- X_sim[t,]
   lagcover <- growD[which(growD$Year==yearid[t]),"CoverLag"]
[…]

### 2015-09-29 devel/ms_scripts.R
updates ms_scripts with colonization algorithm

@@ -4,13 +4,15 @@
 rm(list=ls(all=TRUE))
 
 ####
-####  Set some file paths, etc.
+####  Set some file paths, etc. ------------------------------------------------
 ####
 datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
 knotpath <- "../results/"
 
+
+
 ####
-####  Load required libraries
+####  Load required libraries --------------------------------------------------
 ####
 library(rstan)
 library(reshape2)
@@ -23,7 +25,7 @@ library(RColorBrewer)
 
 
 ####
-####  Get data
+####  Get data -----------------------------------------------------------------
 ####
 fullD <- read.csv(paste0(datapath,"wy_sagecover_subset_noNA.csv"))
 if(length(which(is.na(fullD$Cover))) > 0) stop("data contains NA values")
@@ -37,6 +39,32 @@ climD <- read.csv(paste(datapath,
                         "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
                         sep=""))
 
+
+
+####
+####  Load knot data and MCMC results ------------------------------------------
+####
+load(paste0(knotpath,"Knot_cell_distances_subset.Rdata"))
+K <- K.data$K
+outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
+
+
+
+####
+####  Fit colonization logistic ------------------------------------------------
+####
+colD <- growD[which(growD$CoverLag == 0), ]
+colD$colonizes <- ifelse(colD$Cover==0, 0, 1)
+col.fit <- glm(colonizes ~ 1, data=colD, family = "binomial")
+col.intercept <- as.numeric(coef(col.fit))
+antilogit <- function(x) { exp(x) / (1 + exp(x) ) }
+avg.new.cover <- round(mean(colD[which(colD$Cover>0),"Cover"]),0)
+
+
+
+####
+####  Set up plotting themes ---------------------------------------------------
+####
 tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),
                 strip.text=element_text(face="bold"),
                 axis.title=element_text(size=16),text=element_text(size=16),
[…]

### 2015-10-01 docs/ms_sageAbundance.Rmd
updates ms with correct figures
updates ms with glm for colonization probabilities

@@
     fig_caption: true
-    keep_tex: true
+    keep_tex: false
 fontsize: 12pt
@@
 #### Remotely-sensed time series
-To demonstrate our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
+To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
 Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
@@
 To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
-This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Figure 1).
+This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
 In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
-For our purposes we use a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year.
+For our purposes we use a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
 Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).
@@
 However, given the relative largeness of our dataset and computational limitations, we defined kernels around 271 knots (Figure 2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix A).
-
 The Bayesian posterior distribution can be expressed as
@@
 \end{align}
+
+### Accomodating zeros
+Since our process model (Eq. ) includes a log transformation of the observations ($\text{log}(y_{t-1}$), our model does not accomodate zeros. 
+Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47). 
+Thus, we simply excluded those pixels from the model fitting process. 
+However, when it came time to simulate the model, we needed to include possible transitions from zero to non-zero percent cover. 
+We fit an intercept-only logistic model to estimate the probability of a pixel going from zero to non-zero cover
+
+\begin{align}
+\text{logit}(\mu_{i}) &= \alpha \\
+y_{i} &\sim \text{Bernoulli}(\mu_{i})
+\end{align}
+
+where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($\alpha$).
+We fit this simple model using the `lm' command in R [@R2013].
 

### 2015-10-06 docs/ms_sageAbundance.Rmd
updates table to include just the one study area

@@
 \hline
- & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
-Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
+Emissions Scenario & Absolute change in temperature & Percentage change in precipitation\tabularnewline
 \hline
-RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
-RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
-RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
+RCP 4.5 & 2.98\textdegree & 8.94\%\tabularnewline
+RCP 6.0 & 3.13\textdegree & 8.64\%\tabularnewline
+RCP 8.5 & 4.79\textdegree & 11.0\%\tabularnewline
 \hline

### 2015-10-06 devel/ms_scripts.R
adds data folder with climD; adds HPC script for paramvary simulations

@@ -504,14 +504,18 @@ ggsave("../results/clim_change_densities.png", height=4, width=5, dpi = 200)
 
 
 
+
+
+###########################################################
+###########################################################
+###########################################################
+####                                                   ####
+####  Run population model with parameter uncertainty  ####
+####                                                   ####
+###########################################################
+###########################################################
+###########################################################
 # 
-# 
-# ###################################################
-# ###################################################
-# ###################################################
-# ####
-# ####  Run population model with parameter uncertainty
-# ####
 # colnames(outs)[which(colnames(outs)=="chain")] <- "Chain"
 # 
 # time.steps <- 10
@@ -540,12 +544,17 @@ ggsave("../results/clim_change_densities.png", height=4, width=5, dpi = 200)
 #     eta <- K%*%as.numeric(unlist(alphas))
 #     for(t in 2:time.steps){
 #       Xtmp <- X_sim[sample(c(1:nrow(X_sim)), 1),]
-#       tmp.mu <- mean_params[mean_params$Parameter=="int_mu","..1"] + mean_params[mean_params$Parameter=="beta_mu","..1"]*ex.mat[t-1,] + 
-#         sum(betas*Xtmp)
+#       dens.dep <- beta_mu*log(ex.arr[counter,t-1,])
+#       tmp.mu <- int_mu + dens.dep + sum(betas*Xtmp)
 #       tmp.mu <- exp(tmp.mu + eta)
-#       #   print(max(tmp.mu))
-#       tmp.out <- rpois(ncol(ex.mat), lambda = tmp.mu)
-#       tmp.out[which(tmp.out>100)] <- mean(ex.mat[t-1,])
+#       tmp.out <- rpois(ncol(ex.arr[counter,,]), lambda = tmp.mu)
+#       
+#       #Colonization
+#       zeros <- which(ex.arr[counter,t-1,]==0)
+#       colonizers <- rbinom(length(zeros), size = 1, antilogit(col.intercept))
+#       colonizer.cover <- colonizers*avg.new.cover
+#       tmp.out[zeros] <- colonizer.cover
+#       
 #       ex.arr[counter,t,] <- tmp.out
 #     }
 #     counter <- counter+1

### 2015-10-06 scripts/sim_sageAbundance_HPC.R

@@ -1,0 +1,226 @@
+###########################################################
+##  Supporting Material for Tredennick et al. 201x       ##  
+##  Code to reproduce results displayed in figures x,x,x ##
+###########################################################
+
+##  Script to simulate fitted sage population abundance model
+##  using observed and projected climate. This script runs simulations
+##  for each MCMC iteration combination of parameters. That way we get
+##  the full uncertainty around model forecasts.
+
+##  Since this is a large job, this script is run on the Utah State
+##  University super computer. Thus, file paths are different from
+##  other scripts. This also means that you should probably not run
+##  this as is on your PC. It will take forever...
+
+##  Author:       Andrew Tredennick
+##  Email:        atredenn@gmail.com
+##  Date created: 10-06-2015
+
+
+
+####
+#### Clear workspace -----------------------------------------------------------
+####
+rm(list=ls(all=TRUE))
+
+
+
+####
+####  Load required libraries --------------------------------------------------
+####
+library(rstan)
+library(reshape2)
+library(plyr)
+
+
+####
+####  Set global simulation parameters -----------------------------------------
+####
+time.steps <- 600
+
+
+
+####
+####  Set climate driver id programmatically -----------------------------------
+####
+##  For HPC runs
+args <- commandArgs(trailingOnly = F)
+myargument <- args[length(args)]
+myargument <- sub("-","",myargument)
+climate_id <- as.numeric(myargument)
+
+##  For PC tests (comment out for HPC runs!!)
+# climate_id <- 1
+
+
+
+####
+####  Set some file paths, etc. ------------------------------------------------
+####
+##  For HPC runs
+datapath <- ""
+datapath2 <- ""
+resultspath <- ""
+
+##  For PC runs (comment out for HPC runs!!)
+# datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
+# datapath2 <- "../data/"
+# resultspath <- "../results/"
+
+
+
+####
+####  Get data -----------------------------------------------------------------
+####
[…]

### 2015-10-06 scripts/sim_sageAbundance_HPC.R
adds counter in loop for tracking

@@ -160,7 +160,7 @@ alpha_data <- outs[grep("alpha", outs$Parameter),]
 climeffs <- c("beta.1.", "beta.2.", "beta.3.", "beta.4.", "beta.5.")
 
 # Set number of chains and iterations
-nchains <- length(unique(outs$chain))
+nchains <- length(unique(outs$Chain))
 niters <- length(unique(outs$mcmc_iter))
 totsims <- nchains*niters
 
@@ -179,7 +179,8 @@ ex.arr[,1,] <- 1 # initialize all pixels at 1% cover
 counter <- 1 # for keeping track of full number of iterations
 for(i in 1:nchains){ # loop through chains
   
-  for(j in 1:niters){ # loop throug iterations within chain
+  for(j in 1:niters){ # loop through iterations within chain
+    
     chain <- i
     iter <- j
     int_mu <- as.numeric(subset(outs, Chain==chain & mcmc_iter==iter & Parameter=="int_mu")[,"value"])
@@ -194,7 +195,7 @@ for(i in 1:nchains){ # loop through chains
       dens.dep <- beta_mu*log(ex.arr[counter,t-1,])
       tmp.mu <- int_mu + dens.dep + sum(betas*Xtmp)
       tmp.mu <- exp(tmp.mu + eta)
-      tmp.out <- rpois(ncol(ex.arr[counter,,]), lambda = tmp.mu)
+      tmp.out <- rpois(pixels, lambda = tmp.mu)
       
       #Colonization
       zeros <- which(ex.arr[counter,t-1,]==0)
@@ -205,6 +206,8 @@ for(i in 1:nchains){ # loop through chains
       ex.arr[counter,t,] <- tmp.out
     } # next time step
     
+  counter <- counter+1  
+  
   } # next MCMC iteration
   
 } # next MCMC chain

### 2015-10-06 scripts/sim_sageAbundance_HPC.R
changes to 100 random parameter sets for computational tractability

@@ -37,7 +37,8 @@ library(plyr)
 ####
 ####  Set global simulation parameters -----------------------------------------
 ####
-time.steps <- 600
+totsims <- 100 # number of random parameter sets to simulate
+time.steps <- 500
 
 
 
@@ -51,7 +52,7 @@ myargument <- sub("-","",myargument)
 climate_id <- as.numeric(myargument)
 
 ##  For PC tests (comment out for HPC runs!!)
-# climate_id <- 1
+climate_id <- 1
 
 
 
@@ -64,9 +65,9 @@ datapath2 <- ""
 resultspath <- ""
 
 ##  For PC runs (comment out for HPC runs!!)
-# datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
-# datapath2 <- "../data/"
-# resultspath <- "../results/"
+datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
+datapath2 <- "../data/"
+resultspath <- "../results/"
 
 
 
@@ -159,11 +160,6 @@ alpha_data <- outs[grep("alpha", outs$Parameter),]
 # Make vector for extracting climate effect names in MCMC dataframe
 climeffs <- c("beta.1.", "beta.2.", "beta.3.", "beta.4.", "beta.5.")
 
-# Set number of chains and iterations
-nchains <- length(unique(outs$Chain))
-niters <- length(unique(outs$mcmc_iter))
-totsims <- nchains*niters
-
 # Get number of pixels to simulate
 pixels <- nrow(subset(growD, Year==1985))
 
@@ -171,46 +167,40 @@ pixels <- nrow(subset(growD, Year==1985))
 ex.arr <- array(NA, dim=c(totsims, time.steps, pixels))
 ex.arr[,1,] <- 1 # initialize all pixels at 1% cover
 
-
+# Get vectors for random parameter samples
+nchains <- length(unique(outs$Chain))
+niters <- length(unique(outs$mcmc_iter))
 
 ####
 ####  Main simulation loop -----------------------------------------------------
 ####
-counter <- 1 # for keeping track of full number of iterations
-for(i in 1:nchains){ # loop through chains
+for(i in 1:totsims){ # loop through chains
+  chain <- sample(nchains, 1) # sample random chain
[…]

### 2015-10-06 scripts/sim_sageAbundance_HPC.R
comments out stuff for HPC runs

@@ -52,7 +52,7 @@ myargument <- sub("-","",myargument)
 climate_id <- as.numeric(myargument)
 
 ##  For PC tests (comment out for HPC runs!!)
-climate_id <- 1
+# climate_id <- 1
 
 
 
@@ -65,9 +65,9 @@ datapath2 <- ""
 resultspath <- ""
 
 ##  For PC runs (comment out for HPC runs!!)
-datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
-datapath2 <- "../data/"
-resultspath <- "../results/"
+# datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
+# datapath2 <- "../data/"
+# resultspath <- "../results/"
 
 
 

### 2015-10-08 R/clim_proj_format_funcs.R
adds new function for formatting prjected climate into proper df for simulating

@@ -1,0 +1,74 @@
+##  Function to create a climate dataframe from CMIP5 projections
+##  Takes long dataframe of temperature and precipitation with yearly
+##  values by model and scenario. Returns a data frame for a particular
+##  model and scenario.
+
+#' Produce a climate dataframe for sageAbundance simulations
+#' @param tdata subset of temperature dataframe for a single model and scenario
+#' @param pdata subset of precipitation dataframe for a single model and scenario
+#' @param years vector of desired years in output
+#' @return clim_data formatted climate dataframe for simulating population model
+#' @author Andrew Tredennick \email{atredenn@gmail.com}
+format_climate <- function(tdata, pdata, years){
+  ### Add in 2 extra years to get lag precip != NA
+  orig_years <- years                           # save original year vector
+  years <- c(min(years)-2, min(years)-1, years) # add in 2 extra years
+  
+  ### Temperature
+  t_tmp <- subset(tdata, Year %in% years)   # get appropriate years
+  t_tmp <- t_tmp[ ,c("Year", "TmeanSpr")]   # subset relevant columns
+  colnames(t_tmp) <- c("year", "TmeanSpr2") # rename TmeanSpr column
+  t_tmp2 <- t_tmp                           # make duplicate dataframe
+  t_tmp2$year <- t_tmp2$year+1              # add year for lag temperature
+  colnames(t_tmp2)[2] <- "TmeanSpr1"        # rename for lag temperature
+  t_tmp <- merge(t_tmp,t_tmp2,all.x=T)      # merge together
+  
+  ### Precipitation
+  p_tmp <- subset(pdata, Year %in% years)      # get appropriate years
+  p_tmp <- p_tmp[ ,c("Year", "season", "ppt")] # subset relevant columns
+  
+  # Lag (t-1) cumulative precipitation
+  p_lag <- ddply(p_tmp, .(Year), summarise,
+                 pptLag = sum(ppt))      # create summary df with cumulative ppt
+  p_lag$Year <- p_lag$Year + 2           # make it lag by 2 years
+  colnames(p_lag) <- c("year", "pptLag") # rename for consistency
+  
+  # Summer precipitation
+  p_summer <- subset(p_tmp, season=="summer")     # subset for just summer
+  p_summer <- p_summer[ ,c("Year", "ppt")]        # get relevant columns
[…]

### 2015-10-08 R/clim_proj_format_funcs.R
adds intermodel comparison forecasts script

@@ -1,15 +1,16 @@
-##  Function to create a climate dataframe from CMIP5 projections
-##  Takes long dataframe of temperature and precipitation with yearly
-##  values by model and scenario. Returns a data frame for a particular
-##  model and scenario.
-
 #' Produce a climate dataframe for sageAbundance simulations
 #' @param tdata subset of temperature dataframe for a single model and scenario
 #' @param pdata subset of precipitation dataframe for a single model and scenario
 #' @param years vector of desired years in output
 #' @return clim_data formatted climate dataframe for simulating population model
-#' @author Andrew Tredennick \email{atredenn@gmail.com}
+#' @author Andrew Tredennick
 format_climate <- function(tdata, pdata, years){
+  ### Load 'plyr', if not already loaded
+#   if("plyr" %in% rownames(installed.packages()) == FALSE){
+#     stop("you need to install the 'plyr' package")
+#   }
+  if("package:plyr" %in% search() == FALSE) library(plyr)
+  
   ### Add in 2 extra years to get lag precip != NA
   orig_years <- years                           # save original year vector
   years <- c(min(years)-2, min(years)-1, years) # add in 2 extra years

### 2015-10-08 scripts/sim_sage_byYear_byModel.R

@@ -1,0 +1,123 @@
+##  Script to run sage abundance model through time, starting
+##  at the last observation. Requires time series of climate
+##  projections.
+
+##  Author:       Andrew Tredennick
+##  Email:        atredenn@gmail.com
+##  Date created: 10-08-2015
+
+
+### Clear the workspace
+rm(list=ls())
+
+
+
+####
+####  Set some global simulation settings --------------------------------------
+####
+parameter_reps <- 2
+clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
+
+
+####
+####  Load necessary libraries -------------------------------------------------
+####
+library(sageAbundance)
+library(plyr)
+library(reshape2)
+
+#TODO: get sage Abundance functions to load!! Until then...
+source("../R/clim_proj_format_funcs.R")
+
+
+
+####
+####  Read in percent cover file, climate projections, MCMC output -------------
+####
+obs_data <- read.csv("../data/wy_sagecover_subset_noNA.csv")
+obs_data <- subset(obs_data, Year>1984) # subsets out NA CoverLag values
+temp_projs <- readRDS("../data/CMIP5_yearly_project_temperature.RDS")
+ppt_projs <- readRDS("../data/CMIP5_yearly_project_precipitation.RDS")
+mcmc_outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
+
+
+
+####
+####  Subset out observed climate; get scaling mean and sd ------------------------
+####
+obs_clim <- obs_data[,c("Year",clim_vars)]
+obs_clim <- unique(obs_clim)
+obs_clim_means <- colMeans(obs_clim[,clim_vars])
+obs_clim_sds <- apply(obs_clim[,clim_vars], 2, sd)
+obs_clim_scalers <- data.frame(variable = clim_vars,
+                               means = obs_clim_means,
+                               sdevs = obs_clim_sds)
+
+####
+####  Define simulation function -----------------------------------------------
+####
+iterate_sage <- function(N, int, beta.dens, beta.clim, eta, weather){
+  dens.dep <- beta.dens*log(N)
+  clim.effs <- sum(beta.clim*weather)
+  Nout <- exp(int + dens.dep + clim.effs + eta)
+  return(Nout)
+}
+
+
+
+####
+####  Begin simulation set up --------------------------------------------------
[…]

### 2015-10-08 scripts/sim_sage_byYear_byModel.R
updates yearly forecast script

@@ -15,7 +15,7 @@ rm(list=ls())
 ####
 ####  Set some global simulation settings --------------------------------------
 ####
-parameter_reps <- 2
+parameter_reps <- 100
 clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
 
 
@@ -32,13 +32,15 @@ source("../R/clim_proj_format_funcs.R")
 
 
 ####
-####  Read in percent cover file, climate projections, MCMC output -------------
+####  Read in percent cover file, climate projections, MCMC output, K.data -----
 ####
 obs_data <- read.csv("../data/wy_sagecover_subset_noNA.csv")
 obs_data <- subset(obs_data, Year>1984) # subsets out NA CoverLag values
 temp_projs <- readRDS("../data/CMIP5_yearly_project_temperature.RDS")
 ppt_projs <- readRDS("../data/CMIP5_yearly_project_precipitation.RDS")
 mcmc_outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
+load("../results/Knot_cell_distances_subset.Rdata")
+K <- K.data$K
 
 
 
@@ -99,20 +101,37 @@ for(do_model in all_models){
     climate_now["TmeanSpr2"] <- (climate_now["TmeanSpr2"] - obs_clim_means["TmeanSpr2"])/obs_clim_sds["TmeanSpr2"]
     
     # Create storage matrix for population
-    n_save <- array(dim = c(parameter_reps, num_sims, nrow(last_obs)))
+    n_save <- array(dim = c(num_sims+1, parameter_reps, nrow(last_obs)))
+    n_save[1,,] <- last_obs$Cover # set first record to last observation
     
-    for(t in 1:num_sims){
+    for(t in 2:num_sims){
       
-      weather <- climate_now[t, clim_vars]
+      weather <- climate_now[t-1, clim_vars] #t-1 since clim data starts in 2012
       
       for(i in 1:parameter_reps){
         randchain <- sample(1:nchains, 1)
         randiter <- sample(1:niters, 1)
         params_now <- subset(mcmc_outs, chain==randchain & mcmc_iter==randiter)
+        alphas <- params_now[grep("alpha", params_now$Parameter), "value"]
+        int_mu <- params_now[grep("int_mu", params_now$Parameter), "value"]
+        beta_mu <- params_now[grep("beta_mu", params_now$Parameter), "value"]
[…]

### 2015-10-08 scripts/sim_sage_byYear_byModel.R
updates yearly forecast script with counter and looping prints

@@ -83,6 +83,7 @@ niters <- length(unique(mcmc_outs$mcmc_iter))
 ####
 ####  Begin looping: parameters within years within scenario within model ------
 ####
+count <- 1
 for(do_model in all_models){
   
   for(do_scenario in all_scenarios){
@@ -133,8 +134,11 @@ for(do_model in all_models){
     path <- "../results/yearlyforecasts/"
     saveRDS(n_save, paste0(path,file))
     
+    print(paste("Done with", do_model, do_scenario))
+    print(paste(count, "of", length(all_models)))
+    
   } # next scenario
-  
+  count <- count+1
 } # next model
 
 

### 2015-10-09 scripts/sim_sage_byYear_byModel.R
updates projection scenario vector

@@ -73,7 +73,7 @@ iterate_sage <- function(N, int, beta.dens, beta.clim, eta, weather){
 last_year <- max(obs_data$Year)
 last_obs <- subset(obs_data, Year == last_year)
 all_models <- unique(temp_projs$model)
-all_scenarios <- unique(temp_projs$scenario)
+all_scenarios <- c("rcp45", "rcp60", "rcp85")
 sim_years <- c((last_year+1):max(temp_projs$Year))
 num_sims <- length(sim_years)
 

### 2015-10-09 scripts/plot_temporal_forecasts.R
adds script for plotting the temporal forecasts

@@ -1,0 +1,157 @@
+##  Script to plot results from all model temporal runs
+
+##  Author: Andrew Tredennick
+##  Email: atredenn@gmail.com
+##  Date created: 10-09-2015
+
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(ggplot2)
+library(plyr)
+library(reshape2)
+
+
+
+####
+####  Read in model simulations, 1 by 1, and stack -----------------------------
+####
+setwd("../results/yearlyforecasts/")
+all_files <- list.files()
+
+longdf <- data.frame(model=NA,
+                     scenario=NA,
+                     param_set=NA,
+                     timestep=NA,
+                     cover=NA)
+
+for(i in 1:length(all_files)){
+  ttt <- readRDS(all_files[i])
+  ttt2 <- adply(ttt, c(1,3))
+  ttt3 <- melt(ttt2, id.vars = c("X1", "X2"))
+  ttt3$variable <- as.numeric(ttt3$variable)
+  colnames(ttt3) <- c("param_set", "id", "timestep", "cover")
+  ttt3agg <- ddply(ttt3, .(param_set, timestep), summarise,
+                   cover = mean(cover, rm.na=TRUE))
+  ttt3agg$model <- strsplit(all_files[i], "_")[[1]][1]
+  ttt3agg$scenario <- strsplit(all_files[i], "_")[[1]][2]
+  tmpout <- ttt3agg[,c("model", "scenario", "param_set", "timestep", "cover")]
+  longdf <- rbind(longdf, tmpout)
+}
+longdf <- longdf[2:nrow(longdf),]
+longall <- longdf
+longall$parammod <- paste(longall$param_set, longall$model)
+
+
+# longdf <- subset(longall, model %in% c("bcc-csm1-1", "ccsm4"))
+longdf <- longall
+meandf <- ddply(longdf, .(timestep, scenario), summarise,
+                avgcover=mean(cover, na.rm=TRUE),
+                upcover=quantile(cover, 0.95, na.rm=TRUE),
+                locover=quantile(cover, 0.05, na.rm=TRUE))
+meandf$timestep <- rep(c(2011:2098), each=3)
+
+
+
+####
+####  Read in a format observation data ----------------------------------------
+####
+obs_data <- read.csv("../../data/wy_sagecover_subset_noNA.csv")
[…]

### 2015-10-09 scripts/sim_sage_byYear_byModel.R

@@ -15,7 +15,7 @@ rm(list=ls())
 ####
 ####  Set some global simulation settings --------------------------------------
 ####
-parameter_reps <- 100
+parameter_reps <- 2
 clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
 
 
@@ -42,7 +42,22 @@ mcmc_outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
 load("../results/Knot_cell_distances_subset.Rdata")
 K <- K.data$K
 
+### Only keep models with rcp45, 60, and 85 scenarios
+allmods <- unique(temp_projs$model)
+keeps <- character(length(allmods))
+my_scens <- c("rcp45", "rcp60", "rcp85")
+for(i in 1:length(allmods)){
+  tmp <- subset(temp_projs, model==allmods[i])
+  tmp.scns <- unique(tmp$scenario)
+  flag <- length(which(my_scens %in% tmp.scns == FALSE))
+  ifelse(flag>0, keeps[i]<-"no", keeps[i]<-"yes")
+}
+modelkeeps <- data.frame(model=allmods,
+                         allscenarios=keeps)
+my_mods <- modelkeeps[which(modelkeeps$allscenarios=="yes"),"model"]
 
+temp_projs <- subset(temp_projs, model %in% my_mods)
+ppt_projs <- subset(ppt_projs, model %in% my_mods)
 
 ####
 ####  Subset out observed climate; get scaling mean and sd ------------------------
@@ -102,33 +117,34 @@ for(do_model in all_models){
     climate_now["TmeanSpr2"] <- (climate_now["TmeanSpr2"] - obs_clim_means["TmeanSpr2"])/obs_clim_sds["TmeanSpr2"]
     
     # Create storage matrix for population
-    n_save <- array(dim = c(num_sims+1, parameter_reps, nrow(last_obs)))
-    n_save[1,,] <- last_obs$Cover # set first record to last observation
+    n_save <- array(dim = c(parameter_reps, num_sims+1, nrow(last_obs)))
+    n_save[,1,] <- last_obs$Cover # set first record to last observation
     
-    for(t in 2:num_sims){
       
-      weather <- climate_now[t-1, clim_vars] #t-1 since clim data starts in 2012
-      
-      for(i in 1:parameter_reps){
-        randchain <- sample(1:nchains, 1)
-        randiter <- sample(1:niters, 1)
[…]

### 2015-10-09 docs/ms_sageAbundance.Rmd
updates ms figures

@@
 
-```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=T}
+```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=F}
 library(png)
@@
 
-```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=T}
+```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=F}
 library(png)
@@
 
-```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=T}
+```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=F}
 library(png)
@@
 
-```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=T}
+```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=F}
 library(png)
@@
 img <- readPNG("../results/clim_change_densities.png")
+grid.raster(img)
+```
+
+```{r figure_8,dependson="plot-options", fig.cap="Model forecasts under future climate for three IPCC emissions scenarios.
+Shaded regions show 90/% quantiles for simulations conducted using different sets of parameters from the MCMC output.
+Lines show mean trajectories.", eval=T}
+library(png)
+library(grid)
+img <- readPNG("../results/temporal_forecast.png")
 grid.raster(img)

### 2015-10-09 scripts/plot_temporal_forecasts.R

@@ -5,6 +5,10 @@
 ##  Date created: 10-09-2015
 
 
+##  Clear the workspace...
+rm(list=ls())
+
+
 ####
 ####  Load libraries -----------------------------------------------------------
 ####
@@ -22,23 +26,25 @@ all_files <- list.files()
 
 longdf <- data.frame(model=NA,
                      scenario=NA,
-                     param_set=NA,
-                     timestep=NA,
+                     paramset=NA,
+                     year=NA,
                      cover=NA)
 
 for(i in 1:length(all_files)){
   ttt <- readRDS(all_files[i])
-  ttt2 <- adply(ttt, c(1,3))
-  ttt3 <- melt(ttt2, id.vars = c("X1", "X2"))
-  ttt3$variable <- as.numeric(ttt3$variable)
-  colnames(ttt3) <- c("param_set", "id", "timestep", "cover")
-  ttt3agg <- ddply(ttt3, .(param_set, timestep), summarise,
-                   cover = mean(cover, rm.na=TRUE))
-  ttt3agg$model <- strsplit(all_files[i], "_")[[1]][1]
-  ttt3agg$scenario <- strsplit(all_files[i], "_")[[1]][2]
-  tmpout <- ttt3agg[,c("model", "scenario", "param_set", "timestep", "cover")]
+  ttt2 <- as.data.frame(apply(ttt, c(2,1), mean))
+  colnames(ttt2) <- paste0("params", 1:ncol(ttt2))
+  ttt2$year <- c(2011:2098)
+  ttt3 <- melt(ttt2, id.vars="year")
+  ttt3$model <- strsplit(all_files[i], "_")[[1]][1]
+  ttt3$scenario <- strsplit(all_files[i], "_")[[1]][2]
+  colnames(ttt3) <- c("year", "paramset", "cover", "model", "scenario")
+  tmpout <- ttt3[,c("model", "scenario", "paramset", "year", "cover")]
   longdf <- rbind(longdf, tmpout)
+  print(paste(i, "out of", length(all_files)))
 }
+
+
 longdf <- longdf[2:nrow(longdf),]
 longall <- longdf
 longall$parammod <- paste(longall$param_set, longall$model)
@@ -46,11 +52,10 @@ longall$parammod <- paste(longall$param_set, longall$model)
 
 # longdf <- subset(longall, model %in% c("bcc-csm1-1", "ccsm4"))
 longdf <- longall
-meandf <- ddply(longdf, .(timestep, scenario), summarise,
[…]

### 2015-10-09 scripts/sim_sage_byYear_byModel.R

@@ -15,7 +15,7 @@ rm(list=ls())
 ####
 ####  Set some global simulation settings --------------------------------------
 ####
-parameter_reps <- 2
+parameter_reps <- 50
 clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
 
 

### 2015-10-09 scripts/ms_scripts.R
increase dpi on spatial figs

@@ -6,7 +6,7 @@ rm(list=ls(all=TRUE))
 ####
 ####  Set some file paths, etc. ------------------------------------------------
 ####
-datapath <- "/Users/atredenn/Dropbox/sageAbundance_data/"
+datapath <- "../data/"
 knotpath <- "../results/"
 
 
@@ -35,9 +35,8 @@ growD <- subset(fullD, Year>1984) # get rid of NA lagcover years
 growD$Cover <- round(growD$Cover,0) # round for count-like data
 growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
 
-climD <- read.csv(paste(datapath,
-                        "/studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv",
-                        sep=""))
+climD <- read.csv(paste0(datapath,
+                        "/climate/DAYMET/FormattedClimate_WY_SA1.csv"))
 
 
 
@@ -249,7 +248,7 @@ tmp.theme3=theme(axis.ticks = element_blank(), axis.text = element_blank(),
                 axis.title=element_text(size=12),text=element_text(size=12),
                 legend.text=element_text(size=10))
 
-png("../results/obs_pred_spatial_subset.png", width = 8, height=8, units = "in", res=100)
+png("../results/obs_pred_spatial_subset.png", width = 8, height=8, units = "in", res=300)
 g1 <- ggplot(subset(sp.equil2, variable=="Observed"), aes(x=Lon, y=Lat))+
   geom_raster(aes(z=value, fill=value))+
   scale_fill_gradientn(colours=myPalette(200), name="% Cover", limit=c(0,25))+

### 2015-10-10 docs/ms_sageAbundance.Rmd
updates ms discussion

@@
 In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation (Fig.
-3).
+Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
 
-The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space (Fig. 6). 
-However, several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
+Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
 We decided to work at the native spatial resolution of the remote sensing data ($30\times30$ meters). 
@@
 In other words, climate only varied through time and not over space.
-This is a of course a limitation of our study, and helps to explain the weak effect of climate covariates when averaged over space (Fig. 3).
+This is a of course a limitation of our study, and helps to explain the weak effect of climate covariates when averaged over space (Fig. 2).
 In the future we plan to work at a lower spatial resolution but a larger spatial extent so that we can use spatially-explicit climate covariates that match the resolution of the percent cover data.
 
-These decisions ultimately affect our results, in positive and negative ways, but also show the flexibility of the approach we developed.
-The underlying model we described and its subsequent use for projecting population changes remains unchanged in the face of individual choices regarding spatial extent and resolution.
-Likewise, the approach can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
+The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space and time.
+We applied the fitted model to three IPCC emissions scenarios and forecast modest increases in sagebrush percent cover (Figs. 4, 5).
+The forecasted increases are primarily driven by a positive response to precipitation (Fig. 2).
+A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
+The descrepency between Homer et al.'s [@Homer2015] results and ours appears to be due to the climate projections used for forecasting future changes.
+The Homer et al.
+study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
+Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of 90% confidence (Fig. 5).
+
+Our spatiotemporal modeling framework is very flexible.
+It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
 As such datasets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
@@
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=T}
+Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=F}
 library(png)
@@
 ```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover, and prediction bias and standard deviation (C, D).
-All simulations used mean parameter values and no random year effects.", fig.height=8, eval=T}
+All simulations used mean parameter values and no random year effects.", fig.height=8, eval=F}
 library(png)
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=T}
+Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=F}
 library(png)
@@
 Shaded regions show 90/% quantiles for simulations conducted using different sets of parameters from the MCMC output.
-Lines show mean trajectories.", eval=T}
+Lines show mean trajectories.", eval=F}
 library(png)

### 2015-10-10 scripts/fit_sageAbundance_loyo.R

@@ -1,0 +1,148 @@
+##  Script for leave-one-year-out cross validation of the
+##  sagebrush abundance model. Here we use a simplified version
+##  with no spatial random effect to reduce computational overhead.
+
+##  Author:       Andrew Tredennick
+##  Email:        atredenn@gmail.com
+##  Date created: 10-09-2015
+
+##  Clear the workspace...
+rm(list=ls())
+
+
+####
+####  Set MCMC configurations --------------------------------------------------
+####
+niters <- 2000
+nburns <- 1000
+nchains <- 3
+
+
+####
+####  Get leave out year id programmatically -----------------------------------
+####
+##  For HPC runs
+args <- commandArgs(trailingOnly = F)
+myargument <- args[length(args)]
+myargument <- sub("-","",myargument)
+year_id <- as.numeric(myargument)
+
+##  For PC tests (comment out for HPC runs!!)
+year_id <- 1
+
+
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(rstan)
+library(ggmcmc)
+
+
+
+####
+####  Read in observation data -------------------------------------------------
+####
+obs_data <- read.csv("../wy_sagecover_subset_noNA.csv")
+
+# Get data structure right
+growD <- subset(obs_data, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Leave out a year
+all_years <- unique(growD$Year)
+growD <- subset(growD, Year != all_years[year_id])
+
+
+
+####
+####  Write the STAN model -----------------------------------------------------
+####
+model_string <- "
+data{
+  int<lower=0> nobs; // number of observations
+  int<lower=0> nyrs; // number of years
+  int<lower=0> ncovs; // number of climate covariates
+  int<lower=0> yrid[nobs]; // year id
+  int y[nobs]; // observation vector
+  vector[nobs] lag; // lag cover vector
+  matrix[nobs,ncovs] X; // climate covariates
+}
+parameters{
[…]

### 2015-10-12 docs/appendix_a.Rmd
adds appendix with CMIP5 model table

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{X\arabic{figure}}

Appendix X: Table of CMIP5 models used in this study
=======================================

\scriptsize

| Modeling Center | Model | Institution |
| --------------- | ----- | ----------- |
| BCC | BCC-CSM1.1 | Beijing Climate Center, China Meteorological Administration | 
| BCC | BCC-CSM1.1(m) | Beijing Climate Center, China Meteorological Administration | 
| CCCma | CanAM4 | Canadian Centre for Climate Modelling and Analysis |
| CCCma | CanCM4 | Canadian Centre for Climate Modelling and Analysis |
| CCCma | CanESM2 | Canadian Centre for Climate Modelling and Analysis |
| CCCMC | CMCC-CESM | Centro Euro-Mediterraneo per I Cambiamenti Climatici |
| CCCMC | CMCC-CM | Centro Euro-Mediterraneo per I Cambiamenti Climatici |
| CCCMC | CMCC-CMS | Centro Euro-Mediterraneo per I Cambiamenti Climatici |
| CNRM-CERFACS | CNRM-CM5 | Centre National de Recherches Meteorologiques / |
|  |  | Centre Europeen de Recherche et Formation Avancees en Calcul Scientifique |
| CNRM-CERFACS | CNRM-CM5-2 | Centre National de Recherches Meteorologiques / |
|  |  | Centre Europeen de Recherche et Formation Avancees en Calcul Scientifique |
| COLA and NCEP | CFSv2-2011 | Center for Ocean-Land-Atmosphere Studies and |
|  |  | National Centers for Environmental Prediction |
| CSIRO-BOM | ACCESS1.0 | CSIRO (Commonwealth Scientific and Industrial Research |
|  |  | Organisation, Australia), and BOM (Bureau of Meteorology, Australia) |

### 2015-10-12 docs/ms_sageAbundance.Rmd
adds appendix with CMIP5 model table

@@
 We downloaded the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
-The CMIP5 provides projections from a suite of global circulations models.
+The CMIP5 provides projections from a suite of global circulations models (Table X1).
 For our purposes, we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
@@
 Compute, storage and other re-sources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
+We acknowledge the World Climate Research Programme's Working Group on Coupled Modelling, which is responsible for CMIP, and we thank the climate modeling groups (listed in Table X1) for producing and making available their model output.
+For CMIP the U.S.
+Department of Energy's Program for Climate Model Diagnosis and Intercomparison provides coordinating support and led development of software infrastructure in partnership with the Global Organization for Earth System Science Portals.
 

### 2015-10-14 scripts/fit_sageAbundance_loyo_lm.R
adds glm no random effects cross validation

@@ -1,0 +1,81 @@
+##  Script for leave-one-year-out cross validation of the
+##  sagebrush abundance model. Here we use a simplified version
+##  with no spatial random effect to reduce computational overhead.
+##  We use a frequentist approach (lm).
+
+##  Author:       Andrew Tredennick
+##  Email:        atredenn@gmail.com
+##  Date created: 10-09-2015
+
+##  Clear the workspace...
+rm(list=ls())
+
+
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(ggplot2)
+library(reshape2)
+library(plyr)
+library(lme4)
+
+
+
+####
+####  Read in observation data -------------------------------------------------
+####
+obs_data <- read.csv("../data/wy_sagecover_subset_noNA.csv")
+
+# Get data structure right
+growD <- subset(obs_data, Year>1984) # get rid of NA lagcover years
+growD$Cover <- round(growD$Cover,0) # round for count-like data
+growD$CoverLag <- round(growD$CoverLag,0) # round for count-like data
+
+# Get year information
+all_years <- unique(growD$Year)
+num_years <- length(all_years)
+
+"%w/o%" <- function(x, y) x[!x %in% y] # x without y
+
+####
+####  Loop over leave out years and predict left out year ----------------------
+####
+rmse <- numeric(num_years)
+counter <- 1
+for(omit_year in all_years){
+  tmpD <- subset(growD, Year != omit_year)
+  # Remove pixels that have 0s in the time series
+  zids <- tmpD[which(tmpD$Cover==0),"ID"]
+  zids <- unique(c(zids, tmpD[which(tmpD$CoverLag==0),"ID"]))
+  pixels_to_keep <- tmpD$ID %w/o%  zids
+  modelD <- tmpD[which(tmpD$ID %in% pixels_to_keep),]
+  obs_clim_means <- colMeans(modelD[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")])
+  obs_clim_sds <- apply(modelD[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")], 2, sd)
+  
[…]

### 2015-10-15 scripts/fit_sageAbundance_loyo_lm.R
adds complete fit for in-sample RMSE comparison

@@ -38,6 +38,29 @@ num_years <- length(all_years)
 
 "%w/o%" <- function(x, y) x[!x %in% y] # x without y
 
+
+
+####
+####  Fit and predict total in sample observations -----------------------------
+####
+zids <- growD[which(growD$Cover==0),"ID"]
+zids <- unique(c(zids, growD[which(growD$CoverLag==0),"ID"]))
+pixels_to_keep <- growD$ID %w/o%  zids
+modelD <- growD[which(growD$ID %in% pixels_to_keep),]
+
+X <- modelD[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
+X <- scale(X, center = TRUE, scale = TRUE)
+modelD[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")] <- X
+# tmp.glm <- glmer(y~lag+X+(1|year), family="poisson")
+tmp.lm <- glm(Cover~log(CoverLag)+pptLag+ppt1+ppt2+TmeanSpr1+TmeanSpr2, 
+              family="poisson", data=modelD)
+all.pred <- predict(tmp.lm, type="response")
+sq.error.all <- as.numeric((all.pred-modelD$Cover)^2)
+rmse.all <- sqrt(mean(sq.error.all))
+
+
+
+
 ####
 ####  Loop over leave out years and predict left out year ----------------------
 ####

### 2015-10-15 docs/ms_sageAbundance.Rmd
adds new map figure and updates results text

@@
 Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B.
-Hooten\textsuperscript{2,3,4,5}, Cameron L.
-Aldridge\textsuperscript{5,6,7}, Colin G.
-Homer\textsuperscript{8}, David T.
+Hooten\textsuperscript{2,3,4}, Cameron L.
+Aldridge\textsuperscript{5}, Colin G.
+Homer\textsuperscript{6}, David T.
 Iles\textsuperscript{1}, Andrew Kleinhesselink\textsuperscript{1}, Eric LaMalfa\textsuperscript{1}, Rebecca Mann\textsuperscript{1}, and Peter B.
@@
 
-\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\*
-\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\*
-\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\*
-\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}} \\*
-\textit{\small{\textsuperscript{5}Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80523 USA}} \\*
-\textit{\small{\textsuperscript{6}Department of Ecosystem Science and Sustainability and Natural Resource Ecology Laboratory, Colorado State University, Fort Collins, CO 80523 USA}} \\*
-\textit{\small{\textsuperscript{7}U.S. Geological Survey, Fort Collins Science Center, Fort Collins, CO 80526 USA}} \\*
-\textit{\small{\textsuperscript{8}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD 57198 USA}}
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\
+\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{5}Natural Resource Ecology Laboratory and Department of Ecosystem Science and Sustainability, Colorado State University, Fort Collins, CO, in cooperation with US Geological Survey, Fort Collins Science Center, Fort Collins, CO}}\\
+\textit{\small{\textsuperscript{6}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD 57198 USA}}
 
@@
 Here we advance spatiotemporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
-We use a 27 year time series of sagebrush (*Artemesia* spp.) percent cover over a $5\times2.5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
+We use a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
 We then use the model to forecast future population states (percent cover) under projected climate scenarios.
-Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to spring temperature.
+Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to previous year's precipitation.
 However, gains in cover are not homogeneous in space.
@@
 
-*Key words: population model, climate change, forecasting, spatiotemporal model, remote sensing, sagebrush*, Artemesia, *dimension reduction*
+*Key words: population model, climate change, forecasting, spatiotemporal model, remote sensing, sagebrush*, Artemisia, *dimension reduction*
 
@@
 Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales.
-SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance.
+In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance.
 Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
-Longterm trends in species' occurance can be detected and correlated with climate variables, but these may or may not be representative of a species' latent sensitivity climate.
-Likewise, since SDMs typically rely on occurance data, their projections of habitat suitability or probability of occurance provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
-Despite these limitations, SDMs remain one of the few ways to make population forecasts at large spatial extents.
+Longterm trends in species' occurence can be detected and correlated with climate variables, but these may or may not be representative of a species' sensitivity to climate.
+Likewise, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
+Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.
 
-It seems the ideal tool for forecasting future population states under climate change is a large scale dynamic population model.
+It seems the ideal tool for forecasting future population states under climate change is a large-scale dynamic population model.
 Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary to fit high-dimensional spatial models.
-Fortunately, two recent technological advances put us in the position to develop large scale population models.
-First, remotely-sensed (RS) time series of species-specific plant cover provide data on interannual population dynamics theat can be correlated with yearly climate fluctuations at large spatial scales.
-RS time series have been used to detect long term trends in population abundance, but they have yet to be used to drive a dynamic model.
+Fortunately, two recent technological advances put us in the position to develop large-scale population models.
+First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
+RS time series have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic model.
 Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
-Here, too, these methods have yet to be use in a dynamic setting.
+However, these methods have yet to be use in a dynamic setting.
 
 Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent growth.
-Our approach leverages a unique remotely-sensed time series of *Artemesia* species (sagbrush) to fit a spatiotemporal model of plant population dynamics at the landscape scale.
+Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatiotemporal model of plant population dynamics at the landscape scale.
 We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
@@
 
+*PARAGRAPH FROM ANDY ON SAGEBRUSH AND CLIMATE; EITHER HERE OR IN INTRODUCTION*
+
 #### Climate covariates
 Our approach is based on modeling how changes in plant populations relate to climate variables.
+We downloaded daily historic weather data for our study site from the NASA Daymet data set (*available online*)[^link].
+The Daymet weather data is interpolated between coarse observation units and does capture some spatial variation.
+However, instead of further downscaling the 1 kilometer resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
+We calculated five climate variables from the Daymet data for the time period of our remotely sensed data (1984 to 2011).
+
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
-The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}, (2) year \emph{t} fall through summer precipitation, (3) year \emph{t}-1 fall through summer precipitation, (4) year \emph{t} average spring temperature, and (5) year \emph{t}-1 average spring temperature.
+The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t} (*lagPpt*), (2) year \emph{t} fall through summer precipitation (*ppt2*), (3) year \emph{t}-1 fall through summer precipitation (*ppt1*), (4) year \emph{t} average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-1 average spring temperature (*TmeanSpr1*).
 We selected these variables based on previous studies, even though not all emerge as important predictors in our model.
 Several model selection techniques exist that could be applied to our approach [@Hooten2014], but we avoid model selection here to maximize focus on the general approach.
-
-We downloaded daily historic weather data from the NASA Daymet data set (*available online*)[^link].
-The Daymet weather data is interpolated between coarse observation units and does capture some spatial variation.
-However, instead of further downscaling the 1 kilometer resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
-We calculated the four climate variables outline above from the PRISM data for the time period of our remotely sensed data (1984 to 2011).
 
@@
 The CMIP5 provides projections from a suite of global circulations models (Table X1).
-For our purposes, we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
+To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
 We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
-To keep the temporal variability from observed climate consistent, for climate change scenarios we adjusted observed climate according to projected changes.
+To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
 For temperature we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
@@
 [^link3]: http://tntcat.iiasa.ac.at/RcpDb/
+
+In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make annual forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
+We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.
 
@@
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
-The intercept, $\beta{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
+The intercept, $\beta_{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
 We chose to use a Poisson likelihood because integer percent cover values can be considered a form of count data.
 We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$).
+There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we dealt with the small number of zero percent cover observations.
 
@@
 ### Accomodating zeros
-Since our process model (Eq. ) includes a log transformation of the observations ($\text{log}(y_{t-1}$), our model does not accomodate zeros. 
+Since our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$), our model does not accomodate zeros. 
 Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47). 
@@
 However, the Stan sampler is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
+Each MCMC chain was initialized with unique parameter values and then run in parallel on the Utah State University supercomputer.
+Model fitting required five days.
 We assessed convergence visually and made sure all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
@@
 ####Simulating the model
-We performed three sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, and (3) to forecast future equilibrium population states under projected climate change.
+We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make yearly forecasts of sagebrush cover starting the final year of our observations and ending in 2098.
 Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
@@
 We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output, after discarding and initial 100 time steps.
-For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step.
+For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.
+For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
 
@@
 -------
+Averaged across all GCMs, precipitation and temperature in our study area are projected to increase by 2050-2100 relative to observed climate from 1950-2000, with the magnitude depending on the RCP scenario (Table 1).
+Trajectories of our climate covariates from GCM projections, used to make yearly forecasts, show similar trends (Fig.
+2).
+GCM projections are variable for our study area, highlighting the importance of making forefasts based on ensemble projections.
+
 All parameters in our model converged on stable posterior distributions (Appendix A).
-Only the *lagPpt* climate covariate can be considered important at 90% confidence, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 2).
-The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. 3).
+Only the *lagPpt* climate covariate can be considered important at 90% confidence, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 3).
+The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. B#).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
@@
 This is due to the fact that the model spatial effect in low cover regions can be so strong that it prevents an increase from zero percent cover even in favorable climate years.
-The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed cover in the previous year for each pixel (Fig. 5)
+The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel (Fig. 5)
 
-When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 6, 7).
-The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 6).
-Thus, our model predicts an increase in sagebrush cover heterogeneity since we do not lose low cover pixels and gain higher cover pixels (Fig. 7).
-Without the spatial random effect we would have observed only a directional shift in the distribution of cover, not an expansion as seen in Fig. 7.
+When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
+The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
+Thus, our model predicts an increase in sagebrush cover heterogeneity since we do not lose low cover pixels and gain higher cover pixels (Fig. 5).
+For the yearly forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
+6).
+Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
+This is expected given the positive effect of precipitation on sagebrush cover change (Fig. 3).
 
@@
 -----------
-Landscape scale population models for plant species have been elusive.
+Landscape-scale population models for plant species have been elusive.
 Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
@@
 First, we used species-specific remote sensing data as a response variable.
-Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent. 
+Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
 In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
@@
 \begin{table}[h]
-\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
-average projections.}
+\protect\caption{\label{tab:Changes-in-climate}Projected changes in temperature and precipitation at our study area from CMIP5 average GCM projections for 2050-2100 relative to average temperature and precipitation from 1950-2000.}
 \begin{tabular}{llllll}
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the study area in southwester Wyoming and a snapshot of the percent cover data in 1984 for each site (a and b).
-Green shading on the map shows projected distribution of sagebrush from Kuchler (19xx).
-Latitude and longitude are in decimal degrees.", fig.height=6, eval=FALSE}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5$\\times$5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
 library(png)
 library(grid)
-img <- readPNG("components/figure/StudyAreasMap_1and2.png")
+img <- readPNG("components/figure/studyarea_map.png")
 grid.raster(img)
@@
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=F}
+Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=T}
 library(png)

### 2015-10-15 scripts/make_study_map.R

@@ -1,0 +1,131 @@
+##  Script to make study area map
+
+rm(list=ls()) #wipe workspace clean
+
+require(ggplot2)
+require(ggthemes)
+require(raster)
+require(maptools)
+require(shapefiles)
+require(sp)
+require(rgdal)
+library(colorRamps)
+
+
+#Get multiplot function
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  require(grid)
+  
+  # Make a list from the ... arguments and plotlist
+  plots <- c(list(...), plotlist)
+  
+  numPlots = length(plots)
+  
+  # If layout is NULL, then use 'cols' to determine layout
+  if (is.null(layout)) {
+    # Make the panel
+    # ncol: Number of columns of plots
+    # nrow: Number of rows needed, calculated from # of cols
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                     ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+  
+  if (numPlots==1) {
+    print(plots[[1]])
+    
+  } else {
+    # Set up the page
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+    
+    # Make each plot, in the correct location
+    for (i in 1:numPlots) {
+      # Get the i,j matrix positions of the regions that contain this subplot
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+      
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+
+####
+####  Read in and format map data ----------------------------------------------
+####
+gpclibPermit()
+US.shp <- readShapeSpatial("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/GIS/US_Shapefiles/Western_USt.shp")
+US.shp@data$id <- rownames(US.shp@data)
+US.pts <- fortify(US.shp, region="id")
+WY.shp <- readShapeSpatial("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/GIS/US_Shapefiles/Wyoming.shp")
+WY.shp@data$id <- rownames(WY.shp@data)
+WY.pts <- fortify(WY.shp, region="id")
+
+
+
+####
[…]

### 2015-10-16 scripts/plot_projected_weather.R
adds weather plotting script

@@ -1,0 +1,77 @@
+##  Script to plot GCM weather projections through time
+
+##  Author: Andrew Tredennick
+##  Email: atredenn@gmail.com
+##  Date created: 10-09-2015
+
+
+##  Clear the workspace...
+rm(list=ls())
+
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(ggplot2)
+library(plyr)
+library(reshape2)
+library(gridExtra)
+
+
+
+####
+####  Read in GCM weather ------------------------------------------------------
+####
+precip <- readRDS("../data/CMIP5_yearly_project_precipitation.RDS")
+precip <- subset(precip, scenario!="rcp26" & season!="summer")
+temp <- readRDS("../data/CMIP5_yearly_project_temperature.RDS")
+temp <- subset(temp, scenario!="rcp26")
+
+ppt_agg <- ddply(precip, .(Year, scenario, season), summarise,
+                 avg_value = mean(ppt),
+                 upval=quantile(ppt, 0.95, na.rm=TRUE),
+                 loval=quantile(ppt, 0.05, na.rm=TRUE))
+
+temp_agg <- ddply(temp, .(Year, scenario, season), summarise,
+                 avg_value = mean(TmeanSpr),
+                 upval=quantile(TmeanSpr, 0.95, na.rm=TRUE),
+                 loval=quantile(TmeanSpr, 0.05, na.rm=TRUE))
+
+
+g1 <- ggplot(ppt_agg)+
+  geom_ribbon(aes(x=Year, ymin=loval, ymax=upval, fill=scenario),
+              alpha=0.5)+
+  geom_line(aes(x=Year, y=avg_value, color=scenario))+
+  stat_smooth(method="lm", se=FALSE, aes(x=Year, y=avg_value), color="black", linetype=2)+
+  geom_vline(aes(xintercept=2011))+
+  scale_fill_manual(values=c("tan","coral","darkred"), 
+                    name="IPCC \nScenario",
+                    labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
+  scale_color_manual(values=c("tan","coral","darkred"), 
+                     name="IPCC \nScenario",
+                     labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
+  ylab("Fall through Spring Ppt. (mm)")+
+  ggtitle("A) GCM Precipitation Projections")+
+  theme_bw()
+
[…]

### 2015-10-16 scripts/plot_projected_weather.R
updates colors to bw for weather plots

@@ -37,6 +37,8 @@ temp_agg <- ddply(temp, .(Year, scenario, season), summarise,
                  upval=quantile(TmeanSpr, 0.95, na.rm=TRUE),
                  loval=quantile(TmeanSpr, 0.05, na.rm=TRUE))
 
+# cols <- c("tan","coral","darkred")
+cols <- c("grey10", "grey30", "grey60")
 
 g1 <- ggplot(ppt_agg)+
   geom_ribbon(aes(x=Year, ymin=loval, ymax=upval, fill=scenario),
@@ -44,10 +46,10 @@ g1 <- ggplot(ppt_agg)+
   geom_line(aes(x=Year, y=avg_value, color=scenario))+
   stat_smooth(method="lm", se=FALSE, aes(x=Year, y=avg_value), color="black", linetype=2)+
   geom_vline(aes(xintercept=2011))+
-  scale_fill_manual(values=c("tan","coral","darkred"), 
+  scale_fill_manual(values=cols, 
                     name="IPCC \nScenario",
                     labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
-  scale_color_manual(values=c("tan","coral","darkred"), 
+  scale_color_manual(values=cols, 
                      name="IPCC \nScenario",
                      labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
   ylab("Fall through Spring Ppt. (mm)")+
@@ -60,10 +62,10 @@ g2 <- ggplot(temp_agg)+
   geom_line(aes(x=Year, y=avg_value, color=scenario))+
   stat_smooth(method="lm", se=FALSE, aes(x=Year, y=avg_value), color="black", linetype=2)+
   geom_vline(aes(xintercept=2011))+
-  scale_fill_manual(values=c("tan","coral","darkred"), 
+  scale_fill_manual(values=cols, 
                     name="IPCC \nScenario",
                     labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
-  scale_color_manual(values=c("tan","coral","darkred"), 
+  scale_color_manual(values=cols, 
                      name="IPCC \nScenario",
                      labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
   ylab("Mean Spring Temperature (deg. C)")+

### 2015-10-17 docs/ms_sageAbundance.Rmd
updates ms with bw figures

@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5$\\times$5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 x 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
 library(png)
@@
 
+```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather projections for our climate covariates.", eval=T}
+library(png)
+library(grid)
+img <- readPNG("components/figure/weather_projections.png")
+grid.raster(img)
+```
+
 ```{r figure_2,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time t is important (shown in blue; 90% CI does not overlap zero).", fig.height=4, eval=T}
+Only cumulative precipitation at time t is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=T}
 library(png)
@@
 Shaded regions show 90/% quantiles for simulations conducted using different sets of parameters from the MCMC output.
-Lines show mean trajectories.", eval=F}
+Lines show mean trajectories.", eval=T}
 library(png)

### 2015-10-17 scripts/ms_scripts.R

@@ -170,7 +170,7 @@ png("../results/post_climate_covariates.png", height=4, width=4, units="in", res
 ggplot(clim_posts, aes(x=value, fill=col))+
   geom_vline(aes(xintercept=0), linetype=2)+
   geom_density(alpha=0.5,size=0.5, col=NA, adjust=4)+
-  scale_fill_manual(values=c("grey", "blue", "red"))+
+  scale_fill_manual(values=c("grey70", "black", "red"))+
   ylab("Posterior density")+
   xlab("Standardized coefficient value")+
   facet_grid(rank~., labeller=my_labeller)+

### 2015-10-19 docs/ms_sageAbundance.Rmd
adds Andy's climate paragraph

@@
 We then use the model to forecast the impacts of climate change on sagebrush percent cover.
+
+Sagebrush ecosystems offer an ideal testing ground for our new modeling approach.
+Their distribution is widespread, sagebrush species are sensitive to climate, and forecasts of future sagebrush ecosystems are in high demand due to the precarious status of sage-grouse.
+Nearly all species distribution models predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
+Models suggest that warming could drive sagebrush losses at low elevations and southern parts of its distribution by making snowmelt earlier and increasing evaporation, leading to soil water deficits in the spring [@Schlaepfer2012; @Schlaepfer2014].
+In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
+Thus, the response of sagebrush to climate will likely vary in space depending on elevation, among other factors.
+Direct observations of individual plants and cover on experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season and an increase in sagebrush growth and cover [@Perfors2003; @Harte2015].
+
 The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but we are most excited about its application in cases where remotely-sensed estimates exist.
@@
 
-*PARAGRAPH FROM ANDY ON SAGEBRUSH AND CLIMATE; EITHER HERE OR IN INTRODUCTION*
-
 #### Climate covariates
@@
 The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t} (*lagPpt*), (2) year \emph{t} fall through summer precipitation (*ppt2*), (3) year \emph{t}-1 fall through summer precipitation (*ppt1*), (4) year \emph{t} average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-1 average spring temperature (*TmeanSpr1*).
-We selected these variables based on previous studies, even though not all emerge as important predictors in our model.
+We selected these variables based on previous studies (see *Introduction*), even though not all emerge as important predictors in our model.
 Several model selection techniques exist that could be applied to our approach [@Hooten2014], but we avoid model selection here to maximize focus on the general approach.
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2014].
-However, given the relative largeness of our dataset and computational limitations, we defined kernels around 271 knots (Figure 2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix A).
+However, given the relative largeness of our dataset and computational limitations, we defined kernels around 121 knots (Fig.
+X#) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix X).
 The Bayesian posterior distribution can be expressed as
@@
 We fit the model in R [@R2013] using the 'No-U-Turn' MCMC sampler in Stan [@stan2014] and the RStan package [@rstan2014].
-We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1 000 iterations each, after discarding an initial 1 000 iterations as burn in.
+We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
 Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
@@
 
-All parameters in our model converged on stable posterior distributions (Appendix A).
+All parameters in our model converged on stable posterior distributions (Appendix X).
 Only the *lagPpt* climate covariate can be considered important at 90% confidence, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 3).
-The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. B#).
+The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. X#).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
@@
 Predictions are most confident, even if biased, in low percent cover pixels (Fig. 4D).
-This is due to the fact that the model spatial effect in low cover regions can be so strong that it prevents an increase from zero percent cover even in favorable climate years.
-The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel (Fig. 5)
+The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
+When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = #.##.
 
@@
 Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
+This means our model is statistical rigorous by explicitly accounting for spatial correlation.
+While important statistically, ecologically it means we can be confident in our estimates of climate effects on sagebrush cover change.
+
+As expected, the amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
+Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance, we found little evidence for a temperature effect (Fig. 2).
+Our estimates of temperature effects broadly overlapped zero, but they did trend positive.
+Thus,
 
@@
 ATT was supported by an NSF Postdoctoral Research Fellowship in Biology (DBI-1400370) and AK was supported by an NSF Graduate Research Fellowship.
-We are grateful to Mary LASTNAME at USGS EROS for exctracting the dataset used in this paper.
-Compute, storage and other re-sources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
+We are grateful to Debra K.
+Meyer at USGS EROS for exctracting the dataset used in this paper.
+Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
 We acknowledge the World Climate Research Programme's Working Group on Coupled Modelling, which is responsible for CMIP, and we thank the climate modeling groups (listed in Table X1) for producing and making available their model output.

### 2015-10-19 docs/ms_sageAbundance.Rmd
updates ms discussion

@@
 #### Remotely-sensed time series
-To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
+To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming[^link0] [@Homer2012].
 Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
@@
 Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).
+
+[^link0]: *link to USGS data repo*
 
@@
 For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
-We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output, after discarding and initial 100 time steps.
+We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding and initial 100 time steps.
 For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.
 For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
+We used the posterior mean of each parameter for all simulations, except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
+The random year effects were not used in any simulation.
 
@@
 -----------
-Landscape-scale population models for plant species have been elusive.
+Landscape-scale population models for plant species have been elusive and challenging [@Schurr2012a; @Merow].
 Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
@@
 Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data.
-In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
+In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location (Tredennick and Adler, in review).
 Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
 This means our model is statistical rigorous by explicitly accounting for spatial correlation.
-While important statistically, ecologically it means we can be confident in our estimates of climate effects on sagebrush cover change.
+While important statistically, ecologically it means we can be confident in our estimates of climate effects on sagebrush cover change, given the data in-hand.
+It also means we can make spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).
 
 As expected, the amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
-Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance, we found little evidence for a temperature effect (Fig. 2).
-Our estimates of temperature effects broadly overlapped zero, but they did trend positive.
-Thus,
+Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig.
+2).
+Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
+Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases at higher elevations.
+As seen when comparing our temporal forecasts (Fig. 5) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
+Although our estimates of climate effects were small, they are in agreement with previous studies, indicating our model does an adequate job of capturing the ecological process and the link between sagebrush growth and climate.
+
+The relatively weak climate effects means we should expect small changes when we apply the model to projected climate scenarios.
+Our model projects modest increases in sagebrush cover under all RCP scenarios (Figs. 4,5).
+The forecasted increases are primarily driven by a positive response to precipitation (Fig. 2), but weak effects of temperature also influenced results.
+A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
+The descrepency between Homer et al.'s [-@Homer2015] results and ours is likely due to the climate projections used for forecasting future changes.
+The Homer et al.
+study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
+Thus, the decline in precipitation led to a decline in sagebrush cover.
+Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of 90% confidence (Fig. 5).
+
+We want to stress that there is no clear answer to what climate model to use when making forecasts.
+Here we chose to use all possible projections.
+However, much like weather forecasters [@Bauer2015], ecological forecasters should use expert judgement to choose a climate projection.
+If a certain GCM tends to better predict weather in a specific study area, then that GCM should be given more weight than others.
+We plan to outline such an approach in our future work.
 
@@
 
-The major advantage of our approach is that, once fit, the model can be used to project future changes in population abundance (here, percent cover) in space and time.
-We applied the fitted model to three IPCC emissions scenarios and forecast modest increases in sagebrush percent cover (Figs. 4, 5).
-The forecasted increases are primarily driven by a positive response to precipitation (Fig. 2).
-A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
-The descrepency between Homer et al.'s [@Homer2015] results and ours appears to be due to the climate projections used for forecasting future changes.
-The Homer et al.
-study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
-Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of 90% confidence (Fig. 5).
-
-Our spatiotemporal modeling framework is very flexible.
+Conclusion
+----------
+We have introduced a new approach to fitting and simulating population models at large spatial extents.
+But, this is just a start.
+More work is needed to apply our approach over even larger spatial extents, and at higher spatial resolution.
+Once working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.
+Given the dependence of forecasts on the weather projections used, we urge ecologists to adopt methods from weather forecasters to up- and down-weight GCMs based on study area-specific reliability.
+Lastly, it will be important to include the effects of species interactions [@Clark2014].
+Fortunately, our spatiotemporal modeling framework is very flexible.
 It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
-As such datasets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
+The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or and entirely different functional form can be used depending on the ecological context.
+As such time-series data sets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
 
@@
 We are grateful to Debra K.
-Meyer at USGS EROS for exctracting the dataset used in this paper.
+Meyer at USGS EROS for exctracting the data set used in this paper.
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 x 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
 library(png)
@@
 
-```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather projections for our climate covariates.", eval=T}
+```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area.", eval=T}
 library(png)
@@
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time t is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=T}
+Only cumulative precipitation at time \\emph{t} is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=T}
 library(png)
@@
 ```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover, and prediction bias and standard deviation (C, D).
-All simulations used mean parameter values and no random year effects.", fig.height=8, eval=F}
+All simulations used mean parameter values and no random year effects.", fig.height=8, eval=T}
 library(png)
@@
 Top panels shows equilibrium cover based on simulations using observed cliamte.
-Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.", fig.height=8, eval=F}
+Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.
+Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the model.
+We used mean parameter estimates.", fig.height=8, eval=T}
 library(png)
@@
 ```{r figure_8,dependson="plot-options", fig.cap="Model forecasts under future climate for three IPCC emissions scenarios.
-Shaded regions show 90/% quantiles for simulations conducted using different sets of parameters from the MCMC output.
-Lines show mean trajectories.", eval=T}
+Shaded regions show 90% quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
+Lines show mean trajectories.
+Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
+Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=T}
 library(png)

### 2015-10-19 docs/ms_sageAbundance.Rmd
adds rmse and correlation for predicitons

@@
 
-Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales.
-In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance.
+Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
+In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009].
 Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
@@
 
-It seems the ideal tool for forecasting future population states under climate change is a large-scale dynamic population model.
+It seems the ideal tool for forecasting future population states under climate change is a large-scale dynamic population model [@Schurr2012a; @Merow2014].
 Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary to fit high-dimensional spatial models.
@@
 First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
-RS time series have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic model.
+RS time series have revolutionized models of ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
 Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
@@
 The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
-When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = #.##.
+When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
+The correlation between observations and predictions is 0.62.
 
@@
 -----------
-Landscape-scale population models for plant species have been elusive and challenging [@Schurr2012a; @Merow].
+Landscape-scale population models for plant species have been elusive and challenging [@Schurr2012a; @Merow2014].
 Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
@@
 Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data.
-In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location (Tredennick and Adler, in review).
+In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
 Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
@@
 A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
-The descrepency between Homer et al.'s [-@Homer2015] results and ours is likely due to the climate projections used for forecasting future changes.
+The descrepency between Homer et al.'s (2015) results and ours is likely due to the climate projections used for forecasting future changes.
 The Homer et al.
@@
 It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
-The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or and entirely different functional form can be used depending on the ecological context.
+The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or an entirely different functional form can be used depending on the ecological context.
 As such time-series data sets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.

### 2015-10-19 scripts/ms_scripts.R

@@ -314,9 +314,23 @@ for(t in 1:time.steps){
 }
 
 obs.cover <- growD[,c("Year","Cover","ID")]
-pred.cover <- melt(ex.mat)
-plot(obs.cover$Cover, pred.cover$value, xlim=c(0,20), ylim=c(0,20))
-abline(0,1, col="red")
+
+ex.df <- as.data.frame(ex.mat)
+colnames(ex.df) <- unique(obs.cover$ID)
+ex.df$year <- yearid
+ex.melt <- melt(ex.df, id.vars = "year")
+ex.melt <- ex.melt[with(ex.melt, order(year)), ]
+colnames(ex.melt) <- c("Year", "ID", "cover.pred")
+error.df <- merge(obs.cover, ex.melt)
+error.df$error <- with(error.df, Cover - cover.pred)
+rmse <- sqrt(mean(error.df$error^2))
+cor(error.df$Cover, error.df$cover.pred)
+
+# pred.cover <- melt(ex.mat)
+# plot(obs.cover$Cover, pred.cover$value, xlim=c(0,20), ylim=c(0,20))
+# abline(0,1, col="red")
+
+
 
 brks <- 30
 y <- hist(growD$Cover, freq=FALSE, breaks=brks)

### 2015-10-23 docs/ms_sageAbundance.Rmd
updates ms references
adds in Mevin's edits to ms

@@
 \begin{centering}
-\textbf{\large{A spatiotemporal modeling framework for forecasting climate change impacts on plant populations over large spatial extents}}
+\textbf{\large{A spatio-temporal modeling framework for forecasting climate change impacts on plant populations over large spatial extents}}
 
@@
 Hooten\textsuperscript{2,3,4}, Cameron L.
-Aldridge\textsuperscript{5}, Colin G.
+Aldridge\textsuperscript{5}, Collin G.
 Homer\textsuperscript{6}, David T.
@@
 
-Here we advance spatiotemporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
-We use a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatiotemporal population model that includes climate covariates.
+Here we advance spatio-temporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
+We use a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates.
 We then use the model to forecast future population states (percent cover) under projected climate scenarios.
@@
 Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
-Longterm trends in species' occurence can be detected and correlated with climate variables, but these may or may not be representative of a species' sensitivity to climate.
+Long-term trends in species' occurence can be detected and correlated with climate variables, but these may or may not be representative of a species' sensitivity to climate.
 Likewise, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
@@
 Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent growth.
-Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatiotemporal model of plant population dynamics at the landscape scale.
+Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
 We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
@@
 Sagebrush ecosystems offer an ideal testing ground for our new modeling approach.
-Their distribution is widespread, sagebrush species are sensitive to climate, and forecasts of future sagebrush ecosystems are in high demand due to the precarious status of sage-grouse.
-Nearly all species distribution models predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+Their distribution is widespread, sagebrush species are sensitive to climate, and forecasts of future sagebrush ecosystems are in high demand due to the precarious status of sage-grouse (*Centrocercus* spp.).
+Nearly all species distribution models predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
 Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
@@
 
-The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but we are most excited about its application in cases where remotely-sensed estimates exist.
-In those cases, we think our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.
+The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
+In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.
 
@@
 To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming[^link0] [@Homer2012].
-Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
+As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
 The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
@@
 In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
-For our purposes we use a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
+In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
 Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).
@@
 We downloaded daily historic weather data for our study site from the NASA Daymet data set (*available online*)[^link].
-The Daymet weather data is interpolated between coarse observation units and does capture some spatial variation.
-However, instead of further downscaling the 1 kilometer resolution weather data to match our 30 meter resolution percent cover data, we chose to simply extract weather data for the centroid of our study area.
-We calculated five climate variables from the Daymet data for the time period of our remotely sensed data (1984 to 2011).
+The Daymet weather data are interpolated between coarse observation units and captures some spatial variation.
+We relied on weather data for the centroid of our study area.
+We calculated five climate variables from the Daymet data for the time period coinciding with our remotely sensed data (1984 to 2011).
 
-We narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model projections.
+We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
 The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t} (*lagPpt*), (2) year \emph{t} fall through summer precipitation (*ppt2*), (3) year \emph{t}-1 fall through summer precipitation (*ppt1*), (4) year \emph{t} average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-1 average spring temperature (*TmeanSpr1*).
 We selected these variables based on previous studies (see *Introduction*), even though not all emerge as important predictors in our model.
-Several model selection techniques exist that could be applied to our approach [@Hooten2014], but we avoid model selection here to maximize focus on the general approach.
+Several model selection techniques exist that could be applied to our approach [@Hooten2015], but we avoid model selection here to maximize focus on the general approach.
 
@@
 To forecast future population states, we required future projections of climate for our study area.
-We downloaded the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
+We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
 The CMIP5 provides projections from a suite of global circulations models (Table X1).
@@
 To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
-For temperature we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
-For precipitation we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation.
-We did this for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+For temperature, we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
+For precipitation, we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation.
+We calculate precipitation and temperature changes for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
 The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
@@
 ### Additive spatio-temporal model for sagebrush cover
-We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects [@Conn2015].
+We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects similar to that described by @Conn2015.
 We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally Poisson
 
+\vspace{-4em}
 \begin{align}
@@
 \end{align}
+\vspace{-4em}
 
@@
 
+\vspace{-4em}
 \begin{align}
@@
 \end{align}
+\vspace{-3em}
 
-Our model of percent cover change includes a temporal effect of cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
-Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to speed up convergence during the model fitting stage (see below).
+Our model of percent cover change includes a temporal effect of log-transformed cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
+Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to improve convergence during the model fitting stage and to allow for easier prior specification.
 The intercept, $\beta_{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
-We chose to use a Poisson likelihood because integer percent cover values can be considered a form of count data.
+We used a Poisson likelihood because integer percent cover values in the sagebrush data product can be considered a form of count data.
 We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$).
@@
 Fitting models that appropriately account for spatial autocorrelation over large spatial extents would be virtually impossible without these modern techniques.
-We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [@Hooten2007a; @Conn2015].
+We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
 We can then define the spatial effect as
 
+\vspace{-4em}
 \begin{align}
@@
 
+\vspace{-4em}
 \begin{align}
-K_{s,m} &= w_{s,m} / \sum_{s} w_{s,m}
+K_{s,m} &= w_{s,m} / \sum_{s=1}^{S} w_{s,m}
 \end{align}
+\vspace{-4em}
+
 where
+
+\vspace{-4em}
 \begin{align}
@@
 \end{align}
+\vspace{-4em}
 
 where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
-It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2014].
-However, given the relative largeness of our dataset and computational limitations, we defined kernels around 121 knots (Fig.
+It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
+However, given the relative size of our dataset and computational limitations, we defined kernels around 121 knots (Fig.
 X#) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix X).
@@
 
+\vspace{-4em}
 \begin{align}
 [\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
-&\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2][\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2].
+&\left(\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2].
 \end{align}
+\vspace{-4em}
 
 ### Accomodating zeros
-Since our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$), our model does not accomodate zeros. 
-Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47). 
-Thus, we simply excluded those pixels from the model fitting process. 
-However, when it came time to simulate the model, we needed to include possible transitions from zero to non-zero percent cover. 
+Our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$). 
+Thus, our model does not accomodate zeros. 
+Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47, which is 0.01% of the data set). 
+Thus, we excluded those pixels from the model fitting process. 
+However, when simulating the process, we needed to include possible transitions from zero to non-zero percent cover. 
 We fit an intercept-only logistic model to estimate the probability of a pixel going from zero to non-zero cover
 
+\vspace{-4em}
 \begin{align}
@@
 ####Fitting the model
-We fit the model in R [@R2013] using the 'No-U-Turn' MCMC sampler in Stan [@stan2014] and the RStan package [@rstan2014].
+We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
 We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
 Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
-However, the Stan sampler is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
-Each MCMC chain was initialized with unique parameter values and then run in parallel on the Utah State University supercomputer.
+However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
+Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
 Model fitting required five days.
-We assessed convergence visually and made sure all scale reduction factors were less than 1.01 [@Gelman1992; @Gelman2007].
+We assessed convergence visually and calculated scale reduction factors [< 0.01, @Gelman1992; @Gelman2009].
 
-####Simulating the model
-We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make yearly forecasts of sagebrush cover starting the final year of our observations and ending in 2098.
+####Simulating the process
+We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make annual forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
 Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
@@
 2).
-GCM projections are variable for our study area, highlighting the importance of making forefasts based on ensemble projections.
+GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections.
 
 All parameters in our model converged on stable posterior distributions (Appendix X).
-Only the *lagPpt* climate covariate can be considered important at 90% confidence, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 3).
-The mean posterior estimate for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. X#).
+Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush annual percent cover change (Fig. 3).
+The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. X#).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
@@
 Nonetheless, observations and predictions are qualitatively similar (Fig. 4A, B).
-Predictions are most confident, even if biased, in low percent cover pixels (Fig. 4D).
+Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
 The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
@@
 The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in sagebrush cover heterogeneity since we do not lose low cover pixels and gain higher cover pixels (Fig. 5).
-For the yearly forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
+Thus, our model predicts an increase in sagebrush cover heterogeneity because we do not lose low cover pixels and gain higher cover pixels (Fig. 5).
+For the annual forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
 6).
@@
 -----------
-Landscape-scale population models for plant species have been elusive and challenging [@Schurr2012a; @Merow2014].
+Landscape-scale population models for plant species have been elusive and challenging to implement [@Schurr2012a; @Merow2014].
 Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
-Here we introduced a new statistical approach to modeling plant populations that overcame two major hurdles for spatially-explicit population models.
+We introduced a new statistical approach to modeling plant populations that overcame two major hurdles for spatially-explicit population models.
 First, we used species-specific remote sensing data as a response variable.
-Thus, we had at our disposal spatially-continuous estimates of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data.
-In so doing we overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from the vast literature on spatiotemporal modeling of animal abundance [@Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
-This means our model is statistical rigorous by explicitly accounting for spatial correlation.
-While important statistically, ecologically it means we can be confident in our estimates of climate effects on sagebrush cover change, given the data in-hand.
-It also means we can make spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).
+Thus, we had a spatially-continuous data product of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
+In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
+Second, borrowing from the vast literature on spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
+Accounting for spatial autocorrelation allows for statistically rigorous inference.
+While important statistically, in terms of ecological inference, we can be confident in our estimates of climate effects on sagebrush cover change in our study region.
+It also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).
 
-As expected, the amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
+The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
 Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig.
@@
 
-The relatively weak climate effects means we should expect small changes when we apply the model to projected climate scenarios.
+The relatively weak climate effects suggests that we should expect small changes when we apply the model to projected climate scenarios.
 Our model projects modest increases in sagebrush cover under all RCP scenarios (Figs. 4,5).
-The forecasted increases are primarily driven by a positive response to precipitation (Fig. 2), but weak effects of temperature also influenced results.
+The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
 A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
-The descrepency between Homer et al.'s (2015) results and ours is likely due to the climate projections used for forecasting future changes.
+The discrepancy between Homer et al.'s (2015) results and ours is likely due to the climate projections used for forecasting future changes.
 The Homer et al.
-study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
+(2015) study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
 Thus, the decline in precipitation led to a decline in sagebrush cover.
-Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of 90% confidence (Fig. 5).
+Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig.
+5).
 
-We want to stress that there is no clear answer to what climate model to use when making forecasts.
+There is no clear answer to what climate model to use when making forecasts.
 Here we chose to use all possible projections.
@@
 If a certain GCM tends to better predict weather in a specific study area, then that GCM should be given more weight than others.
-We plan to outline such an approach in our future work.
+Extensions of these ideas are the subject of ongoing research.
 
 Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
-We decided to work at the native spatial resolution of the remote sensing data ($30\times30$ meters). 
-This constrained the extent that we could reasonably model because we needed to place the knots for the spatial ranom effect close enough together to detect the underlying spatial structure. 
-As more and more knots are used, computation time increases substantially [@Wikle2010].
+We focused the native spatial resolution of the remote sensing data ($30\times30$ meters). 
+This constrained the extent that we could reasonably model because of the computational challenges in using a massive spatial random effect. 
+We used dimension reduction techniques, bu as more and more knots are used, computation time increases substantially [@Wikle2010].
 Had we decided to aggregate the original data to a coarser resolution, we would have been able to model a larger spatial extent using the same number of knots, and therefore computation time.
 
-Our decision on spatial extent and resolution also impacted our incorporation of climate covariates.
+Our spatial extent and resolution also impacted our incorporation of climate covariates.
 We decided against further downscaling Daymet data to match the spatial resolution of the sagebrush data.
-This means that in each year, all pixels share the same climate covariates.
-In other words, climate only varied through time and not over space.
-This is a of course a limitation of our study, and helps to explain the weak effect of climate covariates when averaged over space (Fig. 2).
-In the future we plan to work at a lower spatial resolution but a larger spatial extent so that we can use spatially-explicit climate covariates that match the resolution of the percent cover data.
+Thus, in each year all pixels share the same climate covariates.
+In other words, climate informatino in our model only varied through time and not over space.
+This is a potential limitation of our study, and could explain the weak effect of climate covariates when averaged over space (Fig. 3).
+Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit climate covariates that match the resolution of the percent cover data.
 
@@
 We have introduced a new approach to fitting and simulating population models at large spatial extents.
-But, this is just a start.
-More work is needed to apply our approach over even larger spatial extents, and at higher spatial resolution.
-Once working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.
+As methodological and conceptual advancements are made we will be able to apply our approach over even larger spatial extents, and at higher spatial resolution.
+Working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.
 Given the dependence of forecasts on the weather projections used, we urge ecologists to adopt methods from weather forecasters to up- and down-weight GCMs based on study area-specific reliability.
-Lastly, it will be important to include the effects of species interactions [@Clark2014].
-Fortunately, our spatiotemporal modeling framework is very flexible.
+Lastly, it may also be important to include the effects of species interactions [@Clark2014].
+Fortunately, our spatio-temporal modeling framework is very flexible.
 It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
 The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or an entirely different functional form can be used depending on the ecological context.
-As such time-series data sets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
+As large scale time series data sets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
 
@@
 Department of Energy's Program for Climate Model Diagnosis and Intercomparison provides coordinating support and led development of software infrastructure in partnership with the Global Organization for Earth System Science Portals.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
+government.
 
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=F}
 library(png)
@@
 
-```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area.", eval=T}
+```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area.", eval=F}
 library(png)
@@
 The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time \\emph{t} is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=T}
+Only cumulative precipitation at time \\emph{t} is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=F}
 library(png)
@@
 Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the model.
-We used mean parameter estimates.", fig.height=8, eval=T}
+We used mean parameter estimates.", fig.height=8, eval=F}
 library(png)
@@
 Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
-Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=T}
+Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=F}
 library(png)

### 2015-11-02 scripts/rhat_calcs.R
new script for plotting rhats

@@ -1,0 +1,69 @@
+##  Script to read in MCMC results and calculate
+##  scale reduction factors
+
+rm(list=ls())
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(ggplot2)
+
+
+
+####
+####  Define Rhat function; stolen from `ggmcmc` ggs_Rhats() fxn ---------------
+####
+get_rhats <- function (D, family = NA, scaling = 1.5) 
+{
+  if (attributes(D)$nChains < 2) {
+    stop("At least two chains are required")
+  }
+  if (!is.na(family)) {
+    D <- get_family(D, family = family)
+  }
+  psi.dot <- D %>% group_by(Parameter, Chain) %>% dplyr::summarize(psi.dot = mean(value))
+  psi.j <- D %>% group_by(Parameter) %>% dplyr::summarize(psi.j = mean(value))
+  b.df <- dplyr::inner_join(psi.dot, psi.j, by = "Parameter")
+  B <- b.df %>% group_by(Parameter) %>% dplyr::summarize(B = var(psi.j - 
+                                                                   psi.dot) * attributes(D)$nIterations)
+  B <- unique(B)
+  s2j <- D %>% group_by(Parameter, Chain) %>% dplyr::summarize(s2j = var(value))
+  W <- s2j %>% group_by(Parameter) %>% dplyr::summarize(W = mean(s2j))
+  BW <- dplyr::inner_join(B, W, by = "Parameter") %>% dplyr::mutate(wa = (((attributes(D)$nIterations - 
+                                                                              1)/attributes(D)$nIterations) * W) + ((1/attributes(D)$nIterations) * 
+                                                                                                                      B), Rhat = sqrt(wa/W))
+  BW$Rhat[is.nan(BW$Rhat)] <- NA
+  f <- ggplot(BW, aes(x = Rhat, y = Parameter)) + geom_point() + 
+    xlab(expression(hat("R"))) + ggtitle("Potential Scale Reduction Factor")
+  if (!is.na(scaling)) {
+    scaling <- ifelse(scaling > max(BW$Rhat, na.rm = TRUE), 
+                      scaling, max(BW$Rhat, na.rm = TRUE))
[…]

### 2015-11-02 scripts/plot_projected_weather.R
adds scetion to script to compare hindcasts and observations

@@ -77,3 +77,32 @@ png("../docs/components/figure/weather_projections.png", width = 6, height = 8, 
 outplot <- grid.arrange(g1, g2, ncol=1)
 dev.off()
 
+
+
+####
+####  Compare weather hindcasts to average weather at site ---------------------
+####
+obs_weather <- read.csv("../data/FormattedClimate_WY_SA1.csv")
+obs_weather <- melt(obs_weather, id.vars = "year")
+obs_ppt <- subset(obs_weather, variable=="ppt2")
+model_weather <- ddply(ppt_agg, .(Year, season), summarise,
+                       value = mean(avg_value))
+model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
+precip_compare <- data.frame(year = model_weather$Year, 
+                             precip = c(obs_ppt$value, model_weather$value),
+                             value_source = rep(c("observed", "modeled"), each=34))
+ggplot(precip_compare)+
+  geom_boxplot(aes(x=value_source, y=precip))
+
+
+obs_temp <- subset(obs_weather, variable=="TmeanSpr2")
+model_weather <- ddply(temp_agg, .(Year, season), summarise,
+                       value = mean(avg_value))
+model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
+temp_compare <- data.frame(year = model_weather$Year, 
+                             precip = c(obs_temp$value, model_weather$value),
+                             value_source = rep(c("observed", "modeled"), each=34))
+ggplot(temp_compare)+
+  geom_boxplot(aes(x=value_source, y=precip))
+
+

### 2015-11-03 docs/appendix_a.Rmd
updates model table

@@
 |  |  | Organisation, Australia), and BOM (Bureau of Meteorology, Australia) |
+| CSIRO-BOM | ACCESS1.3 | CSIRO (Commonwealth Scientific and Industrial Research |
+|  |  | Organisation, Australia), and BOM (Bureau of Meteorology, Australia) |

### 2015-11-03 docs/appendixB.Rmd
updates knot appendix B

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{B\arabic{figure}}

Appendix B: Detailed methods for knot choice and placement
==========================================================
Our dimension reduction approach requires fitting random effects associated with knots distributed nonrandomly in space. By distributing *m* knots across space, we can fit random effects for each knot and then use process convolution to interpolate the random effect between the knots. Increasing the number of knots results in a more finely-resolved spatial effect, but it also increases processe time. Thus, there exists a tension between the resolution of the spatial effect and computational effort. Without regard to computation time, it is possible to use common model selection techniques to guide the number of knots one should use. With regard to computation time, it is sensible to choose knot placement and number based on the observed residual spatial variation.

To do so, we fit a generalized linear model of the form

\vspace{-4em}
\begin{align}
y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
\end{align}
\vspace{-4em}

where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*

\vspace{-4em}
\begin{align}
\text{log}(\mu_{i,t}) &= \beta_{0} + \beta_{1}y_{i,t-1} + \textbf{x}_{t}'\boldsymbol{\gamma}
\end{align}
\vspace{-3em}

that includes a density-dependence effect of log-transformed cover in the previous year ($y_{i,t-1}$) and climate effects ($\textbf{x}_t$). In other words, a non-hierarchial version of the model presented in our main text. We then examined a spatial variogram of the residuals from the above model to identify the the distance at which spatial dependence disappears.

Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig. B1). Thus, we set the kernal bandwidth ($\sigma$ in Eq. X of the main text) to 500/3 = `r round(500/3,2)`. We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately $\sigma$ = `r round(500/3,2)`. This resulted in 121 knots (Fig. B2).

![Spatial variogram of model residuals showing where spatial dependence disappears (dark red line at ~500 meters), and the value chosen for the kernal bandwidth in Eq. X of the main text (500/3; dashed black line).](components/avgresidual_variogram_withrange_subset.png)

![Map of our study area showing percent cover in 1984 and the location of the 121 evenly-spaced knots.](components/SAGE_Grid_wKnots_subset.png)

### 2015-11-03 docs/appendixB.Rmd
updates appendix B with mention of knot spacing

@@
 We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately $\sigma$ = `r round(500/3,2)`.
-This resulted in 121 knots (Fig.
+Knots were placed evenly across space in the sense that all pixels are equally-spaced along latitude and longitude, but the latitudinal distance and longitudinal distance are not exactly equal.
+This was done to achieve a symmetric grid of knots.
+This resulted in a 11 $\times$ 11 knot grid (121 knots; Fig.
 B2).

### 2015-11-03 docs/ms_sageAbundance.Rmd
updates author list
updates climate projectsion for just 18 models with all three RCP scenarios

@@
 Aldridge\textsuperscript{5}, Collin G.
-Homer\textsuperscript{6}, David T.
-Iles\textsuperscript{1}, Andrew Kleinhesselink\textsuperscript{1}, Eric LaMalfa\textsuperscript{1}, Rebecca Mann\textsuperscript{1}, and Peter B.
+Homer\textsuperscript{6}, Andrew Kleinhesselink\textsuperscript{1}, and Peter B.
 Adler\textsuperscript{1}}}
@@
 Population modeling inference has traditionally been limited to small scale observational studies.
-Yet, global environmental changes, like climate change, tend to play out at the landscape and regional scales, creating a mismatch between current tools and current problems.
-New technology is allowing for data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information.
-For example, plant ecologists now have access to species-specific estimates of plant percent cover through time at spatial extents ranging from landscapes to continents.
-The real strength to be gained using this new form of ecological information will be to move beyond simple descriptive summaries of the data, and couple them with sophisticated model-based statistical approaches that can accommodate more realistic ecological mechanisms.
-A major hurdle in properly analyzing these types of data is in building and fitting statistical models that have valid assumptions.
-Rigorously accounting for spatial autocorrelation has been a crux in former analyses at large scales, but we can now fit appropriate models using modern, computationally efficient model reparameterizations and dimension reduction techniques.
-
-Here we advance spatio-temporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
-We use a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates.
+Yet, global environmental changes, like climate change, play out at landscape and regional scales, creating a mismatch between available tools and real world problems.
+New technology is generating data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information like species-specific estimates of plant percent cover.
+Unlike some ecological sub-disciplines, population modelers have yet to fully embrace the use of remotely-sensed data as estimates of population abundance.
+Such reluctance likely stems from challenges in fitting statistically appropriate models that rigorously account for spatial autocorrelation in large datasets.
+Here, we use modern, computationally efficient model reparameterizations to fit a dynamic population model to a remotely-sensed dataset.
+We use a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates and spatial random effects.
 We then use the model to forecast future population states (percent cover) under projected climate scenarios.
@@
 Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
-Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers.
+Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
 However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
@@
 Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
-Long-term trends in species' occurence can be detected and correlated with climate variables, but these may or may not be representative of a species' sensitivity to climate.
-Likewise, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
+Furthermore, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
 Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.
 
-It seems the ideal tool for forecasting future population states under climate change is a large-scale dynamic population model [@Schurr2012a; @Merow2014].
-Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary to fit high-dimensional spatial models.
+The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014].
+Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary needed to fit high-dimensional spatial models.
 Fortunately, two recent technological advances put us in the position to develop large-scale population models.
@@
 Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
-However, these methods have yet to be use in a dynamic setting.
+However, these methods have yet to be used in a dynamic setting.
 
-Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent growth.
+Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent population growth.
 Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
@@
 
-Sagebrush ecosystems offer an ideal testing ground for our new modeling approach.
-Their distribution is widespread, sagebrush species are sensitive to climate, and forecasts of future sagebrush ecosystems are in high demand due to the precarious status of sage-grouse (*Centrocercus* spp.).
-Nearly all species distribution models predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+Sagebrush ecosystems offer an ideal testing ground for our new modeling approach: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Species distribution models typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
 Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
@@
 In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
-Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).
+Thus, the full dataset contains 369,603 observations consisting of 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
 
@@
 We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
-The CMIP5 provides projections from a suite of global circulations models (Table X1).
+The CMIP5 provides projections from a suite of global circulations models (GCMs), and we used projections from 18 GCMs (Table A1) that produced weather projections for all three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
 To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
@@
 For precipitation, we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation.
-We calculate precipitation and temperature changes for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
-The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
+We calculate precipitation and temperature changes for RCP 4.5, RCP 6.0, and RCP 8.5 scenarios.
 We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.
@@
 
-In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make annual forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
+In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
 We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.
@@
 \begin{align}
-\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
+\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal + dens.
+dep} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
 \end{align}
@@
 
-Our model of percent cover change includes a temporal effect of log-transformed cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
+Our model of percent cover change includes a density-dependent effect of log-transformed cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to improve convergence during the model fitting stage and to allow for easier prior specification.
@@
 Reducing the number of effective spatial random effect parameters speeds up estimation by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
-Fitting models that appropriately account for spatial autocorrelation over large spatial extents would be virtually impossible without these modern techniques.
+Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
 We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
@@
 \begin{align}
-w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right).
+w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right)
 \end{align}
@@
 However, given the relative size of our dataset and computational limitations, we defined kernels around 121 knots (Fig.
-X#) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix X).
+B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix B).
 The Bayesian posterior distribution can be expressed as
@@
 Model fitting required five days.
-We assessed convergence visually and calculated scale reduction factors [< 0.01, @Gelman1992; @Gelman2009].
+We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.01 for all paramters) [@Gelman1992; @Gelman2009].
 
 ####Simulating the process
-We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make annual forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
+We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
 Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
@@
 When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
-The correlation between observations and predictions is 0.62.
+The correlation between observations and predictions was 0.62.
 
@@
 The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in sagebrush cover heterogeneity because we do not lose low cover pixels and gain higher cover pixels (Fig. 5).
-For the annual forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
+Thus, our model predicts an increase in sagebrush cover heterogeneity because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
+For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
 6).
 Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
-This is expected given the positive effect of precipitation on sagebrush cover change (Fig. 3).
+This result reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).
 
@@
 Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
-We introduced a new statistical approach to modeling plant populations that overcame two major hurdles for spatially-explicit population models.
+We introduced a new statistical approach to modeling plant populations that overcomes two major hurdles for spatially-explicit population models.
 First, we used species-specific remote sensing data as a response variable.
@@
 Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
-Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases at higher elevations.
-As seen when comparing our temporal forecasts (Fig. 5) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
+Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases colder parts of their range.
+As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
 Although our estimates of climate effects were small, they are in agreement with previous studies, indicating our model does an adequate job of capturing the ecological process and the link between sagebrush growth and climate.
@@
 We used dimension reduction techniques, bu as more and more knots are used, computation time increases substantially [@Wikle2010].
-Had we decided to aggregate the original data to a coarser resolution, we would have been able to model a larger spatial extent using the same number of knots, and therefore computation time.
+Had we decided to aggregate the original data to a coarser resolution, we would have been able to model a larger spatial extent using the same number of knots without increasing computation time.
 
@@
 In other words, climate informatino in our model only varied through time and not over space.
-This is a potential limitation of our study, and could explain the weak effect of climate covariates when averaged over space (Fig. 3).
+This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).
 Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit climate covariates that match the resolution of the percent cover data.
+It would also be possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play important roles.
 
@@
 The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or an entirely different functional form can be used depending on the ecological context.
-As large scale time series data sets become more available, we anticipate the approach we developed here to complement studies using species distribution models and local-scale demographic population models.
+As large scale time series data sets become more available, we anticipate the approach we developed here to fill the gap in spatial scales between species distribution models and local-scale demographic population models.
 
@@
 ATT was supported by an NSF Postdoctoral Research Fellowship in Biology (DBI-1400370) and AK was supported by an NSF Graduate Research Fellowship.
+Additional support came from the Utah Agricultural Experiment Station, Utah State University, and this article is approved as journal paper number XXXX.
 We are grateful to Debra K.
@@
 government.
+David T.
+Iles, Eric LaMalfa, and Rebecca Mann participated in project conception as part of the distributed graduate seminar and provided comments that improved the manuscript.
 

### 2015-11-03 scripts/climate_formatting/Climate_Proj_Calcs_CMIP5.R

@@ -29,7 +29,7 @@ ppt <- as.data.frame(read.csv("pr.csv", header=FALSE))
 tmp<-read.table("COLS_pr.txt")
 tmp<-as.character(tmp[,1])
 colnames(ppt) <- c("Year", "Month",tmp)
-ppt <- melt(ppt, id.vars=c("Year", "Month"),)
+ppt <- melt(ppt, id.vars=c("Year", "Month"))
 ppt[,4] <- as.numeric(ppt[,4]) #NAs coerced for a couple December 2099 null values 
 tmp<-unlist(strsplit(x=as.character(ppt$variable),split=".",fixed=T))
 tmp<-matrix(tmp,nrow=length(tmp)/3,ncol=3,byrow=T)
@@ -39,7 +39,22 @@ ppt$period<-cut(ppt$Year,breaks=c(1950,2000,2050,2100),
   include.lowest=T,labels=c("past","present","future"))
 ppt$season<-ifelse(ppt$Month > 6 & ppt$Month < 10,"summer","fall2spr")
 
-pptMeans<-aggregate(value~period+season+scenario,data=ppt,FUN=mean)
+pptMeans<-aggregate(value~period+season+scenario+model,data=ppt,FUN=mean)
+allmods <- unique(pptMeans$model)
+keeps <- character(length(allmods))
+my_scens <- c("rcp45", "rcp60", "rcp85")
+for(i in 1:length(allmods)){
+  tmp <- subset(pptMeans, model==allmods[i])
+  tmp.scns <- unique(tmp$scenario)
+  flag <- length(which(my_scens %in% tmp.scns == FALSE))
+  ifelse(flag>0, keeps[i]<-"no", keeps[i]<-"yes")
+}
+modelkeeps <- data.frame(model=allmods,
+                         allscenarios=keeps)
+my_mods <- modelkeeps[which(modelkeeps$allscenarios=="yes"),"model"]
+ppt_projs <- subset(pptMeans, model %in% my_mods)
+
+pptMeans<-aggregate(value~period+season+scenario,data=ppt_projs,FUN=mean)
 colnames(pptMeans) <- c("period","season","scenario","value")
 pptMeans$days<-ifelse(pptMeans$season=="summer",92,365-92)
 pptMeans$ppt<-pptMeans$value*pptMeans$days
@@ -63,7 +78,23 @@ Tavg$period<-cut(Tavg$Year,breaks=c(1950,2000,2050,2100),
   include.lowest=T,labels=c("past","present","future"))
[…]

### 2015-11-03 scripts/ms_scripts.R
updates projected climate call

@@ -352,9 +352,14 @@ ggsave("../results/obs_pred_hist.png", width = 4, height = 4, units="in")
 ####
 ####  Run climate simulations --------------------------------------------------
 ####
+# Read in climate projected changes
+ppt_projs <- subset(read.csv("../data/precipitation_projections.csv"), 
+                    scenario!="rcp26" & season=="fall2spr")
+temp_projs <- subset(read.csv("../data/temperature_projections.csv"),
+                     scenario!="rcp26" & season=="spring")
 projC<-data.frame("scenario"=c("rcp45","rcp60","rcp85"),
-                  "deltaPpt"=c(1.0894,1.0864,1.110),
-                  "deltaTspr"=c(2.975,3.134,4.786))
+                  "deltaPpt"=1+ppt_projs$change,
+                  "deltaTspr"=temp_projs$change)
 mean_params <- ddply(outs, .(Parameter), summarise,
                      mean(value))
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"..1"]

### 2015-11-04 scripts/ms_scripts.R
updates ms_script.R with new climate scenarios

@@ -187,7 +187,7 @@ dev.off()
 ####
 ####  Run equilibrium population simulation ------------------------------------
 ####
-time.steps <- 600
+time.steps <- 2000
 burn.in <- 100
 mean_params <- ddply(outs, .(Parameter), summarise,
                      value = mean(value))
@@ -332,18 +332,18 @@ cor(error.df$Cover, error.df$cover.pred)
 
 
 
-brks <- 30
-y <- hist(growD$Cover, freq=FALSE, breaks=brks)
-yhat <- hist(pred.cover$value, freq=FALSE, breaks=brks)
-hist_ydata <- data.frame(x = y$breaks, y=c(y$density,0))
-hist_yhatdata <- data.frame(x = yhat$breaks, y=c(yhat$density,0))
-
-ggplot()+
-  geom_step(data=hist_ydata, aes(x=x, y=y), size=1.5)+
-  geom_step(data=hist_yhatdata, aes(x=x, y=y), col="darkorange", size=1)+
-  xlab("Cover (%)")+
-  ylab("Density")
-ggsave("../results/obs_pred_hist.png", width = 4, height = 4, units="in")
+# brks <- 30
+# y <- hist(growD$Cover, freq=FALSE, breaks=brks)
+# yhat <- hist(pred.cover$value, freq=FALSE, breaks=brks)
+# hist_ydata <- data.frame(x = y$breaks, y=c(y$density,0))
+# hist_yhatdata <- data.frame(x = yhat$breaks, y=c(yhat$density,0))
+# 
+# ggplot()+
+#   geom_step(data=hist_ydata, aes(x=x, y=y), size=1.5)+
+#   geom_step(data=hist_yhatdata, aes(x=x, y=y), col="darkorange", size=1)+
+#   xlab("Cover (%)")+
+#   ylab("Density")
+# ggsave("../results/obs_pred_hist.png", width = 4, height = 4, units="in")
 
 
 
@@ -365,7 +365,7 @@ mean_params <- ddply(outs, .(Parameter), summarise,
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"..1"]
 betas <- mean_params[grep("beta", mean_params$Parameter),"..1"][2:6]
 eta <- K%*%alphas
-time.steps <- 600
+time.steps <- 2000
 burn.in <- 100
 
 p.climD<-climD[climD$year %in% unique(growD$Year),c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
[…]

### 2015-11-06 docs/ms_sageAbundance.Rmd
Andy's changes to the abstract

@@
 --------
-Population modeling inference has traditionally been limited to small scale observational studies.
-Yet, global environmental changes, like climate change, play out at landscape and regional scales, creating a mismatch between available tools and real world problems.
-New technology is generating data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information like species-specific estimates of plant percent cover.
-Unlike some ecological sub-disciplines, population modelers have yet to fully embrace the use of remotely-sensed data as estimates of population abundance.
-Such reluctance likely stems from challenges in fitting statistically appropriate models that rigorously account for spatial autocorrelation in large datasets.
-Here, we use modern, computationally efficient model reparameterizations to fit a dynamic population model to a remotely-sensed dataset.
-We use a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates and spatial random effects.
-We then use the model to forecast future population states (percent cover) under projected climate scenarios.
-Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to previous year's precipitation.
-However, gains in cover are not homogeneous in space.
-Such spatially explicit estimates of increases/decreases in plant abundance have, until now, been elusive.
-The spatiotemporal model we describe can be applied to any spatially explicit time series of plant cover.
-As such datasets become increasingly available and precise, we expect the methods we use here to become an appealing alternative to species distribution models that lack population dynamics and, generally, estimates of species abundance.
+New remote sensing technology is producing ecological data at unprecedented spatial scales.
+These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots.
+However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time.
+Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.
+We fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming.
+We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
+(include more on the methodological advancement/novelty..
+knots etc??) We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change.
+Our analysis provides spatially explicit estimates of increases and decreases in a plant population that have, until now, been impossible to produce at this scale.
+(any new insights about the statistical method?) Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all.
+We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.
 

### 2015-11-07 docs/ms_sageAbundance.Rmd
Andys edits to the introduction

@@
 
-Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
-In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009].
-Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
-Furthermore, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
-Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.
+The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014]; however, developing useful models at this scale has been limited by the lack of time series data at large spatial extents and the lack of statistical machinary needed to fit high-dimensional spatial models.
+Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
+First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.
+These data can be fit with dynamic population models which include yearly fluctuation in climate as covariates.
+Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
+Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review].
+These new statistical methods have yet to be applied to RS derived plant populations data at large scales.
 
-The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014].
-Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary needed to fit high-dimensional spatial models.
-Fortunately, two recent technological advances put us in the position to develop large-scale population models.
-First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
-RS time series have revolutionized models of ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
-Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
-However, these methods have yet to be used in a dynamic setting.
+Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.
+Currently most assesments of how plant and animal populations will respond to climate change rely on Species distribution models (SDMs).
+SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
+However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs.
+Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [add EhrlenANDMorris].
+Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes.
+Large-scale dynamic population models could overcome both of these limitations.
+They would produce spatially expicit estimates of species abundance within the species range [Ehrlen and Morris] and would provide testable predictions of how populations should respond to short-term climate perturbations.
+These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models (cite a prediction paper).
 
-Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent population growth.
-Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
-We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
-We then use the model to forecast the impacts of climate change on sagebrush percent cover.
+Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014].
+In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014].
+In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
+A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
 
-Sagebrush ecosystems offer an ideal testing ground for our new modeling approach: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
-Species distribution models typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
-Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
-Models suggest that warming could drive sagebrush losses at low elevations and southern parts of its distribution by making snowmelt earlier and increasing evaporation, leading to soil water deficits in the spring [@Schlaepfer2012; @Schlaepfer2014].
-In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
-Thus, the response of sagebrush to climate will likely vary in space depending on elevation, among other factors.
-Direct observations of individual plants and cover on experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season and an increase in sagebrush growth and cover [@Perfors2003; @Harte2015].
+(Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].) <-- Too much detail?
+could move to discussion or cite as evidence that "sagebrush is sensitive to climate" above?
+
+Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.
+Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012].
+We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
 

### 2015-11-07 docs/ms_sageAbundance.Rmd
methods edits

@@
 This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
-To overcome these computational constraints, we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
-Reducing the number of effective spatial random effect parameters speeds up estimation by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
+Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation.
+This greatly speeds up model fitting by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
 Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
@@
 For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
-We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding and initial 100 time steps.
+We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
 For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.

### 2015-11-07 docs/ms_sageAbundance.Rmd
Andy edits to discussion

@@
 We introduced a new statistical approach to modeling plant populations that overcomes two major hurdles for spatially-explicit population models.
-First, we used species-specific remote sensing data as a response variable.
+First, we used high resolution species-specific remote sensing data as a response variable.
 Thus, we had a spatially-continuous data product of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
 In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from the vast literature on spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
-Accounting for spatial autocorrelation allows for statistically rigorous inference.
-While important statistically, in terms of ecological inference, we can be confident in our estimates of climate effects on sagebrush cover change in our study region.
-It also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).
+Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy that rigorously accounted for spatial autocorrelation while also allowing for feasible computational time in model fitting.
+Accounting for spatial autocorrelation in this large dataset allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region.
+The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e.
+within the study region; Figs.
+4,5).
 
@@
 Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
-Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases colder parts of their range.
+Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases in this colder part of the sagebrush range.
 As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
-Although our estimates of climate effects were small, they are in agreement with previous studies, indicating our model does an adequate job of capturing the ecological process and the link between sagebrush growth and climate.
+Although our estimates of climate effects were small, they are in agreement with previous studies, improving our confidence that our model captured a realistic link between sagebrush growth and climate.
 
-The relatively weak climate effects suggests that we should expect small changes when we apply the model to projected climate scenarios.
-Our model projects modest increases in sagebrush cover under all RCP scenarios (Figs. 4,5).
+As expected given the relatively weak climate effects in this study area, our model projects only small changes in sagebrush cover in this area with future climate change.
+Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs.
+4,5).
 The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
 A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
-The discrepancy between Homer et al.'s (2015) results and ours is likely due to the climate projections used for forecasting future changes.
+The discrepancy between Homer et al.'s (2015) results and ours reflects a differences in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
 The Homer et al.
@@
 
-There is no clear answer to what climate model to use when making forecasts.
-Here we chose to use all possible projections.
-However, much like weather forecasters [@Bauer2015], ecological forecasters should use expert judgement to choose a climate projection.
-If a certain GCM tends to better predict weather in a specific study area, then that GCM should be given more weight than others.
-Extensions of these ideas are the subject of ongoing research.
+(not sure this next paragraph is needed--it also sounds a bit too prescriptive about climate projection which is probably outside the realm of the paper) 
+There is no clear answer to what climate model to use when making ecological forecasts.
+Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
+If a certain GCM tends to better predict weather in a specific study area, then that GCM may be given more weight than others.
 
@@
 This constrained the extent that we could reasonably model because of the computational challenges in using a massive spatial random effect. 
-We used dimension reduction techniques, bu as more and more knots are used, computation time increases substantially [@Wikle2010].
-Had we decided to aggregate the original data to a coarser resolution, we would have been able to model a larger spatial extent using the same number of knots without increasing computation time.
+Even with our dimension reduction technique modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
+In order to overcome this, future investigators modeling a larger spatial extent may need to aggregate their original remote-sensing time series data to a coarse spatial resoluation.
+This would allow them to model a much greater spatial extent with a similar number of knots and a similar amount of computation time.
+A coarser scale model would lose some fine-scale detail but could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
 
 Our spatial extent and resolution also impacted our incorporation of climate covariates.
-We decided against further downscaling Daymet data to match the spatial resolution of the sagebrush data.
+We did not downscale Daymet data to match the spatial resolution of the sagebrush data.
 Thus, in each year all pixels share the same climate covariates.
-In other words, climate informatino in our model only varied through time and not over space.
 This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).

### 2015-11-07 docs/ms_sageAbundance.Rmd
adding a ending paragraph to discussion about prediction and validation

@@
 
+While we focused on sagebrush projections at the end of the current century, one important feature of our aproach is that it could in theory be applied to forecast the effects of short-term climate events on sagebrush.
+For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Germino paper on Yellowstone ecosystem drought).
+Validating the model against short-term predictions would give us a way to assess the amount of confidence we should have on any longer-term forecasts produced by the model.
+It would also provide opportunities for refinement of the model itself.
+This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced.
+
 Conclusion
 ----------
-We have introduced a new approach to fitting and simulating population models at large spatial extents.
-As methodological and conceptual advancements are made we will be able to apply our approach over even larger spatial extents, and at higher spatial resolution.
+We have introduced a new approach to fitting and simulating population models at large spatial extents with state of the art population data derived from remote sensing.
+As remote sensing datasets and advances in computing power are made this approach will become useful for a wider number of species and be applicaple at even larger spatial extents, and higher resolution.
 Working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.

### 2015-11-09 scripts/plot_temporal_forecasts_ObsPeriod.R
adds scripts for hindcasts based on GCM weather; no mean change observed

@@ -1,0 +1,98 @@
+##  Script to plot results from all model temporal runs
+
+##  Author: Andrew Tredennick
+##  Email: atredenn@gmail.com
+##  Date created: 10-09-2015
+
+
+##  Clear the workspace...
+rm(list=ls())
+
+
+####
+####  Load libraries -----------------------------------------------------------
+####
+library(ggplot2)
+library(plyr)
+library(reshape2)
+
+
+
+####
+####  Read in model simulations, 1 by 1, and stack -----------------------------
+####
+setwd("../results/yearlyforecasts_ObsPeriod//")
+all_files <- list.files()
+
+longdf <- data.frame(model=NA,
+                     scenario=NA,
+                     paramset=NA,
+                     year=NA,
+                     cover=NA)
+
+for(i in 1:length(all_files)){
+  ttt <- readRDS(all_files[i])
+  ttt2 <- as.data.frame(apply(ttt, c(2,1), mean))
+  colnames(ttt2) <- paste0("params", 1:ncol(ttt2))
+  ttt2$year <- c(1984:2011)
+  ttt3 <- melt(ttt2, id.vars="year")
+  ttt3$model <- strsplit(all_files[i], "_")[[1]][1]
+  ttt3$scenario <- strsplit(all_files[i], "_")[[1]][2]
+  colnames(ttt3) <- c("year", "paramset", "cover", "model", "scenario")
+  tmpout <- ttt3[,c("model", "scenario", "paramset", "year", "cover")]
+  longdf <- rbind(longdf, tmpout)
+  print(paste(i, "out of", length(all_files)))
+}
+
+
+longdf <- longdf[2:nrow(longdf),]
+longall <- longdf
+longall$parammod <- paste(longall$param_set, longall$model)
+
+
+# longdf <- subset(longall, model %in% c("bcc-csm1-1", "ccsm4"))
+longdf <- longall
+meandf <- ddply(longdf, .(year, scenario), summarise,
+                avgcover=mean(cover, na.rm=TRUE),
+                upcover=quantile(cover, 0.95, na.rm=TRUE),
+                locover=quantile(cover, 0.05, na.rm=TRUE))
+
+
+
+####
+####  Read in a format observation data ----------------------------------------
+####
+obs_data <- read.csv("../../data/wy_sagecover_subset_noNA.csv")
+obs_data <- subset(obs_data, Year>1984) # subsets out NA CoverLag values
[…]

### 2015-11-09 scripts/sim_sage_byYear_byModel_ObsPeriod.R

@@ -1,0 +1,162 @@
+##  Script to run sage abundance model through time, starting
+##  at the last observation. Requires time series of climate
+##  projections.
+
+##  Author:       Andrew Tredennick
+##  Email:        atredenn@gmail.com
+##  Date created: 10-08-2015
+
+
+### Clear the workspace
+rm(list=ls())
+
+
+
+####
+####  Set some global simulation settings --------------------------------------
+####
+parameter_reps <- 50
+clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
+
+
+####
+####  Load necessary libraries -------------------------------------------------
+####
+library(sageAbundance)
+library(plyr)
+library(reshape2)
+
+#TODO: get sage Abundance functions to load!! Until then...
+source("../R/clim_proj_format_funcs.R")
+
+
+
+####
+####  Read in percent cover file, climate projections, MCMC output, K.data -----
+####
+obs_data <- read.csv("../data/wy_sagecover_subset_noNA.csv")
+obs_data <- subset(obs_data, Year>1984) # subsets out NA CoverLag values
+temp_projs <- readRDS("../data/CMIP5_yearly_project_temperature.RDS")
+ppt_projs <- readRDS("../data/CMIP5_yearly_project_precipitation.RDS")
+mcmc_outs <- readRDS("../results/poissonSage_randYear_mcmc.RDS")
+load("../results/Knot_cell_distances_subset.Rdata")
+K <- K.data$K
+
+### Only keep models with rcp45, 60, and 85 scenarios
+allmods <- unique(temp_projs$model)
+keeps <- character(length(allmods))
+my_scens <- c("rcp45", "rcp60", "rcp85")
+for(i in 1:length(allmods)){
+  tmp <- subset(temp_projs, model==allmods[i])
+  tmp.scns <- unique(tmp$scenario)
+  flag <- length(which(my_scens %in% tmp.scns == FALSE))
+  ifelse(flag>0, keeps[i]<-"no", keeps[i]<-"yes")
+}
+modelkeeps <- data.frame(model=allmods,
+                         allscenarios=keeps)
+my_mods <- modelkeeps[which(modelkeeps$allscenarios=="yes"),"model"]
+
+temp_projs <- subset(temp_projs, model %in% my_mods)
+ppt_projs <- subset(ppt_projs, model %in% my_mods)
+
+####
+####  Subset out observed climate; get scaling mean and sd ------------------------
+####
[…]

### 2015-11-09 docs/ms_sageAbundance.Rmd

--- 2015-11-07 docs/ms_sageAbundance.Rmd
@@
 
-Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
 SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
@@
 
-The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
-Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig.
+The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig.
+2) which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
+The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig.
 2).
-Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
-Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases in this colder part of the sagebrush range.
-As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
-Although our estimates of climate effects were small, they are in agreement with previous studies, improving our confidence that our model captured a realistic link between sagebrush growth and climate.
+Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
+As seen when comparing our temporally-explicit forecasts (Fig.
+6) to climate projections (Fig.
+2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
 
@@
 
+While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.
+For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al.
+2010).
+Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.
+It would also provide opportunities for refinement of the model itself.
+This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015 prediction paper -- is there a better reference for this idea?].
+
 Conclusion
 ----------
-We have introduced a new approach to fitting and simulating population models at large spatial extents.
-As methodological and conceptual advancements are made we will be able to apply our approach over even larger spatial extents, and at higher spatial resolution.
-Working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.
-Given the dependence of forecasts on the weather projections used, we urge ecologists to adopt methods from weather forecasters to up- and down-weight GCMs based on study area-specific reliability.
-Lastly, it may also be important to include the effects of species interactions [@Clark2014].
-Fortunately, our spatio-temporal modeling framework is very flexible.
-It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
-The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or an entirely different functional form can be used depending on the ecological context.
-As large scale time series data sets become more available, we anticipate the approach we developed here to fill the gap in spatial scales between species distribution models and local-scale demographic population models.
+We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
+As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
+Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].
+For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford and Lauenroth  2009 ].
+Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe.
+The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.
 

### 2015-11-13 scripts/ms_scripts.R
adds coparison of DEM slope to eta

@@ -131,6 +131,18 @@ dev.off()
 
 
 ####
+####  Compare spatial field to landscape features ------------------------------
+####
+slope <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/ancillary_data/ls3731_slope_subset.img")
+slopeDf <- as.data.frame(rasterToPoints(slope))
+eta_slope <- merge(oneyear, slopeDf, by.x=c("Lon", "Lat"), by.y=c("x", "y"))
+ggplot(eta_slope, aes(x=ls3731_slope_subset, y=eta))+
+  geom_point()+
+  stat_smooth(se=FALSE, color="black", method="gam", formula=y ~ s(x, bs = "cs"))+
+  theme_bw()
+
+
+####
 ####  Plot climate covariate posterior densities -------------------------------
 ####
 clim_vars <- c("beta.1.", "beta.2.", "beta.3.", "beta.4.", "beta.5.")

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's adits

@@
 Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review].
-These new statistical methods have yet to be applied to RS derived plant populations data at large scales.
+These new statistical methods have yet to be applied to RS-derived plant populations data at large scales.
 
@@
 In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
-A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
-
-(Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].) <-- Too much detail?
-could move to discussion or cite as evidence that "sagebrush is sensitive to climate" above?
+Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
+A large-scale spatially-explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
 
@@
 We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
-
 The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
@@
 Averaged across all GCMs, precipitation and temperature in our study area are projected to increase by 2050-2100 relative to observed climate from 1950-2000, with the magnitude depending on the RCP scenario (Table 1).
-Trajectories of our climate covariates from GCM projections, used to make yearly forecasts, show similar trends (Fig.
-2).
+Trajectories of our climate covariates from GCM projections, used to make yearly forecasts, show similar trends (Fig. 2). 
 GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections.
@@
 Accounting for spatial autocorrelation in this large dataset allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 
-The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e.
-within the study region; Figs.
-4,5).
+The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
 
 The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig.
-2) which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
-The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig.
-2).
+2), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
+The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 2).  
 Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].  
-As seen when comparing our temporally-explicit forecasts (Fig.
-6) to climate projections (Fig.
-2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
+As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
 
 As expected given the relatively weak climate effects in this study area, our model projects only small changes in sagebrush cover in this area with future climate change. 
-Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs.
-4,5).
+Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 4,5).
 The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
 A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
-The discrepancy between Homer et al.'s (2015) results and ours reflects a differences in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
+The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
 The Homer et al.
@@
 
-(not sure this next paragraph is needed--it also sounds a bit too prescriptive about climate projection which is probably outside the realm of the paper) 
 There is no clear answer to what climate model to use when making ecological forecasts.
@@
 If a certain GCM tends to better predict weather in a specific study area, then that GCM may be given more weight than others.
+Extensions of these ideas are the subject of ongoing research.
 
@@
 While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.  
-For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al.
-2010).
+For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al. 2010).  
 Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.  
 It would also provide opportunities for refinement of the model itself.  
-This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015 prediction paper -- is there a better reference for this idea?].
+This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015].
 
@@
 Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].  
-For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford and Lauenroth  2009 ].
+For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2009].  
 Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe. 

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's abstract edits

@@
 Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.  
-We fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming.
+Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
 We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
-(include more on the methodological advancement/novelty..
-knots etc??) We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change.
-Our analysis provides spatially explicit estimates of increases and decreases in a plant population that have, until now, been impossible to produce at this scale.
-(any new insights about the statistical method?) Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all.
+We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. 
+Our analysis provides spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
+Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all.
 We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's abstract edits
updates refs for Andy's edits

@@
 However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time.
-Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.  
-Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
+Here we demonstrate how recent advances in the modeling of animal populations can be applied to remotely-sensed, species-specific estimates of plant cover at the landscape scale.
+Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
 We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
@@
 Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all.
-We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.
+We expect the approach we develop here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.
 
@@
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
-Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].
+Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011; @Tredennick2015].
 
@@
 Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.
-Currently most assesments of how plant and animal populations will respond to climate change rely on Species distribution models (SDMs).  
+Currently most assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
 SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
 However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
-Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [add EhrlenANDMorris].
+Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015].
 Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes.
 Large-scale dynamic population models could overcome both of these limitations.
-They would produce spatially expicit estimates of species abundance within the species range [Ehrlen and Morris] and would provide testable predictions of how populations should respond to short-term climate perturbations.
-These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models (cite a prediction paper).
+They would produce spatially explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
+These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models [@Luo2011a].
 
@@
 
-Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.
-Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012]. 
+Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.
+Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015].
 We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
@@
 Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].
-For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2009].  
+For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2006].
 Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe.

### 2015-11-13 docs/appendixC.Rmd
appendix C added for traceplotes

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{C\arabic{figure}}

Appendix C: Model convergence diagnostics
=========================================
Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure C1).
Visual inspection of traceplots indicates convergence (Figures C2-C10).

Table:  Definition of parameters and model notation for Figures C1-C10.

| Parameter | Definition | Notation in figures |
| --------- | ---------- | -------------------------------------- |
| $\alpha_{i}$ | spatial random effect for for knot *i* | ``alpha.i`` where ``i`` is the numeric knot id |
| $\gamma_{x}$ | climate effect for covariate *x* | ``beta.x`` where ``x`` is the numeric climate covariate id |
| $\beta_{0t}$ | random effect for year *t* on the intercept | ``int_yr.t`` where ``t`` is the numeric year id |
| $\bar{\beta_{0}}$ | mean intercept | ``int_mu`` |
| $\beta_{1}$ | density-dependent effect | ``beta_mu`` |
| $\sigma_{\beta_{0}}^2$ | variance of year random effect | ``sig_yr`` |
| $\sigma_{\eta}^2$ | variance of spatial random effect | ``sig_a`` |

![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../results/rhat_plot.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_1.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_2.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_3.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_4.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_5.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_6.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_7.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_8.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_9.png)

References
----------

### 2015-11-13 scripts/rhat_calcs.R

@@ -7,7 +7,7 @@ rm(list=ls())
 ####  Load libraries -----------------------------------------------------------
 ####
 library(ggplot2)
-
+library(dplyr)
 
 
 ####
@@ -64,6 +64,6 @@ ggplot(rhats, aes(x=Rhat, y=Parameter))+
   xlab(expression(hat("R"))) + 
   ggtitle("Potential Scale Reduction Factors")
   
-ggsave("rhat_plot.pdf", height = 10)
+ggsave("rhat_plot.png", height = 10, dpi = 100)
 
 

### 2015-11-16 docs/appendixD.Rmd
adds in Collins edits and appendix D for spatial field

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{D\arabic{figure}}

Appendix D: Posterior estimate of spatial random effects ($\boldsymbol{\eta}$)
==============================================================================

```{r figure_D1,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=TRUE, echo=FALSE}
library(png)
library(grid)
img <- readPNG("./components/spatialfield.png")
grid.raster(img)
```

### 2015-11-16 docs/ms_sageAbundance.Rmd
adds in Collins edits and appendix D for spatial field

@@
 We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
-The CMIP5 provides projections from a suite of global circulations models (GCMs), and we used projections from 18 GCMs (Table A1) that produced weather projections for all three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+The CMIP5 provides projections from a suite of global circulations models (GCMs), and we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
 The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
-To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
+We chose these three RCP scenarios to demonstrate our approach because they have close analogues to previous emissions scenarios used by the IPCC.
+To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged the 18 GCM projections for each of the three emissions scenarios for the centroid of our study area over the time period of 2050 to 2100.
 We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
@@
 Model fitting required five days.
-We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.01 for all paramters) [@Gelman1992; @Gelman2009].
+We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all paramters) [@Gelman1992; @Gelman2009].
 
@@
 
-All parameters in our model converged on stable posterior distributions (Appendix X).
+All parameters in our model converged on stable posterior distributions (Appendix C).
 Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush annual percent cover change (Fig. 3).
-The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. X#).
+The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
@@
 The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
-A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline in the future [@Homer2015].
+A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
 The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
-The Homer et al.
-(2015) study also found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
+First, Homer et al. (2015) used downscaled IPCC projections that allowed for spatial variation in future climate, unlike our analysis here.
+Therefore, Homer et al. (2015) did predict some pixel-level increases in sagebrush cover since each pixel was subject to its own regression between climate and cover.
+Second, like our study, Homer et al.
+(2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
 Thus, the decline in precipitation led to a decline in sagebrush cover.

### 2015-11-16 docs/ms_sageAbundance.Rmd
adds in Eric's comments

@@
 #### Climate covariates
-Our approach is based on modeling how changes in plant populations relate to climate variables.
+Our approach is based on modeling how interannual changes in plant cover relate to seasonal climate variables.
 We downloaded daily historic weather data for our study site from the NASA Daymet data set (*available online*)[^link].
@@
 All parameters in our model converged on stable posterior distributions (Appendix C).
-Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush annual percent cover change (Fig. 3).
+Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 3).
 The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
@@
 In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy that rigorously accounted for spatial autocorrelation while also allowing for feasible computational time in model fitting.
+Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy that accounted for spatial autocorrelation while also allowing for feasible computational time in model fitting.
 Accounting for spatial autocorrelation in this large dataset allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 

### 2015-11-17 docs/ms_sageAbundance.Rmd
updates with some of Dave's comments
adds in some of Cam's edits; new scale_bar fxn

@@
 geometry: margin=1in
-
+linkcolor: black
+urlcolor: black
 ---
@@
 Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
-However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
+However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across broad species ranges.
 Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
@@
 
-Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
 SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
@@
 #### Remotely-sensed time series
-To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming[^link0] [@Homer2012].
+To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
@@
 
-[^link0]: *link to USGS data repo*
-
 #### Climate covariates
 Our approach is based on modeling how interannual changes in plant cover relate to seasonal climate variables.
-We downloaded daily historic weather data for our study site from the NASA Daymet data set (*available online*)[^link].
+We downloaded daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
 The Daymet weather data are interpolated between coarse observation units and captures some spatial variation.
@@
 We selected these variables based on previous studies (see *Introduction*), even though not all emerge as important predictors in our model.
-Several model selection techniques exist that could be applied to our approach [@Hooten2015], but we avoid model selection here to maximize focus on the general approach.
+To identify top covariates, we could have applied a variety of model selection techniques [@Hooten2015], but we avoid model selection here to maximize focus on the general approach.
 
@@
 We chose these three RCP scenarios to demonstrate our approach because they have close analogues to previous emissions scenarios used by the IPCC.
-To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged the 18 GCM projections for each of the three emissions scenarios for the centroid of our study area over the time period of 2050 to 2100.
+To simulate equilibrium sagebrush cover under projected future climate, we averaged the 18 GCM projections for each of the three emissions scenarios for the centroid of our study area over the time period of 2050 to 2100.
 We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
@@
 For temperature, we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
-For precipitation, we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation.
-We calculate precipitation and temperature changes for RCP 4.5, RCP 6.0, and RCP 8.5 scenarios.
+For precipitation, we calculated the proportional change: we subtracted average historic (1950-2000) precipitation from average projected (2050-2100) precipitation, and then divided by average historic (1950-2000) precipitation.
+We calculated precipitation and temperature changes for RCP 4.5, RCP 6.0, and RCP 8.5 scenarios because they are close analogues to previous ICPP scenarios.
 We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.
@@
 The last remaining obstacle is to parameterize \textbf{K}.
-Here we use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
+We use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
 This means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
@@
 Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
-For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
+For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
 We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
@@
 A coarser scale model would lose some fine-scale detail but could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
+However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger and larger pixels.
 
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=F}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
 library(png)
@@
 
-```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area.", eval=F}
+```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
+1).", eval=T}
 library(png)
@@
 ```{r figure_2,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.
-The x-axis is the standardized coefficient value because we fit the model using standardized covariate values.
-Only cumulative precipitation at time \\emph{t} is important (shown in dark grey; 90% CI does not overlap zero).", fig.height=4, eval=F}
+The x-axis is the standardized coefficient value because we fit the statistical model for sagebrush cover change (Eq.
+7) using standardized covariate values.
+Only cumulative precipitation at time \\emph{t} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t}; TmeanSpr1 = year \\emph{t-1} average spring temperature; ppt2 = year \\emph{t} fall through summer precipitation; TmeanSpr2 = year \\emph{t} average spring temperature; ppt1 = year \\emph{t-1} fall through summer precipitation.", fig.height=4, eval=T}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover, and prediction bias and standard deviation (C, D).
-All simulations used mean parameter values and no random year effects.", fig.height=8, eval=F}
+```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and standard deviation (C, D) for the extent of our spatial area at 30-m resolution.
+Observed equilibrium sagebrush cover (A) is the temporal mean of each pixel from the 28 year time series.
+Prediction results are from simulations that use mean parameter values and no random year effects.
+Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=T}
 library(png)
@@
 
-```{r figure_6,dependson="plot-options", fig.cap="Projected equilibrium cover under climate change scenarios.
-Top panels shows equilibrium cover based on simulations using observed cliamte.
+```{r figure_6,dependson="plot-options", fig.cap="Projected equilibrium cover under three IPCC climate change scenarios (RCP = Representative Concentration Pathways) for our study area in southwestern Wyoming.
+The top panel shows equilibrium cover based on simulations using observed climate.
 Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.
-Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the model.
-We used mean parameter estimates.", fig.height=8, eval=F}
+Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the statistical model.
+We used mean parameter estimates for all simulations.
+Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=T}
 library(png)
@@
 
-```{r figure_8,dependson="plot-options", fig.cap="Model forecasts under future climate for three IPCC emissions scenarios.
+```{r figure_8,dependson="plot-options", fig.cap="Observed (black line before 2011) and forecasted (colored lines after 2011) sagebrush percent cover.
+Forecasts were made for three IPCC emissions scenarios (RCPs 4.5, 6.0, and 8.5) and are for the period of 2012 to 2098.
 Shaded regions show 90% quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
@@
 Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
-Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=F}
+Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=T}
 library(png)

### 2015-11-17 R/ggplot_scale_bar.R

@@ -1,0 +1,129 @@
+##  Function for scale bar
+
+#--------#
+# Return a list whose elements are :
+# 	- rectangle : a data.frame containing the coordinates to draw the first rectangle ;
+# 	- rectangle2 : a data.frame containing the coordinates to draw the second rectangle ;
+# 	- legend : a data.frame containing the coordinates of the legend texts, and the texts as well.
+#
+# Arguments : #
+#-------------#
+# lon, lat : longitude and latitude of the bottom left point of the first rectangle to draw ;
+# distanceLon : length of each rectangle ;
+# distanceLat : width of each rectangle ;
+# distanceLegend : distance between rectangles and legend texts ;
+# dist.units : units of distance "km" (kilometers) (default), "nm" (nautical miles), "mi" (statute miles).
+createScaleBar <- function(lon,lat,distanceLon,distanceLat,distanceLegend, dist.units = "km"){
+  # First rectangle
+  bottomRight <- gcDestination(lon = lon, lat = lat, bearing = 90, dist = distanceLon, dist.units = dist.units, model = "WGS84")
+  
+  topLeft <- gcDestination(lon = lon, lat = lat, bearing = 0, dist = distanceLat, dist.units = dist.units, model = "WGS84")
+  rectangle <- cbind(lon=c(lon, lon, bottomRight[1,"long"], bottomRight[1,"long"], lon),
+                     lat = c(lat, topLeft[1,"lat"], topLeft[1,"lat"],lat, lat))
+  rectangle <- data.frame(rectangle, stringsAsFactors = FALSE)
+  
+  # Second rectangle t right of the first rectangle
+  bottomRight2 <- gcDestination(lon = lon, lat = lat, bearing = 90, dist = distanceLon*2, dist.units = dist.units, model = "WGS84")
+  rectangle2 <- cbind(lon = c(bottomRight[1,"long"], bottomRight[1,"long"], bottomRight2[1,"long"], bottomRight2[1,"long"], bottomRight[1,"long"]),
+                      lat=c(lat, topLeft[1,"lat"], topLeft[1,"lat"], lat, lat))
+  rectangle2 <- data.frame(rectangle2, stringsAsFactors = FALSE)
+  
+  # Now let's deal with the text
+  onTop <- gcDestination(lon = lon, lat = lat, bearing = 0, dist = distanceLegend, dist.units = dist.units, model = "WGS84")
[…]

### 2015-11-17 scripts/make_study_map.R

@@ -11,6 +11,8 @@ require(sp)
 require(rgdal)
 library(colorRamps)
 
+#Source scale_bar function
+source("/Users/atredenn/Desktop/ggplot_scale_bar.R")
 
 #Get multiplot function
 multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
@@ -105,11 +107,13 @@ ggplot()+
         panel.grid.major=element_blank(),
         panel.grid.minor=element_blank(),
         plot.background=element_blank()
-  )
+  )+
+  scaleBar(lon = -130, lat = 26, distanceLon = 500, distanceLat = 100, 
+           distanceLegend = 200, dist.unit = "km")
 ggsave("../docs/components/figure/us_wyo_studyarea_mark.png")
 
 ggplot()+
-  geom_raster(data=sage_oneyear, aes(y=Lat, x=Lon, fill=Cover))+
+  geom_raster(data=shrub.df, aes(y=y, x=x, fill=Cover))+
   theme_bw() +
   coord_equal() +
   scale_fill_gradient(high="white", low="black") +
@@ -118,10 +122,10 @@ ggplot()+
         panel.background=element_rect(fill="white"),
         panel.grid.major=element_blank(),
         panel.grid.minor=element_blank(),
-        plot.background=element_blank(),
-        axis.text.x=element_blank(),
-        axis.text.y=element_blank(),
-        axis.ticks=element_blank()
+        plot.background=element_blank()
+#         axis.text.x=element_blank(),
+#         axis.text.y=element_blank(),
+#         axis.ticks=element_blank()
   )
 ggsave("../docs/components/figure/oneyear_cover_example.png")
 

### 2015-11-18 docs/ms_sageAbundance.Rmd
adds new study area map with scale bar

@@
 Aldridge\textsuperscript{5}, Collin G.
-Homer\textsuperscript{6}, Andrew Kleinhesselink\textsuperscript{1}, and Peter B.
+Homer\textsuperscript{6}, Andrew R.
+Kleinhesselink\textsuperscript{1}, and Peter B.
 Adler\textsuperscript{1}}}
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming and a snapshot of the percent cover data in 1984.", fig.height=6, eval=T}
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
+Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=T}
 library(png)

### 2015-11-18 scripts/make_study_map.R

@@ -12,7 +12,7 @@ require(rgdal)
 library(colorRamps)
 
 #Source scale_bar function
-source("/Users/atredenn/Desktop/ggplot_scale_bar.R")
+source("../R/ggplot_scale_bar.R")
 
 #Get multiplot function
 multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
@@ -113,11 +113,11 @@ ggplot()+
 ggsave("../docs/components/figure/us_wyo_studyarea_mark.png")
 
 ggplot()+
-  geom_raster(data=shrub.df, aes(y=y, x=x, fill=Cover))+
+  geom_raster(data=sage_oneyear, aes(y=Lat, x=Lon, fill=Cover))+
   theme_bw() +
   coord_equal() +
   scale_fill_gradient(high="white", low="black") +
-  xlab("Longitude") + ylab("Latitude")+ 
+  xlab("Longitude (m)") + ylab("Latitude (m)")+ 
   theme(
         panel.background=element_rect(fill="white"),
         panel.grid.major=element_blank(),

### 2015-11-18 docs/ms_sageAbundance.Rmd
updates with UAESS paper number

@@
 ATT was supported by an NSF Postdoctoral Research Fellowship in Biology (DBI-1400370) and AK was supported by an NSF Graduate Research Fellowship.
-Additional support came from the Utah Agricultural Experiment Station, Utah State University, and this article is approved as journal paper number XXXX.
+Additional support came from the Utah Agricultural Experiment Station, Utah State University, and this article is approved as journal paper number 8856.
 We are grateful to Debra K.

### 2015-11-19 docs/appendixB.Rmd
updates ms with Daves knots comments

--- 2015-11-03 docs/appendixB.Rmd
@@
 X of the main text) to 500/3 = `r round(500/3,2)`.
-We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately $\sigma$ = `r round(500/3,2)`.
-This resulted in 121 knots (Fig.
+We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters.
+Knots were placed evenly across space in the x and y directions.
+This resulted in a 10 $\times$ 20 knot grid (200 knots; Fig.
 B2).
 
-![Spatial variogram of model residuals showing where spatial dependence disappears (dark red line at ~500 meters), and the value chosen for the kernal bandwidth in Eq.
-X of the main text (500/3; dashed black line).](components/avgresidual_variogram_withrange_subset.png)
+![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernal bandwidth in Eq.
+6 of the main text (500/3; dashed black line).](components/avgresidual_variogram_withrange_subset.png)
 
-![Map of our study area showing percent cover in 1984 and the location of the 121 evenly-spaced knots.](components/SAGE_Grid_wKnots_subset.png)
+![Map of our study area showing percent cover in 1984 and the location of the 200 evenly-spaced knots.](components/SAGE_Grid_wKnots_subset.png)

### 2015-11-19 docs/ms_sageAbundance.Rmd
updates ms with Daves knots comments

@@
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
-Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011; @Tredennick2015].
+Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011].
 
@@
 In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
-In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
+In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
 Thus, the full dataset contains 369,603 observations consisting of 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
@@
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
-The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t} (*lagPpt*), (2) year \emph{t} fall through summer precipitation (*ppt2*), (3) year \emph{t}-1 fall through summer precipitation (*ppt1*), (4) year \emph{t} average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-1 average spring temperature (*TmeanSpr1*).
+The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}-1 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt2*), (3) year \emph{t}-2 fall through summer precipitation (*ppt1*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-2 average spring temperature (*TmeanSpr1*), where *t-1* to *t* is the transition of interest.
 We selected these variables based on previous studies (see *Introduction*), even though not all emerge as important predictors in our model.
@@
 
+The spatial random effect ($\boldsymbol{\eta}$) accounts for spatial autocorrelation among pixel that occur closely in space.
+Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$) that essentially creates a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
 Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
-However, given the relative size of our dataset and computational limitations, we defined kernels around 121 knots (Fig.
-B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix B).
-The Bayesian posterior distribution can be expressed as
+However, given the relative size of our dataset and computational limitations, we defined kernels around 200 knots (Fig.
+B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
+An infinite number of knots would result in a perfectly accurate representation of the spatial process and covariance model.
+Since using an infinite number of knots is not possible, any use of the dimension reduction technique is an approximation, where the accuracy increases with the number of knots.
+Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.
+
+The Bayesian posterior distribution of our spatio-temporal model can be expressed as
 
@@
 [\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
-&\left(\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2].
+&\left(\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
 \end{align}
@@
 ### Accomodating zeros
-Our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$). 
+Our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$)). 
 Thus, our model does not accomodate zeros. 
@@
 Landscape-scale population models for plant species have been elusive and challenging to implement [@Schurr2012a; @Merow2014].
-Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
+Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [e.g., @Clark2014].
 We introduced a new statistical approach to modeling plant populations that overcomes two major hurdles for spatially-explicit population models.
-First, we used high resolution species-specific remote sensing data as a response variable.
-Thus, we had a spatially-continuous data product of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
+First, we used moderate resolution species-specific estimates of sagebrush percent cover estimated from remote sensing data as a response variable.
+Thus, we were able to fit a dynamic population model over a large spatial extent.
 In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
@@
 The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig.
-2), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
-The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 2).  
+3), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
+The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 3).
 Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
@@
 
-As expected given the relatively weak climate effects in this study area, our model projects only small changes in sagebrush cover in this area with future climate change. 
-Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 4,5).
+Our model projects only small changes in sagebrush cover in this area with future climate change, which is expected since only precipitation in the previous year emerged as an important covariate (Fig.
+3) and projections of future precipitation show only modest increases (Fig.
+2).
+Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 5,6).
 The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
@@
 Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig.
-5).
+6).
+Lastly, Homer et al.'s (2015) analysis relied on pixel-level regressions of cover through time against climate covariates.
+Our current approach is fundamentally different in that we specifically model the impact of weather on year-to-year changes in sagebrush cover.
+Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.
 
-There is no clear answer to what climate model to use when making ecological forecasts.
+As mentioned above, one of the main differences between our current work and that of Homer et al. (2015) is our choice of climate projections.
+Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
+These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013].
+Thus, there is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
 Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
-If a certain GCM tends to better predict weather in a specific study area, then that GCM may be given more weight than others.
+If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
 Extensions of these ideas are the subject of ongoing research.
@@
 This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).
-Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit climate covariates that match the resolution of the percent cover data.
-It would also be possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play important roles.
+We also did not allow for the possibility that different parts of our study area would respond to climate in different ways.
+This would require spatially-varying climate effects that substantially increase computational time.
+However, in some case, for example over very large regions, allowing the effect of climate to vary over space will be important.
+@Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
+Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit and spatially-varying climate covariates that match the resolution of the percent cover data.
+It is also possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play more important roles than we observed in our study area.
 
-While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.
-For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al. 2010).
+While we presented projections for sagebrush cover at the end of this century, one important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
+For example, our approach could be used to model the effects of a multi-year regional drought on this species [@Debinski2010].
 Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.
@@
 As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
-Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].
+Future modeling could include the effects of non-climate drivers -- including the effects of species interactions and disturbance [@Clark2014].
 For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2006].
@@
 This work is the outcome of a distributed graduate seminar led by PBA and supported by a National Science Foundation CAREER award (DEB-1054040).
+David T.
+Iles, Eric LaMalfa, and Rebecca Mann participated in project conception as part of the distributed graduate seminar and provided comments that improved the manuscript.
 ATT was supported by an NSF Postdoctoral Research Fellowship in Biology (DBI-1400370) and AK was supported by an NSF Graduate Research Fellowship.
@@
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
-We acknowledge the World Climate Research Programme's Working Group on Coupled Modelling, which is responsible for CMIP, and we thank the climate modeling groups (listed in Table X1) for producing and making available their model output.
+We acknowledge the World Climate Research Programme's Working Group on Coupled Modelling, which is responsible for CMIP, and we thank the climate modeling groups (listed in Table A1) for producing and making available their model output.
 For CMIP the U.S.
@@
 government.
-David T.
-Iles, Eric LaMalfa, and Rebecca Mann participated in project conception as part of the distributed graduate seminar and provided comments that improved the manuscript.
 
@@
 Only cumulative precipitation at time \\emph{t} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t}; TmeanSpr1 = year \\emph{t-1} average spring temperature; ppt2 = year \\emph{t} fall through summer precipitation; TmeanSpr2 = year \\emph{t} average spring temperature; ppt1 = year \\emph{t-1} fall through summer precipitation.", fig.height=4, eval=T}
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
 library(png)

### 2015-11-19 scripts/plot_projected_weather.R

@@ -85,8 +85,7 @@ dev.off()
 obs_weather <- read.csv("../data/FormattedClimate_WY_SA1.csv")
 obs_weather <- melt(obs_weather, id.vars = "year")
 obs_ppt <- subset(obs_weather, variable=="ppt2")
-model_weather <- ddply(ppt_agg, .(Year, season), summarise,
-                       value = mean(avg_value))
+ 
 model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
 precip_compare <- data.frame(year = model_weather$Year, 
                              precip = c(obs_ppt$value, model_weather$value),
@@ -100,9 +99,9 @@ model_weather <- ddply(temp_agg, .(Year, season), summarise,
                        value = mean(avg_value))
 model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
 temp_compare <- data.frame(year = model_weather$Year, 
-                             precip = c(obs_temp$value, model_weather$value),
+                             temp = c(obs_temp$value, model_weather$value),
                              value_source = rep(c("observed", "modeled"), each=34))
 ggplot(temp_compare)+
-  geom_boxplot(aes(x=value_source, y=precip))
+  geom_boxplot(aes(x=value_source, y=temp))
 
 

### 2015-11-24 docs/ms_sageAbundance.Rmd
add PBA edits to Introduction

@@
 \begin{centering}
-\textbf{\large{A spatio-temporal modeling framework for forecasting climate change impacts on plant populations over large spatial extents}}
+\textbf{\large{Forecasting climate change impacts on plant populations over large spatial extents}}
 
@@
 --------
-New remote sensing technology is producing ecological data at unprecedented spatial scales. 
-These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots. 
-However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time.
-Here we demonstrate how recent advances in the modeling of animal populations can be applied to remotely-sensed, species-specific estimates of plant cover at the landscape scale.
-Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
-We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
-We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. 
-Our analysis provides spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
-Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all.
-We expect the approach we develop here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.
+Plant populations models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
+Here we overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
+Using computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
+We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of this species.
+We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale.
+Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
+Our approach is an important first step to using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
 
@@
 -------------
-Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
-Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
+Forecasting the impacts of climate change on plant populations and communities is a central challenge for ecology [@Clark2001; @Petchey2015].
+Population models are ideally suited for meeting such a challenge because they provide a way to link climate drivers directly to population dynamics [@Hare2010; @Adler2012; @Ross2015].
 However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across broad species ranges.
-Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
+Almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
 Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
-Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011].
+The real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011].
 
-The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014]; however, developing useful models at this scale has been limited by the lack of time series data at large spatial extents and the lack of statistical machinary needed to fit high-dimensional spatial models.
+The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents aswell as statistical methods for fitting high-dimensional spatial models.
 Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
@@
 Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
-Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review].
-These new statistical methods have yet to be applied to RS-derived plant populations data at large scales.
+Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
+These new statistical methods have yet to be applied to RS-derived plant population data at broad scales.
 
-Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.
+Large-scale, spatially-explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes playing out at landscape and regional scales.
 Currently most assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
@@
 Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015].
-Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes.
-Large-scale dynamic population models could overcome both of these limitations.
-They would produce spatially explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
-These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models [@Luo2011a].
+Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached.
+Furthermore, direct validation of such predictions is extremely rare (cite a paleo paper?).
+Large-scale dynamic population models could overcome these limitations.
+They would produce spatially-explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
+These short-term predictions also would give modelers the opportunity to repeatedly validate and refine their models [@Luo2011a].
 
-Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data.
+Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
 SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
@@
 In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014].
-In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
+In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
 Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
-A large-scale spatially-explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
+A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
 
-Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.
-Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015].
-We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
-The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
-In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.
+Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict population impacts of climate change.
+Our approach relies on a process model motivated by Gompertz density-dependent population growth and a remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015]. 
+We account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
+The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the greatest potential to combine the information of population models (e.g., population status) and the spatial extent of species distribution models.
 

### 2015-11-30 scripts/traceplots.R
adds new results with more knots

@@ -104,3 +104,5 @@ lc.save <- rbind(lc1, lc2, lc3)
 ####
 setwd("/Users/atredenn/Repos/sageAbundance/results/")
 saveRDS(lc.save, outmcmcfile)
+
+

### 2015-11-30 scripts/traceplots.R
updates results with new MCMC results from equal more knots

@@ -21,7 +21,7 @@ library(scales)
 ####
 ####  Read in MCMC chunks from server
 ####
-setwd("/Volumes/A02046115/sage/")
+setwd("/Volumes/A02046115/sage2/")
 maxiter <- 10
 longchain1 <- readRDS(paste0(filestart, "1_chain2.RDS"))
 for(i in 2:maxiter){

### 2015-11-30 scripts/ms_scripts.R
updates ms_scripts to remove aspect test

@@ -133,13 +133,13 @@ dev.off()
 ####
 ####  Compare spatial field to landscape features ------------------------------
 ####
-slope <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/ancillary_data/ls3731_slope_subset.img")
-slopeDf <- as.data.frame(rasterToPoints(slope))
-eta_slope <- merge(oneyear, slopeDf, by.x=c("Lon", "Lat"), by.y=c("x", "y"))
-ggplot(eta_slope, aes(x=ls3731_slope_subset, y=eta))+
-  geom_point()+
-  stat_smooth(se=FALSE, color="black", method="gam", formula=y ~ s(x, bs = "cs"))+
-  theme_bw()
+# slope <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/ancillary_data/ls3731_slope_subset.img")
+# slopeDf <- as.data.frame(rasterToPoints(slope))
+# eta_slope <- merge(oneyear, slopeDf, by.x=c("Lon", "Lat"), by.y=c("x", "y"))
+# ggplot(eta_slope, aes(x=ls3731_slope_subset, y=eta))+
+#   geom_point()+
+#   stat_smooth(se=FALSE, color="black", method="gam", formula=y ~ s(x, bs = "cs"))+
+#   theme_bw()
 
 
 ####

### 2015-11-30 scripts/rhat_calcs.R
updates Rhat plot with multiple panels

@@ -54,12 +54,12 @@ colnames(out.mcmc) <- c("Iteration", "Parameter", "value", "mcmc_iter", "Chain")
 attributes(out.mcmc)$nChains <- length(unique(out.mcmc$Chain))
 attributes(out.mcmc)$nIterations <- length(unique(out.mcmc$Iteration))
 rhats <- get_rhats(out.mcmc)
-rhats$facet_id <- rep(c(1,2), each=nrow(rhats)/2)
+rhats$facet_id <- rep(c(1,2,3,4), each=nrow(rhats)/4)
 
 ggplot(rhats, aes(x=Rhat, y=Parameter))+
   geom_point()+
   geom_vline(aes(xintercept=1.1), col="red")+
-  facet_wrap("facet_id", ncol=2, scales="free_y")+
+  facet_wrap("facet_id", ncol=4, scales="free_y")+
   xlim(min(rhats$Rhat, na.rm = TRUE), 1.5)+
   xlab(expression(hat("R"))) + 
   ggtitle("Potential Scale Reduction Factors")

### 2015-11-30 docs/appendixB.Rmd
updates appendices

@@
 Knots were placed evenly across space in the x and y directions.
-This resulted in a 10 $\times$ 20 knot grid (200 knots; Fig.
+This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig.
 B2).
@@
 ![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernal bandwidth in Eq.
-6 of the main text (500/3; dashed black line).](components/avgresidual_variogram_withrange_subset.png)
+6 of the main text (500/3; dashed black line).](../results/avgresidual_variogram_withrange_subset.png)
 
-![Map of our study area showing percent cover in 1984 and the location of the 200 evenly-spaced knots.](components/SAGE_Grid_wKnots_subset.png)
+![Map of our study area showing percent cover in 1984 and the location of the 231 evenly-spaced knots.](../results/SAGE_Grid_wKnots_subset.png)

### 2015-11-30 docs/appendixD.Rmd
updates appendices

@@
 library(grid)
-img <- readPNG("./components/spatialfield.png")
+img <- readPNG("../results/spatialfield.png")
 grid.raster(img)

### 2015-11-30 scripts/subsetData_getKnots.R

@@ -13,9 +13,8 @@ library(plyr);     library(ggplot2)
 ####
 #### Bring in data
 ####
-setwd("/Users/atredenn/Dropbox/sageAbundance_data/")
-climD<-read.csv("./studyarea1/climate/DAYMET/FormattedClimate_WY_SA1.csv")
-rawD<-read.csv("./WY_SAGECoverData_V2check.csv")
+climD<-read.csv("../data/FormattedClimate_WY_SA1.csv")
+rawD<-read.csv("/Users/atredenn/Dropbox/sageAbundance_data/WY_SAGECoverData_V2check.csv")
 shrub.ras <- raster("/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/studyarea1/sage/ls3731_00sage_subset_est.img")
 
 # merge in climate data 
@@ -82,13 +81,15 @@ dat <- subset(resD, select = c("Lon", "Lat", "resid"))
 coordinates(dat) <- c("Lon", "Lat")
 varMod <- variogram(resid~1, data=dat, width=50)
 
-png("avgresidual_variogram_withrange_subset.png", width = 5, height = 5, 
+results_path <- "~/Repos/sageAbundance/results/"
+png(paste0(results_path,"avgresidual_variogram_withrange_subset.png"), width = 5, height = 5, 
     units = "in", res=200)
 plot(varMod$dist, varMod$gamma, ylim=c(0,max(varMod$gamma)), col="dodgerblue", 
      type="l", main="variogram of model residuals", xlab="distance (m)",
      ylab="semivariance")
 points(varMod$dist, varMod$gamma, ylim=c(0,max(varMod$gamma)), col="dodgerblue", pch=21, bg="white")
 abline(v = 500/3, lwd=3, lty=2)
+abline(v = 500, lwd=3, lty=1, col="darkred")
 dev.off()
 
 range_parameter <- round(500/3) #set to 1/3 of the spatial dependence in the residuals
@@ -164,8 +165,8 @@ tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),
                 axis.title=element_text(size=14),text=element_text(size=16),
                 legend.text=element_text(size=12), legend.title=element_text(size=16))
 
-setwd("/Users/atredenn/Dropbox/sageAbundance_data/")
-png('SAGE_Grid_wKnots_subset.png', width=6, height=3, units = "in", res=200)
[…]

### 2015-12-07 docs/ms_sageAbundance.Rmd
PBA edits on Methods and Results

@@
 #### Remotely-sensed time series
-To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming [@Homer2012].
+To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
@@
 In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
-Thus, the full dataset contains 369,603 observations consisting of 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
+Thus, the full dataset contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
 
@@
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
-The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}-1 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt2*), (3) year \emph{t}-2 fall through summer precipitation (*ppt1*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr2*), and (5) year \emph{t}-2 average spring temperature (*TmeanSpr1*), where *t-1* to *t* is the transition of interest.
-We selected these variables based on previous studies (see *Introduction*), even though not all emerge as important predictors in our model.
-To identify top covariates, we could have applied a variety of model selection techniques [@Hooten2015], but we avoid model selection here to maximize focus on the general approach.
+The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
+We selected these variables a priori based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.
+We could have applied a variety of model selection techniques [@Hooten2015], but chose not to because our focus is on the general approach.
 
@@
 To forecast future population states, we required future projections of climate for our study area.
-We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
-The CMIP5 provides projections from a suite of global circulations models (GCMs), and we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
-The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
-We chose these three RCP scenarios to demonstrate our approach because they have close analogues to previous emissions scenarios used by the IPCC.
+We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
+The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.
 To simulate equilibrium sagebrush cover under projected future climate, we averaged the 18 GCM projections for each of the three emissions scenarios for the centroid of our study area over the time period of 2050 to 2100.
+[ANDREW, I JUST LEARNED PEOPLE
+DON'T LIKE THIS. IS IT COMPUTATIONALLY IMPOSSIBLE TO AVERAGE AFTER SIMULATING ON 
+EACH ENSEMBLE MEMBER?]
 We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
+[THE SECOND HALF OF THIS SENTENCE IS VERY CONFUSING, TRY AGAIN?
+ACTUALLY, THE REST OF THIS PARAGRAPH IS JUST AS CONFUSING!
+WHAT THE HELL??
+I THINK
+YOU NEED TO EXPLAIN THAT WE DID TWO TYPES OF SIMULATIONS, COMPARISON OF BASELINE TO 
+FUTURE EQUILIBRIUM, AND DYNAMIC PROJECTION FROM PRESENT INTO FUTURE. ONLY THEN EXPLAIN HOW YOU USED THE GCM OUTPUT...]
 To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
@@
 We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.
+[DID YOU AGGREGATE ALL THE GCM ENSEMBLE MEMEMBERS AND THEN JUST DID ONE
+RUN OF OUR MODEL? MANY RUNS? PARAMETER UNCERTAINTY? OR WILL YOU EXPLAIN THIS LATER
+IN METHODS?]
 
@@
 Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to improve convergence during the model fitting stage and to allow for easier prior specification.
-The intercept, $\beta_{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
+The intercept, $\beta_{0,t}$, was allowed to vary through time; these random year effects recognize
+that all observations from a particular year share the same climate covariates and thus are not independent.
 We used a Poisson likelihood because integer percent cover values in the sagebrush data product can be considered a form of count data.
 We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$).
-There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we dealt with the small number of zero percent cover observations.
+There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we handled the small number of zero percent cover observations.
 
-The spatial random effect ($\boldsymbol{\eta}$) accounts for spatial autocorrelation among pixel that occur closely in space.
-Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$) that essentially creates a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
-Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
-This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
-Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation.
-This greatly speeds up model fitting by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
+The spatial random effect ($\boldsymbol{\eta}$) accounts for spatial autocorrelation among pixels that occur near each other in space.
+Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$), creating a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
+Fitting the model with a spatial random effect ($\eta$) is computationally for large data sets like ours.
+This is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
+Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation by reducing the size of the spatial covariance matrices that need to be inverted at each MCMC iteration.
 Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
-We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
+Our dimension reduction strategy expresses the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
 We can then define the spatial effect as
@@
 
-where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
+and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
@@
 Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.
+[ARE THESE LAST THREE SENTENCES OF THIS PARAGRAPH REALLY NECESSARY?]
 
@@
 We fit this simple model using the `lm' command in R [@R2013].
+For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied.
 
@@
 ####Simulating the process
-We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
+We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
 Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
@@
 We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
-For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.
+For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
 For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
-We used the posterior mean of each parameter for all simulations, except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
+We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
 The random year effects were not used in any simulation.
@@
 -------
-Averaged across all GCMs, precipitation and temperature in our study area are projected to increase by 2050-2100 relative to observed climate from 1950-2000, with the magnitude depending on the RCP scenario (Table 1).
-Trajectories of our climate covariates from GCM projections, used to make yearly forecasts, show similar trends (Fig. 2). 
-GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections.
+Averaging across all GCMs, precipitation and temperature in our study area are projected to increase; the magnitude of increase depends on the RCP scenario (Table 1). 
+Trajectories of our climate covariates from GCM projections show similar trends (Fig. 2). 
+GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections. [I AM TEMPTED TO CUT THIS LAST LINE]
 
 All parameters in our model converged on stable posterior distributions (Appendix C).
-Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change from year-to-year (Fig. 3).
+Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig.
+3).
+In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
 The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
@@
 
-When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover in space (Fig. 4A,B).
+When we simulated the pixel-based population model based on observed climate it was able to reproduce the spatial pattern of observed percent cover, averaged over time (Fig. 4A,B).
 Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
-Nonetheless, observations and predictions are qualitatively similar (Fig. 4A, B).
 Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
@@
 When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
-The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in sagebrush cover heterogeneity because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
+The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
+Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
 For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
@@
 Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
-This result reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).
+The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).
 

### 2015-12-07 scripts/climate_formatting/Climate_Proj_Calcs_CMIP5_byModel.R
adds climate projections by model

@@ -1,0 +1,111 @@
+## Script to average and aggregate all CMIP5 model projections
+
+##  Relies on data downloaded from the CMIP5 website:
+##  http://cmip-pcmdi.llnl.gov/cmip5/
+
+##  Authors:      Andrew Tredennick and Peter Adler
+##  Email:        atredenn@gmail.com
+##  Date created: 10-10-2013
+
+
+
+### Clean the workspace and set working dir
+rm(list=ls()) 
+setwd("../../data/climate/PROJECTIONS/CMIP5/")
+
+
+####
+####  Load necessary libraries -------------------------------------------------
+####
+library(reshape2)
+library(grid)
+
+
+
+####
+####  Read in and formate precip projections -----------------------------------
+####
+ppt <- as.data.frame(read.csv("pr.csv", header=FALSE))
+tmp<-read.table("COLS_pr.txt")
+tmp<-as.character(tmp[,1])
+colnames(ppt) <- c("Year", "Month",tmp)
+ppt <- melt(ppt, id.vars=c("Year", "Month"))
+ppt[,4] <- as.numeric(ppt[,4]) #NAs coerced for a couple December 2099 null values 
+tmp<-unlist(strsplit(x=as.character(ppt$variable),split=".",fixed=T))
+tmp<-matrix(tmp,nrow=length(tmp)/3,ncol=3,byrow=T)
+colnames(tmp)<-c("model","rep","scenario")
+ppt<-cbind(ppt,tmp)
+ppt$period<-cut(ppt$Year,breaks=c(1950,2000,2050,2100),
+  include.lowest=T,labels=c("past","present","future"))
+ppt$season<-ifelse(ppt$Month > 6 & ppt$Month < 10,"summer","fall2spr")
+
+pptMeans<-aggregate(value~period+season+scenario+model,data=ppt,FUN=mean)
+allmods <- unique(pptMeans$model)
+keeps <- character(length(allmods))
+my_scens <- c("rcp45", "rcp60", "rcp85")
+for(i in 1:length(allmods)){
+  tmp <- subset(pptMeans, model==allmods[i])
+  tmp.scns <- unique(tmp$scenario)
+  flag <- length(which(my_scens %in% tmp.scns == FALSE))
+  ifelse(flag>0, keeps[i]<-"no", keeps[i]<-"yes")
+}
+modelkeeps <- data.frame(model=allmods,
+                         allscenarios=keeps)
[…]

### 2015-12-07 scripts/ms_scripts.R
adds by model climate projections

@@ -321,8 +321,15 @@ for(t in 1:time.steps){
   tmp.mu <- mean_params[mean_params$Parameter=="int_mu","..1"] + dens.dep + sum(betas*Xtmp)
   tmp.mu <- exp(tmp.mu + eta)
   tmp.out <- rpois(ncol(ex.mat), lambda = tmp.mu)
-  tmp.out[which(tmp.out>100)] <- mean(ex.mat[t-1,])
+  
+  #Colonization
+  zeros <- which(lagcover==0)
+  colonizers <- rbinom(length(zeros), size = 1, antilogit(col.intercept))
+  colonizer.cover <- colonizers*avg.new.cover
+  tmp.out[zeros] <- colonizer.cover
+  
   ex.mat[t,] <- tmp.out
+  
 }
 
 obs.cover <- growD[,c("Year","Cover","ID")]
@@ -335,21 +342,20 @@ ex.melt <- ex.melt[with(ex.melt, order(year)), ]
 colnames(ex.melt) <- c("Year", "ID", "cover.pred")
 error.df <- merge(obs.cover, ex.melt)
 error.df$error <- with(error.df, Cover - cover.pred)
-rmse <- sqrt(mean(error.df$error^2))
-cor(error.df$Cover, error.df$cover.pred)
+insamp.rmse <- sqrt(mean(error.df$error^2))
+insamp.cor <- cor(error.df$Cover, error.df$cover.pred)
+out <- data.frame(insamp_rmse = insamp.rmse,
+                  insamp_cor = insamp.cor)
+write.csv(out, "../results/insample_error_cor.csv")
 
 # pred.cover <- melt(ex.mat)
 # plot(obs.cover$Cover, pred.cover$value, xlim=c(0,20), ylim=c(0,20))
 # abline(0,1, col="red")
-
-
-
 # brks <- 30
 # y <- hist(growD$Cover, freq=FALSE, breaks=brks)
 # yhat <- hist(pred.cover$value, freq=FALSE, breaks=brks)
 # hist_ydata <- data.frame(x = y$breaks, y=c(y$density,0))
 # hist_yhatdata <- data.frame(x = yhat$breaks, y=c(yhat$density,0))
-# 
 # ggplot()+
 #   geom_step(data=hist_ydata, aes(x=x, y=y), size=1.5)+
 #   geom_step(data=hist_yhatdata, aes(x=x, y=y), col="darkorange", size=1)+
[…]

### 2015-12-11 docs/ms_sageAbundance.Rmd

@@
 -----------
-Landscape-scale population models for plant species have been elusive and challenging to implement [@Schurr2012a; @Merow2014].
-Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [e.g., @Clark2014].
-We introduced a new statistical approach to modeling plant populations that overcomes two major hurdles for spatially-explicit population models.
-First, we used moderate resolution species-specific estimates of sagebrush percent cover estimated from remote sensing data as a response variable.
-Thus, we were able to fit a dynamic population model over a large spatial extent.
-In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
-Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy that accounted for spatial autocorrelation while also allowing for feasible computational time in model fitting.
-Accounting for spatial autocorrelation in this large dataset allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 
+Despite the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
+Our approach for modeling plant populations overcomes two major hurdles for spatially-explicit population models.
+First, we used moderate resolution, remotely-sensed estimates of sagebrush percent cover as a response variable, enabling us to fit a dynamic population model over a large spatial extent.
+Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we fit the model using a dimension reduction strategy that accounted for spatial autocorrelation within a feasible computational time.
+Accounting for spatial autocorrelation allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 
 The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
@@
 
-Our model projects only small changes in sagebrush cover in this area with future climate change, which is expected since only precipitation in the previous year emerged as an important covariate (Fig.
+Our model projects only small changes in sagebrush cover in this area with future climate change.
+This result reflects the fact that only precipitation in the previous year emerged as an important covariate (Fig.
 3) and projections of future precipitation show only modest increases (Fig.
 2).
-Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 5,6).
+Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs.
+5,6).
 The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
@@
 These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013].
-Thus, there is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
+There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
 Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
@@
 We are grateful to Debra K.
-Meyer at USGS EROS for exctracting the data set used in this paper.
+Meyer at USGS EROS for extracting the data set used in this paper.
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.

### 2015-12-11 docs/ms_sageAbundance.Rmd
updates to discussion section

@@
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
-The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
+The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46 [@Homer2012].
 To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
@@
 
-To forecast future population states, we required future projections of climate for our study area.
+We required future projections of climate for our study area to conduct two sets of simulations: (1) comparisons of equilibrium sagebrush cover under observed climate to forecasted equilibrium cover under future climate scenarios and (2) dynamic forecasts from the present into the future.
 We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
@@
 The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.
-To simulate equilibrium sagebrush cover under projected future climate, we averaged the 18 GCM projections for each of the three emissions scenarios for the centroid of our study area over the time period of 2050 to 2100.
-[ANDREW, I JUST LEARNED PEOPLE
-DON'T LIKE THIS. IS IT COMPUTATIONALLY IMPOSSIBLE TO AVERAGE AFTER SIMULATING ON 
-EACH ENSEMBLE MEMBER?]
-We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
-[THE SECOND HALF OF THIS SENTENCE IS VERY CONFUSING, TRY AGAIN?
-ACTUALLY, THE REST OF THIS PARAGRAPH IS JUST AS CONFUSING!
-WHAT THE HELL??
-I THINK
-YOU NEED TO EXPLAIN THAT WE DID TWO TYPES OF SIMULATIONS, COMPARISON OF BASELINE TO 
-FUTURE EQUILIBRIUM, AND DYNAMIC PROJECTION FROM PRESENT INTO FUTURE. ONLY THEN EXPLAIN HOW YOU USED THE GCM OUTPUT...]
-To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
-For temperature, we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
-For precipitation, we calculated the proportional change: we subtracted average historic (1950-2000) precipitation from average projected (2050-2100) precipitation, and then divided by average historic (1950-2000) precipitation.
-We calculated precipitation and temperature changes for RCP 4.5, RCP 6.0, and RCP 8.5 scenarios because they are close analogues to previous ICPP scenarios.
-We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.
+
+To simulate equilibrium sagebrush cover under projected future climate we applied average projected changes in precipitation and temperature to the observed climate time series.
+For each GCM and RCP scenario combination, we calculated average precipitation and temperature over the 1950-2000 time period and the 2050-2098 time period.
+We then calculated the absolute change in temperature between the two time periods ($\delta T$) and the proportional change in precipitation between the two time periods ($\delta P$) for each GCM and RCP scenario combination.
+Lastly, we applied $\delta T$ and $\delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
+These generated climate time series were used to simulate equilibrium sagebrush cover (see *Simulating the process* below).
+We simulated equilibrium cover separately for each GCM and RCP scenario combination, but we show the average projected climate changes across all models in Table 1.
 
@@
 
-In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
-We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.
-[DID YOU AGGREGATE ALL THE GCM ENSEMBLE MEMEMBERS AND THEN JUST DID ONE
-RUN OF OUR MODEL? MANY RUNS? PARAMETER UNCERTAINTY? OR WILL YOU EXPLAIN THIS LATER
-IN METHODS?]
+For dynamic forecasts we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
+We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
+These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
 
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
-However, given the relative size of our dataset and computational limitations, we defined kernels around 200 knots (Fig.
+However, given the relative size of our dataset and computational limitations, we defined kernels around 221 knots (Fig.
 B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
@@
 Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.
-[ARE THESE LAST THREE SENTENCES OF THIS PARAGRAPH REALLY NECESSARY?]
 
@@
 where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($\alpha$).
-We fit this simple model using the `lm' command in R [@R2013].
+We fit this simple model using the `glm' command in R [@R2013].
 For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied.
@@
 For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
-We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
+We ran equilibrium simulations (1 and 3) for 2,000 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
+To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
 For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
 For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
+We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
 We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
@@
 -----------
-Despite the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-Our approach for modeling plant populations overcomes two major hurdles for spatially-explicit population models.
-First, we used moderate resolution, remotely-sensed estimates of sagebrush percent cover as a response variable, enabling us to fit a dynamic population model over a large spatial extent.
-Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we fit the model using a dimension reduction strategy that accounted for spatial autocorrelation within a feasible computational time.
-Accounting for spatial autocorrelation allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 
-The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
+Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
+We outlined a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
+Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
+We first discuss our results specific to sagebrush and then turn our attention to the general implications of our proposed modeling approach.
 
+### Sagebrush response to climate and climate change
 The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig.
@@
 
+While we presented projections for sagebrush cover at the end of this century, one important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
+For example, our approach could be used to model the effects of a multi-year regional drought on this species [@Debinski2010].
+Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.
+It would also provide opportunities for refinement of the model itself.
+This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015].
+
+### A landscape-scale plant population model
+
+Our approach for modeling plant populations overcomes two major hurdles for spatially-explicit population models.
+First, we used moderate resolution, remotely-sensed estimates of sagebrush percent cover as a response variable, enabling us to fit a dynamic population model over a large spatial extent.
+Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we fit the model using a dimension reduction strategy that accounted for spatial autocorrelation within a feasible computational time.
+Accounting for spatial autocorrelation allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region. 
+The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
+
 Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
@@
 It is also possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play more important roles than we observed in our study area.
-
-While we presented projections for sagebrush cover at the end of this century, one important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
-For example, our approach could be used to model the effects of a multi-year regional drought on this species [@Debinski2010].
-Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.
-It would also provide opportunities for refinement of the model itself.
-This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015].
 
@@
 ```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
-Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=T}
+Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=F}
 library(png)
@@
 ```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
-1).", eval=T}
+1).", eval=F}
 library(png)
@@
 Only cumulative precipitation at time \\emph{t} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=F}
 library(png)
@@
 Prediction results are from simulations that use mean parameter values and no random year effects.
-Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=T}
+Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=F}
 library(png)
@@
 We used mean parameter estimates for all simulations.
-Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=T}
+Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=F}
 library(png)

### 2015-12-14 docs/ms_sageAbundance.Rmd
moves discussion section around

@@
 
-We required future projections of climate for our study area to conduct two sets of simulations: (1) comparisons of equilibrium sagebrush cover under observed climate to forecasted equilibrium cover under future climate scenarios and (2) dynamic forecasts from the present into the future.
-We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
-The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
-The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.
-
-To simulate equilibrium sagebrush cover under projected future climate we applied average projected changes in precipitation and temperature to the observed climate time series.
-For each GCM and RCP scenario combination, we calculated average precipitation and temperature over the 1950-2000 time period and the 2050-2098 time period.
-We then calculated the absolute change in temperature between the two time periods ($\delta T$) and the proportional change in precipitation between the two time periods ($\delta P$) for each GCM and RCP scenario combination.
-Lastly, we applied $\delta T$ and $\delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
-These generated climate time series were used to simulate equilibrium sagebrush cover (see *Simulating the process* below).
-We simulated equilibrium cover separately for each GCM and RCP scenario combination, but we show the average projected climate changes across all models in Table 1.
-
-[^link2]: http://cmip-pcmdi.llnl.gov/cmip5/
-[^link3]: http://tntcat.iiasa.ac.at/RcpDb/
-
-For dynamic forecasts we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
-We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
-These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
-
 ### Additive spatio-temporal model for sagebrush cover
@@
 
-### Computing
-####Fitting the model
+### Fitting the model
 We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
@@
 
-####Simulating the process
+### Simulating the process
 We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098.
@@
 
+We required future projections of climate for our study area to conduct the equilibrium and temporally-explicit forecasts described above.
+We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
+The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
+The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.
+
+To simulate equilibrium sagebrush cover under projected future climate we applied average projected changes in precipitation and temperature to the observed climate time series.
+For each GCM and RCP scenario combination, we calculated average precipitation and temperature over the 1950-2000 time period and the 2050-2098 time period.
+We then calculated the absolute change in temperature between the two time periods ($\delta T$) and the proportional change in precipitation between the two time periods ($\delta P$) for each GCM and RCP scenario combination.
+Lastly, we applied $\delta T$ and $\delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
+These generated climate time series were used to simulate equilibrium sagebrush cover (see *Simulating the process* below).
+We simulated equilibrium cover separately for each GCM and RCP scenario combination, but we show the average projected climate changes across all models in Table 1.
+
+[^link2]: http://cmip-pcmdi.llnl.gov/cmip5/
+[^link3]: http://tntcat.iiasa.ac.at/RcpDb/
+
+For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to make simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
+We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
+These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
+
 Results
@@
 Trajectories of our climate covariates from GCM projections show similar trends (Fig. 2).
-GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections. [I AM TEMPTED TO CUT THIS LAST LINE]
 
@@
 The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
+Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5).
 For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
@@
 ### Sagebrush response to climate and climate change
-The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig.
-3), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013].
-The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 3).
-Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
-As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
+The amount of precipitation in the previous year (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change, followed by spring temperature (Fig. 3).
+These results are concordant with previous research.
+Wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
+However, the mean estimates for the climate effects are relatively small on the standardized scale (Fig. 3).
+Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
+Given findings from previous research on the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate or that our model is poorly specified.
+More likely is that aggregated estimates of plant abundance, like percent cover, mask interannual variability at the level of the individual plant.
+Such averaging inevitably reduces interannual variability, which also makes detecting drivers of internanual variability more difficult.
+Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology.
 
-Our model projects only small changes in sagebrush cover in this area with future climate change.
+Under projected climate, we forecast small positive changes in sagebrush cover in this area for all RCP scenarios (Figs. 5,6).
+Our forecast is a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
+First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
+Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes,
+
 This result reflects the fact that only precipitation in the previous year emerged as an important covariate (Fig.
-3) and projections of future precipitation show only modest increases (Fig.
+3) and projections of future precipitation show only modest increases (Table 1, Fig.
 2).
-Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs.
-5,6).
-The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
-A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
+Likewise, 
+While we forecast an increase in sagebrush cover at our study area, SDM studies tend to project declines in sagebrush abundance in the future due to temperature increases [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+Why do our results differ?
+High elevation site...winter water good...
+
+A previous analysis of a different subset of the remote sensing data set we used also made projections of sagebrush decline, on average, in the future [@Homer2015].
 The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
@@
 Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.
-
-As mentioned above, one of the main differences between our current work and that of Homer et al. (2015) is our choice of climate projections.
-Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
-These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013].
-There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
-Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
-If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
-Extensions of these ideas are the subject of ongoing research.
 
@@
 It is also possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play more important roles than we observed in our study area.
+
+As mentioned above, one of the main differences between our current work and that of Homer et al. (2015) is our choice of climate projections.
+Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
+These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013].
+There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
+Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
+If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
+Extensions of these ideas are the subject of ongoing research.
 

### 2015-12-14 docs/ms_sageAbundance.Rmd
more discussion restructuring

@@
 Forecasting the impacts of climate change on plant populations and communities is a central challenge for ecology [@Clark2001; @Petchey2015].
-Population models are ideally suited for meeting such a challenge because they provide a way to link climate drivers directly to population dynamics [@Hare2010; @Adler2012; @Ross2015].
+Population models are ideally suited for meeting such a challenge because they provide a way to link climate drivers directly to population dynamics [@Hare2010; @Adler2012; @Ross2015; @Shriver2015].
 However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across broad species ranges.
@@
 All parameters in our model converged on stable posterior distributions (Appendix C).
-Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig.
+Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig. 3).
+In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
+Other climate effects strongly overlapped zero but their posterior means were positive, except for fall-through-spring precipitation two years before a cover transition, whose posterior mean was negative (Fig.
 3).
-In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
 The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
@@
 Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
-We first discuss our results specific to sagebrush and then turn our attention to the general implications of our proposed modeling approach.
+The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which inspires confidence in our forecasts.
+More generally, the approach we described is applicable to any spatially-explicit time series of plant population abundance and opens the door to forecasts of plant populations at the landscape-scale.
 
 ### Sagebrush response to climate and climate change
-The amount of precipitation in the previous year (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change, followed by spring temperature (Fig. 3).
-These results are concordant with previous research.
+The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
+All other covariates related to precipitation and temperature were also positive on average, except for the effect of fall-through-spring precipitation two years before a cover transition (Fig. 3).
+These effects, detected at a spatial resolution 30 meters, are consistent with individual-level responses of sagebrush to climate-related variables.
 Wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
@@
 First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
-Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes,
+Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes, while the projected change in temperature was large (Table 1, Fig. 2).
+Lastly, the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
 
-This result reflects the fact that only precipitation in the previous year emerged as an important covariate (Fig.
-3) and projections of future precipitation show only modest increases (Table 1, Fig.
-2).
-Likewise, 
 While we forecast an increase in sagebrush cover at our study area, SDM studies tend to project declines in sagebrush abundance in the future due to temperature increases [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
 Why do our results differ?
-High elevation site...winter water good...
+There are many potential reasons, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in the cold range of the sagebrush distrubtion is the most likely explanation.
+The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b].
+Sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
+In warmer parts the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
+In colder regions, however, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
+Since our study area lies at the cold end of the sagebrush distribution, the weak positive response to temperature detected by our statistical model (Fig.
+3) and carried through to our forecasts (Figs.
+5,6) likely represents the positive effect of earlier snowmelt.
+The positive effect of precipitation we found is consistent with findings from across the sagebrush distribution (CITATIONS), adding further /.//.
 
-A previous analysis of a different subset of the remote sensing data set we used also made projections of sagebrush decline, on average, in the future [@Homer2015].
+Also in contrast to our results, a previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
 The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
@@
 
-While we presented projections for sagebrush cover at the end of this century, one important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
-For example, our approach could be used to model the effects of a multi-year regional drought on this species [@Debinski2010].
-Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.
+We presented projections for sagebrush cover at the end of this century, but an important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
+For example, we could forecast the effects of a multi-year regional drought on this species [@Debinski2010].
+Validating spatial population models against short-term predictions would give ecological forecasters a way to assess the long-term forecasts produced by the model.
 It would also provide opportunities for refinement of the model itself.
-This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015].
+This cycle of prediction, validation, and refinement is missing from most population-level forecasts of the effects of climate change currently produced.
 
-### A landscape-scale plant population model
+### A flexible, landscape-scale plant population modeling approach
 

### 2015-12-14 docs/ms_sageAbundance.Rmd
more updates to discussion structure

@@
 ### Sagebrush response to climate and climate change
+The climate effects estimated in our model, and detected at a spatial resolution of 30 meters, are consistent with individual-level responses of sagebrush to climate-related variables.
+At the level of individual plants, research shows that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
+Mirroring those individual-level responses, posterior means for all climate effects in our cover change model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig.
+3).
 The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
-All other covariates related to precipitation and temperature were also positive on average, except for the effect of fall-through-spring precipitation two years before a cover transition (Fig. 3).
-These effects, detected at a spatial resolution 30 meters, are consistent with individual-level responses of sagebrush to climate-related variables.
-Wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
 However, the mean estimates for the climate effects are relatively small on the standardized scale (Fig. 3).
 Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
-Given findings from previous research on the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate or that our model is poorly specified.
+Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
+Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
+But, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration.
 More likely is that aggregated estimates of plant abundance, like percent cover, mask interannual variability at the level of the individual plant.
 Such averaging inevitably reduces interannual variability, which also makes detecting drivers of internanual variability more difficult.
-Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology.
+Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
 
 Under projected climate, we forecast small positive changes in sagebrush cover in this area for all RCP scenarios (Figs. 5,6).
-Our forecast is a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
+Our forecasts are a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
 First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
@@
 Lastly, the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
+An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the increase in spatial heterogeity we forecast: the distribution of sagebrush cover expands rather than just shifting (Fig.
+D1).
+This effect is seen in the spatial forecasts where low cover pixels are constrained to low cover while high cover pixels increase (Fig. 5).
+Had we not explicitly accounted for spatial-dependence in our model, this result would be missed.
 
@@
 There are many potential reasons, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in the cold range of the sagebrush distrubtion is the most likely explanation.
-The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b].
-Sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
-In warmer parts the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
-In colder regions, however, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
+The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
+The generally positive effect of precipitation we found is consistent with findings from across the sagebrush distribution [@Schlaepfer2012; @Schlaepfer2015].
+However, the effect of temperature may vary depending on location.
+In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
+In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
 Since our study area lies at the cold end of the sagebrush distribution, the weak positive response to temperature detected by our statistical model (Fig.
@@
 5,6) likely represents the positive effect of earlier snowmelt.
-The positive effect of precipitation we found is consistent with findings from across the sagebrush distribution (CITATIONS), adding further /.//.
 
@@
 
-### A flexible, landscape-scale plant population modeling approach
+### A landscape-scale plant population modeling approach: opportunities and limitations
 
@@
 The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
+Our approach is amenable to any spatially-explicit time series of plant abundance, but we see remote-sensing datasets offering the largest opportunity for landscape-scale population models.
+Furthermore, it would be straighforward to include additional covariates related to disturbance (e.g., fire) or biotic interactions.
+Thus, we see our method as a first step toward coupling the mechanistic power of dynamic population models with the spatial extent of SDMs.
+The spatially- and temporally-explicit forecasts made possible by our approach should be especially relevant to land management decisions that need to made based on near-term forecasts.
 
@@
 Even with our dimension reduction technique modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
-In order to overcome this, future investigators modeling a larger spatial extent may need to aggregate their original remote-sensing time series data to a coarse spatial resoluation.
+To overcome this, future investigators modeling a larger spatial extent may need to aggregate their original remote-sensing time series data to a coarse spatial resoluation.
 This would allow them to model a much greater spatial extent with a similar number of knots and a similar amount of computation time.
@@
 We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
+We used the model to forecast future abundances of sagebrush in Wyoming and found that at our relatively cold site sagebrush should be expected to increase in cover.
 As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.

### 2015-12-29 docs/ms_sageAbundance.Rmd
add PBA edits

@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-We outlined a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
+We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
 Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
-The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which inspires confidence in our forecasts.
+The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which imspires some confidence in our forecasts.
 More generally, the approach we described is applicable to any spatially-explicit time series of plant population abundance and opens the door to forecasts of plant populations at the landscape-scale.
@@
 ### Sagebrush response to climate and climate change
-The climate effects estimated in our model, and detected at a spatial resolution of 30 meters, are consistent with individual-level responses of sagebrush to climate-related variables.
-At the level of individual plants, research shows that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
-Mirroring those individual-level responses, posterior means for all climate effects in our cover change model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig.
+The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
+Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
+In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig.
 3).
 The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
-However, the mean estimates for the climate effects are relatively small on the standardized scale (Fig. 3).
+However, the mean estimates for the climate effects are relatively weak (Fig. 3).
 Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
@@
 Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
-But, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration.
-More likely is that aggregated estimates of plant abundance, like percent cover, mask interannual variability at the level of the individual plant.
-Such averaging inevitably reduces interannual variability, which also makes detecting drivers of internanual variability more difficult.
+However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration (e.g. REFERENCE????).
+More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual
+variability.
 Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
 
-Under projected climate, we forecast small positive changes in sagebrush cover in this area for all RCP scenarios (Figs. 5,6).
-Our forecasts are a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
-First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
-Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes, while the projected change in temperature was large (Table 1, Fig. 2).
-Lastly, the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
-An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the increase in spatial heterogeity we forecast: the distribution of sagebrush cover expands rather than just shifting (Fig.
-D1).
-This effect is seen in the spatial forecasts where low cover pixels are constrained to low cover while high cover pixels increase (Fig. 5).
-Had we not explicitly accounted for spatial-dependence in our model, this result would be missed.
+Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios (Figs. 5,6).
+Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
+Precipitation two years before a given year-to-year transition was the strongest standardized effect (Fig.
+3), but precipitation is projected to increase only moderately (Table 1, Fig.
+2) and the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
+In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).
+An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. D1).
+Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig.
+5).
+Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.
 
-While we forecast an increase in sagebrush cover at our study area, SDM studies tend to project declines in sagebrush abundance in the future due to temperature increases [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
-Why do our results differ?
-There are many potential reasons, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in the cold range of the sagebrush distrubtion is the most likely explanation.
+While we forecast an increase in sagebrush cover at our study area, SDM studies typically project dramatic declines in climate suitability for sagebrush with warming [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
+There are many potential explanations for this apparent contrast, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in a cold portion of sagebrush's geographic distribution may be the best.
 The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
-The generally positive effect of precipitation we found is consistent with findings from across the sagebrush distribution [@Schlaepfer2012; @Schlaepfer2015].
-However, the effect of temperature may vary depending on location.
 In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
-Since our study area lies at the cold end of the sagebrush distribution, the weak positive response to temperature detected by our statistical model (Fig.
+Since our study area lies at the cold extreme of the sagebrush range, the weak positive response to temperature that we estimated (Fig.
 3) and carried through to our forecasts (Figs.
@@
 
-Also in contrast to our results, a previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
-The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
-First, Homer et al. (2015) used downscaled IPCC projections that allowed for spatial variation in future climate, unlike our analysis here.
-Therefore, Homer et al. (2015) did predict some pixel-level increases in sagebrush cover since each pixel was subject to its own regression between climate and cover.
-Second, like our study, Homer et al.
-(2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
-Thus, the decline in precipitation led to a decline in sagebrush cover.
-Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig.
-6).
-Lastly, Homer et al.'s (2015) analysis relied on pixel-level regressions of cover through time against climate covariates.
-Our current approach is fundamentally different in that we specifically model the impact of weather on year-to-year changes in sagebrush cover.
-Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.
+A previous analysis of a different subset of the remote sensing data set we used also came to a different conclusion, projecting future sagebrush decline [@Homer2015].
+The discrepancy between Homer et al.'s (2015) results and ours primarily reflects a difference in the climate projections used for projecting future changes rather than differences in our inference about responses to historical variation in weather.
+Consistent with our study, Homer et al.
+(2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the future climate scenario they chose resulted in a mean decrease in precipitation, causing a predicted decline in sagebrush cover.
+A second difference is that Homer et al.'s (2015) analysis relied on regressions of decadal trends in sagebrush cover against decadal trends in climate at the level of individual pixels.
+Our current approach is fundamentally different in that we specifically model the impact of interannual variation in weather on year-to-year changes in sagebrush cover.
+Thus, our model takes advantage of the additional information contained within short-term responses to climate fluctuations.
 
-We presented projections for sagebrush cover at the end of this century, but an important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
-For example, we could forecast the effects of a multi-year regional drought on this species [@Debinski2010].
-Validating spatial population models against short-term predictions would give ecological forecasters a way to assess the long-term forecasts produced by the model.
-It would also provide opportunities for refinement of the model itself.
-This cycle of prediction, validation, and refinement is missing from most population-level forecasts of the effects of climate change currently produced.
+We projected sagebrush cover to the end of this century, but an important feature of our approach is that it could also produce short-term forecasts.
+For example, we could forecast the effects of a multi-year regional drought on sagebrush cover [@Debinski2010].
+Validating spatial population models against short-term predictions would give ecological forecasters a way to assess and improve the performance of their models, which would greatly increase our confidence in long-term forecasts.
+This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change .
 
@@
 Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
-We focused the native spatial resolution of the remote sensing data ($30\times30$ meters).
-This constrained the extent that we could reasonably model because of the computational challenges in using a massive spatial random effect. 
-Even with our dimension reduction technique modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
-To overcome this, future investigators modeling a larger spatial extent may need to aggregate their original remote-sensing time series data to a coarse spatial resoluation.
-This would allow them to model a much greater spatial extent with a similar number of knots and a similar amount of computation time.
-A coarser scale model would lose some fine-scale detail but could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
+We retained the native spatial resolution of the remote sensing data ($30\times30$ meters).
+This constrained the extent that we could reasonably model because of the computational challenges in estimating spatial random effects. 
+Even with our dimension reduction technique, modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
+To model a larger spatial extent, we could aggregate the original remote-sensing time series data to a coarser spatial resoluation.
+This would allow us to model a much greater spatial extent with a similar number of knots and a similar computation time.
+While a coarser scale model would lose some fine-scale detail, it could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
 However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger and larger pixels.
 
-Our spatial extent and resolution also impacted our incorporation of climate covariates.
-We did not downscale Daymet data to match the spatial resolution of the sagebrush data.
-Thus, in each year all pixels share the same climate covariates.
+Our spatial extent and resolution also affected our use of climate covariates.
+We did not downscale Daymet data to match the spatial resolution of the sagebrush data, meaning that in each year all pixels share the same climate covariates.
 This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).
-We also did not allow for the possibility that different parts of our study area would respond to climate in different ways.
-This would require spatially-varying climate effects that substantially increase computational time.
-However, in some case, for example over very large regions, allowing the effect of climate to vary over space will be important.
+We also did not allow different portions of our study area to respond to climate in different ways.
+Doing so would require spatially-varying climate effects and a substantial increase in computational time.
+However, in future applications it will be important to allow climate effects to vary over space.
 @Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
-Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit and spatially-varying climate covariates that match the resolution of the percent cover data.
-It is also possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play more important roles than we observed in our study area.
+We might expect climate effects to interact with spatial covariates such as soil type, slope, and aspect.
+In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents.
 
-As mentioned above, one of the main differences between our current work and that of Homer et al. (2015) is our choice of climate projections.
+The contrast between our conclusions and those of Homer et al. (2015), mentioned above, emphasizes the irole of climate projections in ecological forecasting.
 Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].

### 2015-12-30 docs/ms_sageAbundance.Rmd
small updates

--- 2015-12-14 docs/ms_sageAbundance.Rmd
@@
 --------
-Plant populations models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
+Plant population models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
 Here we overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
@@
 
-The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents aswell as statistical methods for fitting high-dimensional spatial models.
+The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents and statistical methods for fitting high-dimensional spatial models.
 Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
 First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.
-These data can be fit with dynamic population models which include yearly fluctuation in climate as covariates.
+These data can be fit with dynamic population models which include yearly fluctuations in climate as covariates.
 Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
@@
 Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015].
-Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached.
-Furthermore, direct validation of such predictions is extremely rare (cite a paleo paper?).
+Furthermore, because they lack short-term dynamics, SDMs usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached.
+Direct validation of such predictions is extremely rare [@Roberts2012].
 Large-scale dynamic population models could overcome these limitations.
@@
 For each GCM and RCP scenario combination, we calculated average precipitation and temperature over the 1950-2000 time period and the 2050-2098 time period.
-We then calculated the absolute change in temperature between the two time periods ($\delta T$) and the proportional change in precipitation between the two time periods ($\delta P$) for each GCM and RCP scenario combination.
-Lastly, we applied $\delta T$ and $\delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
-These generated climate time series were used to simulate equilibrium sagebrush cover (see *Simulating the process* below).
-We simulated equilibrium cover separately for each GCM and RCP scenario combination, but we show the average projected climate changes across all models in Table 1.
+We then calculated the absolute change in temperature between the two time periods ($\Delta T$) and the proportional change in precipitation between the two time periods ($\Delta P$) for each GCM and RCP scenario combination.
+Lastly, we applied $\Delta T$ and $\Delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
+These generated climate time series were used to simulate equilibrium sagebrush cover.
+We simulated equilibrium cover separately for each GCM and RCP scenario combination before averaging the results, but we show the average projected climate changes across all models in Table 1.
 
@@
 
-For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to make simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011).
-We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
+For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
+We aggregated daily GCM output for each GCM and RCP scenario into the seasonal climate covariates used to fit our model.
 These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
@@
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
-Since our study area lies at the cold end of the sagebrush distribution, the weak positive response to temperature detected by our statistical model (Fig.
+The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from DAYMET estimates).
+Thus, our study area lies at the cold end of the sagebrush steppe distribution.
+Therefore, the weak positive response to temperature detected by our statistical model (Fig.
 3) and carried through to our forecasts (Figs.
@@
 As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
-Future modeling could include the effects of non-climate drivers -- including the effects of species interactions and disturbance [@Clark2014].
+Future modeling could include the effects of non-climate drivers -- including the effects of species interactions and disturbance.
 For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2006].

### 2015-12-30 docs/ms_sageAbundance.Rmd

--- 2015-12-29 docs/ms_sageAbundance.Rmd
@@
 Validating spatial population models against short-term predictions would give ecological forecasters a way to assess and improve the performance of their models, which would greatly increase our confidence in long-term forecasts.
-This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change .
+This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change.
 

### 2015-12-30 docs/ms_sageAbundance.Rmd
adds temperature of study area comparison to discussion

@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
-Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
-The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which imspires some confidence in our forecasts.
-More generally, the approach we described is applicable to any spatially-explicit time series of plant population abundance and opens the door to forecasts of plant populations at the landscape-scale.
+Here, we developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
+We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.
 
@@
 However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration (e.g. REFERENCE????).
-More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual
-variability.
+More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability.
 Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
@@
 An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. D1).
-Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig.
-5).
+Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
 Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.
@@
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
+The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from DAYMET estimates).
 Since our study area lies at the cold extreme of the sagebrush range, the weak positive response to temperature that we estimated (Fig.

### 2015-12-30 docs/appendixC.Rmd
updates app C with correct figures

@@
 \renewcommand\thefigure{C\arabic{figure}}  
+\renewcommand\thetable{C\arabic{table}}
 
@@
 Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure C1).
-Visual inspection of traceplots indicates convergence (Figures C2-C10).
+Visual inspection of traceplots indicates convergence (Figures C2-C16).
 
-Table:  Definition of parameters and model notation for Figures C1-C10.
+\footnotesize
+
+Table:  Definition of parameters and model notation for Figures C1-C16.
 
@@
 
+\normalsize
+
 ![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../results/rhat_plot.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_1.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_01.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_2.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_02.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_3.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_03.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_4.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_04.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_5.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_05.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_6.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_06.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_7.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_07.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_8.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_08.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_9.png)
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_09.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_10.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_11.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_12.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_13.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_14.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_15.png)
 

### 2015-12-30 docs/ms_sageAbundance.Rmd
updates app C with correct figures

@@
    - \doublespacing
+   - \usepackage[document]{ragged2e}
 bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
@@
 Model fitting required five days.
-We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all paramters) [@Gelman1992; @Gelman2009].
+We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].
 
@@
 Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
-However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration (e.g. REFERENCE????).
+However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
 More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability.
@@
 ### A landscape-scale plant population modeling approach: opportunities and limitations
-
 Our approach for modeling plant populations overcomes two major hurdles for spatially-explicit population models.
@@
 ```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
-Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=F}
+Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=T}
 library(png)
@@
 ```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
-1).", eval=F}
+1).", eval=T}
 library(png)
@@
 Only cumulative precipitation at time \\emph{t} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=F}
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
 library(png)
@@
 Prediction results are from simulations that use mean parameter values and no random year effects.
-Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=F}
+Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=T}
 library(png)
@@
 We used mean parameter estimates for all simulations.
-Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=F}
+Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=T}
 library(png)

### 2015-12-30 docs/appendixE.Rmd
adds app E with forecast densities

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{E\arabic{figure}}

Appendix E: Density plots of sagebrush cover forecasts ($\boldsymbol{\eta}$)
==============================================================================

```{r figure_E1,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time. These densities represent the desnity of cover values from the spatial plots in Figure 5 of the main text.", eval=TRUE, echo=FALSE}
library(png)
library(grid)
img <- readPNG("../results/clim_change_densities.png")
grid.raster(img)
```

### 2015-12-30 docs/ms_sageAbundance.Rmd
adds app E with forecast densities
fixes small typos

@@
 We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of this species.
-We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale.
+We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale.
 Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
@@
 We downloaded daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
-The Daymet weather data are interpolated between coarse observation units and captures some spatial variation.
+The Daymet weather data are interpolated between coarse observation units and capture some spatial variation.
 We relied on weather data for the centroid of our study area.
@@
 Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$), creating a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
-Fitting the model with a spatial random effect ($\eta$) is computationally for large data sets like ours.
+Fitting the model with a spatial random effect ($\eta$) is computationally demanding for large data sets like ours.
 This is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
@@
 It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
-However, given the relative size of our dataset and computational limitations, we defined kernels around 221 knots (Fig.
+However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig.
 B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
@@
 Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
-In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig.
+In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation two years before a cover transition (*ppt1*, Fig.
 3).
@@
 In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).
-An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. D1).
+An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. E1).
 Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
@@
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
-The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from DAYMET estimates).
+The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from Daymet estimates).
 Since our study area lies at the cold extreme of the sagebrush range, the weak positive response to temperature that we estimated (Fig.
@@
 
-The contrast between our conclusions and those of Homer et al. (2015), mentioned above, emphasizes the irole of climate projections in ecological forecasting.
+The contrast between our conclusions and those of Homer et al. (2015), mentioned above, emphasizes the role of climate projections in ecological forecasting.
 Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
@@
 7) using standardized covariate values.
-Only cumulative precipitation at time \\emph{t} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t-1}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
+Only cumulative precipitation at time \\emph{t-2} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t-2}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
 library(png)
@@
 
-```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and standard deviation (C, D) for the extent of our spatial area at 30-m resolution.
+```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and precision (C, D) for the extent of our spatial area at 30-m resolution.
 Observed equilibrium sagebrush cover (A) is the temporal mean of each pixel from the 28 year time series.

### 2015-12-31 docs/ms_sageAbundance.Rmd
updates descriptions of climate covariates

@@
 We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
-The five climate variables in our population model are: (1) "water year" precipitation (*description*) for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
+The five climate variables in our population model are: (1) cumulative, "water year" precipitation for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
 We selected these variables a priori based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.
@@
 In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
-Other climate effects strongly overlapped zero but their posterior means were positive, except for fall-through-spring precipitation two years before a cover transition, whose posterior mean was negative (Fig.
+Other climate effects strongly overlapped zero but their posterior means were positive, except for fall-through-spring precipitation the first year of a cover transition (*t*-1), whose posterior mean was negative (Fig.
 3).
@@
 Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
-In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation two years before a cover transition (*ppt1*, Fig.
+In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*, Fig.
 3).
-The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
+The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
 However, the mean estimates for the climate effects are relatively weak (Fig. 3).
@@
 Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
-Precipitation two years before a given year-to-year transition was the strongest standardized effect (Fig.
+Cumulative precipitation the year before before a given year-to-year transition was the strongest standardized effect (Fig.
 3), but precipitation is projected to increase only moderately (Table 1, Fig.
-2) and the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
+2) and the negative effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*) had an offsetting effect.
 In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).

### 2015-12-31 docs/ms_sageAbundance.Rmd
fixes climate covariate definitions

@@
 The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5).
+Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5 and Fig. E1).
 For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
@@
 Only cumulative precipitation at time \\emph{t-2} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t-2}; TmeanSpr1 = year \\emph{t-2} average spring temperature; ppt2 = year \\emph{t-1} fall through summer precipitation; TmeanSpr2 = year \\emph{t-1} average spring temperature; ppt1 = year \\emph{t-2} fall through summer precipitation.", fig.height=4, eval=T}
+Climate covariate codes: pptLag = water year precipitation in year \\emph{t-2}; TmeanSpr1 = year \\emph{t-1} average spring temperature; ppt2 = year \\emph{t} fall through summer precipitation; TmeanSpr2 = year \\emph{t} average spring temperature; ppt1 = year \\emph{t-1} fall through summer precipitation.", fig.height=4, eval=T}
 library(png)

### 2016-01-14 docs/appendixB.Rmd
adds year effect to sims; updates docs with Mevin's comments

@@
 
-![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernal bandwidth in Eq.
+![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernel bandwidth in Eq.
 6 of the main text (500/3; dashed black line).](../results/avgresidual_variogram_withrange_subset.png)

### 2016-01-14 docs/ms_sageAbundance.Rmd
adds year effect to sims; updates docs with Mevin's comments

@@
 Plant population models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
-Here we overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
-Using computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
-We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of this species.
+We overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
+Using computationally efficient model reparameterizations, we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming while formally accounting for spatial autocorrelation.
+We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of sagebrush.
 We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale.
 Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
-Our approach is an important first step to using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
+Our approach is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
 
@@
 Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
-Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
+Second, animal population modelers have developed dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
 These new statistical methods have yet to be applied to RS-derived plant population data at broad scales.
@@
 Large-scale, spatially-explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes playing out at landscape and regional scales.
-Currently most assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
+Most current assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
 SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
 However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
-Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015].
+Because SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015].
 Furthermore, because they lack short-term dynamics, SDMs usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached.
@@
 Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014].
-In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014].
+In warmer parts of its range, increased temperature could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014].
 In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
 Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
-A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.
+A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assessing how sagebrush could respond to climate change in the future.
 
@@
 #### Remotely-sensed time series
-To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
+To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
@@
 #### Climate covariates
-Our approach is based on modeling how interannual changes in plant cover relate to seasonal climate variables.
-We downloaded daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
+Our approach models interannual changes in plant cover as a function of seasonal climate variables.
+We used daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
 The Daymet weather data are interpolated between coarse observation units and capture some spatial variation.
@@
 The five climate variables in our population model are: (1) cumulative, "water year" precipitation for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
-We selected these variables a priori based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.
-We could have applied a variety of model selection techniques [@Hooten2015], but chose not to because our focus is on the general approach.
+We selected these variables *a priori* based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.
 
@@
 Fitting the model with a spatial random effect ($\eta$) is computationally demanding for large data sets like ours.
-This is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
+The computational demand is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
 Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation by reducing the size of the spatial covariance matrices that need to be inverted at each MCMC iteration.
@@
 Our dimension reduction strategy expresses the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
-We can then define the spatial effect as
+We can then approximate the spatial effect as
 
@@
 \begin{align}
-\boldsymbol{\eta} &= \textbf{K}\boldsymbol{\alpha}, \\
+\boldsymbol{\eta} &\approx \textbf{K}\boldsymbol{\alpha}, \\
 \alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2).
@@
 
-The last remaining obstacle is to parameterize \textbf{K}.
+The last remaining obstacle is to parameterize the matrix of basis functions, \textbf{K}.
 We use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
@@
 and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
-It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
+It is possible, through exhaustive model selection and fitting, to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
 However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig.
 B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
-An infinite number of knots would result in a perfectly accurate representation of the spatial process and covariance model.
-Since using an infinite number of knots is not possible, any use of the dimension reduction technique is an approximation, where the accuracy increases with the number of knots.
+An infinite number of knots would result in an exact representation of the spatial process and covariance model.
+Computationally, using an infinite number of knots is not possible, thus the use of dimension reduction techniques serves as an approximation, where the accuracy increases with the number of knots.
 Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.
@@
 ### Accomodating zeros
-Our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$)). 
+Our process model (in Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$)). 
 Thus, our model does not accomodate zeros. 
-Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47, which is 0.01% of the data set). 
+Fortunately, we had very few instances where pixels had 0% cover at time *t-1* (*N* = 47, which is 0.01% of the data set). 
 Thus, we excluded those pixels from the model fitting process. 
@@
 \begin{align}
-\text{logit}(\mu_{i}) &= \alpha \\
-y_{i} &\sim \text{Bernoulli}(\mu_{i})
+y_{i} &\sim \text{Bernoulli}(\mu_{i}) \\
+\text{logit}(\mu_{i}) &= b_0
 \end{align}
@@
 
-where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($\alpha$).
+where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($b_0$).
 We fit this simple model using the `glm' command in R [@R2013].
@@
 ### Fitting the model
-We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
+We fit the spatiotemporal model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
 We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
-Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
-However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
-Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
+Short chains of samples are a hallmark of the Stan algorithm, which is extremely efficient.
+Compared to other samplers, fewer iterations are required to achieve convergence.
+Each chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
 Model fitting required five days.
@@
 To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
-For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
+For simulation (2), we initialized each pixel with its actual percent cover value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
 For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
 We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
-We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
-The random year effects were not used in any simulation.
+We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the chains.
+The random year effects were not used in simulations 1-3 where.
 
 We required future projections of climate for our study area to conduct the equilibrium and temporally-explicit forecasts described above.
-We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
+Thus, we used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
 The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
@@
 We aggregated daily GCM output for each GCM and RCP scenario into the seasonal climate covariates used to fit our model.
-These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
+These yearly climate time series were not aggregated further because we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
 
@@
 
-When we simulated the pixel-based population model based on observed climate it was able to reproduce the spatial pattern of observed percent cover, averaged over time (Fig. 4A,B).
+When we simulated the pixel-based population model based on observed climate, it was able to reproduce the spatial pattern of observed percent cover, averaged over time (Fig. 4A,B).
 Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-Here, we developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
+We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
 We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.
@@
 The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
-However, the mean estimates for the climate effects are relatively weak (Fig. 3).
+However, mean estimates for the climate effects are relatively weak (Fig. 3).
 Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
 Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
-Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
+We used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
 However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
 More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability.
-Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
+Nonetheless, our model was capabable of detecting climate effects that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
 
@@
 In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).
-An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. E1).
+An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. E1).
 Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
@@
 There are many potential explanations for this apparent contrast, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in a cold portion of sagebrush's geographic distribution may be the best.
-The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
+The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt because their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
 In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
@@
 The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from Daymet estimates).
-Since our study area lies at the cold extreme of the sagebrush range, the weak positive response to temperature that we estimated (Fig.
+Our study area lies at the cold extreme of the sagebrush range, thus the weak positive response to temperature that we estimated (Fig.
 3) and carried through to our forecasts (Figs.
@@
 A previous analysis of a different subset of the remote sensing data set we used also came to a different conclusion, projecting future sagebrush decline [@Homer2015].
-The discrepancy between Homer et al.'s (2015) results and ours primarily reflects a difference in the climate projections used for projecting future changes rather than differences in our inference about responses to historical variation in weather.
+The discrepancy between the results of Homer et al.
+(2015) and ours primarily reflects a difference in the climate projections used for projecting future changes rather than differences in our inference about responses to historical variation in weather.
 Consistent with our study, Homer et al.
 (2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the future climate scenario they chose resulted in a mean decrease in precipitation, causing a predicted decline in sagebrush cover.
-A second difference is that Homer et al.'s (2015) analysis relied on regressions of decadal trends in sagebrush cover against decadal trends in climate at the level of individual pixels.
+A second difference is that Homer et al. (2015) relied on regressions of decadal trends in sagebrush cover against decadal trends in climate at the level of individual pixels.
 Our current approach is fundamentally different in that we specifically model the impact of interannual variation in weather on year-to-year changes in sagebrush cover.
@@
 Accounting for spatial autocorrelation allows for statistically rigorous inference on the effects of interannual climate on sagebrush cover change in our study region.
-The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e. within the study region; Figs. 4,5).
+The spatial covariance structure also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (i.e., within the study region; Figs. 4,5).
 Our approach is amenable to any spatially-explicit time series of plant abundance, but we see remote-sensing datasets offering the largest opportunity for landscape-scale population models.
@@
 Thus, we see our method as a first step toward coupling the mechanistic power of dynamic population models with the spatial extent of SDMs.
-The spatially- and temporally-explicit forecasts made possible by our approach should be especially relevant to land management decisions that need to made based on near-term forecasts.
+The spatially- and temporally-explicit forecasts made possible by our approach should be especially relevant to land management decisions based on near-term forecasts.
 
@@
 While a coarser scale model would lose some fine-scale detail, it could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
-However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger and larger pixels.
+However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger pixels.
 
@@
 Doing so would require spatially-varying climate effects and a substantial increase in computational time.
-However, in future applications it will be important to allow climate effects to vary over space.
+However, in future applications, it will be important to allow climate effects to vary over space.
 @Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
@@
 
-The contrast between our conclusions and those of Homer et al. (2015), mentioned above, emphasizes the role of climate projections in ecological forecasting.
+The contrast between our conclusions and those of Homer et al. (2015) emphasizes the role of climate projections in ecological forecasting.
 Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
@@
 ----------
-We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
+We introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
 We used the model to forecast future abundances of sagebrush in Wyoming and found that at our relatively cold site sagebrush should be expected to increase in cover.
@@
 We are grateful to Debra K.
-Meyer at USGS EROS for extracting the data set used in this paper.
-Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
+Meyer at USGS EROS for extracting the data set used in this paper and to David Koons for comments that improved the manuscript.
+Compute, storage, and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
 We acknowledge the World Climate Research Programme's Working Group on Coupled Modelling, which is responsible for CMIP, and we thank the climate modeling groups (listed in Table A1) for producing and making available their model output.

### 2016-01-14 scripts/sim_sage_byYear_byModel.R

@@ -127,6 +127,7 @@ for(do_model in all_models){
       params_now <- subset(mcmc_outs, chain==randchain & mcmc_iter==randiter)
       alphas <- params_now[grep("alpha", params_now$Parameter), "value"]
       int_mu <- params_now[grep("int_mu", params_now$Parameter), "value"]
+      sd_int <- params_now[grep("sig_yr", params_now$Parameter), "value"]
       beta_mu <- params_now[grep("beta_mu", params_now$Parameter), "value"]
       beta_clims <- params_now[grep("beta.", params_now$Parameter), "value"][2:6]
       eta_now <-  K%*%as.numeric(unlist(alphas))
@@ -134,9 +135,9 @@ for(do_model in all_models){
         for(t in 2:num_sims){
           
           weather <- climate_now[t-1, clim_vars] #t-1 since clim data starts in 2012
-        
+          int_now <- rnorm(1, int_mu, sd_int)
           n_save[i,t,] <- iterate_sage(N = n_save[i,t-1,], 
-                                       int = int_mu, 
+                                       int = int_now, 
                                        beta.dens = beta_mu,
                                        beta.clim = beta_clims,
                                        eta = eta_now,

### 2016-01-14 docs/ms_sageAbundance.Rmd
updates with Cam's comments

@@
 These data can be fit with dynamic population models which include yearly fluctuations in climate as covariates.
-Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
+Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long-term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
 Second, animal population modelers have developed dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
@@
 Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data.
-Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; Xian2012; @Harte2015; Homer2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
 SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
@@
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
-The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46 [@Homer2012].
+The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
 To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
@@
 Each chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
-Model fitting required five days.
-We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].
+Model fitting required five days on a four node Central Processing Unit with 2 $\times$ AMD Opteron(tm) Processor 4386 @ 3.10 Ghz, 64GB of RAM per node, 16 cores per node, and each chain launched in parallel on separate cores.
+We assessed convergence visually and calculated scale-reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].
 
@@
 When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
-The correlation between observations and predictions was 0.62.
+The Pearson's correlation between observations and predictions was 0.62.
 
@@
 Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
-The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).
+The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change estimated for our study area (Fig. 3).
 
@@
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
-The average temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from Daymet estimates).
+The average annual temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from Daymet estimates).
 Our study area lies at the cold extreme of the sagebrush range, thus the weak positive response to temperature that we estimated (Fig.
 3) and carried through to our forecasts (Figs.
-5,6) likely represents the positive effect of earlier snowmelt.
+5,6) likely represents the positive effect of earlier snowmelt, and thus higher moisture availability early in the growing season.
 
@@
 A second difference is that Homer et al. (2015) relied on regressions of decadal trends in sagebrush cover against decadal trends in climate at the level of individual pixels.
-Our current approach is fundamentally different in that we specifically model the impact of interannual variation in weather on year-to-year changes in sagebrush cover.
+Our current approach is fundamentally different in that we specifically model the impact of interannual variation in weather on year-to-year changes in sagebrush cover using a dynamic population model.
 Thus, our model takes advantage of the additional information contained within short-term responses to climate fluctuations.
@@
 Doing so would require spatially-varying climate effects and a substantial increase in computational time.
-However, in future applications, it will be important to allow climate effects to vary over space.
+However, in future applications, it will be important to allow climate effects to vary over space to better capture reality.
 @Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
 We might expect climate effects to interact with spatial covariates such as soil type, slope, and aspect. 
-In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents.
+In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents where variability may be greater.
 
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 2.5 $\\times$ 5 kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
+```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 5,070 $\\times$ 2,430 meter kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
 Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=T}
@@
 We used mean parameter estimates for all simulations.
-Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.", fig.height=6, eval=T}
+Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.
+Axes definitions: Lat = latitude; Lon = longitude.", fig.height=6, eval=T}
 library(png)

### 2016-01-14 docs/appendixB.Rmd
update appB

@@
 B1).
-Thus, we set the kernal bandwidth ($\sigma$ in Eq.
-X of the main text) to 500/3 = `r round(500/3,2)`.
+Thus, we set the kernel bandwidth ($\sigma$ in Eq.
+6 of the main text) to 500/3 = `r round(500/3,2)` because the effective spatial range under an exponential covariance structure is equal to $\sigma \times 3$.
 We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters.

### 2016-01-14 docs/ms_sageAbundance.Rmd
adds mention of shared climate variables earlier
update appB

@@
 Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data.
-Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; Xian2012; @Harte2015; Homer2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
+Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Xian2012; @Harte2015; @Homer2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
 SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
@@
 More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability.
+Additionally, we chose not to downscale the Daymet weather data, meaning that in a given year all pixels shared the same climate, which limits our statistical power.
 Nonetheless, our model was capabable of detecting climate effects that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.

### 2016-01-15 scripts/sim_sage_byYear_byModel.R
runs forecasts for every MCMC iteration

@@ -94,6 +94,10 @@ num_sims <- length(sim_years)
 
 nchains <- length(unique(mcmc_outs$chain))
 niters <- length(unique(mcmc_outs$mcmc_iter))
+dogrid <- expand.grid(1:niters, 1:nchains)
+
+
+tmp.time=Sys.time()
 
 ####
 ####  Begin looping: parameters within years within scenario within model ------
@@ -121,9 +125,11 @@ for(do_model in all_models){
     n_save[,1,] <- last_obs$Cover # set first record to last observation
     
       
-    for(i in 1:parameter_reps){
-      randchain <- sample(1:nchains, 1)
-      randiter <- sample(1:niters, 1)
+    for(i in 1:nrow(dogrid)){
+#       randchain <- sample(1:nchains, 1)
+#       randiter <- sample(1:niters, 1)
+      randchain <- dogrid[i,2]
+      randiter <- dogrid[i,1]
       params_now <- subset(mcmc_outs, chain==randchain & mcmc_iter==randiter)
       alphas <- params_now[grep("alpha", params_now$Parameter), "value"]
       int_mu <- params_now[grep("int_mu", params_now$Parameter), "value"]
@@ -159,5 +165,7 @@ for(do_model in all_models){
 } # next model
 
 
-
+time.2=Sys.time()-tmp.time
+sfStop()
+time.2
 

### 2016-01-15 scripts/sim_sage_byYear_byModel.R
fixes forecast holder

@@ -121,7 +121,8 @@ for(do_model in all_models){
     climate_now["TmeanSpr2"] <- (climate_now["TmeanSpr2"] - obs_clim_means["TmeanSpr2"])/obs_clim_sds["TmeanSpr2"]
     
     # Create storage matrix for population
-    n_save <- array(dim = c(parameter_reps, num_sims+1, nrow(last_obs)))
+    # n_save <- array(dim = c(parameter_reps, num_sims+1, nrow(last_obs)))
+    n_save <- array(dim = c(nrow(dogrid), num_sims+1, nrow(last_obs)))
     n_save[,1,] <- last_obs$Cover # set first record to last observation
     
       

### 2016-01-18 devel/approx_posterior_test.R
adds posterior mean approx test

@@ -1,0 +1,16 @@
+n.sim=100000
+mu=2.8
+sig=0.3
+
+lam=exp(rnorm(n.sim,mu,sig))
+y.1=rpois(n.sim,lam)
+y.2=rpois(n.sim,exp(mu))
+
+mean(y.1)
+mean(y.2)
+
+library(plotrix)
+multhist(list(y.1,y.2),breaks=200,col=c(1,2),
+         xlab="percent cover", ylab="frequency", border=c(1,2))
+text(210, 7000,"approx. posterior", col=2)
+text(230, 2000,"true posterior", col=1)

### 2016-01-18 scripts/ms_scripts.R
updates equil sims with random year effects

@@ -203,6 +203,8 @@ time.steps <- 2000
 burn.in <- 100
 mean_params <- ddply(outs, .(Parameter), summarise,
                      value = mean(value))
+yrint_ids <- grep("int_yr", mean_params$Parameter)
+int_yrs <- mean_params[yrint_ids,"value"]
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"value"]
 betas <- mean_params[grep("beta", mean_params$Parameter),"value"][2:6]
 eta <- K%*%alphas
@@ -213,9 +215,10 @@ clim_sim <- climD[climD$year %in% unique(growD$Year),]
 X_sim = clim_sim[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
 X_sim = scale(X_sim, center = TRUE, scale = TRUE)
 for(t in 2:time.steps){
+  int_now <- int_yrs[sample(c(1:length(int_yrs)), 1)]
   Xtmp <- X_sim[sample(c(1:nrow(X_sim)), 1),]
   dens.dep <- mean_params[mean_params$Parameter=="beta_mu","value"]*log(ex.mat[t-1,])
-  tmp.mu <- mean_params[mean_params$Parameter=="int_mu","value"] + dens.dep + sum(betas*Xtmp)
+  tmp.mu <- int_now + dens.dep + sum(betas*Xtmp)
   tmp.mu <- exp(tmp.mu + eta)
   tmp.out <- rpois(ncol(ex.mat), lambda = tmp.mu)
   
@@ -303,6 +306,8 @@ dev.off()
 ####
 mean_params <- ddply(outs, .(Parameter), summarise,
                      mean(value))
+yrint_ids <- grep("int_yr", mean_params$Parameter)
+int_yrs <- mean_params[yrint_ids,"value"]
 alphas <- mean_params[grep("alpha", mean_params$Parameter),"..1"]
 betas <- mean_params[grep("beta", mean_params$Parameter),"..1"][2:6]
 eta <- K%*%alphas
@@ -314,6 +319,7 @@ clim_sim <- climD[climD$year %in% unique(growD$Year),]
 X_sim = clim_sim[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")]
 X_sim = scale(X_sim, center = TRUE, scale = TRUE)
 for(t in 1:time.steps){
+  int_now <- int_yrs[t]
[…]

### 2016-01-18 scripts/plot_temporal_forecasts.R

@@ -23,6 +23,8 @@ library(reshape2)
 ####
 setwd("../results/yearlyforecasts/")
 all_files <- list.files()
+torms <- grep("MEAN", all_files)
+all_files <- all_files[-torms]
 
 longdf <- data.frame(model=NA,
                      scenario=NA,

### 2016-01-18 scripts/sim_equilibrium_forecasts.R

@@ -101,7 +101,7 @@ avg.new.cover <- round(mean(colD[which(colD$Cover>0),"Cover"]),0)
 ####
 sim_sage <- function(init_cover_vector, scaled_clim_matrix, 
                      avg_new_cover, params, time_steps){
-  int_mu <- params[["int_mu"]]
+  int_yrs <- params[["int_yrs"]]
   beta_mu <- params[["beta_mu"]]
   betas <- params[["betas"]]
   eta <- params[["eta"]]
@@ -111,10 +111,11 @@ sim_sage <- function(init_cover_vector, scaled_clim_matrix,
   pb <- txtProgressBar(min=2, max=time_steps, char="+", style=3, width=65)
   for(t in 2:time_steps){
     ## Growth Process
+    int_now <- int_yrs[sample(c(1:length(int_yrs)), 1)]
     Xtmp <- X[sample(c(1:nrow(X)), 1),]
     dens.dep <- beta_mu*log(cover_matrix[t-1,])
     dens.dep[which(dens.dep==-Inf)] <- 0
-    tmp.mu <- int_mu + dens.dep + sum(betas*Xtmp)
+    tmp.mu <- int_now + dens.dep + sum(betas*Xtmp)
     tmp.mu <- exp(tmp.mu + eta)
     tmp.out <- rpois(ncol(cover_matrix), lambda = tmp.mu)
     
@@ -140,8 +141,10 @@ alphas <- mean_params[grep("alpha", mean_params$Parameter),"value"]
 betas <- mean_params[grep("beta", mean_params$Parameter),"value"][2:6]
 eta <- K%*%alphas
 beta_mu <- mean_params[mean_params$Parameter=="beta_mu","value"]
-int_mu <- mean_params[mean_params$Parameter=="int_mu","value"]
-params <- list(int_mu=int_mu, beta_mu=beta_mu, betas=betas, eta=eta)
+# int_mu <- mean_params[mean_params$Parameter=="int_mu","value"]
+yrint_ids <- grep("int_yr", mean_params$Parameter)
+int_yrs <- mean_params[yrint_ids,"value"]
+params <- list(int_yrs=int_yrs, beta_mu=beta_mu, betas=betas, eta=eta)
 
 
 

### 2016-01-18 docs/ms_sageAbundance.Rmd
updates ms with Collin comments and random year effects; sourcing script working; submission-ready

@@
     fig_caption: true
-    keep_tex: false
+    keep_tex: true
 fontsize: 12pt
@@
 Large-scale dynamic population models could overcome these limitations.
-They would produce spatially-explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
+They would produce spatially-explicit estimates of species abundance within the species range [@Ehrlen2015], have the potential to model expansion in abundance outside the range when coupled with dynamic models of dispersal, and would provide testable predictions of how populations should respond to short-term climate perturbations.
 These short-term predictions also would give modelers the opportunity to repeatedly validate and refine their models [@Luo2011a].
@@
 As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
-The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
+The regression tree model was further validated using another 297 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
 To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
@@
 We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects similar to that described by @Conn2015.
+Interannual change in percent cover represents the integrated outcome of recruitment, survival, growth, and retrogression (shrinkage) of individual plants from year to year.
 We model observed integer percent cover (*y*) in cell *i* at time *t* as conditionally Poisson
@@
 [\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
-&\left(\prod_{j=1}^{J}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
+&\left(\prod_{m=1}^{M}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
 \end{align}
@@
 We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
+
 We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the chains.
-The random year effects were not used in simulations 1-3 where.
+Random year effects were included in simulations by randomly drawing a posterior mean year effect ($\beta_{0,t}$) for each iteration (simulations 1 and 3), using the posterior mean year effect for a specific year (simulation 2), or by a drawing a future-year random effect from the posterior mean and standard deviation of the mean intercept (simulation 4, e.g., $\beta_{0,T} \sim \text{normal}(\bar{\beta_{0}}, \sigma_{\beta_{0}}^2)$ for some future year *T*).
+Our simulation approach provides a reasonable and computationally efficient approximation to the true posterior predictive mean when used in these scenarios with our data.
 
@@
 Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
-Cumulative precipitation the year before before a given year-to-year transition was the strongest standardized effect (Fig.
+Cumulative precipitation the year before a given year-to-year transition was the strongest standardized effect (Fig.
 3), but precipitation is projected to increase only moderately (Table 1, Fig.
@@
 In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
-The average annual temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 is 4.6\textdegree C (calculated from Daymet estimates).
+The average annual temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 was 4.6\textdegree C (calculated from Daymet estimates).
 Our study area lies at the cold extreme of the sagebrush range, thus the weak positive response to temperature that we estimated (Fig.
@@
 (2015) and ours primarily reflects a difference in the climate projections used for projecting future changes rather than differences in our inference about responses to historical variation in weather.
+Homer et al. (2015) used downscaled weather projections from a single model from the IPCC 4 whereas we used native-resolution weather projections from a suite of models from the IPCC 5.
 Consistent with our study, Homer et al.
@@
 Thus, our model takes advantage of the additional information contained within short-term responses to climate fluctuations.
+Lastly, the location of Homer et al.'s (2015) study area is, on average, at a lower elevation than our current study area.
+The geographic difference results in different historical and projected climate, and, as discussed above, sagebrush may respond differently to warming depending on geographic location.
 
@@
 In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents where variability may be greater.
-
-The contrast between our conclusions and those of Homer et al. (2015) emphasizes the role of climate projections in ecological forecasting.
-Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
-These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013].
-There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
-Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
-If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
-Extensions of these ideas are the subject of ongoing research.
 
@@
 Observed equilibrium sagebrush cover (A) is the temporal mean of each pixel from the 28 year time series.
-Prediction results are from simulations that use mean parameter values and no random year effects.
+Prediction results are from simulations that use posteior mean parameter values.
+Precision in (D) represents the variability of each pixel over the course of the 2,000 iteration simulation.
 Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=T}
@@
 Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the statistical model.
-We used mean parameter estimates for all simulations.
+We used posterior mean parameter estimates for all simulations.
 Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.

### 2016-01-18 scripts/plot_projected_weather.R

@@ -82,26 +82,26 @@ dev.off()
 ####
 ####  Compare weather hindcasts to average weather at site ---------------------
 ####
-obs_weather <- read.csv("../data/FormattedClimate_WY_SA1.csv")
-obs_weather <- melt(obs_weather, id.vars = "year")
-obs_ppt <- subset(obs_weather, variable=="ppt2")
- 
-model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
-precip_compare <- data.frame(year = model_weather$Year, 
-                             precip = c(obs_ppt$value, model_weather$value),
-                             value_source = rep(c("observed", "modeled"), each=34))
-ggplot(precip_compare)+
-  geom_boxplot(aes(x=value_source, y=precip))
-
-
-obs_temp <- subset(obs_weather, variable=="TmeanSpr2")
-model_weather <- ddply(temp_agg, .(Year, season), summarise,
-                       value = mean(avg_value))
-model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
-temp_compare <- data.frame(year = model_weather$Year, 
-                             temp = c(obs_temp$value, model_weather$value),
-                             value_source = rep(c("observed", "modeled"), each=34))
-ggplot(temp_compare)+
-  geom_boxplot(aes(x=value_source, y=temp))
+# obs_weather <- read.csv("../data/FormattedClimate_WY_SA1.csv")
+# obs_weather <- melt(obs_weather, id.vars = "year")
+# obs_ppt <- subset(obs_weather, variable=="ppt2")
+#  
+# model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
+# precip_compare <- data.frame(year = model_weather$Year, 
+#                              precip = c(obs_ppt$value, model_weather$value),
+#                              value_source = rep(c("observed", "modeled"), each=34))
+# ggplot(precip_compare)+
+#   geom_boxplot(aes(x=value_source, y=precip))
+# 
+# 
+# obs_temp <- subset(obs_weather, variable=="TmeanSpr2")
+# model_weather <- ddply(temp_agg, .(Year, season), summarise,
+#                        value = mean(avg_value))
+# model_weather <- subset(model_weather, Year < 2014 & Year > 1979)
+# temp_compare <- data.frame(year = model_weather$Year, 
+#                              temp = c(obs_temp$value, model_weather$value),
[…]

### 2016-01-19 docs/ms_sageAbundance.Rmd
fixes subscripting for alpha in posterior expression
adds spatial structure v. landscape features to discussion

@@
 [\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
-&\left(\prod_{m=1}^{M}[\alpha_j|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
+&\left(\prod_{m=1}^{M}[\alpha_m|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
 \end{align}
@@
 In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).
+
 An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. E1).
@@
 Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.
+We were unable to attribute the spatial structure apparent in the data (Fig.
+4A) and approximated by our model ($\boldsymbol{\eta}$, Fig.
+D1) to slope, aspect, elevation, or coarse soil type (results not shown).
+The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects related to fire and species invasions.
 

### 2016-01-22 docs/ms_sageAbundance.Rmd
final typos fixes for submission

@@
 \boldsymbol{\eta} &\approx \textbf{K}\boldsymbol{\alpha}, \\
-\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2).
+\alpha_{m} &\sim \text{Normal}(0,\sigma_{\eta}^2).
 \end{align}
@@
 We then calculated the absolute change in temperature between the two time periods ($\Delta T$) and the proportional change in precipitation between the two time periods ($\Delta P$) for each GCM and RCP scenario combination.
-Lastly, we applied $\Delta T$ and $\Delta P$ the the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
+Lastly, we applied $\Delta T$ and $\Delta P$ to the observed 28-year climate time series to generate a future climate time series for each GCM and RCP scenario combination.
 These generated climate time series were used to simulate equilibrium sagebrush cover.
@@
 Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
-Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
+Given findings from previous research demonstrating the importance of precipitation and temperature to sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
 We used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.

### 2016-06-17 devel/randeffect.Rmd
makes revisions in response to reviews

---
title: "Options for including year random effect on intercept in simulations"
output: pdf_document
---

When we simulate equilibrium sagebrush abundance (Figs. 4-5), we use point estimates of model parameters.
Currently, we are only using the year random effects on the intercept for the temporally-explicit forecasts (Fig. 6).
But maybe we should be using them in the equilibrium simulations too; the question is how to incorporate them.
As a reminder, we fit the model with a temporal random effect on the intercept, such that

\begin{align}
\beta_{0,t} \sim \text{normal}(\bar{\beta_{0}}, \sigma_{\beta_{0}}^2)
\end{align}

So, when we use point estimates to simulate equilibrium abundance based on observed climate, do we use the mean of $\bar{\beta_{0}}$ for the intercept? Or, do we use the posterior mean of a randomly chosen $\beta_{0,t}$ at each time step for the intercept? Or, do we draw a random climate year *ct*, and then use the same year for the intercept, e.g., posterior mean of $\beta_{0,ct}$?

Currently, for Figs. 4-5, we are using the posterior mean of $\bar{\beta_{0}}$ as the intercept for all time steps of the simulation. My intuition tells me we want to use the posterior mean of a randomly chosen $\beta_{0,t}$ at each time step for the intercept. But, I want to get a consensus on this.

### 2016-06-17 docs/ms_sageAbundance.Rmd
makes revisions in response to reviews

--- 2016-01-19 docs/ms_sageAbundance.Rmd
@@
    - \usepackage[document]{ragged2e}
+   - \usepackage{color, soul}
 bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
@@
 Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
-Our approach is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
+Our approach \hl{, which brings together several common techniques,} is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
 
@@
 
-Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict population impacts of climate change.
-Our approach relies on a process model motivated by Gompertz density-dependent population growth and a remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015]. 
-We account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
-The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the greatest potential to combine the information of population models (e.g., population status) and the spatial extent of species distribution models.
+Building on recent technological advances in spatial statistics [@Latimer2009; @Conn2015] and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict population impacts of climate change.
+\hl{As a proof-of-concept, we use} a process model motivated by Gompertz density-dependent population growth and a remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015].
+We account for spatial autocorrelation with dimension reduction techniques [@Latimer2009; @Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
+The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the greatest potential to combine the information of population models (e.g., population status \hl{and temporal dynamics}) and the spatial extent of species distribution models.
 
@@
 To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
-As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
+As part of a separate study, \hl{Homer et al.
+2012 estimated} sagebrush percent cover using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
 The regression tree model was further validated using another 297 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
-To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
-This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
-In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
+To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \hl{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
+This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. A1).
+In this \hl{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
 In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
-Thus, the full dataset contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
+Thus, the full \hl{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
 
@@
 There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we handled the small number of zero percent cover observations.
+\hl{We assume that the remotely sensed estimates of percent cover are "true" and free of error.}
+\hl{This need not be the case, and if measurement error is known then it could be included in our Bayesian model as a "data model"} [@Hobbs2015].
 
@@
 We fit this simple model using the `glm' command in R [@R2013].
-For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied.
+For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied \hl{and covariates such as cover of neighboring cells could be included}.
 
@@
 These yearly climate time series were not aggregated further because we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
+\hl{Note that forecasting with our model, which relies on historical correlations between sagebrush cover change and weather, using future climate projections requires assuming such relationships remain in the future.}
 
@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
+We \hl{introduced} a new approach \hl{that combines a} remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling \hl{literature} [e.g., @Conn2015].
+\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming.}
 We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.
@@
 First, we used moderate resolution, remotely-sensed estimates of sagebrush percent cover as a response variable, enabling us to fit a dynamic population model over a large spatial extent. 
+\hl{Species-specific estimates of plant abundance are becoming commonplace as remote sensing technology develops} [e.g., @Baldeck2014; @Colgan2014], \hl{and in a few years several remotely-sensed time series may be available.}
 Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we fit the model using a dimension reduction strategy that accounted for spatial autocorrelation within a feasible computational time.
@@
 In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents where variability may be greater.
+
+\hl{The uncerainty associated with our forecasts highlights several opportunities to improve our approach.
+First, Parameter uncertainty could be reduced by regulating the variance of the posterior distributions of climate covariates via ridge regression} [e.g., @Gerber2015].
+\hl{Second, uncertainty associated with climate projections could be reduced by identifying GCMs that perform exceptionally well for a particular study location} [e.g., @Rupp2013].
+\hl{Such considerations will be important when forecasting in support of particular management objectives.}
 

### 2016-06-17 devel/poisson_test.R

@@ -1,0 +1,23 @@
+n <- 100
+mu <- 3
+sd.mu <- 0.5
+K <- 10000
+
+diff.yhat <- numeric(K)
+yhat1avg <- numeric(K)
+yhat2avg <- numeric(K)
+for(k in 1:K){
+  lambda <- exp(rnorm(1, mu, sd.mu))
+  y.hat1 <- rpois(n, lambda)
+  y.hat2 <- rpois(n, exp(mu))
+  diff.yhat[k] <- mean(y.hat1)-mean(y.hat2)
+  yhat1avg[k] <- mean(y.hat1)
+  yhat2avg[k] <- mean(y.hat2)
+}
+
+par(mfrow=c(1,2))
+hist(diff.yhat)
+plot(yhat1avg, yhat2avg, pch=".")
+abline(0,1)
+
+

### 2016-06-17 devel/stan_lppd_cv_test.R

@@ -1,0 +1,253 @@
+##  Script to simulate a linear model and test regularization
+##  via cross-validation using STAN
+
+library(rstan)
+library(plyr)
+library(ggmcmc)
+
+rm(list=ls())
+
+source("waic_fxns.R")
+
+####  Simulate some data ----
+nobs <- 200
+alpha <- 10
+beta1 <- 2
+beta2 <- -5
+#Simulate correlated predictor variables
+R = matrix(cbind(1,0.80,0.2,0.80,1,0.7,0.2,0.7,1),nrow=3)  #correlation structure for predictors- 3x3, thus 3 beta's
+U = t(chol(R))
+nvars = dim(U)[1]    #3 beta's
+set.seed(1432423)
+random.normal=rbind(rep(1,nobs),rnorm(nobs,2,1),rnorm(nobs,-5,1))
+X = U %*% random.normal
+newX = t(X)
+X=newX
+X[,2:nvars]=scale(X[,2:nvars],center=TRUE, scale=TRUE)  # scaled to mean 0 and variance 1
+# X #Design matrix with intercept and two scaled and centered predictor variables.
+
+y <- rnorm(nobs, X%*%c(alpha,beta1,beta2), 20)
+
+
+####  Write STAN model ----
+model_string <- "
+data{
+  int<lower=0> nobs; // number observations
+  vector[nobs] y; // observation vector
+  int<lower=0> ncovs; // number of covariates (+ intercept)
+  //vector[ncovs] beta_means;
+  real<lower=0> beta_tau;
+  matrix[nobs,ncovs] x; // predictor vector
+  int<lower=0> npreds; // # of points to predict
+  matrix[npreds,ncovs] x_holdout;  // data for which to make predictions
+  vector[npreds] y_holdout;  // observations for predictions
+}
+parameters{
+  real b1;
+  vector[ncovs] b;
+  real<lower=0.000001> tau_model;
+}
+transformed parameters{
+  vector[nobs] mu;
+  mu <- b1+x*b;
+}
+model{
+  //Likelihood
+  y ~ normal(mu, tau_model);
+
+  //Priors
+  b1 ~ normal(0, 1000);
+  b ~ normal(0, beta_tau);
+}
+generated quantities {
+  vector[npreds] y_hat; // predictions
[…]

### 2016-06-17 docs/ms_sageAbundance.Rmd
changes figure imports

@@
    - \usepackage{color, soul}
+   - \usepackage{graphicx}
 bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
@@
 -------
-```{r figure_1_alt,dependson="plot-options", fig.cap="Location of the 5,070 $\\times$ 2,430 meter kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
-Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.", fig.height=6, eval=T}
-library(png)
-library(grid)
-img <- readPNG("components/figure/studyarea_map.png")
-grid.raster(img)
-```
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{./components/figure/studyarea_map.png}
+  \caption{Location of the 5,070 $\times$ 2,430 meter kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
+Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.}
+\end{figure}
 
-```{r figure_1,dependson="plot-options", fig.cap="Percent cover in space and time for our study area in southwestern Wyoming.", eval=F}
-library(png)
-library(grid)
-img <- readPNG("../results/all_years_percCover.png")
-grid.raster(img)
-```
+\newpage{}
 
-```{r figure_2a,dependson="plot-options", fig.cap="GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
-1).", eval=T}
-library(png)
-library(grid)
-img <- readPNG("components/figure/weather_projections.png")
-grid.raster(img)
-```
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{./components/figure/weather_projections.png}
+  \caption{GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
+1).}
+\end{figure}
 
-```{r figure_2,dependson="plot-options", fig.cap="Posterior distributions of climate covariates.
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{../results/post_climate_covariates.png}
+  \caption{Posterior distributions of climate covariates.
 The x-axis is the standardized coefficient value because we fit the statistical model for sagebrush cover change (Eq.
 7) using standardized covariate values.
-Only cumulative precipitation at time \\emph{t-2} (pptLag) is important (shown in dark grey; 90% CI does not overlap zero).
-Climate covariate codes: pptLag = water year precipitation in year \\emph{t-2}; TmeanSpr1 = year \\emph{t-1} average spring temperature; ppt2 = year \\emph{t} fall through summer precipitation; TmeanSpr2 = year \\emph{t} average spring temperature; ppt1 = year \\emph{t-1} fall through summer precipitation.", fig.height=4, eval=T}
-library(png)
-library(grid)
-img <- readPNG("../results/post_climate_covariates.png")
-grid.raster(img)
-```
+Only cumulative precipitation at time \emph{t-2} (pptLag) is important (shown in dark grey; 90\% CI does not overlap zero).
+Climate covariate codes: pptLag = water year precipitation in year \emph{t-2}; TmeanSpr1 = year \emph{t-1} average spring temperature; ppt2 = year \emph{t} fall through summer precipitation; TmeanSpr2 = year \emph{t} average spring temperature; ppt1 = year \emph{t-1} fall through summer precipitation.}
+\end{figure}
 
-```{r figure_3,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=F}
-library(png)
-library(grid)
-img <- readPNG("../results/spatialfield.png")
-grid.raster(img)
-```
+\newpage{}
 
-```{r figure_4,dependson="plot-options", fig.cap="Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and precision (C, D) for the extent of our spatial area at 30-m resolution.
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{../results/obs_pred_spatial_subset.png}
+  \caption{Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and precision (C, D) for the extent of our spatial area at 30-m resolution.
 Observed equilibrium sagebrush cover (A) is the temporal mean of each pixel from the 28 year time series.
@@
 Precision in (D) represents the variability of each pixel over the course of the 2,000 iteration simulation.
-Axes definitions: Lat = latitude; Lon = longitude.", fig.height=8, eval=T}
-library(png)
-library(grid)
-img <- readPNG("../results/obs_pred_spatial_subset.png")
-grid.raster(img)
-```
+Axes definitions: Lat = latitude; Lon = longitude.}
+\end{figure}
 
-```{r figure_5,dependson="plot-options", fig.cap="Histograms of projected equilibrium cover from simulations using observed climate (orange) and average observed percent cover (black).", eval=F}
-library(png)
-library(grid)
-img <- readPNG("../results/obs_pred_hist.png")
-grid.raster(img)
-```
+\newpage{}
 
-```{r figure_6,dependson="plot-options", fig.cap="Projected equilibrium cover under three IPCC climate change scenarios (RCP = Representative Concentration Pathways) for our study area in southwestern Wyoming.
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{../results/clim_change_mean_spatial.png}
+  \caption{Projected equilibrium cover under three IPCC climate change scenarios (RCP = Representative Concentration Pathways) for our study area in southwestern Wyoming.
 The top panel shows equilibrium cover based on simulations using observed climate.
@@
 Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.
-Axes definitions: Lat = latitude; Lon = longitude.", fig.height=6, eval=T}
-library(png)
-library(grid)
-img <- readPNG("../results/clim_change_mean_spatial.png")
-grid.raster(img)
-```
+Axes definitions: Lat = latitude; Lon = longitude.}
+\end{figure}
 
-```{r figure_7,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time.", eval=F}
-library(png)
-library(grid)
-img <- readPNG("../results/clim_change_densities.png")
-grid.raster(img)
-```
+\newpage{}
 
-```{r figure_8,dependson="plot-options", fig.cap="Observed (black line before 2011) and forecasted (colored lines after 2011) sagebrush percent cover.
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{../results/temporal_forecast.png}
+  \caption{Observed (black line before 2011) and forecasted (colored lines after 2011) sagebrush percent cover.
 Forecasts were made for three IPCC emissions scenarios (RCPs 4.5, 6.0, and 8.5) and are for the period of 2012 to 2098.
-Shaded regions show 90% quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
+Shaded regions show 90\% quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
 Lines show mean trajectories.
 Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
-Before calculating the mean and 90% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.", eval=T}
-library(png)
-library(grid)
-img <- readPNG("../results/temporal_forecast.png")
-grid.raster(img)
-```
+Before calculating the mean and 90\% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.}
+\end{figure}
 
-\pagebreak{}
+\newpage{}
 

### 2016-06-17 docs/ms_sageAbundance.Rmd
little clean up; adds short term forecast to figure 6

@@
 \hl{The uncerainty associated with our forecasts highlights several opportunities to improve our approach.
-First, Parameter uncertainty could be reduced by regulating the variance of the posterior distributions of climate covariates via ridge regression} [e.g., @Gerber2015].
+First, parameter uncertainty could be reduced by regulating the variance of the posterior distributions of climate covariates via ridge regression} [e.g., @Gerber2015].
 \hl{Second, uncertainty associated with climate projections could be reduced by identifying GCMs that perform exceptionally well for a particular study location} [e.g., @Rupp2013].
@@
   \centering
-      \includegraphics[height=6in]{./components/figure/studyarea_map.png}
+      \includegraphics[height=6in]{../figures/studyarea_map.png}
   \caption{Location of the 5,070 $\times$ 2,430 meter kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
@@
   \centering
-      \includegraphics[width=4in]{./components/figure/weather_projections.png}
+      \includegraphics[width=4in]{../figures/weather_projections.png}
   \caption{GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
@@
   \centering
-      \includegraphics[width=4in]{../results/post_climate_covariates.png}
+      \includegraphics[width=4in]{../figures/post_climate_covariates.png}
   \caption{Posterior distributions of climate covariates.
@@
 7) using standardized covariate values.
-Only cumulative precipitation at time \emph{t-2} (pptLag) is important (shown in dark grey; 90\% CI does not overlap zero).
+Only cumulative precipitation at time \emph{t-2} (pptLag) is important (shown in blue; 90\% CI does not overlap zero).
 Climate covariate codes: pptLag = water year precipitation in year \emph{t-2}; TmeanSpr1 = year \emph{t-1} average spring temperature; ppt2 = year \emph{t} fall through summer precipitation; TmeanSpr2 = year \emph{t} average spring temperature; ppt1 = year \emph{t-1} fall through summer precipitation.}
@@
   \centering
-      \includegraphics[height=6in]{../results/obs_pred_spatial_subset.png}
+      \includegraphics[height=6in]{../figures/obs_pred_spatial_subset.png}
   \caption{Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and precision (C, D) for the extent of our spatial area at 30-m resolution.
@@
   \centering
-      \includegraphics[height=6in]{../results/clim_change_mean_spatial.png}
+      \includegraphics[height=6in]{../figures/clim_change_mean_spatial.png}
   \caption{Projected equilibrium cover under three IPCC climate change scenarios (RCP = Representative Concentration Pathways) for our study area in southwestern Wyoming.
@@
   \centering
-      \includegraphics[width=4in]{../results/temporal_forecast.png}
+      \includegraphics[width=3in]{../figures/figure6.png}
   \caption{Observed (black line before 2011) and forecasted (colored lines after 2011) sagebrush percent cover.
-Forecasts were made for three IPCC emissions scenarios (RCPs 4.5, 6.0, and 8.5) and are for the period of 2012 to 2098.
-Shaded regions show 90\% quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
+Long-term forecasts (A) were made for three IPCC emissions scenarios (RCPs 4.5, 6.0, and 8.5) and are for the period of 2012 to 2098.
+Shaded regions show limits of the 5th and 95th quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
 Lines show mean trajectories.
 Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
-Before calculating the mean and 90\% quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixles.}
+Before calculating the mean and quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixels.
+Panel (B) shows an example short-term forecast (10 years) using the MIROC5 GCM projections under RCP 8.5.
+Each line shows a forecast from one parameter set.}
 \end{figure}

### 2016-06-17 scripts/ms_scripts.R

@@ -182,7 +182,7 @@ png("../results/post_climate_covariates.png", height=4, width=4, units="in", res
 ggplot(clim_posts, aes(x=value, fill=col))+
   geom_vline(aes(xintercept=0), linetype=2)+
   geom_density(alpha=0.5,size=0.5, col=NA, adjust=4)+
-  scale_fill_manual(values=c("grey70", "black", "red"))+
+  scale_fill_manual(values=c("grey60", "steelblue", "red"))+
   ylab("Posterior density")+
   xlab("Standardized coefficient value")+
   facet_grid(rank~., labeller=my_labeller)+

### 2016-06-17 scripts/plot_temporal_forecasts.R

@@ -86,7 +86,7 @@ ggplot()+
   geom_line(data=meandf, aes(x=year, y=avgcover, color=scenario))+
   geom_vline(aes(xintercept=2011))+
   ylab("Mean sagebrush cover (%)")+
-  guides(color=FALSE)+
+  guides(color=FALSE,fill=FALSE)+
   scale_y_continuous(limits=c(0,30))+
   scale_fill_manual(values=c("tan","coral","darkred"), 
                     name="IPCC \nScenario",
@@ -95,6 +95,21 @@ ggplot()+
                      name="IPCC \nScenario",
                      labels=c("RCP 4.5", "RCP 6.0", "RCP 8.5"))+
   theme_bw()
-ggsave("../temporal_forecast_wpois.png", width = 5, height = 4)
+ggsave("../temporal_forecast_wpois.png", width = 5, height = 4, dpi = 120)
 
+
+
+####
+####  PLOT A SHORT TERM FORECAST FROM ONE GCM
+####
+one_gcm <- subset(longdf, model == "miroc5" & year < 2022 & scenario=="rcp85")
+ggplot()+
+  geom_line(data=obs_agg, aes(x=Year, y=avgcover))+
+  geom_line(data=one_gcm, aes(x=year, y=cover, group=paramset), alpha=0.5, color="darkred")+
+  geom_vline(aes(xintercept=2011))+
+  ylab("Mean sagebrush cover (%)")+
+  guides(color=FALSE)+
+  scale_y_continuous(limits=c(0,30))+
+  theme_bw()
+ggsave("../short_term_miroc.png", width = 5, height = 4, dpi=120)
 

### 2016-06-17 docs/appendixB.Rmd
adds new appendix; includes ms revisions; ready for coauthors

--- 2016-01-14 docs/appendixB.Rmd
@@
    - \doublespacing
+   - \usepackage{graphicx}
 bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
@@
 
-Appendix B: Detailed methods for knot choice and placement
-==========================================================
-Our dimension reduction approach requires fitting random effects associated with knots distributed nonrandomly in space.
-By distributing *m* knots across space, we can fit random effects for each knot and then use process convolution to interpolate the random effect between the knots.
-Increasing the number of knots results in a more finely-resolved spatial effect, but it also increases processe time.
-Thus, there exists a tension between the resolution of the spatial effect and computational effort.
-Without regard to computation time, it is possible to use common model selection techniques to guide the number of knots one should use.
-With regard to computation time, it is sensible to choose knot placement and number based on the observed residual spatial variation.
+Appendix B: Annual maps of sagebrush cover
+==========================================
 
-To do so, we fit a generalized linear model of the form
+We used remotely-sensed estimates of sagebrush cover [@Homer2012].
+Here we show the annual maps of sagebrush cover as hindcasted with the regression tree algorithm described in @Homer2012.
 
-\vspace{-4em}
-\begin{align}
-y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
-\end{align}
-\vspace{-4em}
-
-where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*
-
-\vspace{-4em}
-\begin{align}
-\text{log}(\mu_{i,t}) &= \beta_{0} + \beta_{1}y_{i,t-1} + \textbf{x}_{t}'\boldsymbol{\gamma}
-\end{align}
-\vspace{-3em}
-
-that includes a density-dependence effect of log-transformed cover in the previous year ($y_{i,t-1}$) and climate effects ($\textbf{x}_t$).
-In other words, a non-hierarchial version of the model presented in our main text.
-We then examined a spatial variogram of the residuals from the above model to identify the the distance at which spatial dependence disappears.
-
-Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig.
-B1).
-Thus, we set the kernal bandwidth ($\sigma$ in Eq.
-X of the main text) to 500/3 = `r round(500/3,2)`.
-We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters.
-Knots were placed evenly across space in the x and y directions.
-This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig.
-B2).
-
-![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernel bandwidth in Eq.
-6 of the main text (500/3; dashed black line).](../results/avgresidual_variogram_withrange_subset.png)
-
-![Map of our study area showing percent cover in 1984 and the location of the 231 evenly-spaced knots.](../results/SAGE_Grid_wKnots_subset.png)
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=7in]{../figures/all_years_percCover.png}
+  \caption{Estimated percent cover of 30 x 30 meter pixels in our study area in southwestern Wyoming (see Fig. 1). Maps show percent cover for each year (1984-2011).}
+\end{figure}

### 2016-06-17 docs/appendixC.Rmd
adds year effect to sims; updates docs with Mevin's comments
adds new appendix; includes ms revisions; ready for coauthors

--- 2015-11-13 docs/appendixC.Rmd
@@
 
-Appendix C: Model convergence diagnostics
-=========================================
-Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure C1).
-Visual inspection of traceplots indicates convergence (Figures C2-C10).
+Appendix C: Detailed methods for knot choice and placement
+==========================================================
+Our dimension reduction approach requires fitting random effects associated with knots distributed nonrandomly in space.
+By distributing *m* knots across space, we can fit random effects for each knot and then use process convolution to interpolate the random effect between the knots.
+Increasing the number of knots results in a more finely-resolved spatial effect, but it also increases processe time.
+Thus, there exists a tension between the resolution of the spatial effect and computational effort.
+Without regard to computation time, it is possible to use common model selection techniques to guide the number of knots one should use.
+With regard to computation time, it is sensible to choose knot placement and number based on the observed residual spatial variation.
 
-Table:  Definition of parameters and model notation for Figures C1-C10.
+To do so, we fit a generalized linear model of the form
 
-| Parameter | Definition | Notation in figures |
-| --------- | ---------- | -------------------------------------- |
-| $\alpha_{i}$ | spatial random effect for for knot *i* | ``alpha.i`` where ``i`` is the numeric knot id |
-| $\gamma_{x}$ | climate effect for covariate *x* | ``beta.x`` where ``x`` is the numeric climate covariate id |
-| $\beta_{0t}$ | random effect for year *t* on the intercept | ``int_yr.t`` where ``t`` is the numeric year id |
-| $\bar{\beta_{0}}$ | mean intercept | ``int_mu`` |
-| $\beta_{1}$ | density-dependent effect | ``beta_mu`` |
-| $\sigma_{\beta_{0}}^2$ | variance of year random effect | ``sig_yr`` |
-| $\sigma_{\eta}^2$ | variance of spatial random effect | ``sig_a`` |
+\vspace{-4em}
+\begin{align}
+y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
+\end{align}
+\vspace{-4em}
 
-![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../results/rhat_plot.png)
+where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_1.png)
+\vspace{-4em}
+\begin{align}
+\text{log}(\mu_{i,t}) &= \beta_{0} + \beta_{1}y_{i,t-1} + \textbf{x}_{t}'\boldsymbol{\gamma}
+\end{align}
+\vspace{-3em}
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_2.png)
+that includes a density-dependence effect of log-transformed cover in the previous year ($y_{i,t-1}$) and climate effects ($\textbf{x}_t$).
+In other words, a non-hierarchial version of the model presented in our main text.
+We then examined a spatial variogram of the residuals from the above model to identify the the distance at which spatial dependence disappears.
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_3.png)
+Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig.
+C1).
+Thus, we set the kernel bandwidth ($\sigma$ in Eq.
+6 of the main text) to 500/3 = `r round(500/3,2)` because the effective spatial range under an exponential covariance structure is equal to $\sigma \times 3$.
+We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters.
+Knots were placed evenly across space in the x and y directions.
+This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig.
+C2).
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_4.png)
+![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernel bandwidth in Eq.
+6 of the main text (500/3; dashed black line).](../figures/avgresidual_variogram_withrange_subset.png)
 
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_5.png)
-
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_6.png)
-
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_7.png)
-
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_8.png)
-
-![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_9.png)
-
-References
-----------
+![Map of our study area showing percent cover in 1984 and the location of the 231 evenly-spaced knots.](../figures/SAGE_Grid_wKnots_subset.png)

### 2016-06-17 docs/appendixD.Rmd
adds new appendix; includes ms revisions; ready for coauthors

@@
 \renewcommand\thefigure{D\arabic{figure}}  
+\renewcommand\thetable{D\arabic{table}}
 
-Appendix D: Posterior estimate of spatial random effects ($\boldsymbol{\eta}$)
-==============================================================================
+Appendix D: Model convergence diagnostics
+=========================================
+Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure D1).
+Visual inspection of traceplots indicated convergence.
 
-```{r figure_D1,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=TRUE, echo=FALSE}
-library(png)
-library(grid)
-img <- readPNG("../results/spatialfield.png")
-grid.raster(img)
-```
+\footnotesize
+
+Table:  Definition of parameters and model notation for Figure D1.
+
+| Parameter | Definition | Notation in figures |
+| --------- | ---------- | -------------------------------------- |
+| $\alpha_{i}$ | spatial random effect for for knot *i* | ``alpha.i`` where ``i`` is the numeric knot id |
+| $\gamma_{x}$ | climate effect for covariate *x* | ``beta.x`` where ``x`` is the numeric climate covariate id |
+| $\beta_{0t}$ | random effect for year *t* on the intercept | ``int_yr.t`` where ``t`` is the numeric year id |
+| $\bar{\beta_{0}}$ | mean intercept | ``int_mu`` |
+| $\beta_{1}$ | density-dependent effect | ``beta_mu`` |
+| $\sigma_{\beta_{0}}^2$ | variance of year random effect | ``sig_yr`` |
+| $\sigma_{\eta}^2$ | variance of spatial random effect | ``sig_a`` |
+
+\normalsize
+
+![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../figures/rhat_plot.png)
+
+<!--
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_01.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_02.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_03.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_04.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_05.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_06.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_07.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_08.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_09.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_10.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_11.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_12.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_13.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_14.png)
+
+![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_15.png)
+
+-->
+
+References
+----------

### 2016-06-17 docs/appendixF.Rmd
adds new appendix; includes ms revisions; ready for coauthors

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{F\arabic{figure}}

Appendix F: Density plots of sagebrush cover forecasts ($\boldsymbol{\eta}$)
==============================================================================

```{r figure_F1,dependson="plot-options", fig.cap="Estimated kernel densities for current and projected sagebrush cover across space and simulation time. These densities represent the desnity of cover values from the spatial plots in Figure 5 of the main text.", eval=TRUE, echo=FALSE}
library(png)
library(grid)
img <- readPNG("../figures/clim_change_densities.png")
grid.raster(img)
```

### 2016-06-17 docs/ms_sageAbundance.Rmd
adds new appendix; includes ms revisions; ready for coauthors

@@
 To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \hl{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
-This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. A1).
+This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. B1).
 In this \hl{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
@@
 However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig.
-B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
+C2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix C).
 An infinite number of knots would result in an exact representation of the spatial process and covariance model.
@@
 Model fitting required five days on a four node Central Processing Unit with 2 $\times$ AMD Opteron(tm) Processor 4386 @ 3.10 Ghz, 64GB of RAM per node, 16 cores per node, and each chain launched in parallel on separate cores.
-We assessed convergence visually and calculated scale-reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].
+We assessed convergence visually and calculated scale-reduction factors (Appendix D, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].
 
@@
 
-All parameters in our model converged on stable posterior distributions (Appendix C).
+All parameters in our model converged on stable posterior distributions (Appendix D).
 Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig. 3).
@@
 3).
-The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
+The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. E1).
 This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.
@@
 
-When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
+When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6A).
 The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
-Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5 and Fig. E1).
+Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5 and Fig. F1).
 For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig.
-6).
-Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
+6A).
+Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6A).
 The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change estimated for our study area (Fig. 3).
+We also show how our model is capable of near-term forecasts in Fig 6B.
 
@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-We \hl{introduced} a new approach \hl{that combines a} remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling \hl{literature} [e.g., @Conn2015].
-\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming.}
+We \hl{introduced} a new approach \hl{that uses methods from the dynamic spatio-temporal modeling literature} [e.g., @Conn2015] \hl{to fit a population model to remotely-sensed estimated of plant percent cover}.
+\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming} [@Homer2012].
 We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.
@@
 
-Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios (Figs. 5,6).
+Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios in the long-term (Figs. 5,6A).
 Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
@@
 
-An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. E1).
+An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. F1).
 Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
@@
 4A) and approximated by our model ($\boldsymbol{\eta}$, Fig.
-D1) to slope, aspect, elevation, or coarse soil type (results not shown).
+E1) to slope, aspect, elevation, or coarse soil type (results not shown).
 The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects related to fire and species invasions.
@@
 3) and carried through to our forecasts (Figs.
-5,6) likely represents the positive effect of earlier snowmelt, and thus higher moisture availability early in the growing season.
+5,6A) likely represents the positive effect of earlier snowmelt, and thus higher moisture availability early in the growing season.
 
@@
 
-We projected sagebrush cover to the end of this century, but an important feature of our approach is that it could also produce short-term forecasts.
+We projected sagebrush cover to the end of this century, but an important feature of our approach is that it can also produce short-term forecasts (Fig. 6B).
 For example, we could forecast the effects of a multi-year regional drought on sagebrush cover [@Debinski2010].

### 2016-07-01 docs/ms_sageAbundance.Rmd
add in all but PBA final edits

@@
 Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
-Our approach \hl{, which brings together several common techniques,} is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
+Our approach, \textcolor{blue}{which brings together several common techniques,} is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
 
@@
 Building on recent technological advances in spatial statistics [@Latimer2009; @Conn2015] and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict population impacts of climate change.
-\hl{As a proof-of-concept, we use} a process model motivated by Gompertz density-dependent population growth and a remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015].
+\textcolor{blue}{As a proof-of-concept, we use} a process model motivated by Gompertz density-dependent population growth and a remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015].
 We account for spatial autocorrelation with dimension reduction techniques [@Latimer2009; @Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
-The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the greatest potential to combine the information of population models (e.g., population status \hl{and temporal dynamics}) and the spatial extent of species distribution models.
+\textcolor{blue}{Unlike most SDMs, our approach models the dynamics of plant abundance through time, and thus, is a population model, in the same spirit that models of animal counts through time are population models.}
+The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the greatest potential to combine the information of population models (e.g., population status \textcolor{blue}{and temporal dynamics}) and the spatial extent of species distribution models.
 
@@
 To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
-As part of a separate study, \hl{Homer et al.
+As part of a separate study, \textcolor{blue}{Homer et al.
 2012 estimated} sagebrush percent cover using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
 The regression tree model was further validated using another 297 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
-To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \hl{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
+To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \textcolor{blue}{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
 This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. B1).
-In this \hl{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
+In this \textcolor{blue}{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
 In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
-Thus, the full \hl{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
+Thus, the full \textcolor{blue}{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
 
@@
 There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we handled the small number of zero percent cover observations.
-\hl{We assume that the remotely sensed estimates of percent cover are "true" and free of error.}
-\hl{This need not be the case, and if measurement error is known then it could be included in our Bayesian model as a "data model"} [@Hobbs2015].
+\textcolor{blue}{We assume that the remotely sensed estimates of percent cover are "true" and free of error.}
+\textcolor{blue}{This need not be the case, and if measurement or sampling error is known then it could be included in our Bayesian model as a "sampling model"} [@Hobbs2015].
 
@@
 We fit this simple model using the `glm' command in R [@R2013].
-For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied \hl{and covariates such as cover of neighboring cells could be included}.
+For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied \textcolor{blue}{and covariates such as cover of neighboring cells could be included}.
 
@@
 These yearly climate time series were not aggregated further because we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
-\hl{Note that forecasting with our model, which relies on historical correlations between sagebrush cover change and weather, using future climate projections requires assuming such relationships remain in the future.}
+\textcolor{blue}{Note that forecasting with our model, which relies on historical correlations between sagebrush cover change and weather, using future climate projections requires assuming such relationships remain in the future.}
 
@@
 Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
-We \hl{introduced} a new approach \hl{that uses methods from the dynamic spatio-temporal modeling literature} [e.g., @Conn2015] \hl{to fit a population model to remotely-sensed estimated of plant percent cover}.
-\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming} [@Homer2012].
+We \textcolor{blue}{introduced} a new approach \textcolor{blue}{that uses methods from the dynamic spatio-temporal modeling literature} [e.g., @Conn2015] \textcolor{blue}{to fit a population model to remotely-sensed estimated of plant percent cover}.
+\textcolor{blue}{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming} [@Homer2012].
 We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.
@@
 First, we used moderate resolution, remotely-sensed estimates of sagebrush percent cover as a response variable, enabling us to fit a dynamic population model over a large spatial extent. 
-\hl{Species-specific estimates of plant abundance are becoming commonplace as remote sensing technology develops} [e.g., @Baldeck2014; @Colgan2014], \hl{and in a few years several remotely-sensed time series may be available.}
+\textcolor{blue}{Species-specific estimates of plant abundance are becoming commonplace as remote sensing technology develops} [e.g., @Baldeck2014; @Colgan2014], \textcolor{blue}{and in a few years several remotely-sensed time series may be available.}
 Second, borrowing from new methods in spatio-temporal modeling of animal abundance [e.g., @Conn2015], we fit the model using a dimension reduction strategy that accounted for spatial autocorrelation within a feasible computational time.
@@
 
-\hl{The uncerainty associated with our forecasts highlights several opportunities to improve our approach.
+\textcolor{blue}{The uncertainty associated with our forecasts highlights several opportunities to improve our approach.
 First, parameter uncertainty could be reduced by regulating the variance of the posterior distributions of climate covariates via ridge regression} [e.g., @Gerber2015].
-\hl{Second, uncertainty associated with climate projections could be reduced by identifying GCMs that perform exceptionally well for a particular study location} [e.g., @Rupp2013].
-\hl{Such considerations will be important when forecasting in support of particular management objectives.}
+\textcolor{blue}{Second, uncertainty associated with climate projections could be reduced by identifying GCMs that perform exceptionally well for a particular study location} [e.g., @Rupp2013].
+\textcolor{blue}{Such considerations will be important when forecasting in support of particular management objectives.}
+\textcolor{blue}{However, knowledge of uncertainty is itself important knowledge for management} [@Bradshaw2000].
+\textcolor{blue}{Deciding that no management actions should be taken based on the data at hand is itself a management decision.}
 
@@
 We are grateful to Debra K.
-Meyer at USGS EROS for extracting the data set used in this paper and to David Koons for comments that improved the manuscript.
+Meyer at USGS EROS for extracting the data set used in this paper and to David Koons \textcolor{blue}{and two anonymous reviewers} for comments that improved the manuscript.
 Compute, storage, and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
@@
 Before calculating the mean and quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixels.
-Panel (B) shows an example short-term forecast (10 years) using the MIROC5 GCM projections under RCP 8.5.
-Each line shows a forecast from one parameter set.}
+\textcolor{blue}{Panel (B) shows an example short-term forecast (10 years) using the MIROC5 GCM projections under RCP 8.5.
+Each line shows a forecast from one parameter set.}}
 \end{figure}

### 2016-07-01 docs/ms_sageAbundance.Rmd
adds in PBA final edits

@@
 We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of sagebrush.
-We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale.
+We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of \textcolor{blue}{sagebrush population trajectories} that have, until now, been impossible to produce at this scale.
 Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics.
-Our approach, \textcolor{blue}{which brings together several common techniques,} is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
+Our approach, \textcolor{blue}{which combines several existing techniques in a novel way, demonstrates the use of} remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.
 
@@
 
-The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents and statistical methods for fitting high-dimensional spatial models.
+The ideal tool would be a \textcolor{blue}{broad}-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents and statistical methods for fitting high-dimensional spatial models.
 Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
@@
 In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
-Thus, the full \textcolor{blue}{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
+Thus, the subset of the \textcolor{blue}{remote sensing product} we use contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).
 
@@
 These yearly climate time series were not aggregated further because we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.
-\textcolor{blue}{Note that forecasting with our model, which relies on historical correlations between sagebrush cover change and weather, using future climate projections requires assuming such relationships remain in the future.}
+\textcolor{blue}{The key assumption of our forecasting approach is that the historical correlations between weather and sagebrush cover change will continue to hold in the future.}
 
@@
 However, mean estimates for the climate effects are relatively weak (Fig. 3).
+
 Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
@@
 E1) to slope, aspect, elevation, or coarse soil type (results not shown).
-The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects related to fire and species invasions.
+The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects \textcolor{blue}{of disturbance}.
 
@@
 \textcolor{blue}{However, knowledge of uncertainty is itself important knowledge for management} [@Bradshaw2000].
-\textcolor{blue}{Deciding that no management actions should be taken based on the data at hand is itself a management decision.}
+\textcolor{blue}{Deciding that no actions should be taken based on the data at hand is itself a management decision.}
 

### 2016-07-01 docs/ms_sageAbundance.tex
adds tex verision of ms

@@
             citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
+            urlcolor=black,
+            linkcolor=black,
             pdfborder={0 0 0}}
@@
 \usepackage{titling}
+
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
 \setlength{\droptitle}{-2em}
@@
 \doublespacing
+\usepackage[document]{ragged2e}
+\usepackage{color, soul}
+\usepackage{graphicx}
 
@@
 
-\section{Forecasting climate change impacts on plant population dynamics
-at large spatial extents: a test case with sagebrush (\emph{Artemisia})
-species}\label{forecasting-climate-change-impacts-on-plant-population-dynamics-at-large-spatial-extents-a-test-case-with-sagebrush-species}
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
-\subsubsection[Andrew T. Tredennick, Mevin B. Hooten, Cameron L.
-Aldridge, Colin G. Homer, David T. Iles, Andrew Kleinhesselink, Eric
-LaMalfa, Rebecca Mann, and Peter B. Adler]{Andrew T.
-Tredennick\footnote{Corresponding author:
-  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}},
-Mevin B. Hooten, Cameron L. Aldridge, Colin G. Homer, David T. Iles,
-Andrew Kleinhesselink, Eric LaMalfa, Rebecca Mann, and Peter B.
-Adler}\label{andrew-t.-tredennickcorrauth-mevin-b.-hooten-cameron-l.-aldridge-colin-g.-homer-david-t.-iles-andrew-kleinhesselink-eric-lamalfa-rebecca-mann-and-peter-b.-adler}
+\begin{centering}
+\textbf{\large{Forecasting climate change impacts on plant populations over large spatial extents}}
 
-\emph{Andrew T. Tredennick, Department of Wildland Resources and the
-Ecology Center, Utah State University, Logan, UT}
+\textsc{\small{Andrew T.
+Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B.
+Hooten\textsuperscript{2,3,4}, Cameron L.
+Aldridge\textsuperscript{5}, Collin G.
+Homer\textsuperscript{6}, Andrew R.
+Kleinhesselink\textsuperscript{1}, and Peter B.
+Adler\textsuperscript{1}}}
 
-\emph{Mevin B. Hooten, U.S. Geological Survey, Colorado Cooperative Fish
-and Wildlife Research Unit; Department of Fish, Wildlife, and
-Conservation Biology, Colorado State University; Department of
-Statistics, Colorado State University; Graduate Degree Program in
-Ecology, Colorado State University, Fort Collins, CO}
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\
+\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{5}Natural Resource Ecology Laboratory and Department of Ecosystem Science and Sustainability, Colorado State University, Fort Collins, CO, in cooperation with US Geological Survey, Fort Collins Science Center, Fort Collins, CO}}\\
+\textit{\small{\textsuperscript{6}U.S. Geological Survey (USGS) Earth Resources Observation and Science (EROS) Center, Sioux Falls, SD 57198 USA}}
 
-\emph{Cameron L. Aldridge, Department of Ecosystem Science and
-Sustainability, Colorado State University; Natural Resource Ecology
-Laboratory, Colorado State University; U.S. Geological Survey, Fort
-Collins Science Center; Graduate Degree Program in Ecology, Colorado
-State University, Fort Collins, CO}
+\end{centering}
 
-\emph{Colin G. Homer, U.S. Geological Survey (USGS) Earth Resources
-Observation and Science (EROS) Center, Sioux Falls, SD}
+\renewcommand*{\thefootnote}{\arabic{footnote}}
 
-\emph{David T. Iles, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
+\setcounter{footnote}{0}
 
-\emph{Andrew Kleinhesselink, Department of Wildland Resources and the
-Ecology Center, Utah State University, Logan, UT}
+\subsection{Abstract}\label{abstract}
 
-\emph{Eric LaMalfa, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
+Plant population models are powerful tools for predicting climate change
+impacts in one location, but are difficult to apply at landscape scales.
+We overcome this limitation by taking advantage of two recent advances:
+remotely-sensed, species-specific estimates of plant cover and
+statistical models developed for spatio-temporal dynamics of animal
+populations. Using computationally efficient model reparameterizations,
+we fit a spatiotemporal population model to a 28 year time series of
+sagebrush (\emph{Artemisia} spp.) percent cover over a \(2.5\times5\) km
+landscape in southwestern Wyoming while formally accounting for spatial
+autocorrelation. We include interannual variation in precipitation and
+temperature as covariates in the model to investigate how climate
+affects the cover of sagebrush. We then use the model to forecast the
+future abundance of sagebrush at the landscape scale under projected
+climate change, generating spatially explicit estimates of
+\textcolor{blue}{sagebrush population trajectories} that have, until
+now, been impossible to produce at this scale. Our broad-scale and
+long-term predictions are rooted in small-scale and short-term
+population dynamics and provide an alternative to predictions offered by
+species distribution models that do not include population dynamics. Our
+approach,
+\textcolor{blue}{which combines several existing techniques in a novel way, demonstrates the use of}
+remote sensing data to model population responses to environmental
+change that play out at spatial scales far greater than the traditional
+field study plot.
 
-\emph{Rebecca Mann, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
-
-\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
-
-\subsection{Summary}\label{summary}
-
-\begin{enumerate}
-\item Global environmental changes, like climate change, tend to play out at the landscape and regional scales.
-Thus, it is reasonable that attempts to forecast the impacts of such change do so at similar spatial scales.
-For plant species, this has meant relying primarily on a single tool, species distribution models, that contain little information on population dynamics and states.
-
-\item Plant population models have proven exceedingly useful for projecting population dynamics and states under altered climate regimes, but most models are parameterized using “local” scale data from, generally, 1 $\text{m}^{2}$ plots.
-These models, even if highly predictive, cannot capture the full spatial variability in population responses to climate and thus are difficult to extrapolate from the micro to meso scale.
-
-\item Here we build on density-structured population modeling approaches and describe a population model based on remote sensing pixels -- a `pixel-based' population model.
-At its core the model is simply an individual-based population model where, instead of individuals, we focus on pixels as super individuals.
-To demonstrate the approach, we model the rate of transitions among discrete percent cover classes using a 27 year remotely sensed time series of sagebrush (Artemesia spp.) cover in southwestern Wyoming.
-We use climate covariates in vital rate regressions so that we can forecast future population states under projected climate change.
-
-\item The pixel-based population model successfully recovers the the mean and spread of the observed data when we use observed climate covariates.
-
-\item Our approach...
-\end{enumerate}
-
-\emph{Keywords}: population model, forecasting, integral projection
-model, etc.
+\emph{Key words: population model, climate change, forecasting,
+spatiotemporal model, remote sensing, sagebrush}, Artemisia,
+\emph{dimension reduction}
 
@@
 Forecasting the impacts of climate change on plant populations and
-communities is a central challenge of modern ecology. The challenge lies
-in faithfully representing population dynamics at spatial and temporal
-scales relevant to policy and management decisions. However, almost
+communities is a central challenge for ecology (Clark et al. 2001,
+Petchey et al. 2015). Population models are ideally suited for meeting
+such a challenge because they provide a way to link climate drivers
+directly to population dynamics (Hare et al. 2010, Adler et al. 2012,
+Ross et al. 2015, Shriver 2015). However, inference from population
+models is typically limited to small spatial extents because the data
+required is difficult to collect across broad species ranges. Almost
 every study of plant population dynamics relies on demographic
-observations recorded at the meter to sub-meter scale. Local-scale
-demographic data make building population projection models an easy task
-(Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is
-very difficult to extrapolate small-scale studies to large spatial
-extents with any certainty because the data likely only represent a
-small subset of parameter space and environmental conditions. Also,
-given the costliness (in time and money) of collecting detailed
-demographic data, it is unlikely that ``enough'' local-scale data can be
-collected to accurately project large-scale population dynamics.
+observations recorded at the meter to sub-meter scale (see, e.g.,
+Salguero-G{ó}mez et al. 2015). Local-scale demographic data make
+building population projection models an easy task (Ellner and Rees
+2006, Rees and Ellner 2009, Adler et al. 2012), but it is very difficult
+to extrapolate small-scale studies to large spatial extents with any
+certainty because the data likely only represent a small subset of
+parameter space and environmental conditions (Freckleton et al. 2011,
+Queenborough et al. 2011). The real challenge is not to simply make
+population forecasts, but to do so at spatial scales relevant to policy
+and management decisions (Queenborough et al. 2011).
 
-The fundamental limitation of local-scale data is that they are
-insufficient for estimating the full extent of variation in model
-parameters possible at larger spatial and temporal extents (Freckleton
-et al., 2011; Queenborough et al., 2011). It is possible to estimate the
-variance of population model parameters related to spatial and temporal
-variability, but the observations used to estimate parameter uncertainty
-tend to be clustered in space or time. Thus, extrapolating beyond the
-spatial or temporal resolution of the data used to parameterize models
-can lead to biased and wrong forecasts of future population states.
+The ideal tool would be a \textcolor{blue}{broad}-scale, dynamic
+population model (Schurr et al. 2012, Merow et al. 2014), but developing
+useful models at this scale has been limited by the availability of time
+series data at large spatial extents and statistical methods for fitting
+high-dimensional spatial models. Fortunately, new advances in remote
+sensing and statistics now allow us to overcome both of these
+limitations. First, new remote-sensing (RS) methods are now producing
+accurate time series of species-specific plant cover at landscape
+scales. These data can be fit with dynamic population models which
+include yearly fluctuations in climate as covariates. Such RS time
+series have revolutionized models of how climate affects ecosystem-level
+processes (e.g., Running et al. 2004) and have been used to detect
+long-term trends in plant population abundance (e.g., Homer et al.
+2015), but they have yet to be used to drive a dynamic population model.
+Second, animal population modelers have developed dimension reduction
+and reparameterization techniques to efficiently fit high-dimension
+spatiotemporal models (see Conn et al. 2015 for a review). These new
+statistical methods have yet to be applied to RS-derived plant
+population data at broad scales.
 
-Alternatively, large-scale trends in populations are easily detected
-using widely available monitoring data, but such data are rarely used to
-project future population states. Given the tradeoffs between the
-spatial-scale of data and the inference that can be drawn from the data,
-we suggest combining the best features of local-scale population
-modeling and large-scale monitoring data. Specifically, we propose a new
-approach for modeling population dynamics at large spatial extents based
-on the theory and mechanics of individual-based models (IBMs) as applied
-to remotely-sensed time series data.
+Large-scale, spatially-explicit population models based on RS data could
+offer a valuable new way to investigate the effects of large-scale
+environmental changes playing out at landscape and regional scales. Most
+current assesments of how plant and animal populations will respond to
+climate change rely on species distribution models (SDMs). SDMs rely on
+static associations between contemporary climate and a species'
+distribution or, more rarely, abundance to project future distribution
+or abundance (Elith and Leathwick 2009) and they are easily applied at
+landscape to continental scales (e.g., Maiorano et al. 2013, Clark et
+al. 2014). However, the short-term and small-scale population dynamics
+that actually drive the large-scale distributions of species are not
+represented in most SDMs. Because SDMs typically rely on occurence data,
+their projections of habitat suitability or probability of occurence
+provide little information on the future states of populations in the
+core of their range -- areas where a species exists now and is expected
+to persist in the future (Ehrl{é}n and Morris 2015). Furthermore,
+because they lack short-term dynamics, SDMs usually cannot produce any
+estimate of the rate at which local populations will increase or
+decrease in the near-term and instead project a future equilibrium
+species distribution that may or may not ever be reached. Direct
+validation of such predictions is extremely rare (Roberts and Hamann
+2012). Large-scale dynamic population models could overcome these
+limitations. They would produce spatially-explicit estimates of species
+abundance within the species range (Ehrl{é}n and Morris 2015), have the
+potential to model expansion in abundance outside the range when coupled
+with dynamic models of dispersal, and would provide testable predictions
+of how populations should respond to short-term climate perturbations.
+These short-term predictions also would give modelers the opportunity to
+repeatedly validate and refine their models (Luo et al. 2011).
 
-The approach we describe in this paper flows naturally from recent work
-on density-structured population models. Density-structured models rely
-on modeling the transitions of a population among discrete states rather
-than the traditional approach of modeling the transitions of
-individuals. Relatively few ecologists have taken up the
-density-structured mantle first proposed by Taylor, C. M. et al. (2004),
-but recently Freckleton et al. (2011) and Queenborough et al. (2011)
-have higlighted the potential value the approach.
+Sagebrush (\emph{Artemisia} spp.) ecosystems offer an ideal testing
+ground for new spatially explicit population models derived from RS
+data. Sagebrush species are widely distributed (Kuchler 1964), they are
+sensitive to climate (Perfors et al. 2003, Miglia et al. 2005, Poore et
+al. 2009, Dalgleish et al. 2011, Xian et al. 2012, Apodaca 2013,
+Schlaepfer et al. 2014a, 2014b, Harte et al. 2015, Homer et al. 2015),
+new landscape and regional scale time series of sagebrush cover are now
+being produced from aerial imagery (Homer et al. 2012), and forecasts of
+future sagebrush ecosystems are in high demand due to the precarious
+conservation status of the greater sage-grouse (\emph{Centrocercus
+urophasianus}) (Arnett and Riley 2015). SDMs typically predict that much
+of the area occupied by sagebrush ecosystems today will become
+unsuitable for sagebrush due to climate change, resulting in a dramatic
+loss in the extent of sagebrush habitat by the end of this century
+(Shafer et al. 2001, Neilson et al. 2005, Bradley 2010, Schlaepfer et
+al. 2012, Still and Richardson 2015). Ecohydrology models supply a
+possible mechanism for sagebrush losses predicted by SDMs: climate
+warming could lead to earlier snowmelt, increased evaporation and
+ultimately less recharge of deeper soil layers in the spring (Schlaepfer
+et al. 2012, 2014a). In warmer parts of its range, increased temperature
+could be especially detrimental to sagebrush as it depends on water from
+deeper soil to survive and grow in this arid region (Pechanec et al.
+1937, Schlaepfer et al. 2011, Germino and Reinhardt 2014). In contrast,
+at higher elevations and in colder regions, warming and earlier snowmelt
+could lengthen the growing season and increase sagebrush occurrence
+(Schlaepfer et al. 2012, 2014a). Direct observations of individual
+plants and experimental plots tend to agree with these models: growth
+tends to respond negatively to spring and summer temperatures (Miglia et
+al. 2005, Poore et al. 2009, Apodaca 2013) except at higher elevations
+where earlier snowmelt may allow for a longer growing season (Perfors et
+al. 2003, Harte et al. 2015). A large-scale, spatially-explicit
+population model for sagebrush driven by interannual climate variability
+would provide a valuable new tool for assessing how sagebrush could
+respond to climate change in the future.
+
+Building on recent technological advances in spatial statistics (Latimer
+et al. 2009, Conn et al. 2015) and anticipating ever-increasing
+availability of RS data (He et al. 2015), we demonstrate how large-scale
+plant population models could be used to predict population impacts of
+climate change. \textcolor{blue}{As a proof-of-concept, we use} a
+process model motivated by Gompertz density-dependent population growth
+and a remotely-sensed time series of sagebrush cover from Wyoming (Homer
+et al. 2012, 2015). We account for spatial autocorrelation with
+dimension reduction techniques (Latimer et al. 2009, Conn et al. 2015)
+and produce spatially-explicit estimates of sagebrush percent cover.
+\textcolor{blue}{Unlike most SDMs, our approach models the dynamics of plant abundance through time, and thus, is a population model, in the same spirit that models of animal counts through time are population models.}
+The modeling framework we propose can be applied to any
+spatially-explicit time series of plant cover or density, but its
+application to remotely-sensed data products offers the greatest
+potential to combine the information of population models (e.g.,
+population status \textcolor{blue}{and temporal dynamics}) and the
+spatial extent of species distribution models.
 
@@
 
-\subsubsection{Data: Remotely-sensed time
-series}\label{data-remotely-sensed-time-series}
+\subsubsection{Data}\label{data}
 
-To show our pixel-based modeling approach we use a subset of a
-remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub
-percent cover in Wyoming (Homer et al., 2012) (Figure
-{[}fig:Location-of-the{]}). Sagebrush percent cover was estimated using
-a regression tree to relate ground reflectances retrieved by three
+\paragraph{Remotely-sensed time
+series}\label{remotely-sensed-time-series}
+
+To demonstrate our modeling approach, we use a subset of a
+remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy
+cover in Wyoming (Homer et al. 2012). As part of a separate study,
+\textcolor{blue}{Homer et al. 2012 estimated} sagebrush percent cover
+using a regression tree to relate ground reflectances retrieved by three
 sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780
 field observations of sagebrush cover distributed across Wyoming. The
-regression tree model was further validated using another 500 field
+regression tree model was further validated using another 297 field
 observations. For Wyoming sagebrush, the model achieved an \(R^{2}\) =
-0.65. To hind-cast sagebrush cover the regression tree model was applied
-to historical remote sensing images. This resulted in an annual time
-series of sagebrush cover at 30 m resolution from 1984 to 2011. In this
-dataset, values represent the percentage of a 30 \(\times\) 30 m pixel
-covered by sagebrush. For our purposes we use two 5 \(\times\) 5 km
-subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each
-full dataset contains 754,596 observations (27 years \(\times\) 27,948
-pixels).
+0.65 and an out-of-sample RMSE of 5.46\% (Homer et al. 2012). To
+hind-cast sagebrush cover the regression tree model was applied to
+historical remote sensing images
+\textcolor{blue}{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}.
+This resulted in an annual time series of sagebrush cover at 30 meter
+resolution from 1984 to 2011 (Fig. B1). In this
+\textcolor{blue}{remote sensing product}, values represent the
+percentage of a 30 \(\times\) 30 meter pixel covered by sagebrush. In
+our study, we focused on a 5,070 \(\times\) 2,430 meter subset totaling
+13,689 30 \(\times\) 30 meter pixels each year (Fig. 1). Thus, the
+subset of the \textcolor{blue}{remote sensing product} we use contains
+369,603 observations spanning 27 year-to-year transitions (27 years
+\(\times\) 13,689 pixels).
 
-\subsubsection{Data: Climate covariates}\label{data-climate-covariates}
+\paragraph{Climate covariates}\label{climate-covariates}
 
-Our approach is based on modeling how changes in plant populations
-relate to climate variables. We a priori narrowed our focus to climate
-covariates we know are important for sagebrush and that could be easily
-calculated from general circulation model future projections. In all
-vital rate regressions we focus on the demographic processes driving
-changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t},
-and how these processes are mediated by climate. The four climate
-variables in our vital rate regressions are: (1) year \emph{t} fall
-through summer precipitation, (2) year \emph{t}-1 fall through summer
-precipitation, (3) year \emph{t} average spring temperature, and (4)
-year \emph{t}-1 average spring temperature (Table
-{[}tab:Definition-of-climate{]}). Climate data to parameterize vital
-rate regressions (`observed' climate) came from the PRISM climate
-database. Climate data were extracted for the centroid our study area.
+Our approach models interannual changes in plant cover as a function of
+seasonal climate variables. We used daily historic weather data for the
+center of our study site from the NASA Daymet data set (\emph{available
+online})\footnote{\url{http://daymet.ornl.gov/}}. The Daymet weather
+data are interpolated between coarse observation units and capture some
+spatial variation. We relied on weather data for the centroid of our
+study area. We calculated five climate variables from the Daymet data
+for the time period coinciding with our remotely sensed data (1984 to
+2011).
 
-Climate data used to inform model projections of future sagebrush cover
-came from average outputs of all models used in the most recent IPCC
-report, the Coupled Model Intercomparison Project 5 (CMIP5;
-\url{http://cmip-pcmdi.llnl.gov/cmip5/}). We downloaded model estimates
-of historic climate (1950-2000) and projected climate (2050-2100) for
-the centroid of our study area. We wanted to keep the temporal
-variability from observed climate consistent, so our aim was to adjust
-observed climate according to projected changes. For temperature we
-simply subtracted the average historic temperature from the average
-projected temperature. For precipitation we calculated the proportional
-change: we subtracted average projected precipitation from average
-historic precipitation, and then divided by average historic
-precipitation. We did this for three ``Representative Concentration
-Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
-\url{http://tntcat.iiasa.ac.at/RcpDb/}. Basically they correspond to
-stabilization of radiative forcing before 2100, after 2100, and
-increasing greenhouse gas emissions, respectively. We implemented
-projected temperature and precipitation changes in the model by
-perturbing observed annual values by the values in Table \#.
-
-\begin{table}[h]
-\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
-average projections.}
-\begin{tabular}{llllll}
-\hline
- & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
-Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
-\hline
-RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
-RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
-RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
-\hline
-\end{tabular}
-\end{table}
+We narrowed our focus to climate covariates we know are important for
+sagebrush and that could be calculated from general circulation model
+projections. The five climate variables in our population model are: (1)
+cumulative, ``water year'' precipitation for year \emph{t}-2
+(\emph{lagPpt}), (2) year \emph{t}-1 fall through summer precipitation
+(\emph{ppt1}), (3) year \emph{t} fall through summer precipitation
+(\emph{ppt2}), (4) year \emph{t}-1 average spring temperature
+(\emph{TmeanSpr1}), and (5) year \emph{t} average spring temperature
+(\emph{TmeanSpr2}), where \emph{t-1} to \emph{t} is the transition of
+interest. We selected these variables \emph{a priori} based on previous
+studies (see \emph{Introduction}), though not all emerge as important
+predictors in our model.
 
@@
 We use a descriptive model for sagebrush cover that includes additive
-spatial and temporal effects. Since the spatial and temporal effects are
-additive, we are not considering space-time interactions. Our model aims
-to relate observed sagebrush cover, \(C_{s,t}\), at time \emph{t} and
-spatial location \emph{s} to an underlying process \(\mu_{s,t}\) that is
-defined over all spatial locations and observation times. For
-statistical convenience, we consider discrete values of sagebrush
-percent cover as counts, giving us the formulation
+spatial and temporal effects similar to that described by Conn et al.
+(2015). Interannual change in percent cover represents the integrated
+outcome of recruitment, survival, growth, and retrogression (shrinkage)
+of individual plants from year to year. We model observed integer
+percent cover (\emph{y}) in cell \emph{i} at time \emph{t} as
+conditionally Poisson
 
-\begin{align}
-[\textbf{C}] &= \text{Poisson}(\boldsymbol{\lambda}) \\
-\boldsymbol{\lambda} &= \text{exp}(\boldsymbol{\mu}) 
-\end{align}
+\vspace{-4em}\begin{align}
+y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
+\end{align}\vspace{-4em}
 
-which states that counts (integer percent cover) are Poisson distributed
-with a log link function on abundance intensity
-(\(\boldsymbol{\lambda}\)). We assume the underlying process,
-\(\boldsymbol{\mu}\), is a function of climate covariates (\textbf{X}),
-lagged sagebrush abundace (a temporal process), and a purely spatial
-random effect (\(\boldsymbol{\eta}\)):
+where \(\mu_{i,t}\) is the expected percent cover of pixel \emph{i} in
+year \emph{t}
 
-\begin{align}
-\mu_{s,t} = \underbrace{\beta_{0} + \beta_{1}\mu_{s,t-1}}_\text{temporal} + \underbrace{\mathbf{X}_{t}\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{s}}_\text{spatial} + \underbrace{\varepsilon_{s,t}}_\text{error}
-\end{align}
+\vspace{-4em}\begin{align}
+\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal + dens.
+dep} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
+\end{align}\vspace{-3em}
 
-Fitting the spatial random effect (\(\eta\)) is computationally
-demanding when the observations come from 10s of 1,000s locations. The
-dimensionality is just too large. To overcome these computational
-constraints we employed a dimension reduction strategy to reduce the
-number of parameters that need to be estimated to account for spatial
-variation. A common strategy is to express high dimensional effects,
-like \(\boldsymbol{\eta}\), as the product of an expansion matrix,
-\textbf{K}, and a smaller parameter vector, \(\boldsymbol{\alpha}\). We
-can then define the spatial effect as
+Our model of percent cover change includes a density-dependent effect of
+log-transformed cover in the previous year (\(y_{i,t-1}\)), climate
+effects (\(\textbf{x}_t\)), and a spatial random effect (\(\eta\)) for
+each pixel \emph{i}. Climate effects were standardized
+{[}\((x_{i}-\bar{x}) / \sigma(x)\){]} to improve convergence during the
+model fitting stage and to allow for easier prior specification. The
+intercept, \(\beta_{0,t}\), was allowed to vary through time; these
+random year effects recognize that all observations from a particular
+year share the same climate covariates and thus are not independent. We
+used a Poisson likelihood because integer percent cover values in the
+sagebrush data product can be considered a form of count data. We also
+evaluated a negative binomial model, but found little evidence for
+overdispersion beyond what our model was already accomodating via the
+spatial random effects (\(\boldsymbol{\eta}\)). There was no evidence of
+zero-inflation in our data, but see below (\emph{Accomodating zeros})
+for how we handled the small number of zero percent cover observations.
+\textcolor{blue}{We assume that the remotely sensed estimates of percent cover are "true" and free of error.}
+\textcolor{blue}{This need not be the case, and if measurement or sampling error is known then it could be included in our Bayesian model as a "sampling model"}
+(Hobbs and Hooten 2015).
 
-\begin{align}
-\boldsymbol{\eta} = \textbf{K}\boldsymbol{\alpha}.
-\end{align}
+The spatial random effect (\(\boldsymbol{\eta}\)) accounts for spatial
+autocorrelation among pixels that occur near each other in space. Thus,
+\(\boldsymbol{\eta}\) acts as an offset on the intercept
+(\(\beta_{0,t}\)), creating a spatial field that defines how pixels
+differ from the mean, on average, in space (e.g., areas of perennially
+low or high cover, relative to average cover). Fitting the model with a
+spatial random effect (\(\eta\)) is computationally demanding for large
+data sets like ours. The computational demand is due to the required
+calculations of the spatial covariance matrices, which increase as a
+cubic function of the number of locations (Wikle 2010). Key to our
+approach is a dimension reduction strategy that greatly reduces the
+number of parameters needed to be estimated to account for spatial
+variation by reducing the size of the spatial covariance matrices that
+need to be inverted at each MCMC iteration. Fitting models that
+appropriately account for spatial autocorrelation over large spatial
+extents would not be feasible without these modern techniques. Our
+dimension reduction strategy expresses the high dimensional spatial
+random effect, \(\boldsymbol{\eta}\), as the product of an expansion
+matrix, \textbf{K}, and a smaller parameter vector,
+\(\boldsymbol{\alpha}\) (e.g., Hooten et al. 2003, Hooten and Wikle
+2007, Conn et al. 2015). We can then approximate the spatial effect as
+
+\vspace{-4em}\begin{align}
+\boldsymbol{\eta} &\approx \textbf{K}\boldsymbol{\alpha}, \\
+\alpha_{m} &\sim \text{Normal}(0,\sigma_{\eta}^2).
+\end{align}\vspace{-4em}
 
@@
 space, where \emph{S} is the total number of observed locations. Thus,
-we are able to reduce the effective number of parameters from
-\({S} \rightarrow m\).
+we are able to reduce the effective number of parameters from \({S}\) to
+\(m\).
 
-Obviously, the last remaining obstacle is to parameterize \textbf{K}.
-Here we use kernel convolution to interpolate the spatial random effect
-between \emph{m} ``knots'' that are nonrandomly distributed across the
-space of our study area. These means we are modeling spatial random
-effects at the knot level, and we use \textbf{K} to interpolate those
-effects between knots. We use a simple Gaussian kernel density to define
-the distance-decay function around the knots (\textbf{w}), such that the
-entries of \textbf{K} are
+The last remaining obstacle is to parameterize the matrix of basis
+functions, \textbf{K}. We use kernel convolution (Barry and Hoef 1996,
+Higdon 1998) to interpolate the spatial random effect between \emph{m}
+``knots'' that are nonrandomly distributed across the space of our study
+area. This means we are modeling spatial random effects at the knot
+level, and we use \textbf{K} to interpolate those effects between knots.
+We use an exponential kernel density to define the distance-decay
+function around the knots (\textbf{w}), such that the entries of
+\textbf{K} are
 
-\begin{align}
-K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
-w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
-\end{align}
+\vspace{-4em}\begin{align}
+K_{s,m} &= w_{s,m} / \sum_{s=1}^{S} w_{s,m}
+\end{align}\vspace{-4em}
 
-where \(d_{s,m}\) is the Euclidean distance between the centroid of
-sample cell \emph{s} and the location of knot \emph{m}, and \(\sigma\)
-is the kernel bandwidth. Based on residual spatial structure we found
-after fitting a simple model without spatial effects, we set \(\phi\) to
-about one-third of the range of spatial dependence evident in the
-residual variograms -- 250 meters in this case. In the end, the simplest
-way to understand our dimension reduction strategy is to consider
-\(\alpha\)s as weights for each knot.
+where
 
-\subsubsection*{Computing and model
-estimation}\label{computing-and-model-estimation}
-\addcontentsline{toc}{subsubsection}{Computing and model estimation}
+\vspace{-4em}\begin{align}
+w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right)
+\end{align}\vspace{-4em}
+
+and \(d_{s,m}\) is the Euclidean distance between the centroid of sample
+cell \emph{s} and the location of knot \emph{m}, and \(\sigma\) is the
+kernel bandwidth. It is possible, through exhaustive model selection and
+fitting, to determine the optimal form of the kernel and to estimate
+optimal values for \(\sigma\) (Higdon 2002, Hooten and Hobbs 2015).
+However, given the relative size of our dataset and computational
+limitations, we defined kernels around 231 knots (Fig. C2) whose nearest
+neighbor distances are approximately equal to the range of spatial
+dependence in residuals from a simple GLM fit without climate covariates
+and the spatial random effect (\textasciitilde{}500 meters; Appendix C).
+An infinite number of knots would result in an exact representation of
+the spatial process and covariance model. Computationally, using an
+infinite number of knots is not possible, thus the use of dimension
+reduction techniques serves as an approximation, where the accuracy
+increases with the number of knots. Given the tradeoff between knot
+number and computation time, we chose to base our knot number on the
+spatial dependence as described above.
+
+The Bayesian posterior distribution of our spatio-temporal model can be
+expressed as
+
+\vspace{-4em}\begin{align}
+[\boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\alpha}, \sigma_{\eta}^2 | \textbf{y}] &\propto \left( \prod_{t=1}^{T} \prod_{i=1}^{n} [y_{i,t} | \beta_{0,t}, \beta_{1}, \boldsymbol{\gamma},\boldsymbol{\alpha}] [\beta_{0,t}|\bar{\beta_{0}}, \sigma_{\beta_{0}}^2] \right) \times \\
+&\left(\prod_{m=1}^{M}[\alpha_m|\sigma_{\eta}^2]\right)[\bar{\beta_{0}}][\beta_{1}][\boldsymbol{\gamma}][\sigma_{\beta_{0}}^2][\sigma_{\eta}^2]. \nonumber
+\end{align}\vspace{-4em}
+
+\subsubsection{Accomodating zeros}\label{accomodating-zeros}
+
+Our process model (in Eq. 2) includes a log transformation of the
+observations (\(\text{log}(y_{t-1}\))). Thus, our model does not
+accomodate zeros. Fortunately, we had very few instances where pixels
+had 0\% cover at time \emph{t-1} (\emph{N} = 47, which is 0.01\% of the
+data set). Thus, we excluded those pixels from the model fitting
+process. However, when simulating the process, we needed to include
+possible transitions from zero to non-zero percent cover. We fit an
+intercept-only logistic model to estimate the probability of a pixel
+going from zero to non-zero cover
+
+\vspace{-4em}\begin{align}
+y_{i} &\sim \text{Bernoulli}(\mu_{i}) \\
+\text{logit}(\mu_{i}) &= b_0
+\end{align}\vspace{-4em}
+
+where \textbf{y} is a vector of 0s and 1s corresponding to whether a
+pixel was colonized (\textgreater{}0\% cover) or not (remains at 0\%
+cover) and \(\mu_i\) is the expected probability of colonization as a
+function of the mean probability of colonization (\(b_0\)). We fit this
+simple model using the `glm' command in R (R Core Team 2013). For data
+sets in which zeros are more common and the colonization process more
+important, the same spatial statistical approach we used for our cover
+change model could be applied
+\textcolor{blue}{and covariates such as cover of neighboring cells could be included}.
+
+\subsubsection{Fitting the model}\label{fitting-the-model}
+
+We fit the spatiotemporal model in R (R Core Team 2013) using the
+`No-U-Turn' Hamiltonian Monte Carlo sampler in Stan (Stan Development
+Team 2014a) and the RStan package (Stan Development Team 2014b). We
+obtained posterior distributions of all model parmaters from three MCMC
+chains comprised of 1,000 iterations each, after discarding an initial
+1,000 iterations as burn in. Short chains of samples are a hallmark of
+the Stan algorithm, which is extremely efficient. Compared to other
+samplers, fewer iterations are required to achieve convergence. Each
+chain was initialized with unique parameter values and the model was fit
+in parallel using the Utah State University High-Performance Computing
+facility. Model fitting required five days on a four node Central
+Processing Unit with 2 \(\times\) AMD Opteron(tm) Processor 4386 @ 3.10
+Ghz, 64GB of RAM per node, 16 cores per node, and each chain launched in
+parallel on separate cores. We assessed convergence visually and
+calculated scale-reduction factors (Appendix D, \(\hat{R}\) \textless{}
+1.1 for all parameters) (Gelman and Rubin 1992, Gelman and Hill 2009).
+
+\subsubsection{Simulating the process}\label{simulating-the-process}
+
+We performed four sets of simulations to (1) compare observed and
+simulated equilibrium cover, (2) compare observed and simulated year-
+and location-specific cover, (3) forecast future equilibrium population
+states under projected climate change, and (4) make temporally-explicit
+forecasts of sagebrush cover starting the final year of our observations
+and ending in year 2098. Using the posterior distribution of model
+parameters, we simulated a matrix of pixels equal to the size of the
+study area (13,689 pixels or matrix elements). For simulations (1) and
+(3) we initialized all pixels with arbitrarily low cover (1\%) and then
+projected the model forward by randomly drawing climate covariates from
+the observed climate time series (for 1) or a perturbed climate time
+series (for 3). We ran equilibrium simulations (1 and 3) for 2,000 time
+steps and then compared the output across simulations, after discarding
+an initial 100 time steps. To calculate average future equilibrium
+sagebrush cover, we ran simulation (3) for each GCM and RCP scenario
+separately, and then averaged the results over GCMs. For simulation (2),
+we initialized each pixel with its actual percent cover value for time
+\emph{t} and cell \emph{s} and projected the model forward one time step
+and compared the one-step ahead forecast with the observed value. For
+simulation (4), we initialized each pixel with the final observed value
+in 2011 and then projected the model forward based on GCM yearly weather
+projections. We ran these simulations for each GCM and RCP scenario
+combination separately and then aggregated the results over the GMCs by
+calculating the mean and the 90th percentiles for each RCP scenario.
+
+We used the posterior mean of each parameter for all simulations except
+for (4) where we ran 50 simulations with unique sets of parameters from
+the chains. Random year effects were included in simulations by randomly
+drawing a posterior mean year effect (\(\beta_{0,t}\)) for each
+iteration (simulations 1 and 3), using the posterior mean year effect
+for a specific year (simulation 2), or by a drawing a future-year random
+effect from the posterior mean and standard deviation of the mean
+intercept (simulation 4, e.g.,
+\(\beta_{0,T} \sim \text{normal}(\bar{\beta_{0}}, \sigma_{\beta_{0}}^2)\)
+for some future year \emph{T}). Our simulation approach provides a
+reasonable and computationally efficient approximation to the true
+posterior predictive mean when used in these scenarios with our data.
+
+We required future projections of climate for our study area to conduct
+the equilibrium and temporally-explicit forecasts described above. Thus,
+we used the most recent climate projections from the Intergovernmental
+Panel on Climate Change (IPCC), the Coupled Model Intercomparison
+Project 5 (CMIP5; \emph{available online})\footnote{\url{http://cmip-pcmdi.llnl.gov/cmip5/}}.
+The CMIP5 provides projections from a suite of global circulations
+models (GCMs); we used projections from 18 GCMs (Table A1) that produced
+weather projections for three ``Representative Concentration Pathways'':
+RCP 4.5, RCP 6.0, and RCP 8.5 (\emph{described online})\footnote{\url{http://tntcat.iiasa.ac.at/RcpDb/}}.
+The three RCPs correspond to stabilization of radiative forcing before
+2100, after 2100, and ongoing increase in greenhouse gas emissions,
+respectively.
+
+To simulate equilibrium sagebrush cover under projected future climate
+we applied average projected changes in precipitation and temperature to
+the observed climate time series. For each GCM and RCP scenario
+combination, we calculated average precipitation and temperature over
+the 1950-2000 time period and the 2050-2098 time period. We then
+calculated the absolute change in temperature between the two time
+periods (\(\Delta T\)) and the proportional change in precipitation
+between the two time periods (\(\Delta P\)) for each GCM and RCP
+scenario combination. Lastly, we applied \(\Delta T\) and \(\Delta P\)
+to the observed 28-year climate time series to generate a future climate
+time series for each GCM and RCP scenario combination. These generated
+climate time series were used to simulate equilibrium sagebrush cover.
+We simulated equilibrium cover separately for each GCM and RCP scenario
+combination before averaging the results, but we show the average
+projected climate changes across all models in Table 1.
+
+For the temporally-explicit forecasts we used yearly GCM projections
+from 2012 to 2098 to simulate the process starting from the end point of
+the remotely sensed sagebrush cover data (ends in 2011). We aggregated
+daily GCM output for each GCM and RCP scenario into the seasonal climate
+covariates used to fit our model. These yearly climate time series were
+not aggregated further because we ran simulations for each GCM and RCP
+scenario, rather than one simulation per RCP scenario averaged over
+GCMs.
+\textcolor{blue}{The key assumption of our forecasting approach is that the historical correlations between weather and sagebrush cover change will continue to hold in the future.}
+
+\subsection{Results}\label{results}
+
+Averaging across all GCMs, precipitation and temperature in our study
+area are projected to increase; the magnitude of increase depends on the
+RCP scenario (Table 1). Trajectories of our climate covariates from GCM
+projections show similar trends (Fig. 2).
+
+All parameters in our model converged on stable posterior distributions
+(Appendix D). Only the \emph{lagPpt} climate covariate can be considered
+important based on a 90\% credible interval, and it had a positive
+effect on sagebrush percent cover change (Fig. 3). In other words, if
+the year 2000 water year was wetter than average, sagebrush cover would
+increase from the 2001 to the 2002 growing season. Other climate effects
+strongly overlapped zero but their posterior means were positive, except
+for fall-through-spring precipitation the first year of a cover
+transition (\emph{t}-1), whose posterior mean was negative (Fig. 3). The
+posterior mean for the spatial random effect, \(\boldsymbol{\eta}\),
+captured the overall spatial structure of the observed data (Fig. E1).
+This indicates our choice of knot placement and dimension reduction
+strategy was adequate for describing permanent spatial variation in the
+data.
+
+When we simulated the pixel-based population model based on observed
+climate, it was able to reproduce the spatial pattern of observed
+percent cover, averaged over time (Fig. 4A,B). Our model shows a
+tendency to underpredict perennially-low percent cover pixels (Fig. 4C),
+but does a better job at predicting high cover pixels. Point predictions
+are most confident, though slightly biased, in low percent cover pixels
+(Fig. 4D). The model is also able to adequately reproduce observed
+dynamics when we make one-step-ahead predictions based on observed
+climate and cover in the previous year for each pixel. When we made
+these in-sample, one-step-ahead forecasts, the model achieved an RMSE =
+4.31, in units of percent cover. The Pearson's correlation between
+observations and predictions was 0.62.
+
+When we apply the fitted model to IPCC climate change scenarios, the
+model predicts gains in sagebrush percent cover, on average (Figs. 5,
+6A). The spatial effect remains strong enough in low cover regions to
+counteract the positive effect of projected precipitation increases
+(Fig. 5). Thus, our model predicts an increase in the heterogeneity of
+sagebrush cover because projected cover increases are smaller in low
+cover pixels than in high cover pixels (Fig. 5 and Fig. F1). For the
+temporally-explicit forecasts, we show spatially-averaged values and the
+associated uncertainty due to variability in GCM projections,
+variability in model parameters, and uncertainty in our process model
+(Fig. 6A). Based on our model and GCM projections, we forecast an
+average increase in sagebrush cover at our study area, but a decrease is
+not outside the realm of possibility (shaded regions in Fig. 6A). The
+generally increasing trend reflects the positive effect of precipitation
+on sagebrush cover change estimated for our study area (Fig. 3). We also
+show how our model is capable of near-term forecasts in Fig 6B.
+
+\subsection{Discussion}\label{discussion}
+
+Despite the need to forecast population responses to climate change over
+large spatial extents, as demonstrated by the wide application of
+species distribution models (e.g., Clark et al. 2014), landscape-scale
+population models for plant species remain more concept than reality
+(Schurr et al. 2012, Merow et al. 2014). We \textcolor{blue}{introduced}
+a new approach
+\textcolor{blue}{that uses methods from the dynamic spatio-temporal modeling literature}
+(e.g., Conn et al. 2015)
+\textcolor{blue}{to fit a population model to remotely-sensed estimated of plant percent cover}.
+\textcolor{blue}{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming}
+(Homer et al. 2012). We first discuss our results specific to sagebrush
+ecology and response to climate, and then discuss the more general
+implications and limitations of our proposed approach.
+
+\subsubsection{Sagebrush response to climate and climate
+change}\label{sagebrush-response-to-climate-and-climate-change}
+
+The climate effects we estimated, based on cover data at 30 meter
+spatial resolution, are consistent with individual-level responses of
+sagebrush to climate-related variables. Research on individual plants
+has shown that wetter winters are correlated with greater stem growth in
+sagebrush (Poore et al. 2009, Apodaca 2013) and that warmer spring
+temperatures may enhance sagebrush growth in cold climates by advancing
+the date of snowmelt and increasing the length of the growing season
+(Perfors et al. 2003, Harte et al. 2015). In agreement with those
+individual-level responses, posterior means for all precipitation and
+temperature effects in our model were positive, except for the effect of
+fall-through-spring precipitation in the first year of a cover
+transition (\emph{ppt1}, Fig. 3). The cumulative amount of precipitation
+the year before a cover transition (\emph{pptLag} in our model) emerged
+as the strongest predictor of sagebrush cover change (Fig. 3). However,
+mean estimates for the climate effects are relatively weak (Fig. 3).
+
+Such small effects could indicate that sagebrush are not very sensitive
+to interannual climate variability, that our model is poorly specified,
+or that climate responses are difficult to detect using coarse-scale
+data. Given findings from previous research demonstrating the importance
+of precipitation and temperature to sagebrush growth (Pechanec et al.
+1937, Schlaepfer et al. 2011, Germino and Reinhardt 2014) and
+regeneration (Schlaepfer et al. 2014b), it is unlikely that sagebrush
+are insensitive to climate. We used aggregated climate covariates that
+may not completely capture the climate-dependence of sagebrush cover
+change. However, the covariates we chose closely match the
+climate-related variables that have been shown to drive sagebrush
+growth, survival, and regeneration (e.g., Dalgleish et al. 2011,
+Schlaepfer et al. 2014b). More likely, aggregated estimates of plant
+abundance, such as percent cover, mask interannual variability at the
+level of the individual plant and makes it more difficult to detect the
+drivers of internanual variability. Additionally, we chose not to
+downscale the Daymet weather data, meaning that in a given year all
+pixels shared the same climate, which limits our statistical power.
+Nonetheless, our model was capabable of detecting climate effects that
+agree with our knowledge of sagebrush ecology and allowed us to make
+forecasts of future sagebrush abundance.
+
+Under projected climate, we forecast modest increases in sagebrush cover
+for all RCP scenarios in the long-term (Figs. 5,6A). Our forecasts
+reflect both the estimated effect size for each climate covariate and
+the amount of change in those covariates projected by the GCMs.
+Cumulative precipitation the year before a given year-to-year transition
+was the strongest standardized effect (Fig. 3), but precipitation is
+projected to increase only moderately (Table 1, Fig. 2) and the negative
+effect of fall-through-spring precipitation in the first year of a cover
+transition (\emph{ppt1}) had an offsetting effect. In contrast, mean
+spring temperature had a weak positive effect on sagebrush cover
+changes, but the projected temperature increase is large (Table 1, Fig.
+2).
+
+An interesting consequence of explicitly modeling the effect of space
+(through \(\boldsymbol{\eta}\)) is the forecasted increase in spatial
+heterogeity (Fig. F1). Our model projects little change in low cover
+pixels but substantial increases in the cover of high cover pixels (Fig.
+5). Had we not explicitly accounted for spatial-dependence in our model,
+we would have missed this result. We were unable to attribute the
+spatial structure apparent in the data (Fig. 4A) and approximated by our
+model (\(\boldsymbol{\eta}\), Fig. E1) to slope, aspect, elevation, or
+coarse soil type (results not shown). The lack of correlation between
+\(\boldsymbol{\eta}\) and landscape factors leads us to conclude that
+the spatial structure in our data set emerges from some combination of
+fine-scale microhabitat associations and legacy effects
+\textcolor{blue}{of disturbance}.
+
+While we forecast an increase in sagebrush cover at our study area, SDM
+studies typically project dramatic declines in climate suitability for
+sagebrush with warming (Shafer et al. 2001, Neilson et al. 2005, Bradley
+2010, Schlaepfer et al. 2012, Still and Richardson 2015). There are many
+potential explanations for this apparent contrast, ranging from the type
+of model used to the particular climate covariates considered, but the
+location of our study area in a cold portion of sagebrush's geographic
+distribution may be the best. The response of plant species to weather
+varies along climatic gradients (e.g., Clark et al. 2011, Vanderwel et
+al. 2013), and sagebrush are especially sensitive to the timing of
+snowmelt because their growth depends on recharge of deep soil water
+(Schlaepfer et al. 2012, 2014a). In warmer parts of the sagebrush range,
+earlier snowmelt is detrimental to growth and survival (Pechanec et al.
+1937, Schlaepfer et al. 2011, Germino and Reinhardt 2014). In colder
+regions, earlier snowmelt due to temperature increases can lengthen the
+growing season and increase sagebrush occurence and cover (Schlaepfer et
+al. 2012, 2014a). The average annual temperature across the sagebrush
+steppe biome is 6.9\textdegree C (sd = 1.6; Schlaepfer et al. 2011),
+whereas average temperature at our study area from 1980 to 2013 was
+4.6\textdegree C (calculated from Daymet estimates). Our study area lies
+at the cold extreme of the sagebrush range, thus the weak positive
+response to temperature that we estimated (Fig. 3) and carried through
+to our forecasts (Figs. 5,6A) likely represents the positive effect of
+earlier snowmelt, and thus higher moisture availability early in the
+growing season.
+
+A previous analysis of a different subset of the remote sensing data set
+we used also came to a different conclusion, projecting future sagebrush
+decline (Homer et al. 2015). The discrepancy between the results of
+Homer et al. (2015) and ours primarily reflects a difference in the
+climate projections used for projecting future changes rather than
+differences in our inference about responses to historical variation in
+weather. Homer et al. (2015) used downscaled weather projections from a
+single model from the IPCC 4 whereas we used native-resolution weather
+projections from a suite of models from the IPCC 5. Consistent with our
+study, Homer et al. (2015) found a generally positive relationship
+between pixel-level sagebrush cover and precipitation, but the future
+climate scenario they chose resulted in a mean decrease in
+precipitation, causing a predicted decline in sagebrush cover. A second
+difference is that Homer et al. (2015) relied on regressions of decadal
+trends in sagebrush cover against decadal trends in climate at the level
+of individual pixels. Our current approach is fundamentally different in
+that we specifically model the impact of interannual variation in
+weather on year-to-year changes in sagebrush cover using a dynamic
+population model. Thus, our model takes advantage of the additional
+information contained within short-term responses to climate
+fluctuations. Lastly, the location of Homer et al.'s (2015) study area
+is, on average, at a lower elevation than our current study area. The
+geographic difference results in different historical and projected
+climate, and, as discussed above, sagebrush may respond differently to
+warming depending on geographic location.
+
+We projected sagebrush cover to the end of this century, but an
+important feature of our approach is that it can also produce short-term
+forecasts (Fig. 6B). For example, we could forecast the effects of a
+multi-year regional drought on sagebrush cover (Debinski et al. 2010).
+Validating spatial population models against short-term predictions
+would give ecological forecasters a way to assess and improve the
+performance of their models, which would greatly increase our confidence
+in long-term forecasts. This cycle of prediction, validation, and
+refinement is missing from most currently available population-level
+forecasts of the effects of climate change.
+
+\subsubsection{A landscape-scale plant population modeling approach:
+opportunities and
+limitations}\label{a-landscape-scale-plant-population-modeling-approach-opportunities-and-limitations}
+
+Our approach for modeling plant populations overcomes two major hurdles
+for spatially-explicit population models. First, we used moderate
+resolution, remotely-sensed estimates of sagebrush percent cover as a
+response variable, enabling us to fit a dynamic population model over a
+large spatial extent.
+\textcolor{blue}{Species-specific estimates of plant abundance are becoming commonplace as remote sensing technology develops}
+(e.g., Baldeck and Asner 2014, Colgan and Asner 2014),
+\textcolor{blue}{and in a few years several remotely-sensed time series may be available.}
+Second, borrowing from new methods in spatio-temporal modeling of animal
+abundance (e.g., Conn et al. 2015), we fit the model using a dimension
+reduction strategy that accounted for spatial autocorrelation within a
+feasible computational time. Accounting for spatial autocorrelation
+allows for statistically rigorous inference on the effects of
+interannual climate on sagebrush cover change in our study region. The
+spatial covariance structure also provided a way to obtain
+spatially-explicit predictions at a resolution below that of the climate
+covariates (i.e., within the study region; Figs. 4,5). Our approach is
+amenable to any spatially-explicit time series of plant abundance, but
+we see remote-sensing datasets offering the largest opportunity for
+landscape-scale population models. Furthermore, it would be
+straighforward to include additional covariates related to disturbance
+(e.g., fire) or biotic interactions. Thus, we see our method as a first
+step toward coupling the mechanistic power of dynamic population models
+with the spatial extent of SDMs. The spatially- and temporally-explicit
+forecasts made possible by our approach should be especially relevant to
+land management decisions based on near-term forecasts.
+
+Several \emph{a priori} modeling decisions determined the spatial extent
+and resolution of our results. We retained the native spatial resolution
+of the remote sensing data (\(30\times30\) meters). This constrained the
+extent that we could reasonably model because of the computational
+challenges in estimating spatial random effects. Even with our dimension
+reduction technique, modeling a larger area at this resolution would
+require a greater number of spatial knots, and computation time would
+increase substantially (Wikle 2010). To model a larger spatial extent,
+we could aggregate the original remote-sensing time series data to a
+coarser spatial resoluation. This would allow us to model a much greater
+spatial extent with a similar number of knots and a similar computation
+time. While a coarser scale model would lose some fine-scale detail, it
+could be applied to a much larger area, potentially gaining some
+strength in estimating climate effects by spanning a greater range of
+climate variation. However, gains made by incorporating greater regional
+variability by modeling at a coarser resolution could be offset by the
+loss of information inherent when aggregating plant responses into
+larger pixels.
+
+Our spatial extent and resolution also affected our use of climate
+covariates. We did not downscale Daymet data to match the spatial
+resolution of the sagebrush data, meaning that in each year all pixels
+share the same climate covariates. This is a potential limitation of our
+study, and could explain the weak effect of climate covariates that we
+observed (Fig. 3). We also did not allow different portions of our study
+area to respond to climate in different ways. Doing so would require
+spatially-varying climate effects and a substantial increase in
+computational time. However, in future applications, it will be
+important to allow climate effects to vary over space to better capture
+reality. Conn et al. (2015) provide examples of how such spatiotemporal
+interactions can be included in abundance models. We might expect
+climate effects to interact with spatial covariates such as soil type,
+slope, and aspect. In our relative small study area, we did not observe
+important effects of these factors, but it is possible to include such
+abiotic data layers as predictors when fitting models at larger spatial
+extents where variability may be greater.
+
+\textcolor{blue}{The uncertainty associated with our forecasts highlights several opportunities to improve our approach.
+First, parameter uncertainty could be reduced by regulating the variance of the posterior distributions of climate covariates via ridge regression}
+(e.g., Gerber et al. 2015).
+\textcolor{blue}{Second, uncertainty associated with climate projections could be reduced by identifying GCMs that perform exceptionally well for a particular study location}
+(e.g., Rupp et al. 2013).
+\textcolor{blue}{Such considerations will be important when forecasting in support of particular management objectives.}
+\textcolor{blue}{However, knowledge of uncertainty is itself important knowledge for management}
+(Bradshaw and Borchers 2000).
+\textcolor{blue}{Deciding that no actions should be taken based on the data at hand is itself a management decision.}
+
+\subsection{Conclusion}\label{conclusion}
+
+We introduced a new approach to fitting and simulating population models
+at large spatial extents with plant population data derived from state
+of the art remote sensing. We used the model to forecast future
+abundances of sagebrush in Wyoming and found that at our relatively cold
+site sagebrush should be expected to increase in cover. As more
+species-level remote sensing datasets become available and computing
+power increases this approach will be applicable to a wider number of
+species and even larger spatial extents. Future modeling could include
+the effects of non-climate drivers -- including the effects of species
+interactions and disturbance. For sagebrush, including fire and
+competition with non-native annual grasses in the model may be
+especially important for a complete assesment of the effects of climate
+change (Bradford and Lauenroth 2006). Fortunately, our spatio-temporal
+modeling framework could easily be extended to model additional species
+and dynamic processes as the data become availabe. The approach we have
+developed here fills an important gap in spatial scales between species
+distribution models and local-scale demographic population models.
+
+\subsection{Acknowledgments}\label{acknowledgments}
+
+This work is the outcome of a distributed graduate seminar led by PBA
+and supported by a National Science Foundation CAREER award
+(DEB-1054040). David T. Iles, Eric LaMalfa, and Rebecca Mann
+participated in project conception as part of the distributed graduate
+seminar and provided comments that improved the manuscript. ATT was
+supported by an NSF Postdoctoral Research Fellowship in Biology
+(DBI-1400370) and AK was supported by an NSF Graduate Research
+Fellowship. Additional support came from the Utah Agricultural
+Experiment Station, Utah State University, and this article is approved
+as journal paper number 8856. We are grateful to Debra K. Meyer at USGS
+EROS for extracting the data set used in this paper and to David Koons
+\textcolor{blue}{and two anonymous reviewers} for comments that improved
+the manuscript. Compute, storage, and other resources from the Division
+of Research Computing in the Office of Research and Graduate Studies at
+Utah State University are gratefully acknowledged. We acknowledge the
+World Climate Research Programme's Working Group on Coupled Modelling,
+which is responsible for CMIP, and we thank the climate modeling groups
+(listed in Table A1) for producing and making available their model
+output. For CMIP the U.S. Department of Energy's Program for Climate
+Model Diagnosis and Intercomparison provides coordinating support and
+led development of software infrastructure in partnership with the
+Global Organization for Earth System Science Portals. Any use of trade,
+firm, or product names is for descriptive purposes only and does not
+imply endorsement by the U.S. government.
+
+\pagebreak{}
+
+\subsection{Tables}\label{tables}
+
+\begin{table}[h]
+\protect\caption{\label{tab:Changes-in-climate}Projected changes in temperature and precipitation at our study area from CMIP5 average GCM projections for 2050-2100 relative to average temperature and precipitation from 1950-2000.}
+\begin{tabular}{llllll}
+\hline
+Emissions Scenario & Absolute change in temperature & Percentage change in precipitation\tabularnewline
+\hline
+RCP 4.5 & 2.98\textdegree & 8.94\%\tabularnewline
+RCP 6.0 & 3.13\textdegree & 8.64\%\tabularnewline
+RCP 8.5 & 4.79\textdegree & 11.0\%\tabularnewline
+\hline
+\end{tabular}
+\end{table}
+
+\pagebreak{}
+
+\subsection{Figures}\label{figures}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{../figures/studyarea_map.png}
+  \caption{Location of the 5,070 $\times$ 2,430 meter kilometer study area in southwestern Wyoming (black rectangle) and a snapshot of the percent cover data in 1984 (detailed inset).
+Scale bar is relevant for US map only; refer to axes labels on the detailed inset of sagebrush percent cover for scale of the study area.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{../figures/weather_projections.png}
+  \caption{GCM yearly weather hindcasts (before solid line at 2011) and projections (after solid line at 2011) for precipitation (A) and temperature (B) at our study area in southwestern Wyoming (see Fig.
+1).}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{../figures/post_climate_covariates.png}
+  \caption{Posterior distributions of climate covariates.
+The x-axis is the standardized coefficient value because we fit the statistical model for sagebrush cover change (Eq.
+7) using standardized covariate values.
+Only cumulative precipitation at time \emph{t-2} (pptLag) is important (shown in blue; 90\% CI does not overlap zero).
+Climate covariate codes: pptLag = water year precipitation in year \emph{t-2}; TmeanSpr1 = year \emph{t-1} average spring temperature; ppt2 = year \emph{t} fall through summer precipitation; TmeanSpr2 = year \emph{t} average spring temperature; ppt1 = year \emph{t-1} fall through summer precipitation.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{../figures/obs_pred_spatial_subset.png}
+  \caption{Observed and predicted (A, B) equilibrium percent cover of sagebrush, and prediction bias and precision (C, D) for the extent of our spatial area at 30-m resolution.
+Observed equilibrium sagebrush cover (A) is the temporal mean of each pixel from the 28 year time series.
+Prediction results are from simulations that use posteior mean parameter values.
+Precision in (D) represents the variability of each pixel over the course of the 2,000 iteration simulation.
+Axes definitions: Lat = latitude; Lon = longitude.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{../figures/clim_change_mean_spatial.png}
+  \caption{Projected equilibrium cover under three IPCC climate change scenarios (RCP = Representative Concentration Pathways) for our study area in southwestern Wyoming.
+The top panel shows equilibrium cover based on simulations using observed climate.
+Subsequent panels show equilibrium cover based on perturbed climate for each RCP scenario.
+Forecasts are based on the projected climate changes in Table 1 applied to the observed climate time series used to fit the statistical model.
+We used posterior mean parameter estimates for all simulations.
+Color bar indicates percent cover of sagebrush in each 30x30 meter pixel.
+Axes definitions: Lat = latitude; Lon = longitude.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=3in]{../figures/figure6.png}
+  \caption{Observed (black line before 2011) and forecasted (colored lines after 2011) sagebrush percent cover.
+Long-term forecasts (A) were made for three IPCC emissions scenarios (RCPs 4.5, 6.0, and 8.5) and are for the period of 2012 to 2098.
+Shaded regions show limits of the 5th and 95th quantiles for simulations conducted using 50 different sets of parameters from the MCMC output.
+Lines show mean trajectories.
+Uncertainty in forecasts arises from uncertainty in GCM projections, uncerainty around the ecological process, and uncertainty around parameter estimates.
+Before calculating the mean and quantiles for each year across parameter sets and GCMs, we averaged percent cover over the 13,689 pixels.
+\textcolor{blue}{Panel (B) shows an example short-term forecast (10 years) using the MIROC5 GCM projections under RCP 8.5.
+Each line shows a forecast from one parameter set.}}
+\end{figure}
+
+\newpage{}
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
+community impacts of climate variability and change: when do competitive
+interactions matter? Journal of Ecology 100:478--487.
+
+Apodaca, L. F. 2013. Assessing Growth Response to Climate Controls in a
+Great Basin Artemisia Tridentata Plant Community. PhD thesis, University
+of Nevada Las Vegas.
+
+Arnett, E. B., and T. Z. Riley. 2015. Science, policy, and the fate of
+the greater sage-grouse. Frontiers in Ecology and the Environment
+13:235.
+
+Baldeck, C. A., and G. P. Asner. 2014. Improving remote species
+identification through efficient training data collection. Remote
+Sensing 6:2682--2698.
+
+Barry, R. P., and J. M. V. Hoef. 1996. Blackbox Kriging: Spatial
+Prediction without Specifying Variogram Models.
+
+Bradford, J. B., and W. K. Lauenroth. 2006. Controls over invasion of
+Bromus tectorum: The importance of climate, soil, disturbance, and seed
+availability. Journal of Vegetation Science 17:693--704.
+
+Bradley, B. A. 2010. Assessing ecosystem threats from global and
+regional change: hierarchical modeling of risk to sagebrush ecosystems
+from climate change, land use and invasive species in Nevada, USA.
+Ecography 33:198--208.
+
+Bradshaw, G. A., and J. G. Borchers. 2000. Uncertainty as information:
+Narrowing the science-policy gap. Ecology and Society 4.
+
+Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011. Climate
+change vulnerability of forest biodiversity: Climate and competition
+tracking of demographic rates. Global Change Biology 17:1834--1849.
+
+Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
+Foley, D. M. Lodge, M. Pascual, R. Pielke, W. Pizer, C. Pringle, W. V.
+Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
+2001. Ecological forecasts: an emerging imperative. Science (New York,
+N.Y.) 293:657--660.
+
+Clark, J. S., A. E. Gelfand, C. W. Woodall, and K. Zhu. 2014. More than
+the sum of the parts: Forest climate response from joint species
+distribution models. Ecological Applications 24:990--999.
+
+Colgan, M. S., and G. P. Asner. 2014. Coexistence and environmental
+filtering of species-specific biomass in an African savanna. Ecology
+95:1579--1590.
+
+Conn, P. B., D. S. Johnson, J. M. V. Hoef, M. B. Hooten, J. M. London,
+and P. L. Boveng. 2015. Using spatiotemporal statistical models to
+estimate animal abundance and infer ecological dynamics from survey
+counts. Ecological Monographs 85:235--252.
+
+Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
+Adler. 2011. Climate influences the demography of three dominant
+sagebrush steppe plants. Ecology 92:75--85.
+
+Debinski, D. M., H. Wickham, K. Kindscher, J. C. Caruthers, and M.
+Germino. 2010. Montane meadow change during drought varies with
+background hydrologic regime and plant functional group. Ecology
+91:1672--81.
+
+Ehrl{é}n, J., and W. F. Morris. 2015. Predicting changes in the
+distribution and abundance of species under environmental change.
+Ecology Letters 18:303--314.
+
+Elith, J., and J. R. Leathwick. 2009. Species Distribution Models:
+Ecological Explanation and Prediction Across Space and Time.
+
+Ellner, S. P., and M. Rees. 2006. Integral projection models for species
+with complex demography. The American naturalist 167:410--428.
+
+Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
+Queenborough. 2011. Density-structured models for plant population
+dynamics. American Naturalist 177:1--17.
+
+Gelman, A., and J. Hill. 2009. Data analysis using regression and
+multilevel/hierarchical models. Cambridge University Press, Cambridge.
+
+Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation
+Using Multiple Sequences. Statistical Science 7:457--472.
+
+Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
+Drewien. 2015. Optimal population prediction of sandhill crane
+recruitment based on climate-mediated habitat limitations. Journal of
+Animal Ecology 84:1299--1310.
+
+Germino, M. J., and K. Reinhardt. 2014. Desert shrub responses to
+experimental modification of precipitation seasonality and soil depth:
+Relationship to the two-layer hypothesis and ecohydrological niche.
+Journal of Ecology 102:989--997.
+
+Hare, J. a, M. a Alexander, M. J. Fogarty, E. H. Williams, and J. D.
+Scott. 2010. Forecasting the dynamics of a coastal fishery species using
+a coupled climate--population model. Ecological applications : a
+publication of the Ecological Society of America 20:452--464.
+
+Harte, J., S. R. Saleska, and C. Levy. 2015. Convergent ecosystem
+responses to 23-year ambient and manipulated warming link advancing
+snowmelt and shrub encroachment to transient and long-term climate-soil
+carbon feedback. Global change biology 21:2349--56.
+
+He, K. S., B. A. Bradley, A. F. Cord, D. Rocchini, M.-N. Tuanmu, S.
+Schmidtlein, W. Turner, M. Wegmann, and N. Pettorelli. 2015. Will remote
+sensing shape the next generation of species distribution models? Remote
+Sensing in Ecology and Conservation 1:4--18.
+
+Higdon, D. 1998. A process-convolution approach to modelling
+temperatures in the North Atlantic Ocean. Environmental and Ecological
+Statistics 5:173--190.
+
+Higdon, D. M. 2002. Space and space-time modeling using process
+convolutions. Pages 37--56 \emph{in} C. Anderson, V. Barnett, P.
+Chatwin, and A. El-Shaarawi, editors. Quantitative methods for current
+environmental issues. Springer, London.
+
+Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
+Primer for Ecologists. Princeton University Press, Princeton.
+
+Homer, C. G., C. L. Aldridge, D. K. Meyer, and S. J. Schell. 2012.
+Multi-scale remote sensing sagebrush characterization with regression
+trees over Wyoming, USA: Laying a foundation for monitoring.
+International Journal of Applied Earth Observation and Geoinformation
+14:233--244.
+
+Homer, C. G., G. Xian, C. L. Aldridge, D. K. Meyer, T. R. Loveland, and
+M. S. O'Donnell. 2015. Forecasting sagebrush ecosystem components and
+greater sage-grouse habitat for 2050: Learning from past climate
+patterns and Landsat imagery to predict the future. Ecological
+Indicators 55:131--145.
+
+Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
+selection for ecologists. Ecological Monographs 85:3--28.
+
+Hooten, M. B., and C. K. Wikle. 2007. Shifts in the spatio-temporal
+growth dynamics of shortleaf pine. Environmental and Ecological
+Statistics 14:207--227.
+
+Hooten, M. B., D. R. Larsen, and C. K. Wikle. 2003. Predicting the
+spatial distribution of ground flora on large domains using a
+hierarchical Bayesian model. Landscape Ecology 18:487--502.
+
+Kuchler, A. 1964. Potential Natural Vegetation of the Conterminous
+United States. American Geographical Society, Special Publication No.
+36.
+
+Latimer, A. M., S. Banerjee, H. Sang, E. S. Mosher, and J. A. Silander.
+2009. Hierarchical models facilitate spatial analysis of large data
+sets: A case study on invasive plant species in the northeastern United
+States. Ecology Letters 12:144--154.
+
+Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
+D. S. Schimel. 2011. Ecological forecasting and data assimilation in a
+data-rich era. Ecological Applications 21:1429--1442.
+
+Maiorano, L., R. Cheddadi, N. E. Zimmermann, L. Pellissier, B.
+Petitpierre, J. Pottier, H. Laborde, B. I. Hurdu, P. B. Pearman, A.
+Psomas, J. S. Singarayer, O. Broennimann, P. Vittoz, A. Dubuis, M. E.
+Edwards, H. A. Binney, and A. Guisan. 2013. Building the niche through
+time: using 13,000 years of data to predict the effects of climate
+change on three tree species in Europe. Global Ecology and Biogeography
+22:302--317.
+
+Merow, C., A. M. Latimer, A. M. Wilson, S. M. McMahon, A. G. Rebelo, and
+J. A. Silander. 2014. On using integral projection models to generate
+demographically driven predictions of species' distributions:
+development and validation using sparse data. Ecography 37:1167--1183.
+
+Miglia, K., E. Mcarthur, W. Moore, H. Wang, J. Graham, and D. Freeman.
+2005. Nine-year reciprocal transplant experiment in the gardens of the
+basin and mountain big sagebrush (Artemisia tridentata: Asteraceae)
+hybrid zone of Salt Creek Canyon: the importance of multiple-year
+tracking of fitness Title. Biological Journal of the Linnean
+Society:213--225.
+
+Neilson, R., J. Lenihan, D. Bachelet, and R. Drapek. 2005. Climate
+change implications for sagebrush ecosystems. Page 145 \emph{in} North
+american wildlife and natural resources conference.
+
+Pechanec, J., G. Pickford, and G. Stewart. 1937. Effects of the 1934
+Drought on Native Vegetation of the Upper Snake River Plans, Idaho.
+Ecology:490--505.
+
+Perfors, T., J. Harte, and S. E. Alter. 2003. Enhanced growth of
+sagebrush (Artemisia tridentata) in response to manipulated ecosystem
+warming. Global Change Biology 9:736--742.
+
+Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
+Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
+Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
+Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
+horizon, and examples of its uses and determinants. Ecology Letters
+18:597--611.
+
+Poore, R. E., C. A. Lamanna, J. J. Ebersole, and B. J. Enquist. 2009.
+Controls on Radial Growth of Mountain Big Sagebrush and Implications for
+Climate Change. Western North American Naturalist 69:556--562.
+
+Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
+and R. P. Freckleton. 2011. From meso- to macroscale population
+dynamics: A new density-structured approach. Methods in Ecology and
+Evolution 2:289--302.
+
+R Core Team. 2013. R: A language and environment for statistical
+computing.
+
+Rees, M., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
+Roberts, D. R., and A. Hamann. 2012. Predicting potential climate change
+impacts with bioclimate envelope models: A palaeoecological perspective.
+Global Ecology and Biogeography 21:121--133.
+
+Ross, B. E., M. B. Hooten, J.-M. DeVink, and D. N. Koons. 2015. Combined
+effects of climate, predation, and density dependence on Greater and
+Lesser Scaup population dynamics. Ecological Applications 25:1606--1617.
+
+Running, S. W., R. R. Nemani, F. A. Heinsch, M. Zhao, M. Reeves, and H.
+Hashimoto. 2004. A Continuous Satellite-Derived Measure of Global
+Terrestrial Primary Production. BioScience 54:547.
+
+Rupp, D. E., J. T. Abatzoglou, K. C. Hegewisch, and P. W. Mote. 2013.
+Evaluation of CMIP5 20th century climate simulations for the Pacific
+Northwest US. Journal of Geophysical Research 118:1--23.
+
+Salguero-G{ó}mez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.
+Che-Castaldo, H. Caswell, D. Hodgson, A. Scheuerlein, D. A. Conde, E.
+Brinks, H. de Buhr, C. Farack, F. Gottschalk, A. Hartmann, A. Henning,
+G. Hoppe, G. R{ö}mer, J. Runge, T. Ruoff, J. Wille, S. Zeh, R. Davison,
+D. Vieregg, A. Baudisch, R. Altwegg, F. Colchero, M. Dong, H. de Kroon,
+J.-D. Lebreton, C. J. E. Metcalf, M. M. Neel, I. M. Parker, T. Takada,
+T. Valverde, L. A. V{é}lez-Espino, G. M. Wardle, M. Franco, and J. W.
+Vaupel. 2015. The compadrePlant Matrix Database: an open online
+repository for plant demography. Journal of Ecology 103:202--218.
+
+Schlaepfer, D. R., W. K. Lauenroth, and J. B. Bradford. 2011.
+Ecohydrological niche of sagebrush ecosystems. Ecohydrology:n/a--n/a.
+
+Schlaepfer, D. R., W. K. Lauenroth, and J. B. Bradford. 2014a. Modeling
+regeneration responses of big sagebrush (Artemisia tridentata) to
+abiotic conditions. Ecological Modelling 286:66--77.
+
+Schlaepfer, D., W. K. Lauenroth, and J. B. Bradford. 2012. Effects of
+ecohydrological variables on current and future ranges, local
+suitability patterns, and model accuracy in big sagebrush. Ecography
+5:453--466.
+
+Schlaepfer, D., W. K. Lauenroth, and J. B. Bradford. 2014b. Natural
+Regeneration Processes in Big Sagebrush (Artemisia tridentata).
+Rangeland Ecology \& Management 67:344--357.
+
+Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova, R. B.
+O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
+Schr{ö}der, A. Singer, and N. E. Zimmermann. 2012. How to understand
+species' niches and range dynamics: A demographic research agenda for
+biogeography. Journal of Biogeography 39:2146--2162.
+
+Shafer, S. L., P. J. Bartlein, and R. S. Thompson. 2001. Potential
+changes in the distributions of western North America tree and shrub
+taxa under future climate scenarios. Ecosystems 4:200--215.
+
+Shriver, R. K. 2015. Quantifying how short-term environmental variation
+leads to long-term demographic responses to climate change. Journal of
+Ecology:n/a--n/a.
+
+Stan Development Team. 2014a. Stan: A C++ Library for Probability and
+Sampling, Version 2.5.0.
+
+Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
+2.5.0.
+
+Still, S., and B. Richardson. 2015. Projections of Contemporary and
+Future Climate Niche for Wyoming Big Sagebrush (Artemisia tridentata
+subsp. wyomingensis): A Guide for Restoration. Natural Areas Journal
+35:30--43.
+
+Vanderwel, M. C., V. S. Lyutsarev, and D. W. Purves. 2013.
+Climate-related variation in mortality and recruitment determine
+regional forest-type distributions. Global Ecology and Biogeography
+22:1192--1203.
+
+Wikle, C. K. 2010. Low-rank representations for spatial processes. Pages
+89--106 \emph{in} A. Gelfand, P. Diggle, M. Fuentes, and P. Guttopr,
+editors. Handbook of spatial statistics. Chapman; Hill, Upper Saddle
+River, New Jersey, USA.
+
+Xian, G., C. G. Homer, and C. L. Aldridge. 2012. Effects of Land Cover
+and Regional Climate Variations on Long-Term Spatiotemporal Changes in
+Sagebrush Ecosystems. GIScience \& Remote Sensing 49:378--396.
 

### 2017-02-27 docs/ms_sageAbundance.tex
just changes font color

@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
@@
 \usepackage[margin=1in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
             colorlinks=true,
-            citecolor=blue,
+            linkcolor=black,
+            citecolor=Blue,
             urlcolor=black,
-            linkcolor=black,
-            pdfborder={0 0 0}}
+            breaklinks=true}
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-\maketitle
-
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+
+\definecolor{blue}{rgb}{0,0,0}
 
@@
 observations recorded at the meter to sub-meter scale (see, e.g.,
-Salguero-G{ó}mez et al. 2015). Local-scale demographic data make
-building population projection models an easy task (Ellner and Rees
-2006, Rees and Ellner 2009, Adler et al. 2012), but it is very difficult
-to extrapolate small-scale studies to large spatial extents with any
+Salguero-Gómez et al. 2015). Local-scale demographic data make building
+population projection models an easy task (Ellner and Rees 2006, Rees
+and Ellner 2009, Adler et al. 2012), but it is very difficult to
+extrapolate small-scale studies to large spatial extents with any
 certainty because the data likely only represent a small subset of
@@
 core of their range -- areas where a species exists now and is expected
-to persist in the future (Ehrl{é}n and Morris 2015). Furthermore,
-because they lack short-term dynamics, SDMs usually cannot produce any
-estimate of the rate at which local populations will increase or
-decrease in the near-term and instead project a future equilibrium
-species distribution that may or may not ever be reached. Direct
-validation of such predictions is extremely rare (Roberts and Hamann
-2012). Large-scale dynamic population models could overcome these
-limitations. They would produce spatially-explicit estimates of species
-abundance within the species range (Ehrl{é}n and Morris 2015), have the
-potential to model expansion in abundance outside the range when coupled
-with dynamic models of dispersal, and would provide testable predictions
-of how populations should respond to short-term climate perturbations.
-These short-term predictions also would give modelers the opportunity to
+to persist in the future (Ehrlén and Morris 2015). Furthermore, because
+they lack short-term dynamics, SDMs usually cannot produce any estimate
+of the rate at which local populations will increase or decrease in the
+near-term and instead project a future equilibrium species distribution
+that may or may not ever be reached. Direct validation of such
+predictions is extremely rare (Roberts and Hamann 2012). Large-scale
+dynamic population models could overcome these limitations. They would
+produce spatially-explicit estimates of species abundance within the
+species range (Ehrlén and Morris 2015), have the potential to model
+expansion in abundance outside the range when coupled with dynamic
+models of dispersal, and would provide testable predictions of how
+populations should respond to short-term climate perturbations. These
+short-term predictions also would give modelers the opportunity to
 repeatedly validate and refine their models (Luo et al. 2011).
@@
 
+\hypertarget{refs}{}
+\hypertarget{ref-Adler2012}{}
 Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
@@
 
+\hypertarget{ref-Apodaca2013}{}
 Apodaca, L. F. 2013. Assessing Growth Response to Climate Controls in a
@@
 
+\hypertarget{ref-Arnett2015}{}
 Arnett, E. B., and T. Z. Riley. 2015. Science, policy, and the fate of
@@
 
+\hypertarget{ref-Baldeck2014}{}
 Baldeck, C. A., and G. P. Asner. 2014. Improving remote species
@@
 
+\hypertarget{ref-Barry1996}{}
 Barry, R. P., and J. M. V. Hoef. 1996. Blackbox Kriging: Spatial
@@
 
+\hypertarget{ref-Bradford2006}{}
 Bradford, J. B., and W. K. Lauenroth. 2006. Controls over invasion of
@@
 
+\hypertarget{ref-Bradley2010}{}
 Bradley, B. A. 2010. Assessing ecosystem threats from global and
@@
 
+\hypertarget{ref-Bradshaw2000}{}
 Bradshaw, G. A., and J. G. Borchers. 2000. Uncertainty as information:
@@
 
+\hypertarget{ref-Clark2011}{}
 Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011. Climate
@@
 
+\hypertarget{ref-Clark2001}{}
 Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
@@
 Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
-2001. Ecological forecasts: an emerging imperative. Science (New York,
-N.Y.) 293:657--660.
-
+2001. Ecological forecasts: an emerging imperative. Science
+293:657--660.
+
+\hypertarget{ref-Clark2014}{}
 Clark, J. S., A. E. Gelfand, C. W. Woodall, and K. Zhu. 2014. More than
@@
 
+\hypertarget{ref-Colgan2014}{}
 Colgan, M. S., and G. P. Asner. 2014. Coexistence and environmental
@@
 
+\hypertarget{ref-Conn2015}{}
 Conn, P. B., D. S. Johnson, J. M. V. Hoef, M. B. Hooten, J. M. London,
@@
 
+\hypertarget{ref-Dalgleish2011}{}
 Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
@@
 
+\hypertarget{ref-Debinski2010}{}
 Debinski, D. M., H. Wickham, K. Kindscher, J. C. Caruthers, and M.
@@
 
-Ehrl{é}n, J., and W. F. Morris. 2015. Predicting changes in the
+\hypertarget{ref-Ehrlen2015}{}
+Ehrlén, J., and W. F. Morris. 2015. Predicting changes in the
 distribution and abundance of species under environmental change.
@@
 
+\hypertarget{ref-Elith2009}{}
 Elith, J., and J. R. Leathwick. 2009. Species Distribution Models:
@@
 
+\hypertarget{ref-Ellner2006}{}
 Ellner, S. P., and M. Rees. 2006. Integral projection models for species
@@
 
+\hypertarget{ref-Freckleton2011}{}
 Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
@@
 
+\hypertarget{ref-Gelman2009}{}
 Gelman, A., and J. Hill. 2009. Data analysis using regression and
@@
 
+\hypertarget{ref-Gelman1992}{}
 Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation
@@
 
+\hypertarget{ref-Gerber2015}{}
 Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
@@
 
+\hypertarget{ref-Germino2014}{}
 Germino, M. J., and K. Reinhardt. 2014. Desert shrub responses to
@@
 
+\hypertarget{ref-Hare2010}{}
 Hare, J. a, M. a Alexander, M. J. Fogarty, E. H. Williams, and J. D.
@@
 
+\hypertarget{ref-Harte2015}{}
 Harte, J., S. R. Saleska, and C. Levy. 2015. Convergent ecosystem
@@
 
+\hypertarget{ref-He2015}{}
 He, K. S., B. A. Bradley, A. F. Cord, D. Rocchini, M.-N. Tuanmu, S.
@@
 
+\hypertarget{ref-Higdon1998}{}
 Higdon, D. 1998. A process-convolution approach to modelling
@@
 
+\hypertarget{ref-Higdon2002}{}
 Higdon, D. M. 2002. Space and space-time modeling using process
@@
 
+\hypertarget{ref-Hobbs2015}{}
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 
+\hypertarget{ref-Homer2012}{}
 Homer, C. G., C. L. Aldridge, D. K. Meyer, and S. J. Schell. 2012.
@@
 
+\hypertarget{ref-Homer2015}{}
 Homer, C. G., G. Xian, C. L. Aldridge, D. K. Meyer, T. R. Loveland, and
@@
 
+\hypertarget{ref-Hooten2015}{}
 Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
@@
 
+\hypertarget{ref-Hooten2007a}{}
 Hooten, M. B., and C. K. Wikle. 2007. Shifts in the spatio-temporal
@@
 
+\hypertarget{ref-Hooten2003}{}
 Hooten, M. B., D. R. Larsen, and C. K. Wikle. 2003. Predicting the
@@
 
+\hypertarget{ref-Kuchler1964}{}
 Kuchler, A. 1964. Potential Natural Vegetation of the Conterminous
@@
 
+\hypertarget{ref-Latimer2009}{}
 Latimer, A. M., S. Banerjee, H. Sang, E. S. Mosher, and J. A. Silander.
@@
 
+\hypertarget{ref-Luo2011a}{}
 Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
@@
 
+\hypertarget{ref-Maiorano2013}{}
 Maiorano, L., R. Cheddadi, N. E. Zimmermann, L. Pellissier, B.
@@
 
+\hypertarget{ref-Merow2014}{}
 Merow, C., A. M. Latimer, A. M. Wilson, S. M. McMahon, A. G. Rebelo, and
@@
 
+\hypertarget{ref-Miglia2005}{}
 Miglia, K., E. Mcarthur, W. Moore, H. Wang, J. Graham, and D. Freeman.
@@
 
+\hypertarget{ref-Neilson2005}{}
 Neilson, R., J. Lenihan, D. Bachelet, and R. Drapek. 2005. Climate
@@
 
+\hypertarget{ref-Pechanec1937}{}
 Pechanec, J., G. Pickford, and G. Stewart. 1937. Effects of the 1934
@@
 
+\hypertarget{ref-Perfors2003}{}
 Perfors, T., J. Harte, and S. E. Alter. 2003. Enhanced growth of
@@
 
-Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
+\hypertarget{ref-Petchey2015}{}
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
 Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
@@
 
+\hypertarget{ref-Poore2009}{}
 Poore, R. E., C. A. Lamanna, J. J. Ebersole, and B. J. Enquist. 2009.
@@
 
+\hypertarget{ref-Queenborough2011}{}
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 
+\hypertarget{ref-R2013}{}
 R Core Team. 2013. R: A language and environment for statistical
@@
 
+\hypertarget{ref-Rees2009}{}
 Rees, M., and S. P. Ellner. 2009. Integral projection models for
@@
 
+\hypertarget{ref-Roberts2012}{}
 Roberts, D. R., and A. Hamann. 2012. Predicting potential climate change
@@
 
+\hypertarget{ref-Ross2015}{}
 Ross, B. E., M. B. Hooten, J.-M. DeVink, and D. N. Koons. 2015. Combined
@@
 
+\hypertarget{ref-Running2004}{}
 Running, S. W., R. R. Nemani, F. A. Heinsch, M. Zhao, M. Reeves, and H.
@@
 
+\hypertarget{ref-Rupp2013}{}
 Rupp, D. E., J. T. Abatzoglou, K. C. Hegewisch, and P. W. Mote. 2013.
@@
 
-Salguero-G{ó}mez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.
+\hypertarget{ref-Salguero-Gomez2015}{}
+Salguero-Gómez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.
 Che-Castaldo, H. Caswell, D. Hodgson, A. Scheuerlein, D. A. Conde, E.
 Brinks, H. de Buhr, C. Farack, F. Gottschalk, A. Hartmann, A. Henning,
-G. Hoppe, G. R{ö}mer, J. Runge, T. Ruoff, J. Wille, S. Zeh, R. Davison,
-D. Vieregg, A. Baudisch, R. Altwegg, F. Colchero, M. Dong, H. de Kroon,
+G. Hoppe, G. Römer, J. Runge, T. Ruoff, J. Wille, S. Zeh, R. Davison, D.
+Vieregg, A. Baudisch, R. Altwegg, F. Colchero, M. Dong, H. de Kroon,
 J.-D. Lebreton, C. J. E. Metcalf, M. M. Neel, I. M. Parker, T. Takada,
-T. Valverde, L. A. V{é}lez-Espino, G. M. Wardle, M. Franco, and J. W.
+T. Valverde, L. A. Vélez-Espino, G. M. Wardle, M. Franco, and J. W.
 Vaupel. 2015. The compadrePlant Matrix Database: an open online
@@
 
+\hypertarget{ref-Schlaepfer2011}{}
 Schlaepfer, D. R., W. K. Lauenroth, and J. B. Bradford. 2011.
@@
 
+\hypertarget{ref-Schlaepfer2014}{}
 Schlaepfer, D. R., W. K. Lauenroth, and J. B. Bradford. 2014a. Modeling
@@
 
+\hypertarget{ref-Schlaepfer2012}{}
 Schlaepfer, D., W. K. Lauenroth, and J. B. Bradford. 2012. Effects of
@@
 
+\hypertarget{ref-Schlaepfer2014a}{}
 Schlaepfer, D., W. K. Lauenroth, and J. B. Bradford. 2014b. Natural
@@
 
+\hypertarget{ref-Schurr2012a}{}
 Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova, R. B.
 O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
-Schr{ö}der, A. Singer, and N. E. Zimmermann. 2012. How to understand
+Schröder, A. Singer, and N. E. Zimmermann. 2012. How to understand
 species' niches and range dynamics: A demographic research agenda for
@@
 
+\hypertarget{ref-Shafer2001}{}
 Shafer, S. L., P. J. Bartlein, and R. S. Thompson. 2001. Potential
@@
 
+\hypertarget{ref-Shriver2015}{}
 Shriver, R. K. 2015. Quantifying how short-term environmental variation
@@
 
+\hypertarget{ref-stan2014}{}
 Stan Development Team. 2014a. Stan: A C++ Library for Probability and
@@
 
+\hypertarget{ref-rstan2014}{}
 Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
@@
 
+\hypertarget{ref-Still2015}{}
 Still, S., and B. Richardson. 2015. Projections of Contemporary and
@@
 
+\hypertarget{ref-Vanderwel2013b}{}
 Vanderwel, M. C., V. S. Lyutsarev, and D. W. Purves. 2013.
@@
 
+\hypertarget{ref-Wikle2010}{}
 Wikle, C. K. 2010. Low-rank representations for spatial processes. Pages
@@
 
+\hypertarget{ref-Xian2012}{}
 Xian, G., C. G. Homer, and C. L. Aldridge. 2012. Effects of Land Cover
